---
title: "Load data"
author: "Warner van Kersen (adapted by Beatrice Cornu Hewitt)"
date: "`r Sys.time()`"
output: html_document
---
# Load packages
```{r}
library(dplyr); library(labelled); library(gt); library(openxlsx); library(haven); library(lubridate); library(table1); library(rspiro)
```

# Load datasets from VGO2 and VGO3
```{r}
#### VGO3 data ####
getwd()
# Merged spirometry
ds <- read.csv2("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO_3_cleaned_spirometry.csv", row.names = NULL) %>% select(-any_of("X"))

# Spirometry grading file, cleaned
grade <- read.xlsx("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_spirometrygrades_clean_20220913.xlsx")
  
# Online questionnaire
OQ <- read.xlsx("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO3_CP_OnlineQuest_joined_not_cleaned_20221017.xlsx")
  
# Field questionnaire
field_q <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_CP_fieldquest_20220909.csv")
  
# Field form
field_f <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_CP_fieldform_20220909.csv")


#### VGO2 data ####
# Full dataset
old <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Overdracht VGO gezondheid/Sas dataset/lung_sero_scr_longq_emr_09052017.sas7bdat")
    
# Intake file for test dates - in full dataset 12 dates (test_date_pre) are missing
old_intake <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Data intake + medicijngebruik/vgointake.sas7bdat")
    
# Spirometry
old_spiro <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Longfuncties/Dataset analyse/vgo_lf_12feb2016_n2439.sas7bdat")

# Livestock emitted PM10 and endotoxin fraction modelled by Myrna
old_endo <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO-COPD/Microbiome/Warner_R_projects/dada2 analyse COPD microbioom/00_raw_data/VGO_2494_EUpred_WvK.csv") %>% select(-X) 
    
old_endo <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO-COPD/Microbiome/Warner_R_projects/dada2 analyse COPD microbioom/00_raw_data/VGO_2494_PM10CONCpred_WvK.csv") %>% select(-X) %>% left_join(old_endo,.,by="respnr")
    
# Old complete (for continuous exposure proxies)
old_full <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Overdracht VGO gezondheid/Sas dataset/datasetmyrna18072017.sas7bdat")

#Farm child variables
# farmchild,farmchtotyr,farmchanim,farmchcrop,farmchdairy,farmchveal,farmchgoat,farmchsheep,farmchpig,farmchpoultry,farmchhorse,farmchother,farmchothertxt,farmchholiday,farmnone,farmchanim,farmchmanure,farmchstraw,farmchcropcare
    
# 2021 and 2022 air pollutant data from national air quality monitoring network
AP_monitor <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO3student/VGO_Yuhong/Exposure data/Weekly ambient PM10/VGO3 CP YS Weekly ambient PM10 one week before.csv") %>% 
      dplyr::select(respnr,weeklyap,season) %>% 
      dplyr::rename(ID=respnr,wkprPM10_ugm3=weeklyap)
    
AP_monitor$season <- relevel(as.factor(AP_monitor$season),ref = "Winter")
   
# 2021 and 2022 daily ammonia data
# Distance weighted livestock proxies, not used in final analysis
WghtDist <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Livestock_computations/VGO1/VGO1_2494_WghtDist.csv") %>% dplyr::select(-X,-nfuranimsWghtDist.1000m.sum,-nfuranimsWghtDist.3000m.sum,-nFuranimFarmWghtDist.1000m.sum,-nFuranimFarmWghtDist.3000m.sum) %>% rename(ID=respnr)
   
# Week prior NH3 exposures and difference
NH_diff <- readRDS("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/wk_prior_NH3_diff") # Week prior NH3 levels in VGO2 and VGO3 (then also the difference between these)
  
```
# VGO3 spirometry data cleaning
### Prep cleaned spirometry data for analysis
```{r}
#### VGO3 spirometry ####  
# Check for duplicate records and rempove any 
ds1 <- ds %>% distinct(ID, DateOfBirth, Test.Date, FEV1, FVC, .keep_all =T )

# Turn measurement values presented as character strings to numeric
str(ds1)
# ds2 <- ds1 %>% mutate_if(is.character,as.numeric) %>%
#     mutate(Gender=ds1$Gender,
#            DateOfBirth=ds1$DateOfBirth,
#            Ethnicity=ds1$Ethnicity,
#            Test.Date=ds1$Test.Date)

ds2 <- ds1 %>%
  mutate(across(where(is.character) & !any_of(c("Gender", "Ethnicity", "DateOfBirth", "Test.Date")), as.numeric))
str(ds2)

# Merge measurement values with spiro grades
sum(is.na(grade$FEV1_quality)) # no grades missing
sum(is.na(grade$FVC_quality)) # no grades missing

# calculate n accepted trials
accept <- c("A","B","C") # define accepted range
accept_D <- c("A","B","C","D") # define accepted range + D
naccept <- c("D","E","F","U") # define not accepted range
# A = 3 acceptable trials within 100ml
# A = 3 acceptable trials within 150ml
# C = 2 acceptable trials
# D = 1 acceptable trial  - probably false low

AB <- c("A","B")
EFU <- c("E","F","U")

n_accept_tests <-   sum(grade$FEV1_quality %in% accept & grade$FVC_quality %in% accept) # 845
n_accept_tests_D <-   sum(grade$FEV1_quality %in% accept_D & grade$FVC_quality %in% accept_D) # 861

# create dataset with one value per measurement - select highest FEV1 and FVC from reproducible measurements
ds3 <- grade %>% select(ID,Spirometrist,FEV1_quality,FVC_quality) %>%
    left_join(ds2,.,by="ID") %>%
    mutate(FEV1_accepted=case_when(!FEV1_quality %in% accept~0,
                                    FEV1_quality %in% accept~1),
           FVC_accepted=case_when(!FVC_quality %in% accept~0,
                                    FVC_quality %in% accept~1),
           test_accepted=case_when(!(FEV1_accepted==1&FVC_accepted==1)~0,
                                    FEV1_accepted==1&FVC_accepted==1~1),
           ntrial_acc=case_when(FEV1_quality %in% AB & FVC_quality %in% AB~3,
                                FEV1_quality %in% AB & FVC_quality == "C" ~2,
                                FEV1_quality %in% AB & FVC_quality == "D" ~1,
                                FEV1_quality == "C" & FVC_quality %in% accept~2,
                                FEV1_quality == "D" & FVC_quality %in% accept~1,
                                FEV1_quality == "D" & FVC_quality == "D"~1,
                                FEV1_quality == "D" & FVC_quality == "E"~0,
                                FEV1_quality %in% accept & FVC_quality %in% EFU~0,
                                FEV1_quality %in% EFU~0)) %>%
    group_by(ID,ntrial_acc) %>% 
    mutate(trial_acc=case_when(ntrial_acc==0~0,
                               ntrial_acc==1 & Trial_rank==1~1,
                               ntrial_acc==1 & Trial_rank!=1~0,
                               ntrial_acc==2 & Trial_rank %in% c(1,2)~1,
                               ntrial_acc==2 & !Trial_rank %in% c(1,2)~0,
                               ntrial_acc==3 & Trial_rank %in% c(1,2,3)~1,
                               ntrial_acc==3 & !Trial_rank %in% c(1,2,3)~0
                               )
    )
    

sum(ds3$trial_acc) # 2520
  
table(ds3$FEV1_quality,ds3$FVC_quality) # How often each FEV1 quality level appears with each FVC quality level
  #     A    B    C    D    E    F    U
  # A 4292  102  117   39   89    0   37
  # B  196   44   25   13   20    0    5
  # C   41    0   20    0   13    0    0
  # D   22    8   15   11   13    0    0
  # E   59   23   28    0  110    0   17
  # F   17    0    0    0    0   42    0
  # U    6    0    0    0    8    0   30
table(ds3$FEV1_quality,ds3$ntrial_acc,useNA = "ifany") # how each FEV1 quality grade maps to the numeric ntrial_acc classification
  #    0    1    2    3
  # A  126   39  117 4394
  # B   25   13   25  240
  # C   13    0   61    0
  # D   13   56    0    0
  # E  237    0    0    0
  # F   59    0    0    0
  # U   44    0    0    0

# Create a summary table showing the number of participants whose first spirometry trial (Trial_rank == 1) falls into each combination of:FEV1 grade, FVC grade, Number of acceptable trials (ntrial_acc)
ds3$Trial_rank
library(dplyr)
library(gt)

# Create summary table using gt
ds3 %>%
  filter(Trial_rank == 1) %>%
  group_by(FEV1_quality, FVC_quality, ntrial_acc) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(FEV1_quality, FVC_quality) %>%
  gt() %>%
  cols_label(
    FEV1_quality = "FEV₁ Grade",
    FVC_quality = "FVC Grade",
    ntrial_acc = "Number of Acceptable Attempts",
    n = "Number of Participants"
  ) %>%
  tab_header(
    title = "First Trial Spirometry Quality Summary"
  )


# Calculate number of acceptable tests
accept_tests <- ds3 %>% group_by(ID) %>% filter(Trial_rank==1&test_accepted==1) %>%
summarise(Acceptable_tests=n())
  
# check if n accepted tests matches expected
sum(accept_tests$Acceptable_tests) == n_accept_tests # TRUE 

# Calculate number of acceptable trials (attempts within tests)
total_ntrial_acc <- ds3 %>% group_by(ID) %>% select(ID,ntrial_acc) %>% distinct() %>% ungroup() %>%
    summarise(total=across(ntrial_acc,sum)) %>% unlist() %>% as.integer()
  
# should match sum of trial_acc
total_ntrial_acc == sum(ds3$trial_acc)
  
# Check BEV criteria in all acceptable trials
ds_bevcheck <- ds3 %>% ungroup() %>%
    filter(trial_acc==1) %>% 
    mutate(fvc5perc=0.05*FVC,
           bevcheck=case_when(fvc5perc>0.150~BEV-fvc5perc,
                              fvc5perc<=0.150~BEV-0.150)) %>%
    filter(bevcheck>0) %>%
    select(ID,BEV,bevcheck,fvc5perc,FVC_quality,FEV1_quality)# This filters violations (bevcheck > 0) — meaning the BEV is too large, so the trial might be invalid.
  
# Select the 3 best trials for each participant, when selecting highest FEV1 and FVC only use accepted trials
ds_minmax <- ds3 %>% group_by(ID) %>%
    filter(Trial_rank %in% c(1,2,3)&trial_acc==1) %>%
    # select(ID,Trial_rank,FEV1,FVC,PEF,ntrial_acc,FEV1_quality,FVC_quality) %>% (line used for checking, printing less columns)
    arrange(ID,Trial_rank) %>%
    mutate(fev1min=min(FEV1), # Calc min max med FEV1
           fev1med=case_when(ntrial_acc==3~median(FEV1)),
           fev1max=max(FEV1),
           fev1dif=case_when(ntrial_acc==3~fev1max-fev1med,
                             ntrial_acc==2~fev1max-fev1min),
           fev1_d150=case_when(fev1dif<=0.150~0,
                               fev1dif>0.150~1,
                               ntrial_acc==1~1),
           fvcmin=min(FVC), # calc min max med FVC
           fvcmed=case_when(ntrial_acc==3~median(FVC)),
           fvcmax=max(FVC),
           fvcdif=case_when(ntrial_acc==3~fvcmax-fvcmed,
                            ntrial_acc==2~fvcmax-fvcmin),
           fvc_d150=case_when(fvcdif<=0.150~0,
                              fvcdif>0.150~1,
                              ntrial_acc==1~1),
           lf_repr150=case_when(fvc_d150 == 1 | fev1_d150 == 1~0, # is the test reproducible based on both FEV1 & FVC?
                                fvc_d150 == 0 & fev1_d150 == 0~1),
           pefmin=min(PEF), # calc min max med PEF
           pefmed=case_when(ntrial_acc==3~median(PEF)),
           pefmax=max(PEF),
           pefdiff=case_when(ntrial_acc>2&lf_repr150==1~pefmax-pefmed),
           pef_d670=case_when(pefdiff<=0.67~0, # checks PEF reproducibility (>0.67 L/s difference from median)
                              pefdiff>0.67~1)
           )
           
    
ds_minmax$pef_d670[is.na(ds_minmax$pefdiff)]

# Excluding all participants with <1 (so 2 or more) accepted trials we should have 845 in the dataset
length(unique(ds_minmax$ID[ds_minmax$ntrial_acc>1]))  # 845

# Check fraction reproducible vs. non-reproducible in grades A and B (>2 reproducible trials)
table(ds_minmax$fev1_d150[ds_minmax$ntrial_acc>2])  
# 0    1 
# 2409   33 
# So, among those with 3 acceptable trials, almost all have reproducible FEV1.

``` 

## Spirometry - quality grades
```{r}
sumfunc <- function(z) if (is.numeric(z)) sum(z) else ''

FEV1_grade <- grade %>% rename(FEV1=FEV1_quality) %>% group_by(FEV1) %>% summarise(n=n()) # rename the FEV1_quality to FEV1, groups th eFEV1 grade (A,B,C...), counts the number per grade
# FEV1 n
# A	818			
# B	50			
# C	11			
# D	10			
# E	34			
# F	9			
# U	6

FVC_grade <-  grade %>% rename(FVC=FVC_quality) %>% group_by(FVC) %>% summarise(n=n()) 
# FVC n
# A	811			
# B	27			
# C	31			
# D	10			
# E	38			
# F	7			
# U	14

# FEV1
FEV1_grade_sum <- as.data.frame(lapply(FEV1_grade, sumfunc))
FEV1_grade_sum[1] <- "total"
FEV1_grade <- rbind(FEV1_grade,FEV1_grade_sum) # bind to original FEV1 table
# Create and append “acceptable” row
FEV1_grade_sum_cplus <- data.frame(FEV1_quality="acceptable",n=sum(FEV1_grade$n[FEV1_grade$FEV1 %in% accept])) %>% rename(FEV1=FEV1_quality)
FEV1_grade <- rbind(FEV1_grade,FEV1_grade_sum_cplus) # Adds a row counting grades considered acceptable (like A, B, etc.) from the accept vector

# FVC
FVC_grade_sum <- as.data.frame(lapply(FVC_grade, sumfunc))
FVC_grade_sum[1] <- "total"
FVC_grade <- rbind(FVC_grade,FVC_grade_sum) # bind to original FEV1 table
# Create and append “acceptable” row
FVC_grade_sum_cplus <- data.frame(FVC_quality="acceptable",n=sum(FVC_grade$n[FVC_grade$FVC %in% accept])) %>% rename(FVC=FVC_quality)
FVC_grade <- rbind(FVC_grade, FVC_grade_sum_cplus) #  Adds a row counting grades considered acceptable (like A, B, etc.) from the accept vector

# kable(list(FEV1_grade,FVC_grade),align = "l",caption = "Quality grading",padding=0) %>% 
# kableExtra::kable_styling(font_size = 15,full_width = F)

# Subscript FEV1
names(FEV1_grade) <-c("FEV~1~","n") 

# Both FEV1 and FVC acceptable

# kable(FEV1_grade) %>%
#   kable_styling(full_width = FALSE, position = "float_left",font_size = 18)
# kable(FVC_grade) %>%
#   kable_styling(full_width = FALSE, position = "left",font_size = 18)

# kable(list(FEV1_grade,FVC_grade)) %>%
#   kable_styling(full_width = FALSE, position = "left",font_size = 18)

grade_table <- grade %>% select(ID,FEV1_quality,FVC_quality) %>%
  pivot_longer(cols = c(FEV1_quality,FVC_quality),names_to = "variable",values_to = "grade")


grade_table$variable <- str_replace(grade_table$variable,pattern = "_quality",replacement = "")

var_label(grade_table$grade) <- "Quality grading"

table1(~grade|variable, data=grade_table, overall = F)
#	FEV1 (N=938)	FVC (N=938)
# Quality grading		
# A	818 (87.2%)	811 (86.5%)
# B	50 (5.3%)	27 (2.9%)
# C	11 (1.2%)	31 (3.3%)
# D	10 (1.1%)	10 (1.1%)
# E	34 (3.6%)	38 (4.1%)
# F	9 (1.0%)	7 (0.7%)
# U	6 (0.6%)	14 (1.5%)

``` 

## Spirometry - BEV check
```{r}
# BEV should be smaller than 150ml or smaller than 5% of the FVC, whichever is larger
# Print a clean, readable summary table for the subset of data failing or flagged by the BEV criterion.
ds_bevcheck %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  gt() %>%
  tab_header(title = "Print BEV check (BEV - 0.150 or BEV - FVC5%) > 0")


```

## Spirometry - reproducibility
```{r}
d_repro_t <- ds_minmax %>% filter(Trial_rank==1) %>%
    group_by(FEV1_quality,FVC_quality,lf_repr150) %>%
    replace_na(list(FEV1_quality="E or lower",FVC_quality="E or lower")) %>%
    mutate(Quality=paste(FEV1_quality,FVC_quality,sep = "-"),
           lf_repr150=factor(lf_repr150,c(0,1),c("not repr.","repr.")),
           pef_d670=factor(pef_d670,c(0,1),c("<.67 L/sec)",">.67 L/sec"))
    )


var_label(d_repro_t$Quality) <- "FEV<sub>1</sub> - FVC grade"

table1(~Quality|lf_repr150,d_repro_t,caption = "Quality grades and FEV<sub>1</sub>&FVC reproducibility within 150ml")

```

## Spirometry - reproducibility continued with PEF

```{r}
var_label(d_repro_t$pef_d670) <- "PEF diff."

table1(~pef_d670|lf_repr150, d_repro_t,caption = "difference max PEF second best PEF vs. FEV<sub>1</sub> and FVC reproducibility within 150ml")
```

# Create analysis dataset
## Select FEV and FVC variables (volume variables) 
```{r}
# one value per test
# 1. Take FEV<sub>1</sub> and FVC from trial rank 1, if one or both are not reproducible within 150ml
# 2. Take highest FEV<sub>1</sub> and FVC from (possibly different) reproducible curves
ds_prep_forced <- ds_minmax %>% 
filter(!(lf_repr150==0 & Trial_rank>1)) %>% # Keep only trial rank 1 from tests without reproducible trials, but keep all trials when lung function is reproducible
  group_by(ID) %>%
  mutate(FEV1_s=case_when(lf_repr150==0~FEV1, # Creates new variables (FEV1_s, FVC_s, etc.) selecting either: value form the single trial if lung function is not reproducible, or the maximum value (best trial) if it is reproducible ('s' for selected)
                          lf_repr150==1~fev1max),
         FEV25=case_when(lf_repr150==0~FEV.25,
                          lf_repr150==1&FEV1==fev1max~FEV.25), 
         FEV5=case_when(lf_repr150==0~FEV.5,
                          lf_repr150==1&FEV1==fev1max~FEV.5),
         FEV75=case_when(lf_repr150==0~FEV.75,
                          lf_repr150==1&FEV1==fev1max~FEV.75),
         FEV3=case_when(lf_repr150==0~FEV3,
                          lf_repr150==1&FEV1==fev1max~FEV3),
         FEV6=case_when(lf_repr150==0~FEV6,
                          lf_repr150==1&FEV1==fev1max~FEV6),
         FVC_s=case_when(lf_repr150==0~FVC,
                          lf_repr150==1~fvcmax)) %>% 
  filter(!is.na(FEV25)) %>% # Filter empty rows
  select(ID,Spirometrist,Gender,Height,Weight,Ethnicity,Test.Date,lf_repr150,FEV1_quality,FVC_quality,
         test_accepted,FEV1_s,FEV25,FEV5,FEV75,FEV3,FEV6,FVC_s,FEV1.FVC) %>% 
  rename(date=Test.Date,FEV1=FEV1_s,FVC=FVC_s,FEV1_FVC=FEV1.FVC)
      
# Extremely high values exist due to missing decimal separator, introduce decimal separator between first and second digit in those cases.
ds_prep_forced <- transform(ds_prep_forced,
                            FVC=sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",FVC),
                            FEV6=sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",FEV6)
                            ) %>%
                  mutate(FVC=as.numeric(FVC),
                         FEV6=as.numeric(FEV6))

table1(~FEV1+FVC+FEV25+FEV5+FEV75+FEV3+FEV6,data = ds_prep_forced,caption = "Summary of FEV and FVC variables") 


table1(~FEV1+FVC+FEV25+FEV5+FEV75+FEV3+FEV6+FEV1_FVC,data = ds_prep_forced,caption = "Summary of FEV and FVC variables") 
```
## Creating the “flow” variables analysis dataset
```{r}
## Create analysis dataset - Flow variables
# - If not reproducible take all flow variables from Trial_rank 1
# - If reproducible and PEF<sub>max</sub>-PEF<sub>med</sub> **>0.67**
# take all flow variables from Trial_rank 1
# - If reproducible and PEF<sub>max</sub>-PEF<sub>med</sub> **<0.67**
# take all flow variables from Trial_rank 1

options(scipen=999)
ds_prep_flow <- ds_minmax %>% 
filter(!(lf_repr150==0 & Trial_rank>1)) %>% # Keep only trial rank 1 from tests without reproducible trials
  group_by(ID) %>%
  mutate(FEF10_s=case_when(lf_repr150 == 0~FEF10,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF10,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF10),
         FEF25=case_when(lf_repr150 == 0~FEF25.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.),
         FEF25_75=case_when(lf_repr150 == 0~FEF25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.75.),
         FEF40=case_when(lf_repr150 == 0~FEF40.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF40.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF40.),
         FEF50=case_when(lf_repr150 == 0~FEF50.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF50.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF50.),
         FEF60=case_when(lf_repr150 == 0~FEF60.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF60.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF60.),
         FEF75=case_when(lf_repr150 == 0~FEF75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF75.),
         FEF75_85=case_when(lf_repr150 == 0~FEF75.85.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF75.85.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF75.85.),
         FEF80=case_when(lf_repr150 == 0~FEF80.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF80.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF80.),
         FEF25_75=case_when(lf_repr150 == 0~FEF25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.75.),
         FET25_75=case_when(lf_repr150 == 0~FET25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FET25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FET25.75.),
         PEF_s=case_when(lf_repr150 == 0~PEF,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~PEF,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~PEF),
         PEF_Lmin=case_when(lf_repr150 == 0~PEF..L.min.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~PEF..L.min.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~PEF..L.min.)) %>%
  filter(!is.na(FEF10_s)) %>% # Filter empty rows
  select(ID,Trial_rank,pef_d670,FEF10_s,FEF25,FEF25_75,FET25_75,FEF40,FEF50,FEF60,FEF75,FEF75_85,FEF80,FEF25_75,PEF_s,PEF_Lmin) %>%
  rename(FEF10=FEF10_s,PEF=PEF_s)

# ID 14139 has two trials with were pefmax pef, keep flow variables from trial rank 1 (the best trial)

ds_prep_flow <- ds_prep_flow %>% filter(!(ID==14139&Trial_rank==3)) 

table1(~FEF10+FEF25+ FEF25_75+FEF40+FEF50+FEF60+FEF75+FEF75_85+FEF80+FEF25_75+FET25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables") 

# Extremely high values exist, inspect:

ds_prep_flow$PEF[ds_prep_flow$PEF>100] 

# Write function to correct, if first two digits start with 1: decimal

correct_pef <- function(x){
  if (x<100) {
  return(x) 
  } 
  start <- x %>% str_extract(.,"((?=^1)(^\\d{2})|(^\\d{1}))")
  if (str_detect(start,"^\\d{2}$")) {
    output <- sub("(^\\d{2})\\.?(\\d*)","\\1.\\2",x)
  } else {
    output <- sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",x)
  }
  return(as.numeric(output))
}  
  
ds_prep_flow <- ds_prep_flow %>% mutate(PEF=correct_pef(PEF))

# Same problem exists for PEF_Lmin, remove trailing zeroes first, then insert decimal seperator between 3rd and 4th digit
ds_prep_flow$PEF_Lmin <- str_replace(ds_prep_flow$PEF_Lmin,"0+$","") %>%
stringi::stri_sub_replace(.,4,3,value = ".") %>%
  as.numeric()

# Check
table1(~FEF10+FEF25+ FEF25_75+FEF40+FEF50+FEF60+FEF75+FEF75_85+FEF80+FEF25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables - corrected") 

# Check 3 corrected PEF ID's in EasyOne

ds_prep_flow[ds_prep_flow$PEF==9.6217870712280288,] # EasyOne PEF=9.62
ds_prep_flow[ds_prep_flow$PEF==7.020787239074707,]  # EasyOne PEF=7.02
ds_prep_flow[ds_prep_flow$PEF==7.2133121490478512,] # EasyOne PEF=7.21


table1(~FEF10+FEF25+FEF25_75+FEF40+FEF50+FEF60,data = ds_prep_flow,caption = "Summary of flow variables")

table1(~FEF75+FEF75_85+FEF80+FEF25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables - continued")
```

# Metadata cleaning - VGO2 and VGO3
Cleans and prepares the spirometry data:
- Handles missing values in quality grades
- Calculates number of acceptable trials
- Checks for BEV (Back Extrapolated Volume) criteria

Processes questionnaire data:
- Recodes variables
- Handles missing values
- Creates derived variables
```{r}
#### VGO3 ####
# Merge cleaned VGO3 spirometry datasets
# ds_prep_forced - the volume-based dataset - contains lung function volume measurements per individual (ID) - also includes demographics like ID, spirometrist, ethnicity, date...
# ds_prep_flow - the flow-based dataset - contains lung function flow measurements per individual (ID) e.g. FEF & PEF variables 
# Merge analysis datasets
ds_prep <- left_join(ds_prep_forced, ds_prep_flow, by="ID")  %>% 
  select(-any_of(c("Trial_rank","Height","Weight","Gender"))) 
names(ds_prep)

# Process VGO3 online questionnaire
OQ_clean <- OQ %>%
  select(ID, fuquestdate, fugender, fuage, fuasever, fuasdd, ful3mdailycough, ful3mdailyphlegm, 
         fudoctoldcopd, fusmokeever, fuwheezly, fufarmvstfam, fufarmvstjob, fufarmvstoth, 
         fufarmvstpet, fufarmvstshop, fufarmvstno) %>%
  mutate(
    fuquestdate = ymd(fuquestdate), # Convert date to standard format
    fugender = factor(fugender, levels = c(1, 2), labels = c("male", "female")), # Recode gender to "male"/"female"
    fuage = as.numeric(fuage), # Ensure age is numeric
    # Recode binary variables (0 = no, 1 = yes)
    fuwheezly = factor(fuwheezly, levels = c(0, 1), labels = c("no", "yes")), 
    fuasever = factor(fuasever, levels = c(0, 1), labels = c("no", "yes")),
    fuasdd = factor(fuasdd, levels = c(0, 1), labels = c("no", "yes")),
    ful3mdailycough = factor(ful3mdailycough, levels = c(0, 1), labels = c("no", "yes")),
    ful3mdailyphlegm = factor(ful3mdailyphlegm, levels = c(0, 1), labels = c("no", "yes")),
    fudoctoldcopd = factor(fudoctoldcopd, levels = c(0, 1), labels = c("no", "yes")),
    # Recode smoking history into categories
    fusmokeever = factor(fusmokeever, levels = c("no", "yes", "yesstop"), 
                         labels = c("non", "current", "former")),
    # Create a new farm visit flag (yes/no based on any reason)
    fufarmvst = factor(case_when(
      fufarmvstfam == 1 | fufarmvstoth == 1 | fufarmvstpet == 1 | 
        fufarmvstshop == 1 | fufarmvstjob == 1 ~ 1,
      fufarmvstno == 1 ~ 0), levels = c(0, 1), labels = c("no", "yes"))
  ) %>%
  # Add a variable to capture multiple reasons for visiting a farm - to distinguish between people just visiting once vs multiple times
  mutate(
    fufarmvstmult = as.numeric(fufarmvstfam) + as.numeric(fufarmvstoth) + 
      as.numeric(fufarmvstpet) + as.numeric(fufarmvstshop) + as.numeric(fufarmvstjob),
    fufarmvstmult = factor(case_when(
      fufarmvstmult > 1 ~ "yes",
      fufarmvstmult == 1 ~ "no, single reason",
      fufarmvst == "no" ~ "no, did not visit"))
  )

# Process field questionnaire
field_q_clean <- field_q %>%
  select(fuhbid, fuhbgender, fuhbdate, fuhbbdate, fuhblungmed, fuhbcovidinf, 
         fuhbcovposdate, fuhbcovsusdate, fuhbpneumonialy, fuhbtestdiff) %>%
  rename(ID = fuhbid) %>% # rename ID variable 
  mutate(
    fuhbdate = ymd(fuhbdate), # Visit date to Date format (YMD)
    fuhbcovposdate = format(ymd(fuhbcovposdate)), # Convert positive COVID test date to Date, then string
    fuhbcovidinf = factor(fuhbcovidinf, levels = c(0, 1, 2), 
                         labels = c("no", "yes, tested", "yes, suspected")), # Recode COVID infection status
    fuhbcovsusdate = dmy(fuhbcovsusdate) # Suspected COVID date to Date format (DMY)
  ) %>%
  filter(!(ID == 13852 & fuhbdate == "2022-04-26")) %>% # Remove one known duplicate (participant 13852, specific visit date).
  mutate(
    fuhbcovdate = coalesce(ymd(fuhbcovposdate), ymd(fuhbcovsusdate)), # Take the first available non-missing COVID date, using positive test if present, otherwise suspected - this should be the variable used in analyses (unified variable) 
    fuhbcovinterval = interval(fuhbcovdate, fuhbdate) %/% months(1), # Calculate months between COVID and visit date
    fuhbcovl2m = case_when(
      fuhbcovidinf == "no" ~ 0,
      fuhbcovinterval <= 1 ~ 1,
      TRUE ~ 0)  # Flag if COVID occurred in the last 2 months
  )


# Process field form measurements
field_f_clean <- field_f %>%
  select(fuhbid, fuhbfheight1, fuhbfheight2, fuhbfweigth) %>%
  rename(ID = fuhbid, fuhbfweight = fuhbfweigth) %>%
  mutate(
    fuhbheight = (fuhbfheight1 + fuhbfheight2)/2, # Average the two height measurements
    fuhbbmi = fuhbfweight/((fuhbheight/100)^2) # Calculate BMI
  ) %>%
  filter(!(ID == 13852 & fuhbfweight == 100.0)) # remove a known data anomaly

# Handle missing heights
sum(is.na(field_f_clean$fuhbheight))# 4 missing heights from field_f_clean
missing_heights <- ds %>% 
  select(ID, Height, Weight) %>% 
  filter(ID %in% field_f_clean$ID[is.na(field_f_clean$fuhbheight)]) %>% #  Identify records with missing height and from dataset ds, extract IDs, Height, and Weight only for those participants who have NA height in field_f_clean.
  distinct() %>%
  mutate(BMI = Weight/((Height/100)^2)) # Calculate BMI for these entries

# Manually fill in each missing value
for (i in missing_heights$ID) {
  field_f_clean$fuhbheight[field_f_clean$ID == i] <- missing_heights$Height[missing_heights$ID == i]
  field_f_clean$fuhbfweight[field_f_clean$ID == i] <- missing_heights$Weight[missing_heights$ID == i]
  field_f_clean$fuhbbmi[field_f_clean$ID == i] <- missing_heights$BMI[missing_heights$ID == i]
}

# Handle extreme values
field_f_clean <- field_f_clean %>%
  mutate(
    fuhbheight = ifelse(ID == 37532, old_full$Height[old_full$respnr == 37532]*100, fuhbheight), # Correct height for one participant (ID 37532)
    fuhbfweight = case_when(
      ID == 37532 ~ old_full$Weight[old_full$respnr == 37532], # Replaces weight from external data
      ID == 35258 ~ 168,  # Correct extreme BMI
      TRUE ~ fuhbfweight),
    fuhbbmi = fuhbfweight/((fuhbheight/100)^2) # Recalculate BMI using the updated values
  )




#### VGO2 Data Processing ####
old_full_clean <- old_full %>% 
  select(respnr, educationlevel, gender, age, Height, Weight, bmi, Ethnicity, Atopisch, 
         copd_gold, copd_lln, reversibility, smokeever, packyears, asever, pneumonia3y, 
         mindist_bvb12, mindist50geiten_bvb12, mindistpluimvee_bvb12, 
         mindistvarkens_bvb12, mindistpaarden_bvb12, mindistrundvee_bvb12,
         geiten_500_bvb12, geiten_1000_bvb12, geiten_5km_bvb12,
         pluimvee_500_bvb12, pluimvee_1000_bvb12,
         varkens_500_bvb12, varkens_1000_bvb12,
         rundvee_500_bvb12, rundvee_1000_bvb12) %>%
  rename(ID = respnr)

# this part processes intake (enrolment) dates and birthdates from the old_intake dataset
old_intake_clean <- old_intake %>%
  rename(T0_bdate = gebdatum) %>%
  mutate(T0_date = ymd(intakedatum)) %>%
  select(ID, T0_date, T0_bdate)

# This block cleans and processes baseline spirometry data
old_spiro_clean <- old_spiro %>% 
  select(respnr, lf_analist, test_date_pre, FEV1_pre, FVC_pre, fev1_fvc_pre, 
         PEF_pre, FEF25_75_pre, QualityGrade_pre,
         FEV1_pre_pred, FVC_pre_pred, FEV1FVC_pre_pred, FEF25_75_pre_pred) %>%
  rename(ID = respnr, Spirometrist = lf_analist, date = test_date_pre, 
         FEV1 = FEV1_pre, FVC = FVC_pre, FEV1_FVC = fev1_fvc_pre, 
         PEF = PEF_pre, FEF25_75 = FEF25_75_pre, Quality_grade = QualityGrade_pre) %>% 
  mutate(timepoint = "T0") %>%
  filter(!is.na(FEV1))

### AGE VERIFICATION ###
# Compare age between VGO2 and VGO3 
# Calculate expected age at VGO3 visit based on VGO2 data
age_check <- old_intake_clean %>%
  mutate(ID = as.character(ID)) %>%
  left_join(select(old_full_clean, ID, age) %>% mutate(ID = as.character(ID)), by = "ID") %>%
  left_join(select(field_q_clean, ID, fuhbdate, fuhbbdate) %>% mutate(ID = as.character(ID)), by = "ID") %>%
  mutate(
    age_T0 = age,
    age_T1 = interval(ymd(T0_bdate), ymd(fuhbdate)) / years(1),
    expected_age_T1 = age_T0 + interval(ymd(T0_date), ymd(fuhbdate)) / years(1),
    age_discrepancy = age_T1 - expected_age_T1,
    
    age_check_status = case_when(
      is.na(age_T1) ~ "Missing VGO3 data",
      abs(age_discrepancy) <= 1 ~ "Consistent (≤1 year)",
      abs(age_discrepancy) <= 2 ~ "Minor discrepancy (≤2 years)",
      TRUE ~ "Major discrepancy (>2 years)"
    )
  )

# Print discrepancy summary - for any participant with a major discrepancy (>2 years)
age_check %>% 
  count(age_check_status) %>% 
  mutate(percent = round(n/nrow(age_check)*100, 1)) %>% 
  print()

# Correct major discrepancies in birth dates
major_discrepancies <- age_check %>% 
  filter(age_check_status == "Major discrepancy (>2 years)")

if(nrow(major_discrepancies) > 0) {
  message("\nCorrecting birth dates for ", nrow(major_discrepancies), " participants:")
  
  for(id in major_discrepancies$ID) {
    original_dob <- major_discrepancies$fuhbbdate[major_discrepancies$ID == id]
    corrected_dob <- major_discrepancies$T0_bdate[major_discrepancies$ID == id]
    
    message("ID ", id, ": ", original_dob, " → ", corrected_dob)
    
    # Update in field_q_clean
    field_q_clean$fuhbbdate[field_q_clean$ID == id] <- corrected_dob
  }
}
# Replaces their VGO3 birthdate (fuhbbdate) with the correct baseline birthdate (T0_bdate)

# Now update the age in the VGO3 dataset
field_q_clean <- field_q_clean %>%
  mutate(
    fuhbbdate = ymd(fuhbbdate, quiet = TRUE),  # Parse birth date again after corrections
    age_VGO3 = interval(fuhbbdate, fuhbdate) / years(1)
  )

#### GENDER CHECK AND CORRECTION ####
# Convert IDs to character for safety
old_full_clean$ID <- as.character(old_full_clean$ID)
field_q_clean$ID <- as.character(field_q_clean$ID)
ds_prep$ID <- as.character(ds_prep$ID)

# Filter IDs present in both VGO2 and VGO3
common_IDs <- intersect(old_full_clean$ID, field_q_clean$ID)

# Prepare VGO2 gender data for common IDs
gender_vgo2 <- old_full_clean %>% 
  filter(ID %in% common_IDs) %>%
  select(ID, gender_vgo2 = gender)

# Prepare VGO3 gender data for common IDs
gender_vgo3 <- field_q_clean %>%
  filter(ID %in% common_IDs) %>%
  select(ID, gender_vgo3 = fuhbgender)

# Spirometry gender for common IDs
ds_gender <- ds %>%
  filter(ID %in% common_IDs) %>%
  select(ID, Gender) %>%
  mutate(Gender = factor(Gender, levels = c(0,1), labels = c("male", "female")))

# Combine for comparison
gender_vgo2 <- gender_vgo2 %>% mutate(ID = as.character(ID))
gender_vgo3 <- gender_vgo3 %>% mutate(ID = as.character(ID))
ds_gender  <- ds_gender  %>% mutate(ID = as.character(ID))

gender_compare <- gender_vgo2 %>%
  left_join(gender_vgo3, by = "ID") %>%
  left_join(ds_gender, by = "ID") %>%
  filter(!is.na(gender_vgo2) & !is.na(gender_vgo3) & gender_vgo2 != gender_vgo3) %>%
  mutate(match_spiro_vgo2 = (Gender == gender_vgo2))


# Correct VGO3 gender when spirometry gender matches VGO2
for (id in gender_compare$ID[gender_compare$match_spiro_vgo2]) {
  pos <- match(id, gender_vgo2$ID)
  if (!is.na(pos)) {
    field_q_clean$fuhbgender[field_q_clean$ID == id] <- gender_vgo2$gender_vgo2[pos]
    message(paste("Corrected VGO3 gender for ID:", id))
  }
}# VGO3 gender variable field_q_clean$fuhbgender is updated (corrected) to the VGO2 gender only for those participants where the spirometry gender agrees with VGO2, resolving likely errors in VGO3 gender.

# Check remaining mismatches after correction
remaining_mismatches <- gender_vgo2 %>%
  filter(ID %in% common_IDs) %>%
  left_join(field_q_clean %>% select(ID, fuhbgender), by = "ID") %>%
  filter(gender_vgo2 != fuhbgender)

message(paste("Remaining gender mismatches after correction:", nrow(remaining_mismatches))) # 19 mismatches


####### CREATE FINAL DATASETS #######
### 1. VGO3 Standalone Dataset ###
# Ensure ID is character in all datasets used in joins
ds_prep <- ds_prep %>% mutate(ID = as.character(ID))
field_q_clean <- field_q_clean %>% mutate(ID = as.character(ID))
OQ_clean <- OQ_clean %>% mutate(ID = as.character(ID))
field_f_clean <- field_f_clean %>% mutate(ID = as.character(ID))

ds_prep$date <- as.Date(ds_prep$date)

VGO3_metadata <- ds_prep %>%
  left_join(field_q_clean, by = "ID") %>%
  left_join(OQ_clean, by = "ID") %>%
  left_join(field_f_clean, by = "ID") %>%
  mutate(
    study = "VGO3",
    timepoint = "T1",
    date_spiro = date,
    birth_date = ymd(fuhbbdate, quiet = TRUE),
    gender = factor(fuhbgender, levels = c(0, 1), labels = c("male", "female")),
    age = age_VGO3,
    height = as.numeric(fuhbheight),
    weight = as.numeric(fuhbfweight),
    bmi = as.numeric(fuhbbmi),
  
    # Only mutate if raw codes (0/1) are present
    lung_medication = if (!is.factor(fuhblungmed)) factor(fuhblungmed, levels = c(0, 1), labels = c("no", "yes")) else fuhblungmed,
    pneumo_lastyr = if (!is.factor(fuhbpneumonialy)) factor(fuhbpneumonialy, levels = c(0, 1), labels = c("no", "yes")) else fuhbpneumonialy,
    test_difficulty = if (!is.factor(fuhbtestdiff)) factor(fuhbtestdiff, levels = c(0, 1), labels = c("no", "yes")) else fuhbtestdiff,
    wheeze_lastyr = if (!is.factor(fuwheezly)) factor(fuwheezly, levels = c(0, 1), labels = c("no", "yes")) else fuwheezly,
    covid_date = ymd(fuhbcovdate, quiet = TRUE),
    covid_last2mth = if (!is.factor(fuhbcovl2m)) factor(fuhbcovl2m, levels = c(0, 1), labels = c("no", "yes")) else fuhbcovl2m,
    asthma_ever = if (!is.factor(fuasever)) factor(fuasever, levels = c(0, 1), labels = c("no", "yes")) else fuasever,
    asthma_doc_diag = if (!is.factor(fuasdd)) factor(fuasdd, levels = c(0, 1), labels = c("no", "yes")) else fuasdd,
    daily_cough_3mth = if (!is.factor(ful3mdailycough)) factor(ful3mdailycough, levels = c(0, 1), labels = c("no", "yes")) else ful3mdailycough,
    daily_phlegm_3mth = if (!is.factor(ful3mdailyphlegm)) factor(ful3mdailyphlegm, levels = c(0, 1), labels = c("no", "yes")) else ful3mdailyphlegm,
    copd_doc_diag = if (!is.factor(fudoctoldcopd)) factor(fudoctoldcopd, levels = c(0, 1), labels = c("no", "yes")) else fudoctoldcopd,
    # Assume these are already cleaned up earlier, or preserve as-is
    smoke_status = fusmokeever,
    farm_visit = fufarmvst,
    farm_visit_multiple = fufarmvstmult
  ) %>%
  # Remove unused or duplicate variables
  select(
    -fuhbgender, -fuhbdate, -fuhbbdate, -age_VGO3, -fuhblungmed, -fuhbpneumonialy, -fuwheezly,
    -fuhbheight, -fuhbfweight, -fuhbbmi, -fuhbfheight1, -fuhbfheight2,
    -fugender, -fuage, -fuhbcovidinf, -fuhbcovposdate, -fuhbcovsusdate,
    -fuhbcovinterval, -fuquestdate,
    -fufarmvstfam, -fufarmvstjob, -fufarmvstoth, -fufarmvstpet, -fufarmvstno, -fufarmvstshop, -fuhbtestdiff, -fuhbcovdate, -fuhbcovl2m, -fuasever, -fuasdd, -ful3mdailycough, -ful3mdailyphlegm, -fudoctoldcopd, -fusmokeever, -fufarmvst,  -fufarmvstmult, -fuhbdate, -date
  ) %>% 
  ungroup()

names(VGO3_metadata)
str(VGO3_metadata)

# Check NAs for asthma and smokiung status
sum(is.na(VGO3_metadata$asthma_ever)) # 9 missing asthma status
sum(is.na(VGO3_metadata$smoke_status)) # 11 missing smoking status

# Apply corrections of asthma and smoking status from old_full where needed
# Ensure IDs are character for safe matching
old_full <- old_full %>% mutate(respnr = as.character(respnr))
VGO3_metadata <- VGO3_metadata %>% mutate(ID = as.character(ID))

# Apply corrections from old_full where needed
for (i in VGO3_metadata$ID[is.na(VGO3_metadata$asthma_ever)]) {
  val <- old_full$asever[old_full$respnr == i]
  VGO3_metadata$asthma_ever[VGO3_metadata$ID == i] <- factor(val, levels = c(0, 1), labels = c("no", "yes"))
}

for (i in VGO3_metadata$ID[is.na(VGO3_metadata$smoke_status)]) {
  val <- old_full$smokeever[old_full$respnr == i]
  VGO3_metadata$smoke_status[VGO3_metadata$ID == i] <- factor(val, levels = c(0, 1, 2), labels = c("non", "former", "current"))
}

sum(is.na(VGO3_metadata$asthma_ever)) # 0missing asthma status - not in VGO2 data
sum(is.na(VGO3_metadata$smoke_status)) # 0 missing smoking status

# Calculate percentage predicted values for VGO3 (already got the VGO2 % predicted)
# Turn measurement values presented as character strings to numeric
T1 <- VGO3_metadata %>% 
  select(ID, age, Ethnicity, gender, height, FEV1, FVC, FEV1_FVC, FEF25_75)

T1 <- T1 %>%
  mutate(
    gender_num = ifelse(gender == "male", 1, 2)
  )

T1 <- T1 %>%
  mutate(
    ethnic_code = case_when(
      Ethnicity == "Caucasian" ~ 1,
      Ethnicity == "African-American" ~ 2,
      Ethnicity == "NE Asian" ~ 3,
      Ethnicity == "SE Asian" ~ 4,
      TRUE ~ 5  # all others = Other/mixed
    )
  )

# Convert height from cm to m
T1 <- T1 %>%
  mutate(height_m = height / 100)

#FEV1
T1$FEV1_pred <- pctpred_GLI(
    age= T1$age,
    height= T1$height_m,
    gender = T1$gender_num,
    ethnicity = T1$ethnic_code,
    FEV1 = T1$FEV1
    )

# FVC
T1$FVC_pred <- pctpred_GLI(
    age = T1$age,
    height= T1$height_m,
    gender = T1$gender_num,
    ethnicity = T1$ethnic_code,
    FVC = T1$FVC
    )

# FEV1 FVC ratio
T1$FEV1FVC_pred <- pctpred_GLI(
    age = T1$age,
    height= T1$height_m,
    gender = T1$gender_num,
    ethnicity = T1$ethnic_code,
    FEV1FVC = T1$FEV1_FVC
    )

# FEF25_75 (Forced Expiratory Flow between 25% and 75% of FVC (lt/s))
T1$FEF25_75_pred <- pctpred_GLI(
    age = T1$age,
    height= T1$height_m,
    gender = T1$gender_num,
    ethnicity = T1$ethnic_code,
    FEF2575 = T1$FEF25_75
    ) 

str(T1)

T1 <- as.data.frame(T1)

T1_vars <- T1  %>% select(ID, FEV1_pred, FVC_pred, FEV1FVC_pred, FEF25_75_pred)

# Join predicted values to the VGO3_metadata
VGO3_metadata <- VGO3_metadata  %>% 
  left_join(T1_vars, by = "ID")


str(VGO3_metadata)

# Reorder dataframe
VGO3_metadata <- VGO3_metadata %>%
  relocate(ID, study, timepoint, date_spiro, birth_date, age, gender, height, weight, bmi, lung_medication, pneumo_lastyr, test_difficulty, wheeze_lastyr, covid_date, covid_last2mth, asthma_ever, asthma_doc_diag, daily_cough_3mth, daily_phlegm_3mth, copd_doc_diag, smoke_status, farm_visit, farm_visit_multiple)

names(VGO3_metadata)
str(VGO3_metadata)

# Assign variable labels
var_label(VGO3_metadata$ID) <- "VGO participant identifier"
var_label(VGO3_metadata$study) <- "VGO study (VGO2 or VGO3)"
var_label(VGO3_metadata$timepoint) <- "Sample collection timepoint (T0 or T1)"
var_label(VGO3_metadata$date_spiro) <- "Date of VGO3 spirometry"
var_label(VGO3_metadata$birth_date) <- "Date of birth"
var_label(VGO3_metadata$age) <- "Age at visit"
var_label(VGO3_metadata$gender) <- "Participant gender"
var_label(VGO3_metadata$height) <- "Height (cm)"
var_label(VGO3_metadata$bmi) <- "Body Mass Index"
var_label(VGO3_metadata$lung_medication) <- "Use of lung medication"
var_label(VGO3_metadata$pneumo_lastyr) <- "Pneumonia in the last year"
var_label(VGO3_metadata$test_difficulty) <- "Difficulty doing the spirometry test?"
var_label(VGO3_metadata$wheeze_lastyr) <- "Wheezing in the last year"
var_label(VGO3_metadata$covid_date) <- "Date of covid infection (confirmed or suspected)"
var_label(VGO3_metadata$covid_last2mth) <- "Covid infection in the last 2 months?"
var_label(VGO3_metadata$asthma_ever) <- "Ever had asthma"
var_label(VGO3_metadata$asthma_doc_diag) <- "Doctor diagnosed asthma"
var_label(VGO3_metadata$daily_cough_3mth) <- "Daily cough for >3 months"
var_label(VGO3_metadata$daily_phlegm_3mth) <- "Daily phlegm for >3 months"
var_label(VGO3_metadata$copd_doc_diag) <- "Doctor diagnosed COPD"
var_label(VGO3_metadata$smoke_status) <- "Smoking status (non, former, current)"
var_label(VGO3_metadata$farm_visit) <- "Visited a farm in the last year"
var_label(VGO3_metadata$farm_visit_multiple) <- "Multiple visits to a farm in the last year (none, single, multiple)"
# Label spirometry variables
var_label(VGO3_metadata$Spirometrist) <- "Spirometrist initials"
var_label(VGO3_metadata$Ethnicity) <- "Ethnicity for lung function test"
var_label(VGO3_metadata$lf_repr150) <- "Acceptable test, grade C or better"
var_label(VGO3_metadata$FEV1) <- "FEV1 (L)"
var_label(VGO3_metadata$FEV25) <- "FEV 0.25 sec (L) "
var_label(VGO3_metadata$FEV5) <- "FEV 0.5 sec (L)"
var_label(VGO3_metadata$FEV75) <- "FEV 0.75 sec (L)"
var_label(VGO3_metadata$FEV3) <- "FEV 3 sec (L)"
var_label(VGO3_metadata$FEV6) <- "FEV 6 sec (L)"
var_label(VGO3_metadata$FEV1_FVC) <- "FEV1/FVC (Tiffenau index)"
var_label(VGO3_metadata$FVC) <- "FVC (L)"
var_label(VGO3_metadata$FEF10) <-  "FEF at 10% of FVC (L/sec)"
var_label(VGO3_metadata$FEF25) <- "FEF at 25% of FVC (L/sec)"
var_label(VGO3_metadata$FEF25_75) <- "FEF average between 25% and 75% of FVC (L/sec)"
var_label(VGO3_metadata$FET25_75) <- "Forced expiratory time between 25% and 75% of FVC,  (sec)"
var_label(VGO3_metadata$FEF40) <- "FEF at 40% of FVC (L/sec)"
var_label(VGO3_metadata$FEF50) <- "FEF at 50% of FVC (L/sec)"
var_label(VGO3_metadata$FEF60) <- "FEF at 60% of FVC (L/sec)"
var_label(VGO3_metadata$FEF75) <- "FEF at 70% of FVC (L/sec)"
var_label(VGO3_metadata$FEF75_85) <- "FEF average between 75% and 85% of FVC,  (L/sec)"
var_label(VGO3_metadata$FEF80) <- "FEF at 80% of FVC (L/sec)"
var_label(VGO3_metadata$PEF) <- "Peak expiratory flow (L/sec)"
var_label(VGO3_metadata$PEF_Lmin) <- "Peak expiratory flow (L/min)"
var_label(VGO3_metadata$FEV1_pred) <- "Percent predicted FEV1 (25-75% midflow)"
var_label(VGO3_metadata$FVC_pred) <- "Percent predicted FVC (25-75% midflow)"
var_label(VGO3_metadata$FEV1FVC_pred) <- "Percent predicted FEV1/FVC ratio (25-75% midflow)"
var_label(VGO3_metadata$FEF25_75_pred) <- "Percent predicted FEF (25-75% midflow)"

write.csv(VGO3_metadata,"../Output/Participant_data/Datasets/VGO3_data.csv" )



### 2. VGO2 Standalone Dataset ###
old_full_clean <- old_full_clean %>% mutate(ID = as.character(ID))
old_intake_clean <- old_intake_clean %>% mutate(ID = as.character(ID))
old_spiro_clean <- old_spiro_clean %>% mutate(ID = as.character(ID))

old_spiro_clean$date

VGO2_metadata <- old_full_clean %>%
  left_join(old_intake_clean, by = "ID") %>%
  left_join(old_spiro_clean, by = "ID") %>%
  mutate(
    study = "VGO2",
    timepoint = "T0",
    date_spiro = T0_date,
    birth_date = T0_bdate,
    gender = case_when(
    gender == 0 ~ "male",
    gender == 1 ~ "female",
    TRUE ~ NA_character_
  ),
  gender = factor(gender, levels = c("male", "female")),    
  educationlevel = factor(educationlevel, levels = c(0, 1, 2), 
                           labels = c("low", "medium", "high")),
  smoke_status = factor(smokeever, levels = c(0, 1, 2), labels = c("non", "former", "current")),
  asthma_ever = factor(asever, levels = c(0, 1), labels = c("no", "yes")),
  copd_gold = factor(copd_gold, levels = c(0, 1), labels = c("no", "yes"))
  ) %>%
  rename(
    height = Height,
    weight = Weight,
    ethnicity = Ethnicity,
    atopic = Atopisch,
    pneumo_last3yr = pneumonia3y,
    mindist_farm = mindist_bvb12,
    mindist_goat_farm = mindist50geiten_bvb12,
    mindist_poultry_farm = mindistpluimvee_bvb12,
    mindist_pig_farm = mindistvarkens_bvb12,
    mindist_horse_farm = mindistpaarden_bvb12,
    mindist_cattle_farm = mindistrundvee_bvb12,
    ngoats_500m = geiten_500_bvb12,
    ngoats_1000m = geiten_1000_bvb12,
    ngoats_5000m = geiten_5km_bvb12,
    npoultry_500m = pluimvee_500_bvb12,
    npoultry_1000m = pluimvee_1000_bvb12,
    npigs_500m = varkens_500_bvb12,
    npigs_1000m = varkens_1000_bvb12,
    ncattle_500m = rundvee_500_bvb12,
    ncattle_1000m = rundvee_1000_bvb12,
    spirometrist = Spirometrist,
    quality_grade = Quality_grade,
    date_and_time_lf = date,
    FEV1_pred = FEV1_pre_pred,
    FVC_pred = FVC_pre_pred,
    FEV1FVC_pred = FEV1FVC_pre_pred,
    FEF25_75_pred = FEF25_75_pre_pred
  ) %>%
  # Remove unused or duplicate variables
  select(
    -smokeever, -asever, -T0_date, -T0_bdate
  )


VGO2_metadata <- VGO2_metadata %>%
  relocate(ID, study, timepoint, date_spiro, birth_date, age, gender, height, weight, bmi, ethnicity, asthma_ever, smoke_status, smoke_status, packyears, atopic, copd_gold, copd_lln, reversibility)

# Assign variable labels
var_label(VGO2_metadata$ID) <- "VGO participant identifier"
var_label(VGO2_metadata$educationlevel) <- "Education level (0=low, 1=med, 2=high)"
var_label(VGO2_metadata$gender) <- "Participant gender"
var_label(VGO2_metadata$study) <- "VGO study (VGO2 or VGO3)"
var_label(VGO2_metadata$age) <- "Age at visit"
var_label(VGO2_metadata$height) <- "Height (cm)"
var_label(VGO2_metadata$weight) <- "Weight (kg)"
var_label(VGO2_metadata$bmi) <- "Body Mass Index"
var_label(VGO2_metadata$ethnicity) <- "Ethnicity"
var_label(VGO2_metadata$atopic) <- "Atopic history"
var_label(VGO2_metadata$copd_gold) <- "COPD defined by GOLD (doctor diagnosed)"
var_label(VGO2_metadata$copd_lln) <- "COPD defined by LLN"
var_label(VGO2_metadata$reversibility) <- "Bronchodilator reversibility"
var_label(VGO2_metadata$smoke_status) <- "Smoking status (non, former, current)"
var_label(VGO2_metadata$packyears) <- "Smoking pack years"
var_label(VGO2_metadata$timepoint) <- "Sample collection timepoint (T0 or T1)"
var_label(VGO2_metadata$birth_date) <- "Date of birth"
var_label(VGO2_metadata$pneumo_last3yr) <- "Pneumonia in the last 3 years"
var_label(VGO2_metadata$mindist_farm) <- "Distance to nearest livestock farm"
var_label(VGO2_metadata$mindist_goat_farm) <- "Distance to nearest goat farm"
var_label(VGO2_metadata$mindist_poultry_farm) <- "Distance to nearest poultry farm"
var_label(VGO2_metadata$mindist_pig_farm) <- "Distance to nearest pig farm"
var_label(VGO2_metadata$mindist_horse_farm) <- "Distance to nearest horse farm"
var_label(VGO2_metadata$mindist_cattle_farm) <- "Distance to nearest cattle farm"
var_label(VGO2_metadata$ngoats_500m) <- "Number of goats within 500m"
var_label(VGO2_metadata$ngoats_1000m) <- "Number of goats within 1000m"
var_label(VGO2_metadata$ngoats_5000m) <- "Number of goats within 5km"
var_label(VGO2_metadata$npoultry_500m) <- "Number of poultry within 500m"
var_label(VGO2_metadata$npoultry_1000m) <- "Number of poultry within 1000m"
var_label(VGO2_metadata$npigs_500m) <- "Number of pigs within 500m"
var_label(VGO2_metadata$npigs_1000m) <- "Number of pigs within 1000m"
var_label(VGO2_metadata$ncattle_500m) <- "Number of cattle within 500m"
var_label(VGO2_metadata$ncattle_1000m) <- "Number of cattle within 1000m"
var_label(VGO2_metadata$FEV1) <- "Forced Expiratory Volume in 1s (L)"
var_label(VGO2_metadata$FVC) <- "Forced Vital Capacity (L)"
var_label(VGO2_metadata$FEV1_FVC) <- "FEV1/FVC ratio"
var_label(VGO2_metadata$FEF25_75) <- "Forced Expiratory Flow (25–75%)"
var_label(VGO2_metadata$quality_grade) <- "Quality grade of spirometry"
var_label(VGO2_metadata$spirometrist) <- "Name of lung function technician"
var_label(VGO2_metadata$date_spiro) <- "Date of spirometry test"
var_label(VGO2_metadata$FEV1_pred) <- "Percent predicted FEV1"
var_label(VGO2_metadata$FVC_pred) <- "Percent predicted FVC"
var_label(VGO2_metadata$FEV1FVC_pred) <- "Percent predicted FEV1/FVC ratio"
var_label(VGO2_metadata$FEF25_75_pred) <- "Percent predicted FEF (25-75% midflow)"


write.csv(VGO2_metadata, "../Output/Participant_data/Datasets/VGO2_data.csv") 
``` 

# VGO2 and VGO3 combined - longitudinal
```{r}
# Prepare for binding by aligning variable names
# Make sure ID is character
VGO2_metadata <- VGO2_metadata %>% mutate(ID = as.character(ID))
VGO3_metadata <- VGO3_metadata %>% mutate(ID = as.character(ID))

# Check which variables exist in one but not the other
vgo2_only <- setdiff(names(VGO2_metadata), names(VGO3_metadata))
vgo3_only <- setdiff(names(VGO3_metadata), names(VGO2_metadata))

# Harmonize column names if same meaning but different names
# Example: gender is already aligned; if not, rename appropriately
# We'll manually align only relevant fields

# Add missing variables to VGO2 with NA to match VGO3 structure
vars_missing_from_vgo2 <- vgo3_only[!vgo3_only %in% names(VGO2_metadata)]
VGO2_metadata[vars_missing_from_vgo2] <- NA

# Add missing variables to VGO3 with NA to match VGO2 structure
vars_missing_from_vgo3 <- vgo2_only[!vgo2_only %in% names(VGO3_metadata)]
VGO3_metadata[vars_missing_from_vgo3] <- NA

# Make sure the columns are in the same order
VGO2_metadata <- VGO2_metadata %>% select(names(VGO3_metadata))
VGO3_metadata <- VGO3_metadata %>% select(names(VGO2_metadata))

# Combine the datasets
VGO_combined <- bind_rows(VGO2_metadata, VGO3_metadata) %>%
  relocate(ID, study, timepoint, date_spiro, birth_date, age, gender)


# Compute follow up time (fuperiod) and delta lung function values 
# Ensure dates are Date class
VGO_combined <- VGO_combined %>%
  mutate(date_spiro = as.Date(date_spiro))

# Keep only participants with both timepoints
ids_with_both_timepoints <- VGO_combined %>%
  group_by(ID) %>%
  filter(n_distinct(timepoint) == 2) %>%
  pull(ID) %>%
  unique()

VGO_combined_both <- VGO_combined %>%
  filter(ID %in% ids_with_both_timepoints)

# Pivot wider to separate VGO2 and VGO3 values
VGO_wide <- VGO_combined_both %>%
  pivot_wider(
    id_cols = ID,
    names_from = study,
    values_from = c(date_spiro, FEV1, FVC, PEF, FEF25_75, FEV1_FVC)
  )



# Compute follow-up time and lung function changes
VGO_wide <- VGO_wide %>%
  mutate(
    fuperiod = as.numeric(difftime(date_spiro_VGO3, date_spiro_VGO2, units = "days")) / 365.25,

    delta_FEV1 = (FEV1_VGO3 - FEV1_VGO2) * 1000 / fuperiod,
    delta_FVC = (FVC_VGO3 - FVC_VGO2) * 1000 / fuperiod,
    delta_PEF = (PEF_VGO3 - PEF_VGO2) * 1000 / fuperiod,
    delta_FEF25_75 = (FEF25_75_VGO3 - FEF25_75_VGO2) * 1000 / fuperiod,

    FEV1_FVC_perc_VGO2 = FEV1_FVC_VGO2 * 100,
    FEV1_FVC_perc_VGO3 = FEV1_FVC_VGO3 * 100,
    delta_tiff = (FEV1_FVC_perc_VGO3 - FEV1_FVC_perc_VGO2) / fuperiod,

    obstr_VGO2 = ifelse(FEV1_FVC_VGO2 < 0.7, 1, 0),
    obstr_VGO3 = ifelse(FEV1_FVC_VGO3 < 0.7, 1, 0),
    
    new_obstr = case_when(
      obstr_VGO2 == 0 & obstr_VGO3 == 1 ~ 1,
      obstr_VGO2 == 0 & obstr_VGO3 == 0 ~ 0,
      obstr_VGO2 == 1 ~ 2 # classify changes in airway obstruction status between the two timepoints (VGO2 and VGO3). 1= new obstruction, 0 = remained non-obstructed, 2 = obstructed already at baseline (VGO2)
    )
  )

# Check final structure to ensure no nesting
str(VGO_wide)



# Merge wide-form variables back into long-form
VGO_combined_final <- VGO_combined_both %>%
  left_join(
    VGO_wide %>%
      select(ID, fuperiod, starts_with("delta_"), starts_with("obstr_"), new_obstr),
    by = "ID"
  )

# Check final structure
str(VGO_combined_final)


# Assign variable labels to VGO_combined
var_label(VGO_combined_final$ID) <- "VGO participant identifier"
var_label(VGO_combined_final$study) <- "VGO study (VGO2 or VGO3)"
var_label(VGO_combined_final$timepoint) <- "Sample collection timepoint (T0 or T1)"
var_label(VGO_combined_final$date_spiro) <- "Date of spirometry measurements"
var_label(VGO_combined_final$birth_date) <- "Date of birth"
var_label(VGO_combined_final$age) <- "Age at visit"
var_label(VGO_combined_final$gender) <- "Participant gender"
var_label(VGO_combined_final$height) <- "Height (cm)"
var_label(VGO_combined_final$weight) <- "Weight (kg)"
var_label(VGO_combined_final$bmi) <- "Body Mass Index"
var_label(VGO_combined_final$educationlevel) <- "Education level (low, medium, high)"
var_label(VGO_combined_final$ethnicity) <- "Ethnicity"
var_label(VGO_combined_final$atopic) <- "Atopic history"
var_label(VGO_combined_final$smoke_status) <- "Smoking status (non, former, current)"
var_label(VGO_combined_final$packyears) <- "Smoking pack years"
var_label(VGO_combined_final$asthma_ever) <- "Ever had asthma"
var_label(VGO_combined_final$asthma_doc_diag) <- "Doctor diagnosed asthma"
var_label(VGO_combined_final$copd_doc_diag) <- "Doctor diagnosed COPD"
var_label(VGO_combined_final$copd_lln) <- "COPD defined by LLN"
var_label(VGO_combined_final$reversibility) <- "Bronchodilator reversibility"
var_label(VGO_combined_final$pneumo_last3yr) <- "Pneumonia in the last 3 years"
var_label(VGO_combined_final$pneumo_lastyr) <- "Pneumonia in the last year"
var_label(VGO_combined_final$daily_cough_3mth) <- "Daily cough for >3 months"
var_label(VGO_combined_final$daily_phlegm_3mth) <- "Daily phlegm for >3 months"
var_label(VGO_combined_final$wheeze_lastyr) <- "Wheezing in the last year"
var_label(VGO_combined_final$lung_medication) <- "Use of lung medication"
var_label(VGO_combined_final$test_difficulty) <- "Difficulty doing the spirometry test"
var_label(VGO_combined_final$covid_date) <- "Date of covid infection"
var_label(VGO_combined_final$covid_last2mth) <- "Covid infection in the last 2 months"
var_label(VGO_combined_final$mindist_farm) <- "Distance to nearest livestock farm"
var_label(VGO_combined_final$mindist_goat_farm) <- "Distance to nearest goat farm"
var_label(VGO_combined_final$mindist_poultry_farm) <- "Distance to nearest poultry farm"
var_label(VGO_combined_final$mindist_pig_farm) <- "Distance to nearest pig farm"
var_label(VGO_combined_final$mindist_horse_farm) <- "Distance to nearest horse farm"
var_label(VGO_combined_final$mindist_cattle_farm) <- "Distance to nearest cattle farm"
var_label(VGO_combined_final$ngoats_500m) <- "Number of goats within 500m"
var_label(VGO_combined_final$ngoats_1000m) <- "Number of goats within 1000m"
var_label(VGO_combined_final$ngoats_5000m) <- "Number of goats within 5km"
var_label(VGO_combined_final$npoultry_500m) <- "Number of poultry within 500m"
var_label(VGO_combined_final$npoultry_1000m) <- "Number of poultry within 1000m"
var_label(VGO_combined_final$npigs_500m) <- "Number of pigs within 500m"
var_label(VGO_combined_final$npigs_1000m) <- "Number of pigs within 1000m"
var_label(VGO_combined_final$ncattle_500m) <- "Number of cattle within 500m"
var_label(VGO_combined_final$ncattle_1000m) <- "Number of cattle within 1000m"
var_label(VGO_combined_final$spirometrist) <- "Name of lung function technician"
var_label(VGO_combined_final$quality_grade) <- "Quality grade of spirometry"
var_label(VGO_combined_final$date_and_time_lf) <- "Date and time of lung function measurements"
var_label(VGO_combined_final$FEV1) <- "Forced Expiratory Volume in 1s (L)"
var_label(VGO_combined_final$FVC) <- "Forced Vital Capacity (L)"
var_label(VGO_combined_final$FEV1_FVC) <- "FEV1/FVC ratio"
var_label(VGO_combined_final$FEF25_75) <- "Forced Expiratory Flow (25–75%)"
var_label(VGO_combined_final$FEV1_pred) <- "Percent predicted FEV1"
var_label(VGO_combined_final$FVC_pred) <- "Percent predicted FVC"
var_label(VGO_combined_final$FEV1FVC_pred) <- "Percent predicted FEF (25-75% midflow)"
var_label(VGO_combined_final$farm_visit) <- "Visited a farm in the last year"
var_label(VGO_combined_final$farm_visit_multiple) <- "Multiple visits to a farm in the last year (none, single, multiple)"
var_label(VGO_combined_final$fuperiod) <- "Follow-up period in years"
var_label(VGO_combined_final$delta_FEV1) <- "Annual change in FEV1 (mL/year)"
var_label(VGO_combined_final$delta_FVC) <- "Annual change in FVC (mL/year)"
var_label(VGO_combined_final$delta_PEF) <- "Annual change in PEF (mL/year)"
var_label(VGO_combined_final$delta_FEF25_75) <- "Annual change in FEF25-75 (mL/year)"
var_label(VGO_combined_final$delta_tiff) <- "Annual change in FEV1/FVC percentage"
var_label(VGO_combined_final$obstr_VGO2) <- "Obstruction (FEV1/FVC < 0.7) at VGO2"
var_label(VGO_combined_final$obstr_VGO3) <- "Obstruction (FEV1/FVC < 0.7) at VGO3"
var_label(VGO_combined_final$new_obstr) <- "Obstruction trajectory: 0=never, 1=incident, 2=persistent"

# OPTIONAL: Save to disk
write.csv(VGO_combined_final, "../Output/Participant_data/Datasets/VGO_combined.csv", row.names = FALSE)




# Checking FVC pred values
# Create a filtered data frame (without altering the original)
vgo3_table <- data.frame(
  ID = VGO_combined_final$ID[VGO_combined_final$study == "VGO3"],
  FVC_pred = VGO_combined_final$FVC_pred[VGO_combined_final$study == "VGO3"]
)
# Sort by ID
vgo3_table <- vgo3_table[order(vgo3_table$ID), ]
# Show first rows
head(vgo3_table)
# Open full table in RStudio viewer
View(vgo3_table)


vgo3_table <- data.frame(
  ID = VGO_combined_final$ID[VGO_combined_final$study == "VGO3"],
  FEV1_pred = VGO_combined_final$FEV1_pred[VGO_combined_final$study == "VGO3"]
)
# Sort by ID
vgo3_table <- vgo3_table[order(vgo3_table$ID), ]
# Show first rows
head(vgo3_table)
# Open full table in RStudio viewer
View(vgo3_table)


summary(VGO_combined_final$FVC_pred)
str(VGO_combined_final$FVC_pred)
str(VGO_combined_final$FEV1_pred)

```

