---
title: "Cleaning Spirometry data from VGO2 and VGO3 studies"
author: "Warner van Kersen (adapted by Beatrice Cornu Hewitt)"
date: "`r Sys.time()`"
---
# Load packages
```{r}
# Check R version 
R.version.string # R version 4.4.2 (2024-10-31 ucrt)

# Install pacman package
if (!require("pacman")) install.packages("pacman")

# Load all packages
pacman::p_load(
  dplyr, labelled, gt, openxlsx, 
  haven, lubridate, table1, rspiro, stringr, tidyr, readr, stringi
)
```

# Load datasets from VGO2 and VGO3
All files refer to absolute file paths in the UU O-drive. All are also saved in "Input/" folder relative to this R notebook
```{r}
#### VGO3 data ####
# Merged spirometry
ds <- read.csv2("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO_3_cleaned_spirometry.csv", row.names = NULL) %>% select(-any_of("X"))

# Spirometry grading file, cleaned
grade <- read.xlsx("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_spirometrygrades_clean_20220913.xlsx")
  
# Online questionnaire
OQ <- read.xlsx("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO3_CP_OnlineQuest_joined_not_cleaned_20221017.xlsx")
  
# Field questionnaire
field_q <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_CP_fieldquest_20220909.csv")
  
# Field form
field_f <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/VGO-3_CP_fieldform_20220909.csv")

#### VGO2 data ####
# Full dataset
old <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Overdracht VGO gezondheid/Sas dataset/lung_sero_scr_longq_emr_09052017.sas7bdat")

# Intake file for test dates - in full dataset 12 dates (test_date_pre) are missing
old_intake <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Data intake + medicijngebruik/vgointake.sas7bdat")

# Spirometry
old_spiro <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Longfuncties/Dataset analyse/vgo_lf_12feb2016_n2439.sas7bdat")

# Livestock emitted PM10 and endotoxin fraction modelled by Myrna
old_endo <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO-COPD/Microbiome/Warner_R_projects/dada2 analyse COPD microbioom/00_raw_data/VGO_2494_EUpred_WvK.csv") %>% select(-X) 
    
old_endo <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO-COPD/Microbiome/Warner_R_projects/dada2 analyse COPD microbioom/00_raw_data/VGO_2494_PM10CONCpred_WvK.csv") %>% select(-X) %>% left_join(old_endo,.,by="respnr")
    
# Old complete (for continuous exposure proxies)
old_full <- read_sas("O:/DGK/IRAS/EEPI/Projects/VGO/VGO I en II/VGO gezondheid/Overdracht VGO gezondheid/Sas dataset/datasetmyrna18072017.sas7bdat")

# Farm child variables
# farmchild,farmchtotyr,farmchanim,farmchcrop,farmchdairy,farmchveal,farmchgoat,farmchsheep,farmchpig,farmchpoultry,farmchhorse,farmchother,farmchothertxt,farmchholiday,farmnone,farmchanim,farmchmanure,farmchstraw,farmchcropcare
    
# 2021 and 2022 air pollutant data from national air quality monitoring network
AP_monitor <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO3student/VGO_Yuhong/Exposure data/Weekly ambient PM10/VGO3 CP YS Weekly ambient PM10 one week before.csv") %>% 
      dplyr::select(respnr,weeklyap,season) %>% 
      dplyr::rename(ID=respnr,wkprPM10_ugm3=weeklyap)
    
AP_monitor$season <- relevel(as.factor(AP_monitor$season),ref = "Winter")
   
# 2021 and 2022 daily ammonia data
# Distance weighted livestock proxies, not used in final analysis
WghtDist <- read.csv("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Livestock_computations/VGO1/VGO1_2494_WghtDist.csv") %>% dplyr::select(-X,-nfuranimsWghtDist.1000m.sum,-nfuranimsWghtDist.3000m.sum,-nFuranimFarmWghtDist.1000m.sum,-nFuranimFarmWghtDist.3000m.sum) %>% rename(ID=respnr)
   
# Week prior NH3 exposures and difference
NH_diff <- readRDS("O:/DGK/IRAS/EEPI/Projects/VGO/VGO3_2tm3_4__gezondheid/Data/controle_populatie/Data_analyse/VGO3_CP_R_project/Cleaned data/wk_prior_NH3_diff") # Week prior NH3 levels in VGO2 and VGO3 (then also the difference between these)
  
```
# VGO3 spirometry data cleaning
### Prep cleaned spirometry data for analysis
```{r}
#### VGO3 spirometry ####  
# We see that for the FVC values there is an issue in the later samples that the decimal point it missing, 
# How many FVC values are "digits only" (no decimal) right now?
n_bad_before <- sum(!grepl("\\.", ds$FVC) & grepl("^\\d{2,}$", ds$FVC, perl = TRUE), na.rm = TRUE)
message("FVC values with missing decimal BEFORE fix: ", n_bad_before)
# Keep a backup column for audit
ds <- ds %>% mutate(FVC_raw = FVC)
# Normalise decimal commas to dots (if any)
ds <- ds %>% mutate(FVC = gsub(",", ".", FVC, fixed = TRUE))
# Insert a decimal after the first digit where it's clearly missing
needs_dot <- !grepl("\\.", ds$FVC) & grepl("^\\d{2,}$", ds$FVC, perl = TRUE)
ds$FVC[needs_dot] <- sub("^(\\d)(\\d+)$", "\\1.\\2", ds$FVC[needs_dot], perl = TRUE)
# Parse to numeric (drops stray non-numeric chars if any)
ds$FVC <- readr::parse_number(ds$FVC, locale = readr::locale(decimal_mark = "."))
# Quick sanity checks + small audit table
n_bad_after <- sum(!is.na(ds$FVC) & ds$FVC > 20)  # >20 L is implausible
message("FVC values still implausible AFTER fix (>20 L): ", n_bad_after)
# List rows that were changed
fixed_rows <- which(needs_dot)
audit_FVC <- ds[fixed_rows, c("ID", "FVC_raw", "FVC")]
print(audit_FVC)  # show the fixes made



# Check for duplicate records and remove any 
ds1 <- ds %>% distinct(ID, DateOfBirth, Test.Date, FEV1, FVC, .keep_all = T)
# Turn measurement values presented as character strings to numeric
str(ds1)

ds2 <- ds1 %>%
  mutate(across(where(is.character) & !any_of(c("Gender", "Ethnicity", "DateOfBirth", "Test.Date")), as.numeric))
str(ds2)

# Merge measurement values with spiro grades
sum(is.na(grade$FEV1_quality)) # no grades missing
sum(is.na(grade$FVC_quality)) # no grades missing

# calculate n accepted trials
accept <- c("A","B","C") # define accepted range
accept_D <- c("A","B","C","D") # define accepted range + D
naccept <- c("D","E","F","U") # define not accepted range
# A = 3 acceptable trials within 100ml
# A = 3 acceptable trials within 150ml
# C = 2 acceptable trials
# D = 1 acceptable trial  - probably false low

AB <- c("A","B")
EFU <- c("E","F","U")

n_accept_tests <-   sum(grade$FEV1_quality %in% accept & grade$FVC_quality %in% accept) # 845
n_accept_tests_D <-   sum(grade$FEV1_quality %in% accept_D & grade$FVC_quality %in% accept_D) # 861

# create dataset with one value per measurement - select highest FEV1 and FVC from reproducible measurements
ds3 <- grade %>% select(ID,Spirometrist,FEV1_quality,FVC_quality) %>%
    left_join(ds2,.,by="ID") %>%
    mutate(FEV1_accepted=case_when(!FEV1_quality %in% accept~0,
                                    FEV1_quality %in% accept~1),
           FVC_accepted=case_when(!FVC_quality %in% accept~0,
                                    FVC_quality %in% accept~1),
           test_accepted=case_when(!(FEV1_accepted==1&FVC_accepted==1)~0,
                                    FEV1_accepted==1&FVC_accepted==1~1),
           ntrial_acc=case_when(FEV1_quality %in% AB & FVC_quality %in% AB~3,
                                FEV1_quality %in% AB & FVC_quality == "C" ~2,
                                FEV1_quality %in% AB & FVC_quality == "D" ~1,
                                FEV1_quality == "C" & FVC_quality %in% accept~2,
                                FEV1_quality == "D" & FVC_quality %in% accept~1,
                                FEV1_quality == "D" & FVC_quality == "D"~1,
                                FEV1_quality == "D" & FVC_quality == "E"~0,
                                FEV1_quality %in% accept & FVC_quality %in% EFU~0,
                                FEV1_quality %in% EFU~0)) %>%
    group_by(ID,ntrial_acc) %>% 
    mutate(trial_acc=case_when(ntrial_acc==0~0,
                               ntrial_acc==1 & Trial_rank==1~1,
                               ntrial_acc==1 & Trial_rank!=1~0,
                               ntrial_acc==2 & Trial_rank %in% c(1,2)~1,
                               ntrial_acc==2 & !Trial_rank %in% c(1,2)~0,
                               ntrial_acc==3 & Trial_rank %in% c(1,2,3)~1,
                               ntrial_acc==3 & !Trial_rank %in% c(1,2,3)~0
                               )
    )
    

sum(ds3$trial_acc) # 2520
  
table(ds3$FEV1_quality,ds3$FVC_quality) # How often each FEV1 quality level appears with each FVC quality level
  #     A    B    C    D    E    F    U
  # A 4292  102  117   39   89    0   37
  # B  196   44   25   13   20    0    5
  # C   41    0   20    0   13    0    0
  # D   22    8   15   11   13    0    0
  # E   59   23   28    0  110    0   17
  # F   17    0    0    0    0   42    0
  # U    6    0    0    0    8    0   30
table(ds3$FEV1_quality,ds3$ntrial_acc,useNA = "ifany") # how each FEV1 quality grade maps to the numeric ntrial_acc classification
  #    0    1    2    3
  # A  126   39  117 4394
  # B   25   13   25  240
  # C   13    0   61    0
  # D   13   56    0    0
  # E  237    0    0    0
  # F   59    0    0    0
  # U   44    0    0    0

# Create a summary table showing the number of participants whose first spirometry trial (Trial_rank == 1) falls into each combination of:FEV1 grade, FVC grade, Number of acceptable trials (ntrial_acc)
ds3$Trial_rank

# Create summary table using gt
ds3 %>%
  filter(Trial_rank == 1) %>%
  group_by(FEV1_quality, FVC_quality, ntrial_acc) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(FEV1_quality, FVC_quality) %>%
  gt() %>%
  cols_label(
    FEV1_quality = "FEV₁ Grade",
    FVC_quality = "FVC Grade",
    ntrial_acc = "Number of Acceptable Attempts",
    n = "Number of Participants"
  ) %>%
  tab_header(
    title = "First Trial Spirometry Quality Summary"
  )


# Calculate number of acceptable tests
accept_tests <- ds3 %>% group_by(ID) %>% filter(Trial_rank==1&test_accepted==1) %>%
summarise(Acceptable_tests=n())
  
# check if n accepted tests matches expected
sum(accept_tests$Acceptable_tests) == n_accept_tests # TRUE 

# Calculate number of acceptable trials (attempts within tests)
total_ntrial_acc <- ds3 %>% group_by(ID) %>% select(ID,ntrial_acc) %>% distinct() %>% ungroup() %>%
    summarise(total=across(ntrial_acc,sum)) %>% unlist() %>% as.integer()
  
# should match sum of trial_acc
total_ntrial_acc == sum(ds3$trial_acc)
  
# Check BEV criteria in all acceptable trials
ds_bevcheck <- ds3 %>% ungroup() %>%
    filter(trial_acc==1) %>% 
    mutate(fvc5perc=0.05*FVC,
           bevcheck=case_when(fvc5perc>0.150~BEV-fvc5perc,
                              fvc5perc<=0.150~BEV-0.150)) %>%
    filter(bevcheck>0) %>%
    select(ID,BEV,bevcheck,fvc5perc,FVC_quality,FEV1_quality)# This filters violations (bevcheck > 0) — meaning the BEV is too large, so the trial might be invalid.
  
# Select the 3 best trials for each participant, when selecting highest FEV1 and FVC only use accepted trials
ds_minmax <- ds3 %>% group_by(ID) %>%
    filter(Trial_rank %in% c(1,2,3)&trial_acc==1) %>%
    # select(ID,Trial_rank,FEV1,FVC,PEF,ntrial_acc,FEV1_quality,FVC_quality) %>% (line used for checking, printing less columns)
    arrange(ID,Trial_rank) %>%
    mutate(fev1min=min(FEV1), # Calc min max med FEV1
           fev1med=case_when(ntrial_acc==3~median(FEV1)),
           fev1max=max(FEV1),
           fev1dif=case_when(ntrial_acc==3~fev1max-fev1med,
                             ntrial_acc==2~fev1max-fev1min),
           fev1_d150=case_when(fev1dif<=0.150~0,
                               fev1dif>0.150~1,
                               ntrial_acc==1~1),
           fvcmin=min(FVC), # calc min max med FVC
           fvcmed=case_when(ntrial_acc==3~median(FVC)),
           fvcmax=max(FVC),
           fvcdif=case_when(ntrial_acc==3~fvcmax-fvcmed,
                            ntrial_acc==2~fvcmax-fvcmin),
           fvc_d150=case_when(fvcdif<=0.150~0,
                              fvcdif>0.150~1,
                              ntrial_acc==1~1),
           lf_repr150=case_when(fvc_d150 == 1 | fev1_d150 == 1~0, # is the test reproducible based on both FEV1 & FVC?
                                fvc_d150 == 0 & fev1_d150 == 0~1),
           pefmin=min(PEF), # calc min max med PEF
           pefmed=case_when(ntrial_acc==3~median(PEF)),
           pefmax=max(PEF),
           pefdiff=case_when(ntrial_acc>2&lf_repr150==1~pefmax-pefmed),
           pef_d670=case_when(pefdiff<=0.67~0, # checks PEF reproducibility (>0.67 L/s difference from median)
                              pefdiff>0.67~1)
           )
           
    
ds_minmax$pef_d670[is.na(ds_minmax$pefdiff)]

# Excluding all participants with <1 (so 2 or more) accepted trials we should have 845 in the dataset
length(unique(ds_minmax$ID[ds_minmax$ntrial_acc>1]))  # 845

# Check fraction reproducible vs. non-reproducible in grades A and B (>2 reproducible trials)
table(ds_minmax$fev1_d150[ds_minmax$ntrial_acc>2])  
# 0    1 
# 2409   33 
# So, among those with 3 acceptable trials, almost all have reproducible FEV1.

``` 

## Spirometry - quality grades
```{r}
sumfunc <- function(z) if (is.numeric(z)) sum(z) else ''

FEV1_grade <- grade %>% rename(FEV1=FEV1_quality) %>% group_by(FEV1) %>% summarise(n=n()) # rename the FEV1_quality to FEV1, groups th eFEV1 grade (A,B,C...), counts the number per grade
# FEV1 n
# A	818			
# B	50			
# C	11			
# D	10			
# E	34			
# F	9			
# U	6

FVC_grade <-  grade %>% rename(FVC=FVC_quality) %>% group_by(FVC) %>% summarise(n=n()) 
# FVC n
# A	811			
# B	27			
# C	31			
# D	10			
# E	38			
# F	7			
# U	14

# FEV1
FEV1_grade_sum <- as.data.frame(lapply(FEV1_grade, sumfunc))
FEV1_grade_sum[1] <- "total"
FEV1_grade <- rbind(FEV1_grade,FEV1_grade_sum) # bind to original FEV1 table
# Create and append “acceptable” row
FEV1_grade_sum_cplus <- data.frame(FEV1_quality="acceptable",n=sum(FEV1_grade$n[FEV1_grade$FEV1 %in% accept])) %>% rename(FEV1=FEV1_quality)
FEV1_grade <- rbind(FEV1_grade,FEV1_grade_sum_cplus) # Adds a row counting grades considered acceptable (like A, B, etc.) from the accept vector

# FVC
FVC_grade_sum <- as.data.frame(lapply(FVC_grade, sumfunc))
FVC_grade_sum[1] <- "total"
FVC_grade <- rbind(FVC_grade,FVC_grade_sum) # bind to original FEV1 table
# Create and append “acceptable” row
FVC_grade_sum_cplus <- data.frame(FVC_quality="acceptable",n=sum(FVC_grade$n[FVC_grade$FVC %in% accept])) %>% rename(FVC=FVC_quality)
FVC_grade <- rbind(FVC_grade, FVC_grade_sum_cplus) #  Adds a row counting grades considered acceptable (like A, B, etc.) from the accept vector

# kable(list(FEV1_grade,FVC_grade),align = "l",caption = "Quality grading",padding=0) %>% 
# kableExtra::kable_styling(font_size = 15,full_width = F)

# Subscript FEV1
names(FEV1_grade) <-c("FEV~1~","n") 

# Both FEV1 and FVC acceptable

# kable(FEV1_grade) %>%
#   kable_styling(full_width = FALSE, position = "float_left",font_size = 18)
# kable(FVC_grade) %>%
#   kable_styling(full_width = FALSE, position = "left",font_size = 18)

# kable(list(FEV1_grade,FVC_grade)) %>%
#   kable_styling(full_width = FALSE, position = "left",font_size = 18)

grade_table <- grade %>% select(ID,FEV1_quality,FVC_quality) %>%
  pivot_longer(cols = c(FEV1_quality,FVC_quality),names_to = "variable",values_to = "grade")


grade_table$variable <- str_replace(grade_table$variable,pattern = "_quality",replacement = "")

var_label(grade_table$grade) <- "Quality grading"

table1(~grade|variable, data=grade_table, overall = F)
#	FEV1 (N=938)	FVC (N=938)
# Quality grading		
# A	818 (87.2%)	811 (86.5%)
# B	50 (5.3%)	27 (2.9%)
# C	11 (1.2%)	31 (3.3%)
# D	10 (1.1%)	10 (1.1%)
# E	34 (3.6%)	38 (4.1%)
# F	9 (1.0%)	7 (0.7%)
# U	6 (0.6%)	14 (1.5%)

``` 

## Spirometry - BEV check
```{r}
# BEV should be smaller than 150ml or smaller than 5% of the FVC, whichever is larger
# Print a clean, readable summary table for the subset of data failing or flagged by the BEV criterion.
ds_bevcheck %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  gt() %>%
  tab_header(title = "Print BEV check (BEV - 0.150 or BEV - FVC5%) > 0")
```

## Spirometry - reproducibility
```{r}
d_repro_t <- ds_minmax %>% filter(Trial_rank==1) %>%
    group_by(FEV1_quality,FVC_quality,lf_repr150) %>%
    replace_na(list(FEV1_quality="E or lower",FVC_quality="E or lower")) %>%
    mutate(Quality=paste(FEV1_quality,FVC_quality,sep = "-"),
           lf_repr150=factor(lf_repr150,c(0,1),c("not repr.","repr.")),
           pef_d670=factor(pef_d670,c(0,1),c("<.67 L/sec)",">.67 L/sec"))
    )

var_label(d_repro_t$Quality) <- "FEV<sub>1</sub> - FVC grade"

table1(~Quality|lf_repr150,d_repro_t,caption = "Quality grades and FEV<sub>1</sub>&FVC reproducibility within 150ml")
```

## Spirometry - reproducibility continued with PEF
```{r}
var_label(d_repro_t$pef_d670) <- "PEF diff."

table1(~pef_d670|lf_repr150, d_repro_t,caption = "difference max PEF second best PEF vs. FEV<sub>1</sub> and FVC reproducibility within 150ml")
```

# Create analysis dataset
## Select FEV and FVC variables (volume variables) 
```{r}
# one value per test
# 1. Take FEV<sub>1</sub> and FVC from trial rank 1, if one or both are not reproducible within 150ml
# 2. Take highest FEV<sub>1</sub> and FVC from (possibly different) reproducible curves
ds_prep_forced <- ds_minmax %>% 
filter(!(lf_repr150==0 & Trial_rank>1)) %>% # Keep only trial rank 1 from tests without reproducible trials, but keep all trials when lung function is reproducible
  group_by(ID) %>%
  mutate(FEV1_s=case_when(lf_repr150==0~FEV1, # Creates new variables (FEV1_s, FVC_s, etc.) selecting either: value form the single trial if lung function is not reproducible, or the maximum value (best trial) if it is reproducible ('s' for selected)
                          lf_repr150==1~fev1max),
         FEV25=case_when(lf_repr150==0~FEV.25,
                          lf_repr150==1&FEV1==fev1max~FEV.25), 
         FEV5=case_when(lf_repr150==0~FEV.5,
                          lf_repr150==1&FEV1==fev1max~FEV.5),
         FEV75=case_when(lf_repr150==0~FEV.75,
                          lf_repr150==1&FEV1==fev1max~FEV.75),
         FEV3=case_when(lf_repr150==0~FEV3,
                          lf_repr150==1&FEV1==fev1max~FEV3),
         FEV6=case_when(lf_repr150==0~FEV6,
                          lf_repr150==1&FEV1==fev1max~FEV6),
         FVC_s=case_when(lf_repr150==0~FVC,
                          lf_repr150==1~fvcmax)) %>% 
  filter(!is.na(FEV25)) %>% # Filter empty rows
  select(ID,Spirometrist,Gender,Height,Weight,Ethnicity,Test.Date,lf_repr150,FEV1_quality,FVC_quality,
         test_accepted,FEV1_s,FEV25,FEV5,FEV75,FEV3,FEV6,FVC_s,FEV1.FVC) %>% 
  rename(date=Test.Date,FEV1=FEV1_s,FVC=FVC_s,FEV1_FVC=FEV1.FVC)
      
# Extremely high values exist due to missing decimal separator, introduce decimal separator between first and second digit in those cases.
ds_prep_forced <- transform(ds_prep_forced,
                            FVC=sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",FVC),
                            FEV6=sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",FEV6)
                            ) %>%
                  mutate(FVC=as.numeric(FVC),
                         FEV6=as.numeric(FEV6))

table1(~FEV1+FVC+FEV25+FEV5+FEV75+FEV3+FEV6,data = ds_prep_forced,caption = "Summary of FEV and FVC variables") 


table1(~FEV1+FVC+FEV25+FEV5+FEV75+FEV3+FEV6+FEV1_FVC,data = ds_prep_forced,caption = "Summary of FEV and FVC variables") 


```
## Creating the “flow” variables analysis dataset
```{r}
## Create analysis dataset - Flow variables
# - If not reproducible take all flow variables from Trial_rank 1
# - If reproducible and PEF<sub>max</sub>-PEF<sub>med</sub> **>0.67**
# take all flow variables from Trial_rank 1
# - If reproducible and PEF<sub>max</sub>-PEF<sub>med</sub> **<0.67**
# take all flow variables from Trial_rank 1

options(scipen=999)
ds_prep_flow <- ds_minmax %>% 
filter(!(lf_repr150==0 & Trial_rank>1)) %>% # Keep only trial rank 1 from tests without reproducible trials
  group_by(ID) %>%
  mutate(FEF10_s=case_when(lf_repr150 == 0~FEF10,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF10,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF10),
         FEF25=case_when(lf_repr150 == 0~FEF25.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.),
         FEF25_75=case_when(lf_repr150 == 0~FEF25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.75.),
         FEF40=case_when(lf_repr150 == 0~FEF40.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF40.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF40.),
         FEF50=case_when(lf_repr150 == 0~FEF50.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF50.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF50.),
         FEF60=case_when(lf_repr150 == 0~FEF60.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF60.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF60.),
         FEF75=case_when(lf_repr150 == 0~FEF75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF75.),
         FEF75_85=case_when(lf_repr150 == 0~FEF75.85.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF75.85.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF75.85.),
         FEF80=case_when(lf_repr150 == 0~FEF80.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF80.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF80.),
         FEF25_75=case_when(lf_repr150 == 0~FEF25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FEF25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FEF25.75.),
         FET25_75=case_when(lf_repr150 == 0~FET25.75.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~FET25.75.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~FET25.75.),
         PEF_s=case_when(lf_repr150 == 0~PEF,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~PEF,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~PEF),
         PEF_Lmin=case_when(lf_repr150 == 0~PEF..L.min.,
                           lf_repr150 == 1 & (pef_d670==1|is.na(pef_d670)) & Trial_rank==1~PEF..L.min.,
                           lf_repr150 == 1 & pef_d670==0 & PEF==pefmax~PEF..L.min.)) %>%
  filter(!is.na(FEF10_s)) %>% # Filter empty rows
  select(ID,Trial_rank,pef_d670,FEF10_s,FEF25,FEF25_75,FET25_75,FEF40,FEF50,FEF60,FEF75,FEF75_85,FEF80,FEF25_75,PEF_s,PEF_Lmin) %>%
  rename(FEF10=FEF10_s,PEF=PEF_s)

# ID 14139 has two trials with were pefmax pef, keep flow variables from trial rank 1 (the best trial)

ds_prep_flow <- ds_prep_flow %>% filter(!(ID==14139&Trial_rank==3)) 

table1(~FEF10+FEF25+ FEF25_75+FEF40+FEF50+FEF60+FEF75+FEF75_85+FEF80+FEF25_75+FET25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables") 

# Extremely high values exist, inspect:

ds_prep_flow$PEF[ds_prep_flow$PEF>100] 

# Write function to correct, if first two digits start with 1: decimal

correct_pef <- function(x){
  if (x<100) {
  return(x) 
  } 
  start <- x %>% str_extract(.,"((?=^1)(^\\d{2})|(^\\d{1}))")
  if (str_detect(start,"^\\d{2}$")) {
    output <- sub("(^\\d{2})\\.?(\\d*)","\\1.\\2",x)
  } else {
    output <- sub("(^\\d{1})\\.?(\\d*)","\\1.\\2",x)
  }
  return(as.numeric(output))
}  
  
ds_prep_flow <- ds_prep_flow %>% mutate(PEF=correct_pef(PEF))

# Same problem exists for PEF_Lmin, remove trailing zeroes first, then insert decimal seperator between 3rd and 4th digit
ds_prep_flow$PEF_Lmin <- str_replace(ds_prep_flow$PEF_Lmin,"0+$","") %>%
stringi::stri_sub_replace(.,4,3,value = ".") %>%
  as.numeric()

# Check
table1(~FEF10+FEF25+ FEF25_75+FEF40+FEF50+FEF60+FEF75+FEF75_85+FEF80+FEF25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables - corrected") 

# Check 3 corrected PEF ID's in EasyOne

ds_prep_flow[ds_prep_flow$PEF==9.6217870712280288,] # EasyOne PEF=9.62
ds_prep_flow[ds_prep_flow$PEF==7.020787239074707,]  # EasyOne PEF=7.02
ds_prep_flow[ds_prep_flow$PEF==7.2133121490478512,] # EasyOne PEF=7.21


table1(~FEF10+FEF25+FEF25_75+FEF40+FEF50+FEF60,data = ds_prep_flow,caption = "Summary of flow variables")

table1(~FEF75+FEF75_85+FEF80+FEF25_75+PEF+PEF_Lmin,data = ds_prep_flow,caption = "Summary of flow variables - continued")
```

# Metadata cleaning 
Cleans and prepares the spirometry data:
- Handles missing values in quality grades
- Calculates number of acceptable trials
- Checks for BEV (Back Extrapolated Volume) criteria

Processes questionnaire data:
- Recodes variables
- Handles missing values
- Creates derived variables
```{r}
#### VGO3: metadata and spirometry #### 
# Merge cleaned VGO3 spirometry datasets
# ds_prep_forced - volume-based dataset (FEV1, FVC etc.)
# ds_prep_flow   - flow-based dataset (FEF/PEF etc.)

ds_prep <- left_join(ds_prep_forced, ds_prep_flow, by = "ID") %>%
  select(-any_of(c("Trial_rank", "Height", "Weight", "Gender")))

## Online questionnaire (OQ) - VGO3
OQ_clean <- OQ %>%
  select(
    ID, fuquestdate, fugender, fuage,
    fuasever, fuasdd, fuas12m, fuascurmed,
    ful12mdailtycortinhaler, fulyantibiotic,
    fuwheezly, ful3mdailycough, ful3mdailyphlegm,
    fudoctoldcopd, fusmokeever,
    fufarmvstfam, fufarmvstjob, fufarmvstoth,
    fufarmvstpet, fufarmvstshop, fufarmvstno
  )

OQ_clean <- OQ_clean %>%
  mutate(
    fuquestdate = ymd(fuquestdate),
    fugender    = factor(fugender, levels = c(1, 2), labels = c("male", "female")),
    fuage       = suppressWarnings(as.numeric(fuage)),

    # binary recodes
    fuasever         = factor(fuasever, levels = c(0, 1), labels = c("no", "yes")),
    fuwheezly        = factor(fuwheezly, levels = c(0, 1), labels = c("no", "yes")),
    fuasdd           = factor(fuasdd,   levels = c(0, 1), labels = c("no", "yes")),
    fuas12m          = factor(fuas12m,  levels = c(0, 1), labels = c("no", "yes")),
    fuascurmed       = factor(fuascurmed, levels = c(0, 1), labels = c("no", "yes")),
    ful12mdailtycortinhaler = factor(ful12mdailtycortinhaler, levels = c(0, 1), labels = c("no", "yes")),
    fulyantibiotic   = factor(fulyantibiotic, levels = c(0, 1), labels = c("no", "yes")),
    ful3mdailycough  = factor(ful3mdailycough, levels = c(0, 1), labels = c("no", "yes")),
    ful3mdailyphlegm = factor(ful3mdailyphlegm, levels = c(0, 1), labels = c("no", "yes")),
    fudoctoldcopd    = factor(fudoctoldcopd, levels = c(0, 1), labels = c("no", "yes")),

    # smoking
    fusmokeever = factor(
      fusmokeever,
      levels = c("no", "yes", "yesstop"),
      labels = c("non", "current", "former")
    ),

    # farm visit flag
    fufarmvst = factor(
      case_when(
        fufarmvstfam == 1 | fufarmvstoth == 1 | fufarmvstpet == 1 |
          fufarmvstshop == 1 | fufarmvstjob == 1 ~ 1,
        fufarmvstno == 1 ~ 0,
        TRUE ~ NA_real_
      ),
      levels = c(0, 1),
      labels = c("no", "yes")
    )
  ) %>%
  mutate(
    fufarmvstmult = as.numeric(fufarmvstfam) + as.numeric(fufarmvstoth) +
      as.numeric(fufarmvstpet) + as.numeric(fufarmvstshop) + as.numeric(fufarmvstjob),
    fufarmvstmult = factor(
      case_when(
        fufarmvst == "no"   ~ "no, did not visit",
        fufarmvstmult == 1  ~ "no, single reason",
        fufarmvstmult > 1   ~ "yes",
        TRUE ~ NA_character_
      ),
      levels = c("no, did not visit", "no, single reason", "yes")
    )
  )

## Field questionnaire - VGO3
field_q_clean <- field_q %>%
  select(
    fuhbid, fuhbgender, fuhbdate, fuhbbdate, fuhblungmed, fuhbcovidinf,
    fuhbcovposdate, fuhbcovsusdate, fuhbpneumonialy, fuhbtestdiff,
    fuhbantiblydate
  ) %>%
  rename(ID = fuhbid) %>%
  mutate(
    fuhbdate        = lubridate::ymd(fuhbdate),
    fuhbantiblydate = lubridate::ymd(fuhbantiblydate, quiet = TRUE),
    fuhbcovposdate  = lubridate::ymd(fuhbcovposdate),
    fuhbcovidinf    = factor(
      fuhbcovidinf,
      levels = c(0, 1, 2),
      labels = c("no", "yes, tested", "yes, suspected")
    ),
    fuhbcovsusdate  = lubridate::dmy(fuhbcovsusdate)
  ) %>%
  # remove known data anomaly
  filter(!(ID == 13852 & fuhbdate == as.Date("2022-04-26"))) %>%
  mutate(
    fuhbcovdate     = coalesce(fuhbcovposdate, fuhbcovsusdate),
    fuhbcovinterval = lubridate::interval(fuhbcovdate, fuhbdate) %/% months(1),
    covid_lastmonth = dplyr::case_when(
      fuhbcovidinf == "no" ~ 0L,
      fuhbcovinterval <= 1 ~ 1L,
      TRUE ~ 0L
    ),
    # antibiotics within last 4 weeks (28 days)
    abx_days = as.numeric(difftime(fuhbdate, fuhbantiblydate, units = "days")),
    antibiotics_lastmonth = factor(
      dplyr::if_else(
        !is.na(abx_days) & abx_days >= 0 & abx_days <= 28,
        "Yes", "No"
      ),
      levels = c("No", "Yes")
    )
  ) %>%
  # drop intermediate and requested date fields
  select(-any_of(c(
    "abx_days",
    "fuhbcovposdate", "fuhbcovsusdate", "fuhbcovdate", "fuhbantiblydate"
  )))

## Field measurements (height/weight) - VGO3
field_f_clean <- field_f %>%
  select(fuhbid, fuhbfheight1, fuhbfheight2, fuhbfweigth) %>%
  rename(ID = fuhbid, weight = fuhbfweigth) %>%
  mutate(
    height = (fuhbfheight1 + fuhbfheight2) / 2,  # average height
    bmi    = weight / ((height / 100) ^ 2)       # BMI
  ) %>%
  # remove known data anomaly
  filter(!(ID == 13852 & weight == 100.0)) %>%
  # drop intermediates
  select(-any_of(c("fuhbfheight1", "fuhbfheight2")))

# Handle missing heights via spirometry dataset 'ds'
missing_heights <- ds %>%
  select(ID, Height, Weight) %>%
  filter(ID %in% field_f_clean$ID[is.na(field_f_clean$height)]) %>%
  distinct() %>%
  mutate(BMI = Weight / ((Height / 100) ^ 2))

for (i in missing_heights$ID) {
  field_f_clean$height[field_f_clean$ID == i] <- missing_heights$Height[missing_heights$ID == i]
  field_f_clean$weight[field_f_clean$ID == i] <- missing_heights$Weight[missing_heights$ID == i]
  field_f_clean$bmi[field_f_clean$ID == i]    <- missing_heights$BMI[missing_heights$ID == i]
}

# Correct extreme height/weight values using VGO2 data (old_full)
field_f_clean <- field_f_clean %>%
  mutate(
    height = ifelse(
      ID == 37532,
      old_full$Height[old_full$respnr == 37532] * 100,  # m -> cm
      height
    ),
    weight = case_when(
      ID == 37532 ~ old_full$Weight[old_full$respnr == 37532],
      ID == 35258 ~ 168,    # correct extreme BMI
      TRUE ~ weight
    ),
    bmi = weight / ((height / 100) ^ 2)
  )


#### VGO2 baseline (old_full, old_intake, old_spiro) – shared pieces ####
# VGO2 questionnaire baseline data
old_full_clean <- old_full %>%
  select(
    respnr, educationlevel, gender, age, Height, Weight, bmi, Ethnicity, Atopisch,
    copd_gold, copd_lln, sq13, reversibility, smokeever, packyears, asever, ascurmed, as12m, asdd,
    pneumonia3y, farmchild, mindist_bvb12, mindist50geiten_bvb12, mindistpluimvee_bvb12,
    mindistvarkens_bvb12, mindistpaarden_bvb12, mindistrundvee_bvb12,
    geiten_500_bvb12, geiten_1000_bvb12, geiten_5km_bvb12,
    pluimvee_500_bvb12, pluimvee_1000_bvb12,
    varkens_500_bvb12, varkens_1000_bvb12,
    rundvee_500_bvb12, rundvee_1000_bvb12
  )

# VGO2 intake dates / birthdates
old_intake_clean <- old_intake %>%
  rename(T0_bdate = gebdatum) %>%
  mutate(T0_date = ymd(intakedatum)) %>%
  select(ID, T0_date, T0_bdate)

# VGO2 baseline spirometry
old_spiro_clean <- old_spiro %>%
  select(
    respnr, lf_analist, test_date_pre, FEV1_pre, FVC_pre, fev1_fvc_pre,
    PEF_pre, FEF25_75_pre, QualityGrade_pre,
    FEV1_pre_pred, FVC_pre_pred, FEV1FVC_pre_pred, FEF25_75_pre_pred
  ) %>%
  rename(
    ID          = respnr,
    Spirometrist = lf_analist,
    date         = test_date_pre,
    FEV1         = FEV1_pre,
    FVC          = FVC_pre,
    FEV1_FVC     = fev1_fvc_pre,
    PEF          = PEF_pre,
    FEF25_75     = FEF25_75_pre,
    Quality_grade = QualityGrade_pre
  ) %>%
  mutate(timepoint = "T0") %>%
  filter(!is.na(FEV1))


## AGE VERIFICATION: VGO2 vs VGO3
old_full_clean <- old_full_clean %>%
  mutate(ID = respnr)

age_check <- old_intake_clean %>%
  mutate(ID = as.character(ID)) %>%
  left_join(
    select(old_full_clean, ID, age) %>% mutate(ID = as.character(ID)),
    by = "ID"
  ) %>%
  left_join(
    select(field_q_clean, ID, fuhbdate, fuhbbdate) %>% mutate(ID = as.character(ID)),
    by = "ID"
  ) %>%
  mutate(
    age_T0          = age,
    age_T1          = interval(ymd(T0_bdate), ymd(fuhbdate)) / years(1),
    expected_age_T1 = age_T0 + interval(ymd(T0_date), ymd(fuhbdate)) / years(1),
    age_discrepancy = age_T1 - expected_age_T1,
    age_check_status = case_when(
      is.na(age_T1) ~ "Missing VGO3 data",
      abs(age_discrepancy) <= 1 ~ "Consistent (≤1 year)",
      abs(age_discrepancy) <= 2 ~ "Minor discrepancy (≤2 years)",
      TRUE ~ "Major discrepancy (>2 years)"
    )
  )

# Correct major discrepancies in birth dates
major_discrepancies <- age_check %>%
  filter(age_check_status == "Major discrepancy (>2 years)")

if (nrow(major_discrepancies) > 0) {
  message("\nCorrecting birth dates for ", nrow(major_discrepancies), " participants:")

  for (id in major_discrepancies$ID) {
    original_dob  <- major_discrepancies$fuhbbdate[major_discrepancies$ID == id]
    corrected_dob <- major_discrepancies$T0_bdate[major_discrepancies$ID == id]

    message("ID ", id, ": ", original_dob, " \u2192 ", corrected_dob)

    field_q_clean$fuhbbdate[field_q_clean$ID == id] <- corrected_dob
  }
}

# Obvious typo in one birth date
field_q_clean$fuhbbdate[field_q_clean$ID == "16504"] <- as.Date("1950-05-18")

# Update age in VGO3 dataset
field_q_clean <- field_q_clean %>%
  mutate(
    fuhbbdate = ymd(fuhbbdate, quiet = TRUE),
    age_VGO3  = interval(fuhbbdate, fuhbdate) / years(1)
  )

## GENDER CHECK AND CORRECTION
old_full_clean$ID <- as.character(old_full_clean$ID)
field_q_clean$ID  <- as.character(field_q_clean$ID)
ds_prep$ID        <- as.character(ds_prep$ID)

common_IDs <- intersect(old_full_clean$ID, field_q_clean$ID)

# VGO2: gender (0=m, 1=f)
gender_vgo2 <- old_full_clean %>%
  filter(ID %in% common_IDs) %>%
  transmute(
    ID,
    gender_vgo2 = if_else(gender == 1, "female", "male")
  )

# VGO3: fuhbgender, numeric 0/1 or text
gender_vgo3 <- field_q_clean %>%
  filter(ID %in% common_IDs) %>%
  transmute(
    ID,
    gender_vgo3 = case_when(
      is.numeric(fuhbgender) ~ if_else(fuhbgender == 1, "female", "male"),
      TRUE ~ tolower(trimws(as.character(fuhbgender)))
    )
  )

# Spirometry dataset (ds): Gender as text
ds_gender <- ds %>%
  filter(ID %in% common_IDs) %>%
  transmute(
    ID = as.character(ID),
    gender_spiro = tolower(trimws(as.character(Gender)))
  ) %>%
  distinct(ID, .keep_all = TRUE)

gender_compare <- gender_vgo2 %>%
  left_join(gender_vgo3, by = "ID") %>%
  left_join(ds_gender,  by = "ID") %>%
  filter(!is.na(gender_vgo2), !is.na(gender_vgo3), gender_vgo2 != gender_vgo3) %>%
  mutate(match_spiro_vgo2 = (gender_spiro == gender_vgo2))

to01 <- function(lbl) ifelse(lbl == "female", 1L, 0L)

# Correct VGO3 when spiro matches VGO2
for (id in gender_compare$ID[gender_compare$match_spiro_vgo2]) {
  code01 <- to01(gender_vgo2$gender_vgo2[gender_vgo2$ID == id])
  field_q_clean$fuhbgender[field_q_clean$ID == id] <- code01
  message("Corrected VGO3 gender for ID: ", id)
}

remaining_mismatches <- gender_vgo2 %>%
  left_join(field_q_clean %>% select(ID, fuhbgender), by = "ID") %>%
  mutate(
    fuhbgender_lbl = case_when(
      is.numeric(fuhbgender) ~ if_else(fuhbgender == 1, "female", "male"),
      TRUE ~ tolower(trimws(as.character(fuhbgender)))
    )
  ) %>%
  filter(gender_vgo2 != fuhbgender_lbl)

message("Remaining gender mismatches after correction: ", nrow(remaining_mismatches))

#### CREATE FINAL VGO3 DATASET ####
# Ensure ID is character
ds_prep      <- ds_prep      %>% mutate(ID = as.character(ID))
field_q_clean <- field_q_clean %>% mutate(ID = as.character(ID))
OQ_clean     <- OQ_clean     %>% mutate(ID = as.character(ID))
field_f_clean <- field_f_clean %>% mutate(ID = as.character(ID))
NH_diff      <- NH_diff      %>% mutate(ID = as.character(ID))

ds_prep$date <- as.Date(ds_prep$date)

# Keep relevant columns in each component
ds_prep <- ds_prep %>%
  select(any_of(c("ID", "Spirometrist", "Ethnicity", "date",
                  "FEV1", "FVC", "FEV1_FVC", "FEF25_75")))

field_q_clean <- field_q_clean %>%
  select(any_of(c("ID", "fuhbgender", "fuhbdate", "fuhblungmed",
                  "fuhbpneumonialy", "covid_lastmonth",
                  "antibiotics_lastmonth", "age_VGO3")))

OQ_clean <- OQ_clean %>%
  select(any_of(c("ID", "fuasever", "fuas12m", "fuasdd", "fuascurmed",
                  "ful12mdailtycortinhaler", "fulyantibiotic",
                  "fudoctoldcopd", "fusmokeever", "fufarmvst")))

field_f_clean <- field_f_clean %>%
  select(any_of(c("ID", "weight", "height", "bmi")))

NH_diff <- NH_diff %>%
  select(any_of(c("ID", "T0_wkprior_NH3", "T1_wkprior_NH3", "T1_T0_NH3_diff")))

# Merge all into VGO3_metadata (long, T1 only)
VGO3_metadata <- ds_prep %>%
  left_join(field_q_clean, by = "ID") %>%
  left_join(OQ_clean,     by = "ID") %>%
  left_join(field_f_clean, by = "ID") %>%
  left_join(NH_diff,      by = "ID") %>%
  mutate(
    Study     = "VGO3",
    Timepoint = "T1",
    Date      = date,

    # NOTE: make labels lower-case for consistency and GLI derivation later
    Gender = if (!is.factor(fuhbgender)) {
      factor(fuhbgender, levels = c(0, 1), labels = c("male", "female"))
    } else {
      fuhbgender
    },

    Lung_med = if (!is.factor(fuhblungmed)) {
      factor(fuhblungmed, levels = c(0, 1), labels = c("no", "yes"))
    } else {
      fuhblungmed
    },

    Pneumonia_lastyr = if (!is.factor(fuhbpneumonialy)) {
      factor(fuhbpneumonialy, levels = c(0, 1), labels = c("no", "yes"))
    } else {
      fuhbpneumonialy
    },

    Covid_lastmonth = if (!is.factor(covid_lastmonth)) {
      factor(covid_lastmonth, levels = c(0, 1), labels = c("no", "yes"))
    } else {
      covid_lastmonth
    },

    Antibiotics_lastmonth = antibiotics_lastmonth,
    Age          = age_VGO3,
    Asthma_ever  = fuasever,
    Asthma_doc_diag = fuasdd,
    Asthma_attk_lastyr = fuas12m,
    Asthma_med_current = fuascurmed,
    Cortinhal_lastyr   = ful12mdailtycortinhaler,
    Antibiotics_lastyr = fulyantibiotic,
    COPD_emphy_doc_diag = fudoctoldcopd,
    Smoking_status     = fusmokeever,
    Farm_visit_lastyr  = fufarmvst,
    Weight = weight,
    Height = height,
    BMI    = bmi
  ) %>%
  # drop source columns
  select(-any_of(c(
    "date", "fuhbgender", "fuhblungmed", "fuhbpneumonialy",
    "covid_lastmonth", "antibiotics_lastmonth", "age_VGO3",
    "fuasever", "fuas12m", "fuascurmed",
    "ful12mdailtycortinhaler", "fulyantibiotic", "fudoctoldcopd",
    "fusmokeever", "fufarmvst",
    "weight", "height", "bmi", "fuasdd", "farmchild"
  )))

# Fill missing Asthma_ever and Smoking_status from VGO2 (old_full)
old_full <- old_full %>%
  mutate(respnr = as.character(respnr))

VGO3_metadata <- VGO3_metadata %>%
  mutate(ID = as.character(ID))

# Asthma_ever
for (i in VGO3_metadata$ID[is.na(VGO3_metadata$Asthma_ever)]) {
  val <- old_full$asever[old_full$respnr == i]
  VGO3_metadata$Asthma_ever[VGO3_metadata$ID == i] <-
    factor(val, levels = c(0, 1), labels = c("no", "yes"))
}

# Smoking_status
for (i in VGO3_metadata$ID[is.na(VGO3_metadata$Smoking_status)]) {
  val <- old_full$smokeever[old_full$respnr == i]
  VGO3_metadata$Smoking_status[VGO3_metadata$ID == i] <-
    factor(val, levels = c(0, 1, 2), labels = c("non", "former", "current"))
}

# Check missingness (corrected variable names)
sum(is.na(VGO3_metadata$Asthma_ever))      # should be 0
sum(is.na(VGO3_metadata$Smoking_status))   # should be 0

## -----------------------------------------------------------
## Calculate GLI % predicted values for VGO3
## -----------------------------------------------------------

T1 <- VGO3_metadata %>%
  select(ID, Age, Ethnicity, Gender, Height, FEV1, FVC, FEV1_FVC, FEF25_75) %>%
  mutate(
    gender_num = ifelse(Gender == "male", 1, 2),
    ethnic_code = case_when(
      Ethnicity == "Caucasian"        ~ 1,
      Ethnicity == "African-American" ~ 2,
      Ethnicity == "NE Asian"         ~ 3,
      Ethnicity == "SE Asian"         ~ 4,
      TRUE                            ~ 5  # Other/mixed
    ),
    height_m = Height / 100
  )

# FEV1 %pred
T1$FEV1_pred <- pctpred_GLI(
  age      = T1$Age,
  height   = T1$height_m,
  gender   = T1$gender_num,
  ethnicity = T1$ethnic_code,
  FEV1     = T1$FEV1
)

# FVC %pred
T1$FVC_pred <- pctpred_GLI(
  age      = T1$Age,
  height   = T1$height_m,
  gender   = T1$gender_num,
  ethnicity = T1$ethnic_code,
  FVC      = T1$FVC
)

# FEV1/FVC %pred
T1$FEV1FVC_pred <- pctpred_GLI(
  age      = T1$Age,
  height   = T1$height_m,
  gender   = T1$gender_num,
  ethnicity = T1$ethnic_code,
  FEV1FVC  = T1$FEV1_FVC
)

# FEF25-75 %pred (mid-flow)
T1$FEF25_75_pred <- pctpred_GLI(
  age      = T1$Age,
  height   = T1$height_m,
  gender   = T1$gender_num,
  ethnicity = T1$ethnic_code,
  FEF2575  = T1$FEF25_75
)

T1_vars <- T1 %>%
  select(ID, FEV1_pred, FVC_pred, FEV1FVC_pred, FEF25_75_pred)

# Join %pred back to VGO3_metadata
VGO3_metadata <- VGO3_metadata %>%
  left_join(T1_vars, by = "ID")

## -----------------------------------------------------------
## Add farmchild / atopy / education from VGO2 baseline
## -----------------------------------------------------------

old_full_subset <- old_full_clean %>%
  select(respnr, farmchild, Atopisch, educationlevel) %>%
  distinct(respnr, .keep_all = TRUE) %>%
  mutate(ID = as.character(respnr))

VGO3_metadata <- VGO3_metadata %>%
  mutate(ID = as.character(ID)) %>%
  left_join(
    old_full_subset %>%
      select(ID, farmchild, Atopisch, educationlevel),
    by = "ID"
  ) %>%
  mutate(
    Atopic = if (!is.factor(Atopisch))
      factor(Atopisch, levels = c(0, 1), labels = c("no", "yes"))
    else Atopisch,

    Farm_childhood = if (!is.factor(farmchild))
      factor(farmchild, levels = c(0, 1), labels = c("no", "yes"))
    else farmchild,

    Education_level = if (!is.factor(educationlevel))
      factor(educationlevel, levels = c(0, 1, 2), labels = c("low", "medium", "high"))
    else educationlevel
  )

#### Asthma composite variables (VGO3) ####
VGO3_metadata <- VGO3_metadata %>%
  mutate(
    # Asthma_doc_diag -> normalise to 0/1 then factor
    Asthma_doc_diag = dplyr::case_when(
      Asthma_doc_diag %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ 1L,
      Asthma_doc_diag %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ 0L,
      TRUE ~ NA_integer_
    ),
    Asthma_doc_diag = dplyr::case_when(
      is.na(Asthma_doc_diag) &
        Asthma_ever %in% c(0, "0", "no", "No", "NO", FALSE) ~ 0L,
      TRUE ~ Asthma_doc_diag
    ),
    Asthma_doc_diag = factor(
      Asthma_doc_diag,
      levels = c(0, 1),
      labels = c("no", "yes")
    ),

    # Asthma_attk_lastyr
    Asthma_attk_lastyr = dplyr::case_when(
      Asthma_attk_lastyr %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ "yes",
      Asthma_attk_lastyr %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ "no",
      TRUE ~ NA_character_
    ),
    Asthma_attk_lastyr = dplyr::case_when(
      is.na(Asthma_attk_lastyr) &
        Asthma_ever %in% c(0, "0", "no", "No", "NO", FALSE) ~ "no",
      TRUE ~ Asthma_attk_lastyr
    ),
    Asthma_attk_lastyr = factor(
      Asthma_attk_lastyr,
      levels = c("no", "yes")
    ),

    # Asthma_med_current
    Asthma_med_current = dplyr::case_when(
      Asthma_med_current %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ "yes",
      Asthma_med_current %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ "no",
      TRUE ~ NA_character_
    ),
    Asthma_med_current = dplyr::case_when(
      is.na(Asthma_med_current) &
        Asthma_ever %in% c(0, "0", "no", "No", "NO", FALSE) ~ "no",
      TRUE ~ Asthma_med_current
    ),
    Asthma_med_current = factor(
      Asthma_med_current,
      levels = c("no", "yes")
    ),

    # Asthma_current composite
    Asthma_current = factor(
      dplyr::case_when(
        Asthma_doc_diag == "yes" &
          (Asthma_attk_lastyr == "yes" | Asthma_med_current == "yes") ~ "yes",
        TRUE ~ "no"
      ),
      levels = c("no", "yes")
    )
  )

# Drop unnecessary source columns
VGO3_metadata <- VGO3_metadata %>%
  select(-any_of(c("Atopisch", "educationlevel", "farmchild")))

# Reorder columns
VGO3_metadata <- VGO3_metadata %>%
  relocate(
    ID, Study, Timepoint, Date, Spirometrist, Age, Ethnicity, Gender,
    Height, Weight, BMI, Education_level, Smoking_status,
    Pneumonia_lastyr, COPD_emphy_doc_diag,
    Asthma_current, Asthma_ever, Asthma_doc_diag,
    Covid_lastmonth, Asthma_attk_lastyr, Lung_med, Atopic,
    Asthma_med_current, Cortinhal_lastyr,
    Antibiotics_lastmonth, Antibiotics_lastyr,
    Farm_childhood, Farm_visit_lastyr,
    T0_wkprior_NH3, T1_wkprior_NH3, T1_T0_NH3_diff,
    FEV1, FVC, FEV1_FVC, FEF25_75,
    FEV1_pred, FVC_pred, FEV1FVC_pred, FEF25_75_pred
  )

# Labels for VGO3_metadata
var_label(VGO3_metadata$ID)         <- "VGO participant identifier"
var_label(VGO3_metadata$Study)      <- "VGO study (VGO2 or VGO3)"
var_label(VGO3_metadata$Timepoint)  <- "Sample collection timepoint (T0 or T1)"
var_label(VGO3_metadata$Date)       <- "Date of VGO3 spirometry"
var_label(VGO3_metadata$Age)        <- "Age at visit"
var_label(VGO3_metadata$Ethnicity)  <- "Ethnicity for lung function test"
var_label(VGO3_metadata$Spirometrist) <- "Spirometrist initials"
var_label(VGO3_metadata$Gender)     <- "Participant gender"
var_label(VGO3_metadata$Height)     <- "Height (cm)"
var_label(VGO3_metadata$Weight)     <- "Weight (kg)"
var_label(VGO3_metadata$BMI)        <- "Body Mass Index"
var_label(VGO3_metadata$Education_level) <- "Education level (low, medium, high)"
var_label(VGO3_metadata$Smoking_status)  <- "Smoking status (non, former, current)"
var_label(VGO3_metadata$Pneumonia_lastyr) <- "Pneumonia in the last year"
var_label(VGO3_metadata$COPD_emphy_doc_diag) <- "Doctor diagnosed COPD/emphysema"
var_label(VGO3_metadata$Asthma_current)  <- "Current asthma (doc diag + attack/medication)"
var_label(VGO3_metadata$Asthma_ever)     <- "Ever had asthma"
var_label(VGO3_metadata$Asthma_doc_diag) <- "Doctor diagnosed asthma"
var_label(VGO3_metadata$Covid_lastmonth) <- "Covid infection in the last month"
var_label(VGO3_metadata$Asthma_attk_lastyr) <- "Asthma attack in the last year"
var_label(VGO3_metadata$Lung_med)        <- "Use of lung medication"
var_label(VGO3_metadata$Asthma_med_current) <- "Use of asthma medication"
var_label(VGO3_metadata$Cortinhal_lastyr)   <- "Use of corticosteroid inhaler in the last year"
var_label(VGO3_metadata$Antibiotics_lastmonth) <- "Use of antibiotics in the last month"
var_label(VGO3_metadata$Antibiotics_lastyr) <- "Use of antibiotics in the last year"
var_label(VGO3_metadata$Farm_visit_lastyr)  <- "Visited a farm in the last year"

# Spirometry labels
var_label(VGO3_metadata$FEV1)          <- "FEV1 (L)"
var_label(VGO3_metadata$FVC)           <- "FVC (L)"
var_label(VGO3_metadata$FEV1_FVC)      <- "FEV1/FVC (Tiffenau index)"
var_label(VGO3_metadata$FEF25_75)      <- "FEF 25–75% (L/sec)"
var_label(VGO3_metadata$FEV1_pred)     <- "Percent predicted FEV1"
var_label(VGO3_metadata$FVC_pred)      <- "Percent predicted FVC"
var_label(VGO3_metadata$FEV1FVC_pred)  <- "Percent predicted FEV1/FVC ratio"
var_label(VGO3_metadata$FEF25_75_pred) <- "Percent predicted FEF (25–75% midflow)"
var_label(VGO3_metadata$Atopic)        <- "Atopy (>1 positive specific IgE)"
var_label(VGO3_metadata$Farm_childhood) <- "Childhood on farm (no, yes)"
var_label(VGO3_metadata$T0_wkprior_NH3)  <- "T0 (2014/2015) week prior avg NH3 (µg/m3)"
var_label(VGO3_metadata$T1_wkprior_NH3)  <- "T1 (2021/2022) week prior avg NH3 (µg/m3)"
var_label(VGO3_metadata$T1_T0_NH3_diff)  <- "Change in week prior avg NH3 (µg/m3) (T1-T0)"

VGO3_metadata <- VGO3_metadata %>% 
  rename(
    MMEF = FEF25_75,
    MMEF_pred = FEF25_75_pred,
    age = Age,
    spirometrist  = Spirometrist,
    study = Study, 
    timepoint = Timepoint,
    date_spiro = Date,
    gender = Gender,
    height = Height,
    weight = Weight,
    bmi = BMI, 
    ethnicity = Ethnicity,
    educationlevel = Education_level,
    smoke_status = Smoking_status,
    pneumo_lastyr = Pneumonia_lastyr,
    copd_doc_diag = COPD_emphy_doc_diag,
    asthma_current = Asthma_current,
    asthma_ever = Asthma_ever,
    asthma_doc_diag = Asthma_doc_diag,
    covid_lastmonth = Covid_lastmonth, 
    asthma_attk_lastyr = Asthma_attk_lastyr,
    lung_med = Lung_med,
    atopic = Atopic,
    asthma_med_current = Asthma_med_current,
    cortinhal_lastyr = Cortinhal_lastyr,
    antibiotics_lastmonth = Antibiotics_lastmonth,
    antibiotics_lastyr = Antibiotics_lastyr, 
    farm_childhood = Farm_childhood,
    farm_visit_lastyr = Farm_visit_lastyr
  )


VGO3_metadata <- VGO3_metadata %>%
  select(
    ID, study, timepoint, date_spiro, age, ethnicity, spirometrist, gender,
    height, weight, bmi, educationlevel, smoke_status, pneumo_lastyr,
    copd_doc_diag, asthma_current, asthma_ever, asthma_doc_diag,
    covid_lastmonth, asthma_attk_lastyr, lung_med, atopic,
    asthma_med_current, cortinhal_lastyr, antibiotics_lastmonth,
    antibiotics_lastyr, farm_childhood, farm_visit_lastyr,
    T0_wkprior_NH3, T1_wkprior_NH3, T1_T0_NH3_diff,
    FEV1, FVC, FEV1_FVC, MMEF,
    FEV1_pred, FVC_pred, FEV1FVC_pred, MMEF_pred
  )

# Save VGO3 dataset
write.csv(
  VGO3_metadata,
  "../Output/Cross_sectional/Participant_data/VGO3_data.csv",
  row.names = FALSE
)

#### VGO2: metadata and spirometry  (with Asthma_current) #### 

old_full_clean   <- old_full_clean   %>% mutate(ID = as.character(ID))
old_intake_clean <- old_intake_clean %>% mutate(ID = as.character(ID))
old_spiro_clean  <- old_spiro_clean  %>% mutate(ID = as.character(ID))

VGO2_metadata <- old_full_clean %>%
  left_join(old_intake_clean, by = "ID") %>%
  left_join(old_spiro_clean,  by = "ID") %>%
  mutate(
    study      = "VGO2",
    timepoint  = "T0",
    date_spiro = T0_date,
    birth_date = T0_bdate,

    gender = case_when(
      gender == 0 ~ "male",
      gender == 1 ~ "female",
      TRUE        ~ NA_character_
    ),
    gender = factor(gender, levels = c("male", "female")),

    educationlevel = factor(
      educationlevel,
      levels = c(0, 1, 2),
      labels = c("low", "medium", "high")
    ),

    smoke_status = factor(
      smokeever,
      levels = c(0, 1, 2),
      labels = c("non", "former", "current")
    ),

    asthma_ever = factor(asever, levels = c(0, 1), labels = c("no", "yes")),
    copd_gold   = factor(copd_gold, levels = c(0, 1), labels = c("no", "yes")),

    ## ---- Asthma_doc_diag (VGO2: asdd) ----
    Asthma_doc_diag = dplyr::case_when(
      asdd %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ 1L,
      asdd %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ 0L,
      TRUE ~ NA_integer_
    ),
    Asthma_doc_diag = dplyr::case_when(
      is.na(Asthma_doc_diag) &
        asthma_ever %in% c("no", "No", "NO", 0, "0", FALSE) ~ 0L,
      TRUE ~ Asthma_doc_diag
    ),
    Asthma_doc_diag = factor(
      Asthma_doc_diag,
      levels = c(0, 1),
      labels = c("no", "yes")
    ),

    ## ---- Asthma_attk_lastyr (VGO2: as12m) ----
    Asthma_attk_lastyr = dplyr::case_when(
      as12m %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ "yes",
      as12m %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ "no",
      TRUE ~ NA_character_
    ),
    Asthma_attk_lastyr = dplyr::case_when(
      is.na(Asthma_attk_lastyr) &
        asthma_ever %in% c("no", "No", "NO", 0, "0", FALSE) ~ "no",
      TRUE ~ Asthma_attk_lastyr
    ),
    Asthma_attk_lastyr = factor(
      Asthma_attk_lastyr,
      levels = c("no", "yes")
    ),

    ## ---- Asthma_med_current (VGO2: ascurmed) ----
    Asthma_med_current = dplyr::case_when(
      ascurmed %in% c(1, "1", "yes", "Yes", "YES", TRUE)  ~ "yes",
      ascurmed %in% c(0, "0", "no",  "No",  "NO",  FALSE) ~ "no",
      TRUE ~ NA_character_
    ),
    Asthma_med_current = dplyr::case_when(
      is.na(Asthma_med_current) &
        asthma_ever %in% c("no", "No", "NO", 0, "0", FALSE) ~ "no",
      TRUE ~ Asthma_med_current
    ),
    Asthma_med_current = factor(
      Asthma_med_current,
      levels = c("no", "yes")
    ),

    ## ---- Asthma_current (same definition as VGO3) ----
    Asthma_current = factor(
      dplyr::case_when(
        Asthma_doc_diag == "yes" &
          (Asthma_attk_lastyr == "yes" | Asthma_med_current == "yes") ~ "yes",
        TRUE ~ "no"
      ),
      levels = c("no", "yes")
    )
  ) %>%
  rename(
    height    = Height,
    weight    = Weight,
    ethnicity = Ethnicity,
    atopic    = Atopisch,
    copd_doc_diag = sq13, 
    asthma_current = Asthma_current,
    asthma_doc_diag = Asthma_doc_diag,
    asthma_attk_lastyr = Asthma_attk_lastyr,
    asthma_med_current = Asthma_med_current,
    pneumo_lastyr     = pneumonia3y,
    spirometrist       = Spirometrist,
    quality_grade      = Quality_grade,
    date_and_time_lf   = date,
    MMEF               = FEF25_75,
    FEV1_pred          = FEV1_pre_pred,
    FVC_pred           = FVC_pre_pred,
    FEV1FVC_pred       = FEV1FVC_pre_pred,
    MMEF_pred          = FEF25_75_pre_pred
  )

## Labels for VGO2_metadata (additions and existing)
var_label(VGO2_metadata$ID)            <- "VGO participant identifier"
var_label(VGO2_metadata$educationlevel) <- "Education level (0=low, 1=medium, 2=high)"
var_label(VGO2_metadata$gender)        <- "Participant gender"
var_label(VGO2_metadata$study)         <- "VGO study (VGO2 or VGO3)"
var_label(VGO2_metadata$age)           <- "Age at visit"
var_label(VGO2_metadata$height)        <- "Height (cm)"
var_label(VGO2_metadata$weight)        <- "Weight (kg)"
var_label(VGO2_metadata$bmi)           <- "Body Mass Index"
var_label(VGO2_metadata$ethnicity)     <- "Ethnicity"
var_label(VGO2_metadata$asthma_ever)        <- "Ever had asthma"
var_label(VGO2_metadata$asthma_doc_diag)    <- "Doctor diagnosed asthma"
var_label(VGO2_metadata$asthma_attk_lastyr) <- "Asthma attack in the last year"
var_label(VGO2_metadata$asthma_med_current) <- "Use of asthma medication"
var_label(VGO2_metadata$asthma_current)     <- "Current asthma (doc diag + attack/medication)"
var_label(VGO2_metadata$atopic)        <- "Atopic history"
var_label(VGO2_metadata$copd_gold)     <- "COPD defined by GOLD"
var_label(VGO2_metadata$copd_lln)      <- "COPD defined by LLN"
var_label(VGO2_metadata$reversibility) <- "Bronchodilator reversibility"
var_label(VGO2_metadata$smoke_status)  <- "Smoking status (non, former, current)"
var_label(VGO2_metadata$packyears)     <- "Smoking pack years"
var_label(VGO2_metadata$timepoint)     <- "Sample collection timepoint (T0 or T1)"
var_label(VGO2_metadata$birth_date)    <- "Date of birth"
var_label(VGO2_metadata$pneumo_lastyr) <- "Pneumonia in the last year"
var_label(VGO2_metadata$FEV1)           <- "Forced Expiratory Volume in 1s (L)"
var_label(VGO2_metadata$FVC)            <- "Forced Vital Capacity (L)"
var_label(VGO2_metadata$FEV1_FVC)       <- "FEV1/FVC ratio"
var_label(VGO2_metadata$MMEF)       <- "Forced Expiratory Flow (25–75%)"
var_label(VGO2_metadata$quality_grade)  <- "Quality grade of spirometry"
var_label(VGO2_metadata$spirometrist)   <- "Name of lung function technician"
var_label(VGO2_metadata$date_spiro)     <- "Date of spirometry test"
var_label(VGO2_metadata$FEV1_pred)      <- "Percent predicted FEV1"
var_label(VGO2_metadata$FVC_pred)       <- "Percent predicted FVC"
var_label(VGO2_metadata$FEV1FVC_pred)   <- "Percent predicted FEV1/FVC ratio"
var_label(VGO2_metadata$MMEF_pred)  <- "Percent predicted FEF (25–75% midflow)"

VGO2_metadata <- VGO2_metadata %>%
  select(
    ID, study, timepoint, date_spiro, age, ethnicity, spirometrist, gender,
    height, weight, bmi, educationlevel, smoke_status, pneumo_lastyr, copd_doc_diag, copd_gold, copd_lln, asthma_current, asthma_ever, asthma_doc_diag, asthma_attk_lastyr, asthma_med_current, asthma_current, atopic, farmchild, FEV1, FVC, FEV1_FVC, MMEF, FEV1_pred, FVC_pred, FEV1FVC_pred, MMEF_pred)


# Save VGO2 dataset
write.csv(
  VGO2_metadata,
  "../Output/Cross_sectional/Participant_data/VGO2_data.csv",
  row.names = FALSE
)

```
