---
title: "Load data"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

# Load packages 
```{r}
library(phyloseq); library(vegan); library(tidyr); library(dplyr); library(ggplot2);library(tidyverse); library(microbiomer); library(paletteer);library(microbiome); library(devtools); library(forcats); library(haven); library(openxlsx); library(vegan); library(ggsignif); library(patchwork); library(decontam); library(RColorBrewer); library(forcats); library(lubridate)

source("functions.R")

# Theme for plots in script
theme_set(theme_bw())
theme_update(axis.text.x = element_text(angle = 45, hjust=1),
             strip.text = element_text(colour = 'white'),
             strip.background =element_rect(fill="#2F4858", color = "#2F4858"))
```
# Load data
```{r}
#### VGO2 NP #### 
# These samples were sequenced at RIVM, processed using DADA2 at the RIVM by MLO
VGO2_NP_all <- readRDS("../Input/Phyloseq/VGO2_NP/Received_Mari_Lee/ps_raw.Rds") # all time points 

# Check the number of participants included at each time point in the raw dataset (T1, T2, T3)
time_points_NP <- sample_data(VGO2_NP_all) %>%
  data.frame() %>%
  filter(time_period %in% c("T1", "T2", "T3")) %>%
  group_by(time_period) %>%
  summarise(unique_participants = n_distinct(respnr)) %>%
  arrange(time_period) # T1 = 220, T2 = 12, T3 = 12

# Remove samples where subject_identifier contains "T2" or "T3" - we will only keep time point 1 for analysis 
VGO2_NP <- subset_samples(
  VGO2_NP_all,
  !str_detect(subject_identifier, "T2|T3")
)

# # Remove taxa with 0 sum 
# VGO2_NP <- prune_taxa(taxa_sums(VGO2_NP) > 0, VGO2_NP) # 289 samples, 10892 taxa

# Identify all duplicated `respnr` values (only in NP samples)
# Extract full metadata
meta_all <- sample_data(VGO2_NP) %>% data.frame()
meta_all$SampleID <- sample_names(VGO2_NP)
# Split into NP and non-NP
meta_NP <- meta_all %>% filter(sample_type == "NP")
meta_other <- meta_all %>% filter(sample_type != "NP")
# Deduplicate only NP samples by respnr
meta_NP_dedup <- meta_NP %>%
  filter(!is.na(respnr)) %>%
  distinct(respnr, .keep_all = TRUE)
meta_NP_na <- meta_NP %>% filter(is.na(respnr))
# Combine deduplicated NP with other niches
samples_to_keep <- c(meta_NP_dedup$SampleID, meta_NP_na$SampleID, meta_other$SampleID)
# Prune phyloseq object
VGO2_NP_dedup <- prune_samples(samples_to_keep, VGO2_NP)
# Confirm
cat("Original sample count:", nsamples(VGO2_NP), "\n") # 289
cat("After deduplication of NP samples:", nsamples(VGO2_NP_dedup), "\n") # 288


#### VGO2 OP #### 
# These samples were sequenced at the UMCU. Subsequently the V3 region was trimmed by AB, and run through the RIVM DADA2 pipeline by MLO
VGO2_OP_all <- readRDS("../Input/Phyloseq/VGO2_OP/V4/VGO_OP_v4_50_220_3.rds") 
# Extract and modify sample_data
sample_data(VGO2_OP_all) <- sample_data(VGO2_OP_all) %>%
  data.frame() %>%
  mutate(OP_time = case_when(
    str_detect(sample_identifier, "T1") ~ "T1",
    str_detect(sample_identifier, "T2") ~ "T2",
    str_detect(sample_identifier, "T3") ~ "T3",
    TRUE ~ NA_character_  # fallback in case none matched
  )) %>%
  sample_data()

time_points_VGO2_OP <- sample_data(VGO2_OP_all) %>%
  data.frame() %>%
  filter(OP_time %in% c("T1", "T2", "T3")) %>%
  group_by(OP_time) %>%
  summarise(unique_participants = n_distinct(sample_identifier)) %>%
  arrange(OP_time) # T1 = 281, T2 = 21, T3 = 20

# Only keep time point T1 from VGO2 OP data
VGO2_OP <- subset_samples(
  VGO2_OP_all,
  !str_detect(sample_identifier, "T2|T3"))# 340 samples, 1592 taxa (41 removed)
# # Remove taxa with 0 sum 
# VGO2_OP <- prune_taxa(taxa_sums(VGO2_OP) > 0, VGO2_OP) # 340 samples, 1540 taxa

# create respnr column  
VGO2_OP@sam_data$respnr <- sub("T1$", "", VGO2_OP@sam_data$sample_identifier)
VGO2_OP@sam_data$respnr[!grepl("^\\d", VGO2_OP@sam_data$respnr)] <- NA
# number of VGO2 OP samples
length(unique(na.omit(VGO2_OP@sam_data$respnr))) # 281
# Additional metadata for VGO2_OP dataset
VGO2_OP_meta <- read.xlsx("../Input/Phyloseq/VGO2_OP/V4/sample_sheet_VGO_OP.xlsx")
VGO2_OP@sam_data$Sample
VGO2_OP_meta$assay_sample
# First, ensure both identifiers are character type
VGO2_OP@sam_data$Sample <- as.character(VGO2_OP@sam_data$Sample)
VGO2_OP_meta$assay_sample <- as.character(VGO2_OP_meta$assay_sample)
# Subset only relevant metadata column for merging
# Convert sample_data to data.frame and preserve rownames
sam_df <- data.frame(VGO2_OP@sam_data)
sam_df$Sample <- rownames(sam_df)  # create a 'Sample' column for merging
# Make sure meta has correct column names
meta_subset <- VGO2_OP_meta[, c("assay_sample", "pg/ul")]
colnames(meta_subset) <- c("Sample", "pg_ul")  # rename for easier handling
# Merge metadata
merged_meta <- merge(sam_df, meta_subset, by = "Sample", all.x = TRUE)
# Set rownames back using Sample column (which matches phyloseq sample names)
rownames(merged_meta) <- merged_meta$Sample
# Remove redundant 'Sample' column if needed
merged_meta$Sample <- NULL
# Assign merged data back to phyloseq
sample_data(VGO2_OP) <- sample_data(merged_meta)

# Add "niche" column to the sam_data
# Extract sample data as a data.frame
meta_OP_VGO2 <- sample_data(VGO2_OP) %>% data.frame()
meta_OP_VGO2$sample_identifier
# Create the niche column according to your rules:
meta_OP_VGO2$niche <- ifelse(
  !is.na(meta_OP_VGO2$respnr) & grepl("^\\d{5}$", meta_OP_VGO2$respnr),  # VGO3_ID exists & is 5-digit number
  "OP",
  meta_OP_VGO2$sample_identifier  # Otherwise, use sample_ID
)
# Assign updated metadata back to the phyloseq object (make sure niche is a factor)
sample_data(VGO2_OP) <- sample_data(meta_OP_VGO2)

# Identify all duplicated `respnr` values (only in OP samples)
# Extract full metadata
meta_all <- sample_data(VGO2_OP) %>% data.frame()
meta_all$SampleID <- sample_names(VGO2_OP)
# Split into NP and non-NP
meta_OP <- meta_all %>% filter(niche == "OP")
meta_other <- meta_all %>% filter(niche != "OP")
# Deduplicate only NP samples by respnr
meta_OP_dedup <- meta_OP %>%
  filter(!is.na(respnr)) %>%
  distinct(respnr, .keep_all = TRUE)
meta_OP_na <- meta_OP %>% filter(is.na(respnr))
# Combine deduplicated NP with other niches
samples_to_keep <- c(meta_OP_dedup$SampleID, meta_OP_na$SampleID, meta_other$SampleID)
# Prune phyloseq object
VGO2_OP_dedup <- prune_samples(samples_to_keep, VGO2_OP)
# Confirm
cat("Original sample count:", nsamples(VGO2_OP), "\n") # 340
cat("After deduplication of NP samples:", nsamples(VGO2_OP_dedup), "\n") # 340 i.e. none duplicated



#### VGO3 NP & OP #### 
VGO3_NPOP <- readRDS("../Input/Phyloseq/VGO3_NPOP/1_ps_raw.Rds")

# only keep control control participants - exclude GF and GP samples - keep blanks etc.
VGO3_NPOP <- subset_samples(VGO3_NPOP, !(population %in% c("GF", "GP")))
table(VGO3_NPOP@sam_data$niche) 
 #  NP   OP   BD   BP   FB ZMCB ZMCD  UMD   BB 
 # 978  989  152   17  101   87   29   26    3
table(VGO3_NPOP@sam_data$population) # CP 1941

# Exclude duplicated participants
# Extract metadata
meta <- data.frame(sample_data(VGO3_NPOP))
meta$SampleID <- sample_names(VGO3_NPOP)
# Mark duplicates: only within CP population AND niche is OP or NP
meta <- meta %>%
  mutate(dup = if_else(
    population == "CP" & niche %in% c("OP", "NP") & duplicated(paste(VGO3_ID, niche)),
    TRUE, FALSE
  ))
# Keep non-duplicates or samples outside the CP+OP/NP filter
keep_samples <- meta$SampleID[!meta$dup]
# Prune phyloseq object
VGO3_NPOP_dedup <- prune_samples(keep_samples, VGO3_NPOP)
# Optional: confirm counts
cat("Samples before:", nsamples(VGO3_NPOP), "\n") # 2397 
cat("Samples after deduplication:", nsamples(VGO3_NPOP_dedup), "\n") # 2364 


# Realised that some niche values are assigned as NP or OP but they are actually the blanks for these niches 
# Filter for rows where niche is NP or OP but VGO3_ID is NOT a 5-digit number
meta_check <- data.frame(VGO3_NPOP_dedup@sam_data)
bad_niche_assignments <- meta_check %>%
  filter(niche %in% c("NP", "OP") & !grepl("^\\d{5}$", VGO3_ID))
# Print how many problematic rows there are
cat("Number of samples with niche NP or OP but invalid VGO3_ID:", nrow(bad_niche_assignments), "\n") # 22 
# View these rows 
bad_niche_assignments # there are 2 which are not needed, the rest are blanks for the NP so need to be relabelled to be BL

# Extract sample data
meta <- data.frame(VGO3_NPOP_dedup@sam_data)
meta$SampleID <- sample_names(VGO3_NPOP_dedup)
# Remove two specific samples by sample_identifier
samples_to_remove <- meta %>%
  filter(sample_identifier %in% c("ISO_38_OP_TOP_B1_M", "ISO_38_OP_TOP_B2_M")) %>%
  pull(SampleID)
# Prune to remove those samples
VGO3_NPOP_dedup <- prune_samples(!(sample_names(VGO3_NPOP_dedup) %in% samples_to_remove), VGO3_NPOP_dedup)
# Fix niche values where VGO3_ID is not a 5-digit number but niche is NP or OP
# Re-extract updated metadata
meta <- data.frame(VGO3_NPOP_dedup@sam_data)
# Identify the rows to correct
bad_niche_idx <- which(meta$niche %in% c("NP", "OP") & !grepl("^\\d{5}$", meta$VGO3_ID))
# Convert niche to character before assignment
meta$niche <- as.character(meta$niche)
# Now assign "FB" to the bad indices
meta$niche[bad_niche_idx] <- "FB"
# Optionally convert niche back to factor
meta$niche <- factor(meta$niche)
# Assign the modified metadata back to the phyloseq object
sample_data(VGO3_NPOP_dedup) <- sample_data(meta)# Assign corrected metadata back to phyloseq object
table(VGO3_NPOP_dedup@sam_data$niche)

# How many samples have NA in 'niche' 
cat("Number of samples with NA in niche:", sum(is.na(meta$niche)), "\n") #3 have NA
# View the full rows where niche is still NA
meta_with_na_niche <- meta %>%
  filter(is.na(niche))
# Print or inspect these rows
print(meta_with_na_niche)
# Optional: view nicely in RStudio viewer
View(meta_with_na_niche)
# Extract metadata from the phyloseq object
meta <- data.frame(sample_data(VGO3_NPOP_dedup))
meta$SampleID <- sample_names(VGO3_NPOP_dedup)
# Assign "FB" to the two identified samples
fb_samples <- c("ISO_73_NP_TOP_BL_N5_1001", "ISO_73_NP_TOP_BL_N5_1005")
meta$niche[meta$sample_identifier %in% fb_samples] <- "FB"
# Identify the remaining samples that had NA in 'niche'
meta_with_na_niche <- meta %>%
  filter(is.na(niche) & !(sample_identifier %in% fb_samples))
# Set 'niche' to "unknown" for these
# Add "unknown" to the levels of niche if not already present
if (!"unknown" %in% levels(meta$niche)) {
  levels(meta$niche) <- c(levels(meta$niche), "unknown")
}
meta$niche[meta$SampleID %in% meta_with_na_niche$SampleID] <- "unknown"
# Update the phyloseq object with the revised metadata
sample_data(VGO3_NPOP_dedup) <- sample_data(meta)
# Optional: check that no NAs remain
cat("Remaining NAs in niche:", sum(is.na(sample_data(VGO3_NPOP_dedup)$niche)), "\n")


# Check remaining unique participants in NP and OP niches (population CP)
length(unique(VGO3_NPOP_dedup@sam_data$VGO3_ID[VGO3_NPOP_dedup@sam_data$niche == "NP" & VGO3_NPOP_dedup@sam_data$population == "CP"])) # 956
length(unique(VGO3_NPOP_dedup@sam_data$VGO3_ID[VGO3_NPOP_dedup@sam_data$niche == "OP" & VGO3_NPOP_dedup@sam_data$population == "CP"])) # 954

# Remove taxa with 0 sum
# VGO3_NPOP_dedup <- prune_taxa(taxa_sums(VGO3_NPOP_dedup) > 0, VGO3_NPOP_dedup) # 2434  samples, 24134 taxa

# Add 'dataset' column to each phyloseq object's sample_data
sample_data(VGO2_NP_dedup)$dataset <- "VGO2_NP"
sample_data(VGO2_OP_dedup)$dataset <- "VGO2_OP"
sample_data(VGO3_NPOP_dedup)$dataset <- "VGO3_NPOP"

# Add 'timepoint' column to each phyloseq object's sample_data
sample_data(VGO2_NP_dedup)$timepoint <- "T0"
sample_data(VGO2_OP_dedup)$timepoint <- "T0"
sample_data(VGO3_NPOP_dedup)$timepoint <- "T1"

# dna_isolation_date is simply the date, convert this to run numbers to match with DI.Run from the VGO3 dataset
sample_data(VGO2_NP_dedup)$DI.Run <- as.factor(as.numeric(factor(sample_data(VGO2_NP_dedup)$dna_isolation_date)))
# DNA isolation run is not known for VGO2_Op dataset so add it as an empty column in sam_data
sample_data(VGO2_OP_dedup)$DI.Run <- NA  # This creates logical NAs

# filter metadata in each dataset
names(VGO2_NP_dedup@sam_data) # keep: subject_identifier, dna_isolation_date, qPCR_pg_ul, respnr, dataset, timepoint
names(VGO2_OP_dedup@sam_data)# keep: sample_identifier, dataset, timepoint
names(VGO3_NPOP_dedup@sam_data) # keep: sample_identifier, DI.Run, qPCR, VGO3_ID,  dataset, niche, timepoint

# For VGO2_NP: keep subject_identifier, dna_isolation_date, qPCR_pg_ul, respnr, dataset
VGO2_NP_dedup@sam_data
sample_data(VGO2_NP_dedup) <- sample_data(VGO2_NP_dedup)[, c("subject_identifier", "qPCR_pg_ul", "respnr", "dataset", "timepoint", "sample_type", "DI.Run")]

# For VGO2_OP: keep sample_identifier, dataset
sample_data(VGO2_OP_dedup) <- sample_data(VGO2_OP_dedup)[, c("sample_identifier", "pg_ul", "respnr" ,"dataset", "timepoint", "niche", "DI.Run")]

# For VGO3_NPOP: keep DI.Run, qPCR, VGO3_ID, dataset
sample_data(VGO3_NPOP_dedup) <- sample_data(VGO3_NPOP_dedup)[, c("sample_identifier", "qPCR", "VGO3_ID", "dataset", "timepoint", "niche", "DI.Run")]

# rename some of the metadata columns to align with eachother for later merging
# VGO2_NP
colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "subject_identifier"
] <- "sample_ID"

colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "sample_type"
] <- "niche"


colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "qPCR_pg_ul"
] <- "qPCR"

colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "respnr"
] <- "VGO3_ID"

# VGO2_OP
colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "sample_identifier"
] <- "sample_ID"

colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "respnr"
] <- "VGO3_ID"

colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "pg_ul"
] <- "qPCR"

# VGO3_NP
colnames(sample_data(VGO3_NPOP_dedup))[
  colnames(sample_data(VGO3_NPOP_dedup)) == "sample_identifier"
] <- "sample_ID"



####  Check longitudinal data (VGO2 and VGO3) #### 
## NP 
VGO3_NP <- subset_samples(VGO3_NPOP_dedup, niche == "NP")
# VGO3_NP <- prune_taxa(taxa_sums(VGO3_NP) > 0, VGO3_NP) # 957 samples, 11053 taxa

# Extract IDs from sample_data
ids_VGO2 <- sample_data(VGO2_NP_dedup)$VGO3_ID
ids_VGO3 <- sample_data(VGO3_NP)$VGO3_ID

# Find intersecting IDs
shared_ids_NP <- intersect(ids_VGO2, ids_VGO3)
# Number of shared individuals
length(shared_ids_NP) # 112 longitudinal NP samples


## OP 
VGO3_OP <- subset_samples(VGO3_NPOP, niche != "NP")
# VGO3_OP <- prune_taxa(taxa_sums(VGO3_OP) > 0, VGO3_OP) # 1404 samples, 12902 taxa

# Extract IDs from sample_data
ids_VGO2 <- sample_data(VGO2_OP_dedup)$VGO3_ID
ids_VGO3 <- sample_data(VGO3_OP)$VGO3_ID
# Find intersecting IDs
shared_ids_OP <- intersect(ids_VGO2, ids_VGO3)
# Number of shared individuals
length(shared_ids_OP) # 141 longitudinal OP samples
```
# Merge phyloseqs and taxonomy naming 
```{r}
head(taxa_names(VGO2_NP_dedup)) # ASVs named already e.g. "Corynebacterium_1", "Moraxella_2"  etc.
head(taxa_names(VGO2_OP_dedup)) # Named by the sequence
head(taxa_names(VGO3_NPOP_dedup))# ASVs named already e.g. "Dolosigranulum_pigrum_1" "Staphylococcus_2"

ntaxa(VGO2_NP_dedup) # 10892
ntaxa(VGO2_OP_dedup) # 1540
ntaxa(VGO3_NPOP_dedup) # 20687
nsamples(VGO2_NP_dedup) # 289
nsamples(VGO2_OP_dedup) # 340
nsamples(VGO3_NPOP_dedup) # 2039

# Fix VGO2_OP so that the refseq() slot has the sequences and taxa_names are the ASV names
# Get DNA sequences from current taxa_names
seqs <- taxa_names(VGO2_OP_dedup)
# Create ASV names
asv_names <- paste0("ASV", seq_along(seqs))
# Assign ASV names to the phyloseq object
taxa_names(VGO2_OP_dedup) <- asv_names
# Create DNAStringSet with ASV names
dna_seqs <- Biostrings::DNAStringSet(seqs)
names(dna_seqs) <- asv_names
# Add refseq slot
VGO2_OP_dedup <- merge_phyloseq(VGO2_OP_dedup, dna_seqs)
# VGO2_OP_dedup <- prune_taxa(taxa_sums(VGO2_OP_dedup) > 0, VGO2_OP_dedup)
# VGO2_OP_dedup <- prune_samples(sample_sums(VGO2_OP_dedup) > 0, VGO2_OP_dedup)

# Rename ASVs to Sequence Level
VGO2_NP_seq <- rename_ASVs(VGO2_NP_dedup)
VGO2_OP_seq <- rename_ASVs(VGO2_OP_dedup)
VGO3_NPOP_seq <- rename_ASVs(VGO3_NPOP_dedup)

# Check that tax_table is annotated in the same way
colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)

# Change all to lower case
colnames(VGO2_OP_seq@tax_table) <- tolower(colnames(VGO2_OP_seq@tax_table))
colnames(VGO2_NP_seq@tax_table) <- tolower(colnames(VGO2_NP_seq@tax_table))
colnames(VGO3_NPOP_seq@tax_table) <- tolower(colnames(VGO3_NPOP_seq@tax_table))

colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)


table(VGO2_NP_seq@sam_data$niche)
table(VGO2_OP_seq@sam_data$niche)
table(VGO3_NPOP_dedup@sam_data$niche)

# Remove 'Species_one' column from VGO3 phyloseqs
# For VGO3_NP_seq
tax_table(VGO3_NPOP_dedup) <- tax_table(VGO3_NPOP_dedup)[, !colnames(tax_table(VGO3_NPOP_dedup)) %in% c("species_one", "asv")]
colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)


# Check for mismatches in sam_data before merging phyloseqs
# Sam_data variables between objects:
all.equal(colnames(sample_data(VGO2_NP_seq)),
          colnames(sample_data(VGO2_OP_seq)),
          colnames(sample_data(VGO3_NPOP_seq)))
# Get the column names for each phyloseq object
cols_np   <- colnames(sample_data(VGO2_NP_seq))
cols_op   <- colnames(sample_data(VGO2_OP_seq))
cols_vgo3 <- colnames(sample_data(VGO3_NPOP_seq))
# Compare: what's in VGO2 NP but not in VGO2 OP
setdiff(cols_np, cols_op) # nothing
# Compare: what's in VGO2 OP but not in VGO2 NP
setdiff(cols_op, cols_np)
# Compare: what's in VGO2 NP but not in VGO3 OP/NP
setdiff(cols_np, cols_vgo3)

# Get classes for each column in each phyloseq object
classes_np <- sapply(sample_data(VGO2_NP_seq), class)
classes_op <- sapply(sample_data(VGO2_OP_seq), class)
classes_vgo3 <- sapply(sample_data(VGO3_NPOP_seq), class)

# Combine into a comparison table
compare_classes <- data.frame(
  column_name = names(classes_np),
  NP_class    = unname(classes_np),
  OP_class    = unname(classes_op[names(classes_np)]),
  VGO3_class  = unname(classes_vgo3[names(classes_np)])
)

# Convert the mismatches 
sample_data(VGO2_NP_seq)$VGO3_ID <- as.character(sample_data(VGO2_NP_seq)$VGO3_ID)
sample_data(VGO3_NPOP_seq)$niche <- as.character(sample_data(VGO3_NPOP_seq)$niche)
VGO3_NPOP_seq@sam_data$niche

# Merge datasets now that all variables across the phyloseqs are aligned and the same data class
ps_combined <- merge_phyloseq(
  VGO2_NP_seq,
  VGO2_OP_seq,
  VGO3_NPOP_seq
  )

#  Reorder ASVs by Abundance
ps_reordered <- tax_sum_reorder(ps_combined)

# Rename ASVs Based on Taxonomy
ps <- create_phylo_custom(ps_reordered)
table(ps@sam_data$niche)
# Check tax table
head(ps@tax_table)
head(ps@otu_table)
ps@sam_data
ps
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 27827 taxa and 2660 samples ]
# sample_data() Sample Data:       [ 2660 samples by 6 sample variables ]
# tax_table()   Taxonomy Table:    [ 27827 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 27827 reference sequences ]
```

# Join participant metadata/spirometry with microbiome data 
```{r}
# Load metadata
metadata_VGO2 <- read.csv("../Output/Participant_data/Datasets/VGO2_data.csv")
metadata_VGO3 <- read.csv("../Output/Participant_data/Datasets/VGO3_data.csv")
# Combine metadata into one table (optional, only if needed)
metadata_all <- bind_rows(metadata_VGO2, metadata_VGO3)
colnames(metadata_all)[colnames(metadata_all) == "ID"] <- "VGO3_ID"

# Extract sample_data from phyloseq
ps_meta <- data.frame(sample_data(ps))
ps_meta$SampleName <- rownames(ps_meta)  # keep rownames for reassignment

# Match correct metadata to each sample based on ID and timepoint
ps_meta_joined <- ps_meta %>%
  mutate(VGO3_ID = as.character(VGO3_ID)) %>%
  left_join(
    metadata_all %>% mutate(VGO3_ID = as.character(VGO3_ID)),
    by = c("VGO3_ID", "timepoint")
  )

# Remove column X from dataframe
ps_meta_joined$X <- NULL

# Blank samples (with no VGO3_ID) are missing 'study' variable, add this 
# Fill in missing 'study' for blanks based on timepoint
ps_meta_joined <- ps_meta_joined %>%
  mutate(study = ifelse(is.na(study) & timepoint == "T0", "VGO2",
                 ifelse(is.na(study) & timepoint == "T1", "VGO3", study)))

table(VGO2_OP_dedup@sam_data$niche)
table(VGO2_NP_dedup@sam_data$niche)
table(VGO3_NPOP_dedup@sam_data$niche)

# Check niche assignment again
table(ps_meta_joined$dataset)
table(ps_meta_joined$timepoint)
table(ps_meta_joined$niche)

# View the first few rows
sum(is.na(ps_meta_joined$dataset)) # 0
sum(is.na(ps_meta_joined$timepoint)) # 0
sum(is.na(ps_meta_joined$niche)) # 0 

# Replace sample_data with updated metadata (ps_meta_joined)
rownames(ps_meta_joined) <- ps_meta_joined$SampleName
ps2 <- ps
sample_data(ps2) <- sample_data(ps_meta_joined)

# Inspect
ps2@sam_data # Sample Data: [2566 samples by 88 sample variables]:
table(ps2@sam_data$niche)
table(ps2@sam_data$dataset)
table(ps2@sam_data$timepoint)

# Save ps 
saveRDS(ps2, "../Output/Phyloseq/ps2.rds")
```
# Composition plots per dataset - before pre-processing
```{r}
# Subset ps2 for dataset == "VGO3_NP"
ps_vgo2_np <- subset_samples(ps2, dataset == "VGO2_NP" & niche == "NP")
ps_vgo2_np_filt <- prune_taxa(taxa_sums(ps_vgo2_np) > 0, ps_vgo2_np)
ps_vgo2_np_filt <- prune_samples(sample_sums(ps_vgo2_np_filt) > 0, ps_vgo2_np_filt)

# stacked bar plot
VGO2_NP_stackedbar <- create_ordered_bar_n(ps_vgo2_np, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 NP microbiome composition per sample")
# Save plot
ggsave("../Output/Microbiome_descriptives/Before_preprocessing_composition/VGO2_NP_barplot.svg", VGO2_NP_stackedbar, width = 5, height = 6)


# Subset ps2 for dataset == "VGO3_NP"
ps_vgo2_op <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "OP")
ps_vgo2_op_filt <- prune_taxa(taxa_sums(ps_vgo2_op) > 0, ps_vgo2_op)
ps_vgo2_op_filt <- prune_samples(sample_sums(ps_vgo2_op_filt) > 0, ps_vgo2_op_filt)

# stacked bar plot
VGO2_OP_stackedbar <- create_ordered_bar_n(ps_vgo2_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 OP microbiome composition per sample")

# Save plot
ggsave("../Output/Microbiome_descriptives/Before_preprocessing_composition/VGO2_OP_barplot.svg", VGO2_OP_stackedbar, width = 5, height = 6)



# Subset ps2 for dataset == "VGO3_NP"
ps_vgo3_np <- subset_samples(ps2, dataset == "VGO3_NPOP" & niche == "NP")
ps_vgo3_np_filt <- prune_taxa(taxa_sums(ps_vgo3_np) > 0, ps_vgo3_np)
ps_vgo3_np_filt <- prune_samples(sample_sums(ps_vgo3_np_filt) > 0, ps_vgo3_np_filt)

# stacked bar plot
VGO3_NP_stackedbar <- create_ordered_bar_n(ps_vgo3_np_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 NP microbiome composition per sample")
# Save plot
ggsave("../Output/Microbiome_descriptives/Before_preprocessing_composition/VGO3_NP_barplot.svg", VGO3_NP_stackedbar, width = 5, height = 6)


# Subset ps2 for dataset == "VGO3_NP"
ps2@sam_data$dataset
ps_vgo3_op <- subset_samples(ps2, dataset == "VGO3_NPOP" & niche == "OP")
ps_vgo3_op_filt <- prune_taxa(taxa_sums(ps_vgo3_op) > 0, ps_vgo3_op)
ps_vgo3_op_filt <- prune_samples(sample_sums(ps_vgo3_op_filt) > 0, ps_vgo3_op_filt)

# stacked bar plot
VGO3_OP_stackedbar <- create_ordered_bar_n(ps_vgo3_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 OP microbiome composition per sample")
# Save plot
ggsave("../Output/Microbiome_descriptives/Before_preprocessing_composition/VGO3_OP_barplot.svg", VGO3_OP_stackedbar, width = 5, height = 6)



#### Combine plots: NP on top row, OP on bottom row ####
composition_combined <- (VGO2_NP_stackedbar | VGO3_NP_stackedbar) / 
                        (VGO2_OP_stackedbar | VGO3_OP_stackedbar) +
                        plot_annotation(tag_levels = 'A')  # Optional: add panel tags A, B, C, D

print(composition_combined)

# Save combined figure
ggsave("../Output/Microbiome_descriptives/Before_preprocessing_composition/Combined_composition_barplots.svg",
       composition_combined, width = 12, height = 12)

```

# Pre-processing
## Visualise library size
```{r}
df <- meta_to_df(ps2)
df$LibrarySize <- sample_sums(ps2)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))

# Library sizes - all samples including blanks
libsize_niche <- ggplot(data = df, aes(x=Index, y=LibrarySize, color=niche)) + geom_point()
ggsave("../Output/Preprocessing/librarysize_perniche.svg", libsize_niche)

# Library size of NP and OP samples only 
df_filtered <- df %>% filter(niche %in% c("NP", "OP"))
libsize_niche_NPOP <- ggplot(data = df_filtered, aes(x = Index, y = LibrarySize, color = niche)) +
  geom_point()
ggsave("../Output/Preprocessing/librarysize_NPvsOP.svg", libsize_niche_NPOP, width = 6, height = 5)

```
## Visualise bacterial density 
```{r}
bacterial_density_NP_OP <- df %>%
  filter(niche %in% c("NP", "OP")) %>%
  ggplot(aes(x = niche, y = qPCR)) +
  geom_boxplot(alpha = 0.4, fill = "white") +
  geom_jitter(aes(color = niche), width = 0.2, size = 1, alpha = 0.7) +
  labs(x = "Niche", y = "Bacterial density (pg/µl)", color = "Niche") +
  scale_y_continuous(trans = 'log10',
                     breaks = scales::trans_breaks('log10', function(x) 10^x),
                     labels = scales::trans_format('log10', scales::math_format(10^.x)))

ggsave("../Output/Preprocessing/bacterial_density_NPvsOP.svg", bacterial_density_NP_OP)

```
# Decontaminating the microbiome data
## Preparing data for decontamination
```{r}
# We want to decontaminate per sequencing group - i.e. VGO2_NP, VGO2_OP, VGO3_NP and VGO3_OP (VGO3 data was sequenced/processed together for NP and OP)
# Define decontamination groups 
sum(is.na(ps2@sam_data$dataset))# No NAs for dataset

unique(sample_data(ps2)$dataset) #  "VGO2_NP"   "VGO2_OP"   "VGO3_NPOP"
sum(is.na(sample_data(ps2)$decontam_group)) # 0 

sample_data(ps2)$decontam_group <- case_when(
  sample_data(ps2)$dataset == "VGO3_NPOP" ~ "VGO3_NPOP", 
  sample_data(ps2)$dataset == "VGO2_NP" ~ "VGO2_NP",
  sample_data(ps2)$dataset == "VGO2_OP" ~ "VGO2_OP",
  TRUE ~ NA_character_
)

# Count zero-sum samples *before* the split
zero_samples <- sample_sums(ps2) == 0
table(zero_samples, sample_data(ps2)$decontam_group)

table(sample_data(ps2)$decontam_group)
# VGO2_NP   VGO2_OP  VGO3_combined 
# 288       336      2036 

# Split phyloseq by group
# Get the unique group names
groups <- unique(na.omit(sample_data(ps2)$decontam_group))

# Use lapply to avoid `.x` evaluation issues
ps2_groups <- setNames(lapply(groups, function(g) {
  keep_samples <- rownames(sample_data(ps2))[sample_data(ps2)$decontam_group == g]
  prune_samples(keep_samples, ps2)
}), groups)

# Create and assign objects with custom names
for (group in names(ps2_groups)) {
  assign(paste0(group, "_ps_for_decontam"), ps2_groups[[group]])
}

# Save each phyloseq to an individual object
list2env(ps2_groups, envir = .GlobalEnv)

# These are the phyloseqs for decontamination
VGO2_NP_ps_for_decontam
VGO2_OP_ps_for_decontam
VGO3_NPOP_ps_for_decontam

table(VGO2_NP_ps_for_decontam@sam_data$niche) # Niche = Blank, NP, Zymo Mock
table(VGO2_OP_ps_for_decontam@sam_data$niche) # EBL, NEG, NTC, OP, SBL, VBL, ZMCB
table(VGO3_NPOP_ps_for_decontam@sam_data$niche) # BB, BD, BL, BP, FB, NP, OP, UMD, ZMCB, ZMCD

# ZMCD	Zymo mock DNA (miseq PCR)		
# ZMCB	Zymo mock Bacteria (extraction)		
# AMCD	ATCC mock DNA (miseq PCR)		
# UMD	UMCU low density mock DNA (miseq PCR)		
# MSS	Mixed (fecal/NP/OP…) sample control 		
# MSD	Mixed (fecal/NP/OP…) sample control DNA		
# BD	Blank from DNA extraction		
# BP	Blank from PCR		

# For decontam we use the following as negative controls: 
## Blank (n=47)
## For OP it is not clear 
## BD for the VGO3 dataset (NP and OP) (n=152)

# # Check composition of the unknown samples in VGO2_OP: "EBL", "NEG", "NTC", "OP", "SBL", "VBL", "ZMCB"
# # Check OP first
# ps_vgo2_op_op <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "OP")
# ps_vgo2_op_op <- prune_taxa(taxa_sums(ps_vgo2_op_op) > 0, ps_vgo2_op_op)
# ps_vgo2_op_op <- prune_samples(sample_sums(ps_vgo2_op_op) > 0, ps_vgo2_op_op)
# 
# # stacked bar plot
# VGO2_OP_stackedbar <- create_ordered_bar_n(ps_vgo2_op_op, 15) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP microbiome composition per sample")
# VGO2_OP_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_stackedbar.svg", VGO2_OP_stackedbar)
# 
# # NTC
# ps_vgo2_op_ntc <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "NTC")
# ps_vgo2_op_ntc <- prune_taxa(taxa_sums(ps_vgo2_op_ntc) > 0, ps_vgo2_op_ntc)
# ps_vgo2_op_ntc <- prune_samples(sample_sums(ps_vgo2_op_ntc) > 0, ps_vgo2_op_ntc)
# 
# VGO2_OP_NTC_stackedbar <- create_ordered_bar_n(ps_vgo2_op_ntc, 12) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP NTC microbiome composition per sample")
# VGO2_OP_NTC_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_NTC_stackedbar.svg", VGO2_OP_NTC_stackedbar)
# 
# # SBL samples
# ps_vgo2_op_sbl <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "SBL")
# ps_vgo2_op_sbl <- prune_taxa(taxa_sums(ps_vgo2_op_sbl) > 0, ps_vgo2_op_sbl)
# ps_vgo2_op_sbl <- prune_samples(sample_sums(ps_vgo2_op_sbl) > 0, ps_vgo2_op_sbl)
# 
# VGO2_OP_SBL_stackedbar <- create_ordered_bar_n(ps_vgo2_op_sbl, 15) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP SBL microbiome composition per sample")
# VGO2_OP_SBL_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_SBL_stackedbar.svg", VGO2_OP_SBL_stackedbar)
# 
# # VBL samples
# ps_vgo2_op_vbl <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "VBL")
# ps_vgo2_op_vbl <- prune_taxa(taxa_sums(ps_vgo2_op_vbl) > 0, ps_vgo2_op_vbl)
# ps_vgo2_op_vbl <- prune_samples(sample_sums(ps_vgo2_op_vbl) > 0, ps_vgo2_op_vbl)
# 
# VGO2_OP_VBL_stackedbar <- create_ordered_bar_n(ps_vgo2_op_vbl, 15) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP VBL microbiome composition per sample")
# VGO2_OP_VBL_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_VBL_stackedbar.svg", VGO2_OP_VBL_stackedbar)
# 
# 
# # ZMCB samples
# ps_vgo2_op_zmcb <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "ZMCB")
# ps_vgo2_op_zmcb <- prune_taxa(taxa_sums(ps_vgo2_op_zmcb) > 0, ps_vgo2_op_zmcb)
# ps_vgo2_op_zmcb <- prune_samples(sample_sums(ps_vgo2_op_zmcb) > 0, ps_vgo2_op_zmcb)
# 
# VGO2_OP_ZMCB_stackedbar <- create_ordered_bar_n(ps_vgo2_op_zmcb, 8) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP ZMCB microbiome composition per sample")
# VGO2_OP_ZMCB_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_ZMCB_stackedbar.svg", VGO2_OP_ZMCB_stackedbar)
# 
# # EBL samples
# ps_vgo2_op_ebl <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "EBL")
# ps_vgo2_op_ebl <- prune_taxa(taxa_sums(ps_vgo2_op_ebl) > 0, ps_vgo2_op_ebl)
# ps_vgo2_op_ebl <- prune_samples(sample_sums(ps_vgo2_op_ebl) > 0, ps_vgo2_op_ebl)
# 
# VGO2_OP_EBL_stackedbar <- create_ordered_bar_n(ps_vgo2_op_ebl, 15) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP EBL microbiome composition per sample")
# VGO2_OP_EBL_stackedbar
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/VGO2_OP_EBL_stackedbar.svg", VGO2_OP_EBL_stackedbar)

# NEG (there are <2 NEG samples therefore cannot make plot)
# ps_vgo2_op_neg <- subset_samples(ps2, dataset == "VGO2_OP" & niche == "NEG")
# ps_vgo2_op_neg <- prune_taxa(taxa_sums(ps_vgo2_op_neg) > 0, ps_vgo2_op_neg)
# ps_vgo2_op_neg <- prune_samples(sample_sums(ps_vgo2_op_neg) > 0, ps_vgo2_op_neg)
#
# VGO2_OP_NEG_stackedbar <- create_ordered_bar_n(ps_vgo2_op_neg, 15) +
#   theme(axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   geom_bar(stat = "identity", width = 1) +
#   labs(title = "VGO2 OP NEG microbiome composition per sample")
# VGO2_OP_NEG_stackedbar

# # Side by side plots to determine whihc blanks for OP 
# # Combine plots into a single figure
# combined_plot <- (
#   VGO2_OP_stackedbar + 
#   VGO2_OP_NTC_stackedbar + 
#   VGO2_OP_SBL_stackedbar +
#   VGO2_OP_VBL_stackedbar +
#   VGO2_OP_ZMCB_stackedbar +
#   VGO2_OP_EBL_stackedbar
# ) + 
#   plot_layout(ncol = 3) +  # Adjust to ncol = 2 or 3 depending on your screen size
#   plot_annotation(title = "VGO2_OP Microbiome Composition by Niche")
# 
# # Display
# combined_plot
# ggsave("../Output/Decontamination/VGO2_OP_blankcompositions/Blanks_all_VGO2_OP.svg", combined_plot, width = 20, height= 10)

# Use the following samples as the blanks in the decontam analysis for VGO2 OP sequencing: NTC, SBL, VBL, EBL 

# Therefore, 
# For decontam we use the following as negative controls: 
## Blank (n=47)
## NTC, SBL, VBL, EBL (n = 54)
## BL (n=22), BD (n=1), BP (n=1) and FB for VGO3NPOP dataset

sample_data(VGO2_NP_ps_for_decontam)$is_neg <- sample_data(VGO2_NP_ps_for_decontam)$niche == "Blank"
sample_data(VGO2_OP_ps_for_decontam)$is_neg <- sample_data(VGO2_OP_ps_for_decontam)$niche %in% c("SBL", "VBL", "EBL")
sample_data(VGO3_NPOP_ps_for_decontam)$is_neg <- sample_data(VGO3_NPOP_ps_for_decontam)$niche == "BD"


# NP samples were in DI.Run = 39-74
# OP samples were in DI.Run = 1,2, 5-8, 10-12, 33-38, 75-91
## Therefore we only want to select the BD samples from these run to do decontam per niche

# Select BD samples for NP decontam
np_DI_runs <- 39:74
# Subset to NP and the correct BD runs
VGO3_NP_ps_for_decontam <- subset_samples(
  VGO3_NPOP_ps_for_decontam,
  (niche == "NP") | (niche == "BD" & DI.Run %in% np_DI_runs)
)
# Prune taxa with 0 total counts
VGO3_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_NP_ps_for_decontam) > 0, VGO3_NP_ps_for_decontam)

# Define BD runs for OP
op_DI_runs <- c(1, 2, 5:8, 10:12, 33:38, 75:91)
# Subset to OP and the correct BD runs
VGO3_OP_ps_for_decontam <- subset_samples(
  VGO3_NPOP_ps_for_decontam,
  (niche == "OP") | (niche == "BD" & DI.Run %in% op_DI_runs)
)
# Prune taxa with 0 total counts
VGO3_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_OP_ps_for_decontam) > 0, VGO3_OP_ps_for_decontam)


# Remove mocks & other unnecessary samples from the datasets
table(VGO2_NP_ps_for_decontam@sam_data$niche)
table(VGO2_OP_ps_for_decontam@sam_data$niche)
table(VGO3_NP_ps_for_decontam@sam_data$niche)
table(VGO3_OP_ps_for_decontam@sam_data$niche)

VGO2_NP_ps_for_decontam <- subset_samples(VGO2_NP_ps_for_decontam, niche != "Zymo Mock") %>% 
  subset_samples(., qPCR != 0) # 269 samples
table(VGO2_NP_ps_for_decontam@sam_data$niche) # blank = 47, NP = 222

VGO2_OP_ps_for_decontam <- subset_samples(VGO2_OP_ps_for_decontam, niche != "NEG") %>%  subset_samples(., niche != "NTC") %>%  subset_samples(., niche != "ZMCB") %>%
  subset_samples(., qPCR != 0) # 313   samples 
table(VGO2_OP_ps_for_decontam@sam_data$niche) # EBL  = 16, OP = 281, SBL= 8, VBL = 8

VGO3_NP_ps_for_decontam <- subset_samples(VGO3_NP_ps_for_decontam, niche != "BB") %>%  subset_samples(., niche != "BP") %>% subset_samples(., niche != "FB") %>% subset_samples(., niche != "UMD") %>%  subset_samples(., niche != "ZMCB") %>% subset_samples(., niche != "ZMCD") %>% subset_samples(., niche != "unknown") %>%  
  subset_samples(., qPCR != 0) # 1026  samples
table(VGO3_NP_ps_for_decontam@sam_data$niche)

VGO3_OP_ps_for_decontam <- subset_samples(VGO3_OP_ps_for_decontam, niche != "BB") %>%  subset_samples(., niche != "BP") %>% subset_samples(., niche != "FB") %>% subset_samples(., niche != "UMD") %>%  subset_samples(., niche != "ZMCB") %>% subset_samples(., niche != "ZMCD") %>% subset_samples(., niche != "unknown") %>%  
  subset_samples(., qPCR != 0) # 991 samples
table(VGO3_OP_ps_for_decontam@sam_data$niche)
```
## Run decontam combined method
```{r}
rownames(VGO2_NP_ps_for_decontam@tax_table)
rownames(VGO2_OP_ps_for_decontam@tax_table)
rownames(VGO3_NPOP_ps_for_decontam@tax_table)

# VGO2 NP 
VGO2_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO2_NP_ps_for_decontam) > 0, VGO2_NP_ps_for_decontam)
contaminants_VGO2_NP <- isContaminant(VGO2_NP_ps_for_decontam, method="combined", conc="qPCR", neg="is_neg")
contaminants_VGO2_NP <- subset(contaminants_VGO2_NP, contaminant == TRUE)
dim(contaminants_VGO2_NP)[1] # 64 identified in the NP
write.csv(contaminants_VGO2_NP, "../Output/Decontamination/Decontam_combined_method/VGO2_NP_contaminants.csv")

# VGO2 OP
VGO2_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO2_OP_ps_for_decontam) > 0, VGO2_OP_ps_for_decontam)
contaminants_VGO2_OP <- isContaminant(VGO2_OP_ps_for_decontam, method="combined", conc="qPCR", neg="is_neg")
contaminants_VGO2_OP <- subset(contaminants_VGO2_OP, contaminant == TRUE)
dim(contaminants_VGO2_OP)[1] # 24 identified in the NP
write.csv(contaminants_VGO2_OP, "../Output/Decontamination/Decontam_combined_method/VGO2_OP_contaminants.csv")

# VGO3 NP
# Prune taxa with total abundance of 0 across samples
VGO3_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_NP_ps_for_decontam) > 0, VGO3_NP_ps_for_decontam)
# Run decontam on NP subset (using method="combined")
contaminants_VGO3_NP <- isContaminant(VGO3_NP_ps_for_decontam, method = "combined", conc = "qPCR", neg = "is_neg")
contaminants_VGO3_NP <- subset(contaminants_VGO3_NP, contaminant == TRUE)
dim(contaminants_VGO3_NP)[1] # 110 identified in the NP
write.csv(contaminants_VGO3_NP, "../Output/Decontamination/Decontam_combined_method/VGO3_NP_contaminants.csv")

# VGO3 OP
# Prune taxa with total abundance of 0 across samples
VGO3_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_OP_ps_for_decontam) > 0, VGO3_OP_ps_for_decontam)
# Run decontam on NP subset (using method="combined")
contaminants_VGO3_OP <- isContaminant(VGO3_OP_ps_for_decontam, method = "combined", conc = "qPCR", neg = "is_neg")
contaminants_VGO3_OP <- subset(contaminants_VGO3_OP, contaminant == TRUE)
dim(contaminants_VGO3_OP)[1] # 88 identified in the OP
write.csv(contaminants_VGO3_OP, "../Output/Decontamination/Decontam_combined_method/VGO3_OP_contaminants.csv")
```
## Kruskall-Wallis across DNA isolation runs 
We implement a Kruskal-Wallis test to see which ASVs show discrepancies across the DNA isolation runs. These differences might indicate contamination, so the Kruskal-Wallis test is applied to each ASV to assess whether its relative abundance (RA) differs significantly across the DI.Run groups N.B. DI run groups are only available for VGO2 NP and VGO3 NP/OP dataset. 
```{r}
# We only know the DI.Run for VGO2 NP and VGO3 datasets
VGO2_NP_ps_for_decontam  

VGO2_NP_filt_RA_1 <- VGO2_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%
# A total of 792 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 32376 ASVs excluded).
# A total of 412 ASVs were found to be present at or above a level of confident detection (0.01% relative abundance) in at least 15 samples (n = 380 ASVs excluded).

VGO2_NP_filt_RA_2 <- VGO2_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%
# A total of 792 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 32376 ASVs excluded).
# A total of 27 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 40 samples (n = 765 ASVs excluded).

VGO2_NP_filt_RA_1 <- prune_taxa(!taxa_names(VGO2_NP_filt_RA_1) %in% taxa_names(VGO2_NP_filt_RA_2), VGO2_NP_filt_RA_1) # Exclude the high-abundance/core taxa from VGO2_NP_filt_RA_1

# Reshape phyloseq 
df_VGO2 <- ps_to_df(VGO2_NP_filt_RA_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO2_NP_filt_RA_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_VGO2))

# Select DI.Run and ASV columns
df_VGO2 <- df_VGO2 %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_VGO2_pivot <- pivot_longer(df_VGO2, cols = existing_asvs,
                             names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results <- df_VGO2_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results$contaminant <- ifelse(kruskal_results$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo2_np <- subset(kruskal_results, contaminant == TRUE)
KW_contams_vgo2_np # 94 detected 
KW_contams_vgo2_np$ASV

# Filter for significant ASVs
KW_contams_VGO2NP <- ggplot(subset(df_VGO2_pivot, ASV %in% KW_contams_vgo2_np$ASV), 
            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Decontamination/Kruskall_Wallis/KW_VGO2NP.svg", KW_contams_VGO2NP, width = 20, height = 20)


#### VGO3 NP ####
VGO3_NP_ps_for_decontam_1 <- VGO3_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%

VGO3_NP_ps_for_decontam_2 <- VGO3_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%

VGO3_NP_ps_for_decontam_1 <- prune_taxa(!taxa_names(VGO3_NP_ps_for_decontam_1) %in% taxa_names(VGO3_NP_ps_for_decontam_2), VGO3_NP_ps_for_decontam_1) # Exclude the high-abundance/core taxa from ps_vgo3_np_for_decontam_1

# Reshape phyloseq 
df_VGO3_np <- ps_to_df(VGO3_NP_ps_for_decontam_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO3_NP_ps_for_decontam_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_VGO3_np))

# Select DI.Run and ASV columns
df_VGO3_np <- df_VGO3_np %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_VGO3_np_pivot <- pivot_longer(df_VGO3_np, cols = existing_asvs,
                             names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results_vgo3_np <- df_VGO3_np_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results_vgo3_np$contaminant <- ifelse(kruskal_results_vgo3_np$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo3_np <- subset(kruskal_results_vgo3_np, contaminant == TRUE)
KW_contams_vgo3_np # 246 detected 
KW_contams_vgo3_np$ASV

# Filter for significant ASVs
KW_contams_VGO3NP <- ggplot(subset(df_VGO3_np_pivot, ASV %in% KW_contams_vgo3_np$ASV), 
            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Decontamination/Kruskall_Wallis/KW_contams_VGO3NP.svg", KW_contams_VGO3NP, width = 25, height = 25)
ggsave("../Output/Decontamination/Kruskall_Wallis/KW_contams_VGO3NP.png", KW_contams_VGO3NP, width = 25, height = 25)


#### VGO3 OP ####
VGO3_OP_ps_for_decontam_1 <- VGO3_OP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%

VGO3_OP_ps_for_decontam_2 <- VGO3_OP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%

VGO3_OP_ps_for_decontam_1 <- prune_taxa(!taxa_names(VGO3_OP_ps_for_decontam_1) %in% taxa_names(VGO3_OP_ps_for_decontam_2), VGO3_OP_ps_for_decontam_1) # Exclude the high-abundance/core taxa from ps_vgo3_op_for_decontam_1

# Reshape phyloseq 
df_VGO3_op <- ps_to_df(VGO3_OP_ps_for_decontam_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO3_OP_ps_for_decontam_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_VGO3_op))

# Select DI.Run and ASV columns
df_VGO3_op <- df_VGO3_op %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_VGO3_op_pivot <- pivot_longer(df_VGO3_op, cols = existing_asvs,
                                 names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results_vgo3_op <- df_VGO3_op_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results_vgo3_op$contaminant <- ifelse(kruskal_results_vgo3_op$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo3_op <- subset(kruskal_results_vgo3_op, contaminant == TRUE)
KW_contams_vgo3_op #125 detected 
KW_contams_vgo3_op$ASV

# Filter for significant ASVs
KW_contams_VGO3OP <- ggplot(subset(df_VGO3_op_pivot, ASV %in% KW_contams_vgo3_op$ASV), 
                            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Decontamination/Kruskall_Wallis/KW_contams_VGO3OP.svg", KW_contams_VGO3OP, width = 20, height = 20)
```
## Combine contaminants from both methods per dataset
```{r}
# Check which contaminants are the same across decontam method and kruskall wallis
# Extract ASV names from decontam (rownames)
decontam_VGO2_NP_ASVs <- rownames(contaminants_VGO2_NP)
decontam_VGO2_OP_ASVs <- rownames(contaminants_VGO2_OP)
decontam_VGO3_NP_ASVs <- rownames(contaminants_VGO3_NP)
decontam_VGO3_OP_ASVs <- rownames(contaminants_VGO3_OP)

# Extract ASV names from Kruskal-Wallis results
KW_VGO2_NP_ASVs <- KW_contams_vgo2_np$ASV
KW_VGO3_NP_ASVs <- KW_contams_vgo3_np$ASV
KW_VGO3_OP_ASVs <- KW_contams_vgo3_op$ASV

# Function to get counts of overlaps and uniques
compare_contaminants <- function(decontam_asvs, kw_asvs) {
  decontam_set <- unique(decontam_asvs)
  kw_set <- unique(kw_asvs)
  
  shared <- intersect(decontam_set, kw_set)
  decontam_only <- setdiff(decontam_set, kw_set)
  kw_only <- setdiff(kw_set, decontam_set)
  
  list(
    shared_ASVs = shared,
    shared_count = length(shared),
    decontam_only_count = length(decontam_only),
    kw_only_count = length(kw_only)
  )
}
# Comparisons per niche/dataset
comparison_VGO2_NP <- compare_contaminants(decontam_VGO2_NP_ASVs, KW_VGO2_NP_ASVs)
# NO KW test for VGO2 OP
comparison_VGO3_NP <- compare_contaminants(decontam_VGO3_NP_ASVs, KW_VGO3_NP_ASVs)
comparison_VGO3_OP <- compare_contaminants(decontam_VGO3_OP_ASVs, KW_VGO3_OP_ASVs)

comparison_VGO2_NP
comparison_VGO3_NP
comparison_VGO3_OP


# Combine the contaminants identified by decontam with those detected using Kruskall-Wallis test
# VGO2_NP: both decontam and KW
contaminants_VGO2_NP_combined <- unique(c(decontam_VGO2_NP_ASVs, KW_VGO2_NP_ASVs))
length(contaminants_VGO2_NP_combined)# 151
# VGO2_OP: decontam only
contaminants_VGO2_OP_combined <- unique(decontam_VGO2_OP_ASVs)
length(contaminants_VGO2_OP_combined) # 24
# VGO3_NP: both decontam and KW
contaminants_VGO3_NP_combined <- unique(c(decontam_VGO3_NP_ASVs, KW_VGO3_NP_ASVs))
length(contaminants_VGO3_NP_combined)# 299
# VGO3_OP: both decontam and KW
contaminants_VGO3_OP_combined <- unique(c(decontam_VGO3_OP_ASVs, KW_VGO3_OP_ASVs))
length(contaminants_VGO3_OP_combined)# 210 


# Sort ASVs in teh list by their ranking (i.e. the number suffix at the end of their name)
# Sort ASV list by numeric suffix
sort_asvs_by_index <- function(asvs) {
  index <- as.numeric(sub(".*_", "", asvs))  # extract suffix
  asvs[order(index)]
}

# VGO2 NP
contaminants_VGO2_NP_combined_sorted <- sort_asvs_by_index(contaminants_VGO2_NP_combined)
# VGO2 OP
contaminants_VGO2_OP_combined_sorted <- sort_asvs_by_index(contaminants_VGO2_OP_combined)
# VGO3 NP
contaminants_VGO3_NP_combined_sorted <- sort_asvs_by_index(contaminants_VGO3_NP_combined)
# VGO3 OP
contaminants_VGO3_OP_combined_sorted <- sort_asvs_by_index(contaminants_VGO3_OP_combined)



# Create table of contaminants detected by each method and in each dataset 
# Create a union of all contaminant ASVs from all datasets
all_contaminants <- unique(c(
  contaminants_VGO2_NP_combined_sorted,
  contaminants_VGO2_OP_combined_sorted,
  contaminants_VGO3_NP_combined_sorted,
  contaminants_VGO3_OP_combined_sorted
))

# Initialize data frame
final_contaminant_df <- data.frame(
  ASV = all_contaminants,
  VGO2_NP_decontam = all_contaminants %in% contaminants_VGO2_NP_combined_sorted,
  VGO2_OP_decontam = all_contaminants %in% contaminants_VGO2_OP_combined_sorted,
  VGO3_NP_decontam = all_contaminants %in% contaminants_VGO3_NP_combined_sorted,
  VGO3_OP_decontam = all_contaminants %in% contaminants_VGO3_OP_combined_sorted
)

# Optional: sort by numeric ASV suffix for readability
final_contaminant_df <- final_contaminant_df[order(as.numeric(sub(".*_", "", final_contaminant_df$ASV))), ]

# View final table
head(final_contaminant_df)

# Write to CSV
write.csv(final_contaminant_df, "../Output/Decontamination/Final_contaminants_combined.csv", row.names = FALSE)
# Now to do manual checks of ASVs flagged as contaminants
```
## Frequency plot of contaminants
We create a frequency plot of the top 100 identified contaminants. For most of these contaminant ASVs there seems to be a correlation between their relative abundance and the bacterial density (for VGO2 OP the pattern isn't so consistent)
```{r}
# Select top 100 ASVs (or fewer if < 100 available)
top100_VGO2_NP <- head(contaminants_VGO2_NP_combined_sorted, 100)
top100_VGO2_OP <- head(contaminants_VGO2_OP_combined_sorted, 100)
top100_VGO3_NP <- head(contaminants_VGO3_NP_combined_sorted, 100)
top100_VGO3_OP <- head(contaminants_VGO3_OP_combined_sorted, 100)

# Frequency plots for each dataset
# VGO2 NP
plot_VGO2_NP <- plot_frequency(subset_samples(VGO2_NP_ps_for_decontam, niche == "NP"), top100_VGO2_NP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")
plot_VGO2_NP

# VGO2 OP
# plot_VGO2_OP <- plot_frequency(subset_samples(VGO2_OP_ps_for_decontam, niche == "OP"), top100_VGO2_OP, 
#                conc="qPCR") +  xlab("Bacterial density (pg/ul)")
# plot_VGO2_OP
# Doesn't work for OP...
# Subset to OP samples
ps_op <- subset_samples(VGO2_OP_ps_for_decontam, niche == "OP")
# Remove samples with zero total counts
ps_op <- prune_samples(sample_sums(ps_op) > 0, ps_op)
# Get ASVs present in this filtered OP set
asvs_in_op <- taxa_names(ps_op)
# Intersect with top100 ASVs
top100_present_op <- intersect(top100_VGO2_OP, asvs_in_op)
# Plot
plot_VGO2_OP <- plot_frequency(
  ps_op, 
  top100_present_op,
  conc = "qPCR"
) + 
  xlab("Bacterial density (pg/µl)") +
  ggtitle("VGO2 OP")


# VGO3 NP
plot_VGO3_NP <- plot_frequency(subset_samples(ps_vgo3_np_for_decontam, niche == "NP"), top100_VGO3_NP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")

# VGO3 OP
plot_VGO3_OP <- plot_frequency(subset_samples(ps_vgo3_op_for_decontam, niche == "OP"), top100_VGO3_OP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")


ggsave("../Output/Decontamination/freqplot_VGO2_NP_top100.svg", plot_VGO2_NP, width=15, height=15)
ggsave("../Output/Decontamination/freqplot_VGO2_OP_top100.svg", plot_VGO2_OP, width=15, height=15)
ggsave("../Output/Decontamination/freqplot_VGO3_NP_top100.svg", plot_VGO3_NP, width=15, height=15)
ggsave("../Output/Decontamination/freqplot_VGO3_OP_top100.svg", plot_VGO3_OP, width=15, height=15)


# x-axis: Bacterial load (e.g., qPCR concentration, here pg/μl)
# y-axis: Frequency (relative abundance) of an individual ASV

## For true contaminants we expect higher relative abundance at lower bacterial loads, and a declining or flat trend as load increases. There should be a negative correlation between ASV abundance and sample bacterial density. If it is a false contaminant, then there will be no clear correlation with bacterial load. 
```
## Differences in contaminants between niches 
```{r}
# Identify contaminants shared between 
# Compare NP vs OP for VGO2
compare_np_op <- function(np_asvs, op_asvs) {
  shared <- intersect(np_asvs, op_asvs)
  np_only <- setdiff(np_asvs, op_asvs)
  op_only <- setdiff(op_asvs, np_asvs)
  
  list(
    shared_ASVs = shared,
    shared_count = length(shared),
    np_only_ASVs = np_only,
    np_only_count = length(np_only),
    op_only_ASVs = op_only,
    op_only_count = length(op_only)
  )
}

# VGO2: Compare contaminants between NP and OP
np_op_comparison_VGO2 <- compare_np_op(
  contaminants_VGO2_NP_combined_sorted,
  contaminants_VGO2_OP_combined_sorted
)

# VGO3: Compare contaminants between NP and OP
np_op_comparison_VGO3 <- compare_np_op(
  contaminants_VGO3_NP_combined_sorted,
  contaminants_VGO3_OP_combined_sorted
)

# View the results
np_op_comparison_VGO2 # 6 shared, 145 NP only, 18 OP only 
np_op_comparison_VGO3 # 37 shared, 174 NP only, 150 OP only 

```
## Check false contaminants
After screening the list of combined contaminants, we found some that some ASVs may not actually be contaminants.
```{r}
# Specifically we see that there are some high abundant common NP ASVs that are flagged as contaminants in the VGO3 OP dataset. But in the VGO3 sequencing, there were some mixed seuqencing runs (i.e. mixing NP and OP niches). Checking in which runs these common NP ASVs appear to see if we can see signs of spill-over from the NP to OP niches.

# Common NP contaminants to check
# Define the ASVs you want to check for spill-over
contams_check_vgo3_OP <- c("Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3", "Moraxella_4")
# Filter to relevant ASVs in VGO3 OP and keep only non-zero abundance
df_vgo3_op_contams_present <- df_VGO3_op_pivot %>%
  filter(ASV %in% contams_check_vgo3_OP, RA > 0)
# For each ASV, list the unique DI.Runs where it's present
asv_run_presence_simple <- df_vgo3_op_contams_present %>%
  group_by(ASV) %>%
  summarise(
    Runs_present = paste(sort(unique(DI.Run)), collapse = ", "),
    .groups = "drop"
  )
# View result
print(asv_run_presence_simple)
# Optionally write to file
write.csv(asv_run_presence_simple, "../Output/Decontamination/VGO3_OP_suspect_ASV_run_presence.csv", row.names = FALSE) # The ASVs seem to be present across a selection of different DNA isolation runs - therefore not probably related to spill-over from NP niche

# Extra check: 
# Check the relative abundance across DNA isolation runs for these suspicious contaminants identified. 
# Define ASVs of interest
asvs_to_check <- c("Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3", "Moraxella_4")
# Filter dataset for ASVs of interest and ensure DI.Run is ordered
plot_data <- df_VGO3_op_pivot %>%
  filter(ASV %in% asvs_to_check) %>%
  mutate(DI.Run = factor(DI.Run, levels = sort(unique(as.numeric(as.character(DI.Run))))))
# Plot
p <- ggplot(plot_data, aes(x = DI.Run, y = RA)) +
  geom_point(alpha = 0.5, shape = 16, size = 2, position = position_jitter(width = 0.2)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 8, size = 2, color = "red") +
  facet_wrap(~ASV, scales = "free_y") +
  labs(
    x = "DNA Isolation Run (DI.Run)",
    y = "Relative Abundance",
    title = "RA of suspect ASVs across runs (VGO3 OP)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
# Display plot
print(p)
# Save plot 
ggsave("../Output/Decontamination/RA_by_run_VGO3_OP_suspects.svg", p, width = 10, height = 6)# no clear pattern

```
## Non-contaminants after manual inspection
```{r}
# Define non-contaminants manually verified
non_contam <- c(
  "Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3",
  "Moraxella_4", "Veillonella_140", "Prevotella_7_melaninogenica_165",
  "Anaerococcus_provencensis_168", "Acinetobacter_lwoffii_177", "Prevotella_7_179",
  "Cutibacterium_187", "Lactococcus_209", "Actinomyces_225"
)

# Filter contaminants by removing known non-contaminants
all_contaminants_combined_filt <- setdiff(all_contaminants_combined, non_contam)
cat("Number of contaminants after filtering non-contaminants:", length(all_contaminants_combined_filt), "\n") # 415 
```
## Exclude all blanks from phyloseq objects 
```{r}
# Subset phyloseq object to samples from niches "NP" and "OP"
ps2_NPOP <- subset_samples(ps2, niche %in% c("NP", "OP"))

# Remove taxa with zero counts after subsetting samples
ps2_NPOP_f <- prune_taxa(taxa_sums(ps2_NPOP) > 0, ps2_NPOP)
ps2_NPOP; ps2_NPOP_f # 6980 taxa removed
```
## Remove contaminants
```{r}
# Remove contaminant taxa from the filtered phyloseq object
ps2_NPOP_no_contams <- prune_taxa(!(taxa_names(ps2_NPOP_f) %in% all_contaminants_combined_filt), ps2_NPOP_f)

cat("Contaminants removed:", length(taxa_names(ps2_NPOP_f)) - length(taxa_names(ps2_NPOP_no_contams)), "\n") # 415 removed 

# Remove any taxa with zero counts after contaminant removal
ps2_no_zero_taxa <- prune_taxa(taxa_sums(ps2_NPOP_no_contams) > 0, ps2_NPOP_no_contams)

# Final object: ps2_no_zero_taxa contains no contaminants and no zero-count taxa

# Manual inspection of list - "Salmonella_25", "Listeria_26" seem suspicious - not likely in NP and OP, do not appear in the VGO3 OP data, see where they are in the dataset - looking into these: 
# Define the ASVs of interest
asvs_of_interest <- c("Salmonella_25", "Listeria_26")
# Get the OTU table and transpose so rows are samples
otu_df <- as.data.frame(t(otu_table(ps2_NPOP_no_contams)))
# Find samples where either ASV is present (i.e., abundance > 0)
samples_with_asvs <- rownames(otu_df)[rowSums(otu_df[, asvs_of_interest, drop = FALSE] > 0) > 0]
# Extract the sample_data for those samples
sample_data_with_asvs <- sample_data(ps2)[samples_with_asvs, ]
# View 
sample_data_with_asvs
```
# Final list of ASVs remaining
```{r}
# Extract ASV names from phyloseq object
asv_names <- taxa_names(ps2_no_zero_taxa)
# Extract numeric suffix (everything after last underscore)
asv_suffix_numbers <- as.numeric(sub(".*_", "", asv_names))
# Order ASVs by numeric suffix
asv_ordered <- asv_names[order(asv_suffix_numbers)]
# Take only the first 100
asv_first_100 <- head(asv_ordered, 100)
# Print
print(asv_first_100)
# save final ASV list
write.csv(asv_first_100, "../Output/Microbiome_descriptives/Ordered_ASV_names_top100.csv")
```

# Check for duplicated samples
```{r}
# Duplicates were already resolved before but check if any were missed
# Convert sample_data to data.frame
sample_df <- as.data.frame(sample_data(ps2_no_zero_taxa))
# Create a new combined key of VGO3_ID and niche
sample_df$combined_key <- paste(sample_df$VGO3_ID, sample_df$niche, sample_df$study, sep = "_")
# Count how many times each combined_key appears
counts <- table(sample_df$combined_key)
# Keep keys that appear more than once (true duplicates)
dup_keys <- names(counts[counts > 1])
# Filter the data.frame for those duplicated keys
duplicates <- sample_df[sample_df$combined_key %in% dup_keys, ]
# View duplicates
duplicates
```

# Exclude low density samples
Next, we remove all the samples with low bacterial density (lower than 0.095 pg/ul). This includes 20 samples

```{r}
ps3 <- prune_samples(sample_data(ps2_no_zero_taxa)$qPCR >= 0.095, ps2_no_zero_taxa)
ps3 <- prune_taxa(taxa_sums(ps3) > 0, ps3) # Remove taxa with taxa sum of 0

ps2_no_zero_taxa; ps3 # 84 samples were removed 
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25773 taxa and 2415 samples ]
# sample_data() Sample Data:       [ 2415 samples by 88 sample variables ]
# tax_table()   Taxonomy Table:    [ 25773 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25773 reference sequences ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25537 taxa and 2332 samples ]
# sample_data() Sample Data:       [ 2332 samples by 88 sample variables ]
# tax_table()   Taxonomy Table:    [ 25537 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25537 reference sequences ]
```
# Rarefaction curves
```{r}
# First we take a look at the distribution of # reads per sample
read_counts <- sample_sums(ps3) %>% enframe() %>% ggplot(aes(y = 1, x = value)) +
  ggdist::stat_halfeye( adjust = .5, height = .3, justification = -.3,
                        .width = 0, point_colour = NA) + 
  geom_point(size = 1.3, shape = 16, alpha = .2, 
             position = position_jitter(seed = 1, width = 0, height = .05)) +
  geom_boxplot(width = .12, alpha = 0.5, outlier.color = NA) +
  scale_x_log10(breaks = c(1000, 5000, 10000, 30000, 100000, 300000),
                labels = scales::comma) +
  coord_cartesian(ylim = c(0.9, NA), clip = "off") +
  labs(x = "Number of reads/sample", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  geom_vline(xintercept = 10000, linetype = "dashed", color = "red")
read_counts
ggsave("../Output/Preprocessing/read_counts.svg", read_counts, width = 5, height = 4)

# Rarefaction curves - helps evaluate whether the sequencing depth (i.e., number of reads per sample) was sufficient to capture the diversity of microbial communities in the samples.
# X-axis: Number of sequencing reads (sampling effort)
# Y-axis: Number of observed species (or ASVs/OTUs, depending on your level of analysis)
## Each line on the curve represents a sample.
otu_tab_rare <- ps3 %>% prune_samples(sample_sums(.) < 30000, .) %>% # only keep samples with <30000 reads
  otu_table() %>% # extract otu table
  as(., "matrix") %>% # convert it to a matrix
  t # transpose the matrix where rows = samples and cols = taxa

saveRDS(otu_tab_rare, "../Output/Preprocessing/otu_tab_rare.rds")

# Generate rarefaction curves
# rarecurve() (from the vegan package) performs individual-based rarefaction for each sample by randomly subsampling reads without replacement at increasing steps (every 10 reads, here) and calculating the number of unique taxa detected at each step.
# rare_curve_data <- rarecurve(otu_tab_rare, step = 10)

# rare_curves <- map2_dfr(rare_curve_data, 
#                         rownames(otu_tab_rare), ~enframe(.x) %>% 
#                           mutate(sample_id = .y)) %>%
#   mutate(name = str_remove(name, "N") %>% as.numeric) %>%
#   ggplot(aes(x = name, y = value, group = sample_id)) +
#     geom_line(alpha = 0.15) +
#     scale_y_log10() +
#     scale_x_continuous(breaks = c(0, 5000, 10000, 15000, 20000, 25000), 
#                        labels = scales::comma) +
#     geom_vline(aes(xintercept = 10000),  colour="grey50", linetype="dotted") +
#     labs(x = "Sample size", y = "Number of species")
# 
# ggsave("../Output/Microbiome_descriptives/rarefaction_curves.svg", rare_curves)
```

# Exclusion of samples with less than 10,000 reads
```{r}
# Based on these figures, we chose a read cut-off at 10,000 reads. In the rarefaction curves, this was the point where the curves started flattening. 
# Remove the samples with <10,000 reads
ps4 <- prune_samples(sample_sums(ps3) > 10000, ps3)
ps4 <- prune_samples(!is.na(ps4@sam_data$niche), ps4)
ps4 <- prune_taxa(taxa_sums(ps4) > 0, ps4)

ps3; ps4

# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25537 taxa and 2332 samples ]
# sample_data() Sample Data:       [ 2332 samples by 88 sample variables ]
# tax_table()   Taxonomy Table:    [ 25537 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25537 reference sequences ]
# 
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 24355 taxa and 2170 samples ]
# sample_data() Sample Data:       [ 2170 samples by 88 sample variables ]
# tax_table()   Taxonomy Table:    [ 24355 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 24355 reference sequences ]
```

# Composition plots per dataset - after pre-processing
```{r}
# Subset ps2 for dataset == "VGO3_NP"
ps_vgo2_np <- subset_samples(ps4, dataset == "VGO2_NP" & niche == "NP")
ps_vgo2_np_filt <- prune_taxa(taxa_sums(ps_vgo2_np) > 0, ps_vgo2_np)
ps_vgo2_np_filt <- prune_samples(sample_sums(ps_vgo2_np_filt) > 0, ps_vgo2_np_filt)

# stacked bar plot
VGO2_NP_stackedbar <- create_ordered_bar_n(ps_vgo2_np, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 NP microbiome composition per sample (cleaned)")
# Save plot
ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/VGO2_NP_barplot.svg", VGO2_NP_stackedbar, width = 5, height = 6)

# Subset ps2 for dataset == "VGO3_OP"
ps_vgo2_op <- subset_samples(ps4, dataset == "VGO2_OP" & niche == "OP")
ps_vgo2_op_filt <- prune_taxa(taxa_sums(ps_vgo2_op) > 0, ps_vgo2_op)
ps_vgo2_op_filt <- prune_samples(sample_sums(ps_vgo2_op_filt) > 0, ps_vgo2_op_filt)

# stacked bar plot
VGO2_OP_stackedbar <- create_ordered_bar_n(ps_vgo2_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 OP microbiome composition per sample (cleaned)")

# Save plot
ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/VGO2_OP_barplot.svg", VGO2_OP_stackedbar, width = 5, height = 6)


# Subset ps2 for dataset == "VGO3_NP"
ps_vgo3_np <- subset_samples(ps2, dataset == "VGO3_NPOP" & niche == "NP")
ps_vgo3_np_filt <- prune_taxa(taxa_sums(ps_vgo3_np) > 0, ps_vgo3_np)
ps_vgo3_np_filt <- prune_samples(sample_sums(ps_vgo3_np_filt) > 0, ps_vgo3_np_filt)

# stacked bar plot
VGO3_NP_stackedbar <- create_ordered_bar_n(ps_vgo3_np_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 NP microbiome composition per sample (cleaned)")
# Save plot
ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/VGO3_NP_barplot.svg", VGO3_NP_stackedbar, width = 5, height = 6)


# Subset ps2 for dataset == "VGO3_NP"
ps2@sam_data$dataset
ps_vgo3_op <- subset_samples(ps2, dataset == "VGO3_NPOP" & niche == "OP")
ps_vgo3_op_filt <- prune_taxa(taxa_sums(ps_vgo3_op) > 0, ps_vgo3_op)
ps_vgo3_op_filt <- prune_samples(sample_sums(ps_vgo3_op_filt) > 0, ps_vgo3_op_filt)

# stacked bar plot
VGO3_OP_stackedbar <- create_ordered_bar_n(ps_vgo3_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 OP microbiome composition per sample (cleaned)")
# Save plot
ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/VGO3_OP_barplot.svg", VGO3_OP_stackedbar, width = 5, height = 6)



#### Combine plots: NP on top row, OP on bottom row ####
composition_combined <- (VGO2_NP_stackedbar | VGO3_NP_stackedbar) / 
                        (VGO2_OP_stackedbar | VGO3_OP_stackedbar) +
                        plot_annotation(tag_levels = 'A')  # Optional: add panel tags A, B, C, D

print(composition_combined)

# Save combined figure
ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/Combined_composition_barplots.svg", composition_combined, width = 12, height = 12)

```

# Composition plots - average RA per dataset (genus level)
Make plots which collapse ASVs to genera, calculate mean relative abundances for each genus per dataset, and keep genera that have >10% mean RA, create side-by-side bar plot across the datasets
```{r}
get_top_genera_RA <- function(ps_obj, dataset_name) {
  # Aggregate to genus level
  ps_genus <- tax_glom(ps_obj, taxrank = "genus")
  ps_genus_RA <- transform_sample_counts(ps_genus, function(x) x / sum(x))
  
  # Convert to data frame
  genus_df <- psmelt(ps_genus_RA)
  
  # Compute mean relative abundance and select top 10 genera
  genus_summary <- genus_df %>%
    group_by(genus) %>%
    summarise(Mean_RA = mean(Abundance, na.rm = TRUE)) %>%
    arrange(desc(Mean_RA)) %>%
    slice_head(n = 5) %>%
    mutate(Dataset = dataset_name)
  
  return(genus_summary)
}


# Run for each group
RA_vgo2_np <- get_top_genera_RA(ps_vgo2_np_filt, "VGO2_NP")
RA_vgo2_op <- get_top_genera_RA(ps_vgo2_op_filt, "VGO2_OP")
RA_vgo3_np <- get_top_genera_RA(ps_vgo3_np_filt, "VGO3_NP")
RA_vgo3_op <- get_top_genera_RA(ps_vgo3_op_filt, "VGO3_OP")

# Combine
RA_all <- bind_rows(RA_vgo2_np, RA_vgo2_op, RA_vgo3_np, RA_vgo3_op)

# Make Genus a factor ordered by abundance for cleaner plotting
RA_all <- RA_all %>%
  mutate(Genus = fct_reorder(genus, Mean_RA, .fun = max, .desc = TRUE))


# Add residual to each group to make bars stack to 100%
RA_all_with_resid <- RA_all %>%
  group_by(Dataset) %>%
  mutate(Total_RA = sum(Mean_RA),
         Residual_RA = 1 - Total_RA) %>%
  ungroup()

residual_rows <- RA_all_with_resid %>%
  select(Dataset, Residual_RA) %>%
  distinct() %>%
  mutate(genus = "Residual", Mean_RA = Residual_RA) %>%
  select(genus, Mean_RA, Dataset)

RA_all_final <- RA_all %>%
  select(genus, Mean_RA, Dataset) %>%
  bind_rows(residual_rows)

# Reorder genus so Residual is always on top
RA_all_final <- RA_all_final %>%
  mutate(Genus = ifelse(genus == "Residual", "Residual", as.character(genus))) %>%
  mutate(Genus = fct_reorder(Genus, Mean_RA, .fun = max, .desc = TRUE)) %>%
  mutate(Genus = fct_relevel(Genus, "Residual", after = 0))

# Reorder Dataset manually
RA_all_final <- RA_all_final %>%
  mutate(Dataset = factor(Dataset, levels = c("VGO2_NP", "VGO3_NP", "VGO2_OP", "VGO3_OP")))


# Assign colors
genus_levels <- unique(RA_all_final$genus)
n_non_residual <- length(setdiff(genus_levels, "Residual"))
palette_colors <- brewer.pal(min(12, n_non_residual), "Paired")

genus_colors <- c(
  setNames(palette_colors[1:n_non_residual], setdiff(genus_levels, "Residual")),
  Residual = "grey80"
)

# plot
genus_plot_all <- RA_all_final %>%
  ggplot(aes(x = Dataset, y = Mean_RA, fill = Genus)) +
  geom_bar(stat = "identity", color = "white", size = 0.3) +  # white lines between stacked genera
  scale_fill_manual(
    values = genus_colors,
    guide = guide_legend(title = "Genus", label.theme = element_text(face = "italic"))
  ) +
  labs(
    title = "Top 5 genera per dataset",
    y = "Mean relative abundance",
    x = "Dataset"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1), expand = c(0, 0)) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.border = element_rect(color = "grey40", size = 0.5),
    legend.key.size = unit(0.5, "cm")
  )

# View the plot
genus_plot_all

ggsave("../Output/Microbiome_descriptives/After_preprocessing_composition/Genus_composition_plot_all.svg", genus_plot_all, height = 5, width =7)


```

# Rarefy data and compute alpha diversity indices 
```{r}
m <- ps4 %>% meta_to_df()
gc()
set.seed(77)
ps_rarefied <- phyloseq::rarefy_even_depth(ps4, replace = T, trimOTUs = T)
sample_sums(ps_rarefied)# min library size = 10038    

# 2638OTUs were removed because they are no longer present in any sample after random subsampling
# Compute alpha diversity indices: Observed, shannon and simpson
adiv <- data.frame(
  "Observed" = phyloseq::estimate_richness(ps_rarefied, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(ps_rarefied, measures = "Shannon"),
  "Simpson" = phyloseq::estimate_richness(ps_rarefied, measures = "Simpson")
  )
# sample id have been changed to "." instead of "_"
adiv$sample_id <- gsub("\\.", "-", rownames(adiv))
m <- left_join(m, adiv)
m$qPCR_log <- log2(m$qPCR)

# Add alpha diversity values to phyloseq object
rownames(m) <- m$sample_id
m <- select(m, -c("sample_id"))
sample_data(ps4) <- m

# Add read number metadata
ps4@sam_data$final_read_nr <- colSums(ps4@otu_table)
```

# Preprocessing summary
```{r}
stats <- tibble(step = character(),
                samples = numeric(),
                taxa = numeric(),
                reads = numeric())


stats <- ps_stats(ps, stats, "Raw data")
stats <- ps_stats(ps2_NPOP_f, stats, "Removal of blanks and zero sum taxa")
stats <- ps_stats(ps2_no_zero_taxa, stats, "Exclusion of contaminant ASVs")
stats <- ps_stats(ps3, stats, "Exclusion of samples with bacterial density <0.095 pg/uL")
stats <- ps_stats(ps4, stats, "Exclusion of samples with reads <10,000")
stats <- ps_stats(ps_rarefied, stats, "Rarefied for alpha diversity")
stats

write.csv(stats, "../Output/Preprocessing/Processing_steps.csv")
```
# Number samples in each dataset before and after pre-processing
```{r}
# Define a helper function to count samples
count_samples <- function(ps_obj, niche_val, dataset_val) {
  sample_data(ps_obj) %>%
    data.frame() %>%
    filter(niche == niche_val, dataset == dataset_val) %>%
    nrow()
}

# Create the summary table
sample_counts <- tibble(
  Condition = c("NP - VGO2_NP", "NP - VGO3_OPNP", "OP - VGO2_OP", "OP - VGO3_NPOP"),
  
  Before = c(
    count_samples(ps2_NPOP_f, "NP", "VGO2_NP"),
    count_samples(ps2_NPOP_f, "NP", "VGO3_NPOP"),
    count_samples(ps2_NPOP_f, "OP", "VGO2_OP"),
    count_samples(ps2_NPOP_f, "OP", "VGO3_NPOP")
  ),
  
  After = c(
    count_samples(ps4, "NP", "VGO2_NP"),
    count_samples(ps4, "NP", "VGO3_NPOP"),
    count_samples(ps4, "OP", "VGO2_OP"),
    count_samples(ps4, "OP", "VGO3_NPOP")
  )
)

# View the table
print(sample_counts)

write.csv(sample_counts, "../Output/Preprocessing/Sample_counts_before_after_preprocessing.csv")
```

# Metadata additions and fixes
## Categorise lung function variables
```{r}
# Helper functions
# Categorising lung function
categorise_lung_function <- function(df) {
  df$FEV1_category <- cut(df$FEV1_pred,
    breaks = c(-Inf, 49.9, 69.9, 79.9, Inf),
    labels = c("Severe","Moderate","Mild","Normal"), right = TRUE)

  df$FVC_category <- cut(df$FVC_pred,
    breaks = c(-Inf, 80, Inf),
    labels = c("Severely reduced","Normal"), right = FALSE)

  df$FEV1FVC_category <- cut(df$FEV1FVC_pred,
    breaks = c(-Inf, 70, Inf),
    labels = c("Obstructive","Normal"), right = FALSE)

  df$FEF25_75_category <- cut(df$FEF25_75_pred,
    breaks = c(-Inf, 49.9, 69.9, 79.9, Inf),
    labels = c("Severe","Moderate","Mild","Normal"), right = TRUE)

  df$FEV1_category     <- factor(df$FEV1_category,
                                 levels = c("Normal","Mild","Moderate","Severe"), ordered = TRUE)
  df$FVC_category      <- factor(df$FVC_category,
                                 levels = c("Normal","Severely reduced"), ordered = TRUE)
  df$FEV1FVC_category  <- factor(df$FEV1FVC_category,
                                 levels = c("Normal","Obstructive"), ordered = TRUE)
  df$FEF25_75_category <- factor(df$FEF25_75_category,
                                 levels = c("Normal","Mild","Moderate","Severe"), ordered = TRUE)
  df
}

# Add your lung-function categories into sam_data (using your function)
add_lf_categories_to_ps <- function(ps) {
  sd <- as(phyloseq::sample_data(ps), "data.frame")

  # Ensure the inputs to `cut()` are numeric (quietly)
  to_num <- function(x) suppressWarnings(as.numeric(as.character(x)))
  need_num <- intersect(c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred"), names(sd))
  for (v in need_num) sd[[v]] <- to_num(sd[[v]])

  # Use your function to create categories
  sd <- categorise_lung_function(sd)

  phyloseq::sample_data(ps) <- sd
  ps
}

# Add the lung function categories
ps4 <- add_lf_categories_to_ps(ps4)

# Visualise the lung-function category distribution across the different databases 
sd <- as(phyloseq::sample_data(ps4), "data.frame")

# Make a clean facet label
sd$Dataset <- dplyr::case_when(
  sd$dataset %in% c("VGO2_NP","VGO2_OP") ~ sd$dataset,
  sd$dataset == "VGO3_NPOP"              ~ paste0("VGO3_", sd$niche),
  TRUE                                   ~ paste(sd$dataset, sd$niche, sep = "_")
)

lung_cat_vars <- c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")

lung_cat_counts <- sd %>%
  select(any_of(c("Dataset", lung_cat_vars))) %>%
  # FIX: unify types before pivot (avoid mixing ordered factors)
  mutate(across(all_of(lung_cat_vars), ~ as.character(.))) %>%
  pivot_longer(
    cols = all_of(lung_cat_vars),
    names_to = "Variable",
    values_to = "Category",
    values_drop_na = TRUE
    # Alternatively: values_transform = list(Category = as.character)
  ) %>%
  count(Dataset, Variable, Category, name = "n") %>%
  mutate(
    Variable = gsub("_category", "", Variable),
    Variable = factor(Variable, levels = c("FEV1","FVC","FEV1FVC","FEF25_75")),
    # Shared order for all possible categories
    Category = factor(
      Category,
      levels = c("Normal","Mild","Moderate","Severe","Severely reduced","Obstructive")
    )
  )

cat_cols <- c(
  "Normal"           = "#2B83BA",
  "Mild"             = "#ABD9E9",
  "Moderate"         = "#FDAE61",
  "Severe"           = "#D7191C",
  "Severely reduced" = "#8C510A",
  "Obstructive"      = "#762A83"
)

p_counts <- ggplot(lung_cat_counts, aes(x = Variable, y = n, fill = Category)) +
  geom_col(width = 0.85) +
  facet_wrap(~ Dataset, scales = "free_y") +
  scale_fill_manual(values = cat_cols, drop = TRUE) +
  labs(
    title = "Distribution of Lung Function Categories per Dataset",
    x = "Lung Function Metric",
    y = "Number of Samples",
    fill = "Category"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

p_counts

ggsave("../Output/Participant_data/Lung_function_category_barplot.svg",
       p_counts, width = 8, height = 6)
```

# Recode gender and smoking status as factors 
```{r}
# Recoding gender
recode_gender <- function(x) {
  g <- tolower(as.character(x))
  g <- dplyr::case_when(g %in% c("m","male") ~ "male",
                        g %in% c("f","female") ~ "female",
                        TRUE ~ g)
  factor(g, levels = c("male","female"))
}

# Recoding smoking status
recode_smoking <- function(x) {
  s <- tolower(as.character(x))
  s <- dplyr::case_when(
    s %in% c("non","never","no") ~ "never",
    s %in% c("ex","former","previous","prev") ~ "former",
    s %in% c("current","smoker") ~ "current",
    TRUE ~ s
  )
  factor(s, levels = c("never","former","current"))
}

# Recode gender and smoking status inside a phyloseq object's sam_data
recode_gender_smoking_in_ps <- function(ps) {
  sd <- as(phyloseq::sample_data(ps), "data.frame")

  if ("gender" %in% names(sd)) {
    sd$gender <- recode_gender(sd$gender)  # your function
  }
  if ("smoke_status" %in% names(sd)) {
    sd$smoke_status <- recode_smoking(sd$smoke_status)  # your function
  }

  phyloseq::sample_data(ps) <- sd
  ps
}

ps4 <- recode_gender_smoking_in_ps(ps4)
```

# Add season to phyloseq metadata
```{r}
add_season_to_master <- function(ps,
  vgo2_np_raw_rds = "../Input/Phyloseq/VGO2_NP/Received_Mari_Lee/ps_raw.Rds",
  vgo2_op_csv     = "../Input/Phyloseq/VGO2_OP/COPD_microbiome_subjectdata.csv") {

  # Pull sam_data as a plain data.frame
  sd <- data.frame(phyloseq::sample_data(ps), check.names = FALSE, stringsAsFactors = FALSE)
  if (!"sample_id" %in% names(sd)) sd$sample_id <- rownames(sd)

  # Start clean target column
  sd$season_microbiome_sampling <- NA_character_

  # Helpers
  season_from_month <- function(m) {
    ifelse(m %in% c(3,4,5),  "Spring",
    ifelse(m %in% c(6,7,8),  "Summer",
    ifelse(m %in% c(9,10,11),"Autumn", "Winter")))
  }
  code_to_label <- function(z) {
    z <- suppressWarnings(as.integer(as.character(z)))
    c("Winter","Spring","Summer","Autumn")[match(z, 1:4)]
  }
  parse_any_date <- function(x) {
    x <- as.character(x)
    suppressWarnings(lubridate::parse_date_time(
      x, orders = c("Ymd","dmy","mdy","Y-m-d","d/m/Y","m/d/Y"), quiet = TRUE
    ))
  }

  # ===== 1) VGO3 (NP & OP): derive from date_spiro =====
  idx_vgo3 <- sd$dataset == "VGO3_NPOP"
  if (any(idx_vgo3)) {
    d <- parse_any_date(sd$date_spiro[idx_vgo3])
    mo <- as.integer(month(d))
    labs <- season_from_month(mo)
    fill_idx <- idx_vgo3 & !is.na(labs)
    sd$season_microbiome_sampling[fill_idx] <- labs[!is.na(labs)]
  }

  # ===== 2) VGO2 OP: merge season code from CSV (ID -> VGO3_ID) =====
  if (file.exists(vgo2_op_csv)) {
    lk_op <- read.csv(vgo2_op_csv, stringsAsFactors = FALSE) %>%
      mutate(ID = trimws(as.character(ID)),
             season_code_op = season) %>%
      select(ID, season_code_op) %>%
      distinct(ID, .keep_all = TRUE)

    idx_op <- sd$dataset == "VGO2_OP"
    if (any(idx_op)) {
      sd$VGO3_ID_chr <- trimws(as.character(sd$VGO3_ID))
      # Left join only the OP rows, then map back
      op_sub <- sd[idx_op, , drop = FALSE] %>%
        left_join(lk_op, by = c("VGO3_ID_chr" = "ID"))
      labs <- code_to_label(op_sub$season_code_op)
      fill <- !is.na(labs)
      sd$season_microbiome_sampling[idx_op][fill] <- labs[fill]
      sd$VGO3_ID_chr <- NULL
    }
  } else {
    message("VGO2 OP season CSV not found: ", vgo2_op_csv)
  }

  # ===== 3) VGO2 NP: lookup from raw PS after T2/T3 drop + de-dup on respnr =====
  if (file.exists(vgo2_np_raw_rds)) {
    VGO2_NP_all <- readRDS(vgo2_np_raw_rds)
    VGO2_NP_t1 <- phyloseq::subset_samples(VGO2_NP_all, !stringr::str_detect(subject_identifier, "T2|T3"))

    meta_all <- data.frame(phyloseq::sample_data(VGO2_NP_t1), stringsAsFactors = FALSE)
    meta_all$sample_id <- phyloseq::sample_names(VGO2_NP_t1)

    meta_NP    <- meta_all %>% filter(sample_type == "NP")
    meta_other <- meta_all %>% filter(sample_type != "NP")

    meta_NP_dedup <- meta_NP %>% filter(!is.na(respnr)) %>% distinct(respnr, .keep_all = TRUE)
    meta_NP_na    <- meta_NP %>% filter(is.na(respnr))
    keep_ids <- c(meta_NP_dedup$sample_id, meta_NP_na$sample_id, meta_other$sample_id)

    VGO2_NP_dedup <- phyloseq::prune_samples(keep_ids, VGO2_NP_t1)
    np_sd <- data.frame(phyloseq::sample_data(VGO2_NP_dedup), stringsAsFactors = FALSE)
    np_sd$sample_id <- rownames(np_sd)

    lk_np <- np_sd %>%
      transmute(sample_id = as.character(sample_id),
                season_code_np = season) %>%
      distinct(sample_id, .keep_all = TRUE)

    # Join the lookup into ALL rows, fill only VGO2_NP
    sd <- sd %>%
      left_join(lk_np, by = "sample_id")

    idx_np <- sd$dataset == "VGO2_NP" & !is.na(sd$season_code_np)
    sd$season_microbiome_sampling[idx_np] <- code_to_label(sd$season_code_np[idx_np])

    sd$season_code_np <- NULL
  } else {
    message("VGO2 NP raw phyloseq not found: ", vgo2_np_raw_rds)
  }

  # Finalise factor & write back
  sd$season_microbiome_sampling <- factor(sd$season_microbiome_sampling,
                                          levels = c("Winter","Spring","Summer","Autumn"),
                                          ordered = FALSE)

  rownames(sd) <- sd$sample_id
  phyloseq::sample_data(ps) <- sd
  ps
}


# Use it to add season to the phyloseq metadata
ps4 <- add_season_to_master(ps4)

# Check season has been added
ps4@sam_data$season_microbiome_sampling

# Check NAs in season variable
sum(is.na(ps4@sam_data$season_microbiome_sampling)) # 2 NAs - see where these are
df <- data.frame(ps4@sam_data)
print(table(df$dataset, df$season_microbiome_sampling, useNA = "ifany")) # Both NA season values are from the VGO2_NP dataset

# 1 = winter, 2 = spring, 3 = summer, 4 = autumn
```
# Save phyloseq
```{r}
saveRDS(ps4, "../Output/Phyloseq/ps4.rds")
```

