---
title: "Cross-sectional microbiome-lung function analysis"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

Firstly we look at the relationship between teh microbiome and lung function cross-sectionally in 2014/15 (VGO2) and 2021/22 (VGO3)
# Packages
```{r}
library(phyloseq); library(microbiomer); library(ggplot2); library(cowplot); library(patchwork); library(scales); library(tidyverse); library(car); library(lmtest); library(ggsignif); library(forcats); library(rcartocolor); library(broom); library(writexl); library(vegan); library(tibble); library(glue); library(xlsx); library(corrplot); library(reshape2); library(RColorBrewer); library(Hmisc); library(Maaslin2); library(fpc); library(dendextend); library(paletteer); library(ggdendro)

# if(!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("Maaslin2")
```
# Data 
```{r}
microbiome_all <- readRDS("../Output/Phyloseq/ps4.rds")
```
# Separate microbiome data per dataset and niche
```{r}
# Subset per study and niche
# VGO2_NP dataset, NP niche
microbiome_VGO2_NP <- subset_samples(microbiome_all, dataset == "VGO2_NP" & niche == "NP")
nsamples(microbiome_VGO2_NP) # 210 NP samples

# VGO2_OP dataset, OP niche
microbiome_VGO2_OP <- subset_samples(microbiome_all, dataset == "VGO2_OP" & niche == "OP")
nsamples(microbiome_VGO2_OP) # 179 OP samples

# VGO3_NPOP dataset, NP niche
microbiome_VGO3_NP <- subset_samples(microbiome_all, dataset == "VGO3_NPOP" & niche == "NP")
nsamples(microbiome_VGO3_NP) # 799 NP samples

# VGO3_NPOP dataset, OP niche
microbiome_VGO3_OP <- subset_samples(microbiome_all, dataset == "VGO3_NPOP" & niche == "OP")
nsamples(microbiome_VGO3_OP) # 952 OP samples
```
# Filter taxa and convert to relative abundance
```{r}
# VGO2 NP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo2_np_filtered <- microbiome_VGO2_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 893 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22743 ASVs excluded).

# VGO2 OP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo2_op_filtered <- microbiome_VGO2_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 337 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 23299 ASVs excluded).

# VGO3 NP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo3_np_filtered <- microbiome_VGO3_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 883 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22753 ASVs excluded).

# VGO3 OP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo3_op_filtered <- microbiome_VGO3_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 1118 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22518 ASVs excluded).
```
# Extract OTU and metadata 
```{r}
otu_vgo2_np <- t(as.data.frame(otu_table(ps_vgo2_np_filtered)))
meta_vgo2_np <- meta_to_df(ps_vgo2_np_filtered, sample_name = ".row_id")

otu_vgo2_op <- t(as.data.frame(otu_table(ps_vgo2_op_filtered)))
meta_vgo2_op <- meta_to_df(ps_vgo2_op_filtered, sample_name = ".row_id")

otu_vgo3_np <- t(as.data.frame(otu_table(ps_vgo3_np_filtered)))
meta_vgo3_np <- meta_to_df(ps_vgo3_np_filtered, sample_name = ".row_id")

otu_vgo3_op <- t(as.data.frame(otu_table(ps_vgo3_op_filtered)))
meta_vgo3_op <- meta_to_df(ps_vgo3_op_filtered, sample_name = ".row_id")
```
# Check associations between variables
```{r}
# Select lung function variables
lung_vars <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
df_lung <- df[, lung_vars]

# Calculate correlation matrix with p-values
cor_res <- rcorr(as.matrix(df_lung), type = "pearson")  # Hmisc package
cor_matrix <- round(cor_res$r, 3)
p_matrix <- cor_res$P

# Melt the matrices
melted_cor <- melt(cor_matrix)
melted_p <- melt(p_matrix)

# Merge correlation values and p-values
melted_cor$p_value <- melted_p$value

# Function to assign significance stars
melted_cor$signif <- cut(
  melted_cor$p_value,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

# Keep only lower triangle (exclude duplicates & diagonal)
melted_cor <- melted_cor[as.numeric(melted_cor$Var1) > as.numeric(melted_cor$Var2), ]

# Plot enhanced lower triangular heatmap
lung_func_corr <- ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(sprintf("%.2f", value), signif)), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#2166AC",   # blue for negative correlations
    mid = "white",     # neutral = 0
    high = "#B2182B",  # red for positive correlations
    midpoint = 0,
    limits = c(-1, 1),
    name = "Pearson r"
  ) +
  theme_minimal() +
  labs(title = "Correlation Matrix of Lung Function\n Outcomes with Significance",
       x = "", y = "") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

ggsave("../Output/Variable_associations/Lung_function_correlations.svg", width = 5, height = 5)


# Test associations between variables to be included in multivariable models
vars <- c("age", "gender", "season_microbiome_sampling", "smoke_status")

# Check data types
sapply(df[vars], class)

# Initialize results dataframe
results <- data.frame(
  var1 = character(),
  var2 = character(),
  test = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through all pairs
for (i in 1:(length(vars)-1)) {
  for (j in (i+1):length(vars)) {
    v1 <- df[[vars[i]]]
    v2 <- df[[vars[j]]]
    
    # Continuous vs continuous
    if (is.numeric(v1) & is.numeric(v2)) {
      test_res <- cor.test(v1, v2)
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = "Pearson",
        p_value = test_res$p.value
      ))
    }
    
    # Continuous vs factor
    else if ((is.numeric(v1) & is.factor(v2)) | (is.factor(v1) & is.numeric(v2))) {
      if (is.factor(v1)) { tmp <- v1; v1 <- v2; v2 <- tmp; }
      if (nlevels(v2) == 2) {
        test_res <- t.test(v1 ~ v2)
        test_name <- "t-test"
        pval <- test_res$p.value
      } else {
        test_res <- aov(v1 ~ v2)
        test_name <- "ANOVA"
        pval <- summary(test_res)[[1]][["Pr(>F)"]][1]
      }
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = test_name,
        p_value = pval
      ))
    }
    
    # Factor vs factor
    else if (is.factor(v1) & is.factor(v2)) {
      test_res <- chisq.test(table(v1, v2))
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = "Chi-squared",
        p_value = test_res$p.value
      ))
    }
  }
}

# Add significance stars
results <- results %>%
  mutate(
    signif = cut(
      p_value,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("***", "**", "*", "")
    ),
    p_value = signif(round(p_value, 4))
  )

# Arrange results nicely
results <- results %>% arrange(var1, var2)

# View the table
print(results)

# Save the results table to Excel
write_xlsx(results, "../Output/Variable_associations/Covariate_association_results.xlsx")

```

# ALPHA DIVERSITY
We run linear models to assess the association between lung function values (% predicted lung function measures (FEV1, FVC, FEV1/FVC) and these variables categorised) and alpha diversity (Observed, Shannon, Simpson)
## Visualise associations between lung function variables and diversity
```{r}
# Datasets
datasets <- list(
  VGO2_NP = meta_vgo2_np,
  VGO2_OP = meta_vgo2_op,
  VGO3_NP = meta_vgo3_np,
  VGO3_OP = meta_vgo3_op
)

# Variables
cont_vars <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
cat_vars  <- c("FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")
alpha_metrics <- c("Observed", "Shannon", "Simpson")

# Relabel labels
lung_label <- c(
  FEV1_pred      = "FEV1 % predicted",
  FVC_pred       = "FVC % predicted",
  FEV1FVC_pred   = "FEV1/FVC % predicted",
  FEF25_75_pred  = "FEF 25%-75% % predicted",
  FEV1_category  = "FEV1 category",
  FVC_category   = "FVC category",
  FEV1FVC_category = "FEV1/FVC category",
  FEF25_75_category = "FEF 25%-75% category"
)

# Preferred ordinal orders 
ord_levels <- list(
  FEV1_category      = c("Normal","Mild","Moderate","Severe"),
  FVC_category       = c("Normal","Severely reduced"),
  FEV1FVC_category   = c("Normal","Obstructive"),
  FEF25_75_category  = c("Normal","Mild","Moderate","Severe")
)

base_theme <- theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1))

point_col <- "#2c7fb8"

# Helper function to order factor neatly
order_factor_if_possible <- function(x, name) {
  if (!is.null(ord_levels[[name]])) {
    lv <- levels(factor(x))
    ord <- ord_levels[[name]][ord_levels[[name]] %in% lv]
    if (length(ord) >= 2) return(factor(x, levels = ord, ordered = TRUE))
  }
  x
}

# Helper: Spearman rho annotation (returns small grob via ggpubr if installed)
spearman_annot <- function(df, x, y) {
  cc <- stats::complete.cases(df[[x]], df[[y]])
  n  <- sum(cc)
  if (!has_ggpubr || n < 3) return(NULL)
  cor_res <- suppressWarnings(suppressMessages(stats::cor.test(df[[x]][cc], df[[y]][cc], method = "spearman")))
  lbl <- paste0("\u03C1 = ", sprintf("%.2f", cor_res$estimate),
                ", p = ", format.pval(cor_res$p.value, digits = 2))
  ggpubr::annotate_figure(ggplot() + theme_void(),
                          top = ggpubr::text_grob(lbl, size = 10))
}

# Build plots
scatter_plots <- list()
box_plots <- list()

for (name in names(datasets)) {
  df0 <- datasets[[name]]

  # Ensure alpha columns exist and are numeric
  for (a in alpha_metrics) {
    if (a %in% names(df0)) df0[[a]] <- suppressWarnings(as.numeric(df0[[a]]))
  }

  # ------------ SCATTERPLOTS (continuous) ------------
  scatters <- list()
  for (xv in cont_vars) {
    if (!xv %in% names(df0)) next
    # coerce x to numeric if needed
    df0[[xv]] <- suppressWarnings(as.numeric(df0[[xv]]))

    for (alpha in alpha_metrics) {
      if (!alpha %in% names(df0)) next
      dat <- df0 %>% filter(!is.na(.data[[xv]]), !is.na(.data[[alpha]]))
      if (nrow(dat) < 3) next

      p <- ggplot(dat, aes_string(x = xv, y = alpha)) +
        geom_point(alpha = 0.6, size = 1.7, color = point_col) +
        geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
        labs(
          title = paste(alpha, "vs", lung_label[[xv]], "-", name),
          x = lung_label[[xv]],
          y = alpha
        ) +
        base_theme

      # Add Spearman annotation (top-left corner) if ggpubr available
      if (has_ggpubr) {
        rho <- try(suppressWarnings(stats::cor.test(dat[[xv]], dat[[alpha]], method = "spearman")), silent = TRUE)
        if (!inherits(rho, "try-error")) {
          lab <- paste0("\u03C1 = ", sprintf("%.2f", rho$estimate),
                        ", p = ", format.pval(rho$p.value, digits = 2))
          p <- p + ggplot2::annotate("label", x = -Inf, y = Inf, label = lab,
                                     hjust = -0.05, vjust = 1.2, size = 3.2, label.size = NA,
                                     fill = NA)
        }
      }

      scatters[[paste(name, xv, alpha, sep = " :: ")]] <- p
    }
  }

  # ------------ BOXPLOTS (categorical) ------------
  boxes <- list()
  for (xv in cat_vars) {
    if (!xv %in% names(df0)) next
    # order factor if possible
    df0[[xv]] <- order_factor_if_possible(df0[[xv]], xv)

    for (alpha in alpha_metrics) {
      if (!alpha %in% names(df0)) next
      dat <- df0 %>%
        filter(!is.na(.data[[xv]]), !is.na(.data[[alpha]]))

      if (!nrow(dat)) next

      p <- ggplot(dat, aes_string(x = xv, y = alpha, fill = xv)) +
        geom_boxplot(outlier.shape = NA, alpha = 0.9, width = 0.7) +
        geom_jitter(width = 0.15, alpha = 0.35, size = 1.2, color = "grey20") +
        guides(fill = "none") +
        labs(
          title = paste(alpha, "by", lung_label[[xv]], "-", name),
          x = lung_label[[xv]],
          y = alpha
        ) +
        base_theme +
        scale_fill_brewer(palette = "Blues")

      # Kruskal-Wallis p-value at top if >1 level and ggpubr available
      if (has_ggpubr && length(unique(dat[[xv]])) > 1) {
        p <- p + ggpubr::stat_compare_means(method = "kruskal.test",
                                            label.y.npc = "top", size = 3)
      }

      boxes[[paste(name, xv, alpha, sep = " :: ")]] <- p
    }
  }

  # Combine per dataset
  if (length(scatters)) {
    scatter_plots[[name]] <-
      wrap_plots(scatters, ncol = 4, guides = "collect") +
      plot_annotation(title = paste("Scatter plots:", name),
                      theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  }

  if (length(boxes)) {
    box_plots[[name]] <-
      wrap_plots(boxes, ncol = 4, guides = "collect") +
      plot_annotation(title = paste("Boxplots:", name),
                      theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))
  }
}

# print and save

for (nm in names(datasets)) {
  if (!is.null(scatter_plots[[nm]])) {
    print(scatter_plots[[nm]])
    ggsave(paste0("../Output/Alpha_diversity/Boxplots_scatterplots/", nm, "_scatterplots.svg"),
           scatter_plots[[nm]], width = 14, height = 18)
  }
  if (!is.null(box_plots[[nm]])) {
    print(box_plots[[nm]])
    ggsave(paste0("../Output/Alpha_diversity/Boxplots_scatterplots/", nm, "_boxplots.svg"),
           box_plots[[nm]], width = 14, height = 18)
  }
}

```

## Univariable linear regression for alpha diversity 
```{r}
## Univariable linear regression for alpha diversity (signed, standardised coefficients)
# Outcome variables (alpha diversity variables) are all z-score standardised - so I'm modelling a 1 SD chage in the outcome. 
# Input variables:
### Ordinal variables - β tests a linear trend across ordered levels per 1 SD increase in the coded score.
### Binary factors - set the reference categories manually, for 'yes/no' variables - no is automatically made the reference. β is the effect of going from reference (0) to the other category (1), expressed in SD units of that 0/1 variable. (Since a 0/1 variable has SD = sqrt(p(1−p)), standardising makes the coefficient comparable across binaries with different prevalences.)
### Numeric predictors - z-score standardised these, so β = change in outcome (in SD units) per 1 SD increase in the predictor.


## Baseline categories for factor variables in my analysis
# gender: male (baseline).
# asthma_ever: no (baseline).
# copd_doc_diag: no (baseline).
# pneumo_lastyr: no (baseline).
# smoke_status: baseline = never.
# season_microbiome_sampling: baseline = Winter.
# FEV1_category: baseline = Normal.
# FVC_category: baseline = Normal.
# FEV1FVC_category: baseline = Normal.
# FEF25_75_category: baseline = Normal.
# copd_gold: baseline = lowest GOLD stage present in the data (0 if included).


# Inputs 
dataset_list <- list(
  VGO2_NP = meta_vgo2_np,
  VGO2_OP = meta_vgo2_op,
  VGO3_NP = meta_vgo3_np,
  VGO3_OP = meta_vgo3_op
)

alpha_vars <- c("Shannon", "Simpson", "Observed")

test_vars <- c(
  "age","gender","bmi","season_microbiome_sampling","smoke_status",
  "packyears","asthma_ever","copd_gold",
  "copd_doc_diag","pneumo_last3yr","pneumo_lastyr",
  "FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred",
  "FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category"
)

#  Variable typing helpers 
set_binary <- function(x, ref = NULL) {
  x <- as.factor(x)
  if (!is.null(ref) && ref %in% levels(x)) {
    stats::relevel(x, ref = ref)
  } else if (all(c("no","yes") %in% levels(x))) {
    factor(x, levels = c("no","yes"))
  } else x
}

set_ordered <- function(x, pref_levels) {
  if (is.ordered(x)) return(x)
  x  <- as.factor(x)
  lv <- levels(x)
  ord <- pref_levels[pref_levels %in% lv]
  if (length(ord) >= 2L) factor(x, levels = ord, ordered = TRUE)
  else                   factor(x, levels = lv,  ordered = TRUE)
}

set_gold_ordered <- function(x) {
  if (is.ordered(x)) return(x)
  num <- suppressWarnings(as.numeric(as.character(x)))
  if (!all(is.na(num))) {
    ord <- sort(unique(num))
    factor(num, levels = ord, ordered = TRUE)
  } else {
    x <- as.factor(x)
    factor(x, levels = levels(x), ordered = TRUE)
  }
}

# Preferred orders for ordinals
ord_levels <- list(
  smoke_status               = c("never","former","current"),
  season_microbiome_sampling = c("Winter","Spring","Summer","Autumn"),
  FEV1_category              = c("Normal","Mild","Moderate","Severe"),
  FVC_category               = c("Normal","Severely reduced"),
  FEV1FVC_category           = c("Normal","Obstructive"),
  FEF25_75_category          = c("Normal","Mild","Moderate","Severe")
)

# Preprocess datasets (types + refs) 
for (nm in names(dataset_list)) {
  df <- dataset_list[[nm]]

  # Binary
  if ("gender"        %in% names(df)) df$gender        <- set_binary(df$gender,        ref = "male")
  if ("asthma_ever"   %in% names(df)) df$asthma_ever   <- set_binary(df$asthma_ever,   ref = "no")
  if ("copd_doc_diag" %in% names(df)) df$copd_doc_diag <- set_binary(df$copd_doc_diag, ref = "no")
  if ("pneumo_lastyr" %in% names(df)) df$pneumo_lastyr <- set_binary(df$pneumo_lastyr, ref = "no")

  # Ordinal
  if ("smoke_status" %in% names(df))
    df$smoke_status <- set_ordered(df$smoke_status, ord_levels$smoke_status)
  if ("season_microbiome_sampling" %in% names(df))
    df$season_microbiome_sampling <- set_ordered(df$season_microbiome_sampling, ord_levels$season_microbiome_sampling)

  for (v in c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")) {
    if (v %in% names(df)) df[[v]] <- set_ordered(df[[v]], ord_levels[[v]])
  }
  if ("copd_gold" %in% names(df)) df$copd_gold <- set_gold_ordered(df$copd_gold)

  # Numeric
  num_names <- intersect(c("age","bmi","packyears","pneumo_last3yr",
                           "FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred"),
                         names(df))
  for (v in num_names) df[[v]] <- suppressWarnings(as.numeric(as.character(df[[v]])))

  dataset_list[[nm]] <- df
}

# Modelling helper 
get_std_coef <- function(y, varname, df) {
  x <- df[[varname]]

  fit_and_tidy <- function(x_std) {
    if (all(is.na(x_std)) || length(na.omit(unique(x_std))) < 2) return(NULL)
    fit <- tryCatch(lm(y ~ x_std), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    tt <- broom::tidy(fit)
    list(
      beta = tt$estimate[tt$term == "x_std"],
      p    = tt$p.value[tt$term == "x_std"]
    )
  }

  # Ordered factors → integer scores → z
  if (is.factor(x) && is.ordered(x)) {
    x_std <- as.numeric(scale(as.numeric(x) - 1L))
    return(fit_and_tidy(x_std))
  }

  # Binary factors (2 levels) → 0/1 → z
  if (is.factor(x) && nlevels(droplevels(x)) == 2) {
    x01   <- as.numeric(x != levels(x)[1])
    x_std <- as.numeric(scale(x01))
    return(fit_and_tidy(x_std))
  }

  # Numeric → z
  if (is.numeric(x)) {
    if (sd(x, na.rm = TRUE) == 0 || all(is.na(x))) return(NULL)
    x_std <- as.numeric(scale(x))
    return(fit_and_tidy(x_std))
  }

  # Unordered 3+ level factors: skip (no unique signed β)
  NULL
}

# Fit models across datasets/variables 

all_results <- list()

for (dataset_name in names(dataset_list)) {
  df <- dataset_list[[dataset_name]]

  df_std <- df
  for (a in alpha_vars) if (a %in% names(df_std)) df_std[[a]] <- as.numeric(scale(df_std[[a]]))

  res_list <- list()

  for (alpha in alpha_vars) {
    if (!alpha %in% names(df_std)) next
    y <- df_std[[alpha]]

    for (var in test_vars) {
      if (!var %in% names(df_std)) next
      out <- get_std_coef(y, varname = var, df = df_std)
      if (is.null(out)) next

      res_list[[paste(alpha, var, sep = "_")]] <- data.frame(
        dataset     = dataset_name,
        alpha       = alpha,
        variable    = var,
        coefficient = out$beta,
        p_value     = out$p
      )
    }
  }

  all_results[[dataset_name]] <- dplyr::bind_rows(res_list)
}




# Define lung function continuous variables (will flip signs later)
lung_fun_cont <- c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred")

# Nice labels for plotting
pretty_labels <- c(
  "FEV1_pred"      = "FEV1 % predicted (decline)",
  "FVC_pred"       = "FVC % predicted (decline)",
  "FEV1FVC_pred"   = "FEV1/FVC % predicted (decline)",
  "FEF25_75_pred"  = "FEF 25%-75% % predicted (decline)",
  "FEV1_category"  = "FEV1 % predicted category",
  "FVC_category"   = "FVC % predicted category",
  "FEV1FVC_category" = "FEV1/FVC % predicted category",
  "FEF25_75_category" = "FEF 25%-75% % predicted category",
  "age" = "Age",
  "bmi" = "Body Mass Index",
  "packyears"= "Pack years of cigarettes", 
  "smoke_status"= "Smoking status",
  "pneumo_lastyr" = "Pneumonia in last year (number of times)",
  "pneumo_last3yr"= "Pneumonia in last 3 years (number of times)",
  "asthma_ever" = "Asthma status", 
  "copd_gold" = "COPD GOLD definition",
  "copd_doc_diag" = "COPD doctor diagnosed",
  "season_microbiome_sampling" = "Season of sampling",
  "gender" = "Gender"
)

meta_vgo3_np$copd_doc_diag

univariable_results_std <- dplyr::bind_rows(all_results) |>
  dplyr::filter(!is.na(p_value), !is.na(coefficient)) |>
  dplyr::mutate(
    # flip the continuous lung function coefficients so higher = decline
    coefficient = dplyr::if_else(variable %in% lung_fun_cont, -coefficient, coefficient),
    significance = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    variable = factor(variable, levels = rev(test_vars)),
    variable = dplyr::recode(variable, !!!pretty_labels)  # apply nice labels
  )

# UNIVARIABLE: enforce desired y-axis order 
desired_order_univar <- c(
  "Age",
  "Gender",
  "Body Mass Index",
  "Season of sampling",
  "Smoking status",
  "Pack years of cigarettes",
  "Asthma status",
  "COPD GOLD definition",
  "COPD doctor diagnosed",
  "Pneumonia in last year (number of times)",   
  "Pneumonia in last 3 years (number of times)",
  "FEV1 % predicted (decline)",
  "FEV1 % predicted category",
  "FVC % predicted (decline)",
  "FVC % predicted category",
  "FEV1/FVC % predicted (decline)",
  "FEV1/FVC % predicted category",
  "FEF 25%-75% % predicted (decline)",
  "FEF 25%-75% % predicted category"
)

# Make sure the labels exist; keep only those present to avoid warnings
present_levels_univar <- desired_order_univar[desired_order_univar %in% univariable_results_std$variable]

univariable_results_std <- univariable_results_std %>%
  dplyr::mutate(variable = factor(variable, levels = rev(present_levels_univar)))

# Save 
write.csv(univariable_results_std, "../Output/Alpha_diversity/Univariable/Univariable_alpha_diversity_standardised_results.csv")

# Plots 
lim <- max(abs(univariable_results_std$coefficient), na.rm = TRUE)

base_theme <- ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
                 panel.grid = ggplot2::element_blank())

plot_vgo2 <- dplyr::filter(univariable_results_std, dataset %in% c("VGO2_NP","VGO2_OP"))
plot_vgo3 <- dplyr::filter(univariable_results_std, dataset %in% c("VGO3_NP","VGO3_OP"))

pal_low  <- rcartocolor::carto_pal(7, "Geyser")[1]
pal_mid  <- rcartocolor::carto_pal(7, "Geyser")[4]
pal_high <- rcartocolor::carto_pal(7, "Geyser")[7]

heatmap <- function(dat, show_x_axis = TRUE) {
  ggplot2::ggplot(dat, ggplot2::aes(x = alpha, y = variable, fill = coefficient)) +
    ggplot2::geom_tile(color = "white") +
    ggplot2::geom_text(ggplot2::aes(label = significance), size = 3) +
    ggplot2::scale_fill_gradient2(
      low = pal_low, mid = pal_mid, high = pal_high,
      midpoint = 0, limits = c(-lim, lim), name = "Standardised Coefficient"
    ) +
    ggplot2::facet_wrap(~ dataset, ncol = 2) +
    base_theme +
    { if (!show_x_axis)
        ggplot2::theme(axis.title.x = ggplot2::element_blank(),
                       axis.text.x  = ggplot2::element_blank(),
                       axis.ticks.x = ggplot2::element_blank())
      else ggplot2::theme() } +
    ggplot2::labs(x = if (show_x_axis) "Alpha Diversity Metric" else NULL,
                  y = "Explanatory variable")
}

heatmap_vgo2 <- heatmap(plot_vgo2, show_x_axis = FALSE)
heatmap_vgo3 <- heatmap(plot_vgo3, show_x_axis = TRUE)

combined_plot <- (heatmap_vgo2 / heatmap_vgo3) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  ggplot2::theme(legend.position = "right")

univar_alpha_plot <- combined_plot +
  patchwork::plot_annotation(
    title = "Standardised Univariable Associations with Alpha Diversity (signed β)",
    theme = ggplot2::theme(plot.title = ggplot2::element_text(size = 12, face = "bold", hjust = 0.5))
  )

print(univar_alpha_plot)


ggplot2::ggsave(
  "../Output/Alpha_diversity/Univariable/Univariable_alpha_diversity_standardised_heatmap.svg",
  univar_alpha_plot, width = 8, height = 7
)
```
## Multivariable models with lung function (continuous + categorical) 
```{r}
## Multivariable models with lung function (continuous as decline + categorical)
# We standardise the alpha diversity metrics since observed is much higher than the other two, the coefficients are not comparable

# Define variables 
lung_fun_cont <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
lung_fun_cat  <- c("FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")
alpha_vars    <- c("Observed", "Shannon", "Simpson")
covariates    <- c("age", "gender", "smoke_status", "season_microbiome_sampling")

# We want all effects to read as "decline in lung function":
# Continuous predictors: flip sign so +β means LOWER lung function (decline) 
# Categorical predictors: already ordered with 'normal' as reference category 
flip_cont_sign <- TRUE   # flip continuous
flip_cat_sign  <- FALSE  # do NOT flip categorical (they already represent decline)

# Standardise outcomes + create standardised lung-function predictors
for (dataset_name in names(dataset_list)) {
  dataset <- dataset_list[[dataset_name]]

  # 1) z-score alpha metrics
  for (alpha_var in alpha_vars) {
    if (alpha_var %in% names(dataset)) {
      dataset[[alpha_var]] <- as.numeric(scale(dataset[[alpha_var]]))
    }
  }

  # 2) z-score continuous lung-function predictors
  for (v in lung_fun_cont) {
    if (v %in% names(dataset)) {
      dataset[[paste0(v, "_std")]] <- as.numeric(scale(dataset[[v]]))
    }
  }

  # 3) ordinal-trend code + z-score the categorical lung-function predictors
  for (v in lung_fun_cat) {
    if (v %in% names(dataset)) {
      x <- dataset[[v]]
      if (!is.ordered(x)) x <- factor(x, levels = levels(as.factor(x)), ordered = TRUE)
      dataset[[paste0(v, "_trend_std")]] <- as.numeric(scale(as.numeric(x) - 1L))
    }
  }

  dataset_list[[dataset_name]] <- dataset
}

# vectors of the actual std predictor names we'll use in models
lung_fun_cont_std <- paste0(lung_fun_cont, "_std")
lung_fun_cat_std  <- paste0(lung_fun_cat,  "_trend_std")

# pretty labels for plotting/reporting (DECLINE)
pretty_map <- c(
  setNames(c(
    "FEV1 % predicted (decline)",
    "FVC % predicted (decline)",
    "FEV1/FVC % predicted (decline)",
    "FEF 25%-75% % predicted (decline)"
  ), lung_fun_cont_std),
  setNames(c(
    "FEV1 category",
    "FVC category",
    "FEV1/FVC category",
    "FEF 25%-75% category"
  ), lung_fun_cat_std)
)

# Fit models: one LF predictor at a time + covariates
model_results <- list()

for (dataset_name in names(dataset_list)) {
  dataset <- dataset_list[[dataset_name]]
  model_results[[dataset_name]] <- list()

  for (alpha_var in alpha_vars) {
    model_results[[dataset_name]][[alpha_var]] <- list()

    for (lung_var_std in c(lung_fun_cont_std, lung_fun_cat_std)) {
      vars_needed <- c(alpha_var, lung_var_std, covariates)
      if (!all(vars_needed %in% names(dataset))) {
        warning("Skipping: ", dataset_name, " / ", alpha_var, " / ", lung_var_std, " (missing variables)")
        next
      }

      form <- as.formula(paste(alpha_var, "~", paste(c(lung_var_std, covariates), collapse = " + ")))
      fit  <- lm(form, data = dataset)
      model_results[[dataset_name]][[alpha_var]][[lung_var_std]] <- fit
    }
  }
}

# Tidy results (restrict to the lung-function term only) 
res_list <- list()

for (dataset_name in names(model_results)) {
  for (alpha_var in names(model_results[[dataset_name]])) {
    for (lung_var_std in names(model_results[[dataset_name]][[alpha_var]])) {
      fit <- model_results[[dataset_name]][[alpha_var]][[lung_var_std]]
      if (is.null(fit)) next

      td <- broom::tidy(fit)
      td <- td[td$term == lung_var_std, , drop = FALSE]  # keep only the LF term
      if (nrow(td) == 0) next

      td$dataset        <- dataset_name
      td$Alpha_metric   <- alpha_var
      td$lung_var_std   <- lung_var_std
      td$lung_label     <- unname(pretty_map[lung_var_std])

      res_list[[paste(dataset_name, alpha_var, lung_var_std, sep = "_")]] <- td
    }
  }
}

all_results <- dplyr::bind_rows(res_list)

# harmonise direction to represent "decline" for all LF variables 
all_results <- all_results |>
  dplyr::mutate(
    is_cont = lung_var_std %in% lung_fun_cont_std,
    is_cat  = lung_var_std %in% lung_fun_cat_std,
    estimate_decline = dplyr::case_when(
      is_cont & flip_cont_sign ~ -estimate,  # flip continuous to represent decline
      is_cat  & flip_cat_sign  ~ -estimate,  # (kept FALSE above)
      TRUE                     ~  estimate
    )
  )

# Add stars and presentable table
# Add BH correction PER DATASET and build presentable table
multivariable_alpha_diversity_results <- all_results |>
  dplyr::group_by(dataset) |>
  dplyr::mutate(
    # BH within each dataset's set of tests
    p_adj = p.adjust(p.value, method = "BH"),

    # stars based on adjusted p-values (these will drive the plot labels)
    sig_label = dplyr::case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    ),

    # (optional) stars based on raw p-values for reporting in CSV
    sig_label_raw = dplyr::case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    dataset,
    Alpha_metric,
    variable       = lung_label,     # pretty label
    term           = lung_var_std,   # technical term
    estimate       = estimate_decline,
    std.error,
    p.value,                         # raw p-value
    p_adj,                           # BH-adjusted p per dataset
    sig_label,                       # stars from adjusted p (used in plots)
    sig_label_raw,                   # raw p stars (CSV only, optional)
    estimate_with_sig = paste0(signif(estimate, 3), sig_label)
  )


# Desired order (top → bottom in the heatmap)
desired_order_multi <- c(
  "FEV1 % predicted (decline)",
  "FEV1 category",
  "FVC % predicted (decline)",
  "FVC category",
  "FEV1/FVC % predicted (decline)",
  "FEV1/FVC category",
  "FEF 25%-75% % predicted (decline)",
  "FEF 25%-75% category"
)

# Keep only labels that exist (avoid warnings if some are absent)
present_levels_multi <- desired_order_multi[desired_order_multi %in% multivariable_alpha_diversity_results$variable]

# Apply factor ordering (rev() so the first item appears at the TOP of the y-axis)
multivariable_alpha_diversity_results <- multivariable_alpha_diversity_results |>
  dplyr::mutate(variable = factor(variable, levels = rev(present_levels_multi)))



print(multivariable_alpha_diversity_results)

# Write to csv
readr::write_csv(
  multivariable_alpha_diversity_results,
  "../Output/Alpha_diversity/Multivariable/Multivariable_model_results_decline.csv"
)

# Create plots
plot_vgo2_multi <- multivariable_alpha_diversity_results |>
  dplyr::filter(dataset %in% c("VGO2_NP","VGO2_OP")) |>
  dplyr::mutate(
    significance = sig_label,
    estimate     = round(estimate, 3)
  )

plot_vgo3_multi <- multivariable_alpha_diversity_results |>
  dplyr::filter(dataset %in% c("VGO3_NP","VGO3_OP")) |>
  dplyr::mutate(
    significance = sig_label,
    estimate     = round(estimate, 3)
  )

base_theme <- ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30, hjust = 1),
                 panel.grid = ggplot2::element_blank())

legend_breaks <- scales::pretty_breaks(n = 5)(c(plot_vgo2_multi$estimate, plot_vgo3_multi$estimate))
legend_labels <- sprintf("%.3f", legend_breaks)

mk_heatmap <- function(dat, show_x = TRUE) {
  ggplot2::ggplot(dat, ggplot2::aes(x = Alpha_metric, y = variable, fill = estimate)) +
    ggplot2::geom_tile(color = "white") +
    ggplot2::geom_text(ggplot2::aes(label = significance), size = 4) +
    ggplot2::scale_fill_gradient2(
      low  = rcartocolor::carto_pal(7, "Geyser")[1],
      mid  = rcartocolor::carto_pal(7, "Geyser")[4],
      high = rcartocolor::carto_pal(7, "Geyser")[7],
      midpoint = 0,
      name = "Standardised Coefficient (decline)",
      breaks = legend_breaks,
      labels = legend_labels
    ) +
    ggplot2::facet_wrap(~ dataset, ncol = 2) +
    base_theme +
    { if (!show_x)
        ggplot2::theme(axis.title.x = ggplot2::element_blank(),
                       axis.ticks.x = ggplot2::element_blank(),
                       axis.text.x  = ggplot2::element_blank())
      else ggplot2::theme() } +
    ggplot2::labs(x = if (show_x) "Alpha Diversity Metric" else NULL,
                  y = "Lung Function Metric")
}

heatmap_vgo2_multi <- mk_heatmap(plot_vgo2_multi, show_x = FALSE)
heatmap_vgo3_multi <- mk_heatmap(plot_vgo3_multi, show_x = TRUE)

combined_multi_plot <- (heatmap_vgo2_multi / heatmap_vgo3_multi) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  ggplot2::theme(legend.position = "right")

final_multivar_plot <- combined_multi_plot +
  patchwork::plot_annotation(
    title = "Multivariable associations with alpha diversity (lung function treated as decline)",
    theme = ggplot2::theme(plot.title = ggplot2::element_text(size = 12, face = "bold", hjust = 0.5))
  )

print(final_multivar_plot)

ggplot2::ggsave(
  "../Output/Alpha_diversity/Multivariable/Multivariable_alpha_diversity_standardised_heatmap_decline.svg",
  final_multivar_plot, width = 8, height = 5
)

```
# BETA DIVERSITY
## Univariable PERMANOVA functions
```{r}
beta_df <- function(ord, meta, var) { 
  m_sub <- subset(meta, !is.na(meta[var]))
  ord_sub <- ord[rownames(ord) %in% m_sub$sample_id, colnames(ord) %in% 
                   m_sub$sample_id]
  ord_sub <- as.dist(ord_sub)
  return(list(ord_sub = ord_sub, m_sub = m_sub))
}

do_permanova <- function(ord, meta, var) {
  t <- adonis2(as.formula(glue("ord ~ {var}")), 
               permutations = 999, data = meta)
  return(t)
}

do_permanova_com <- function(ord, meta, var) {
  t <- beta_df(ord, meta, var) %>%
    { do_permanova(ord = .$ord_sub, meta = .$m_sub, var) }

  t <- broom::tidy(t)
 
   # (Optional) You might also want to filter out "Total" or "Model" rows if they appear
  t <- t %>% filter(!term %in% c("Residual", "Total"))  

  # Add p-value labels
  t <- t %>% mutate(
    label_p.value = ifelse(p.value < 0.25, "*", NA),
    label_p.value = ifelse(p.value < 0.1, "**", label_p.value),
    label_p.value = ifelse(p.value < 0.05, "***", label_p.value),
    Var = var
  )

  return(t)
}


# Dispersiion check function 
check_dispersion <- function(dist_matrix, meta, var) {
  if (is.factor(meta[[var]])) {
    # Categorical variable — use betadisper
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.dist(d)

    # Skip if <2 groups with ≥2 samples
    if (length(unique(sub_meta[[var]])) < 2 || any(table(sub_meta[[var]]) < 2)) {
      return(tibble(disp_p = NA, disp_note = "Skipped: <2 groups with ≥2 samples"))
    }

    model <- betadisper(d, sub_meta[[var]])
    perm <- permutest(model)
    return(tibble(disp_p = perm$tab[1, "Pr(>F)"], disp_note = NA))
    
  } else {
    # Continuous variable — use regression of average distances
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.matrix(d)
    dist_avg <- apply(d, 1, mean, na.rm = TRUE)

    mod <- lm(dist_avg ~ sub_meta[[var]])
    p <- summary(mod)$coefficients[2, 4]
    return(tibble(disp_p = p, disp_note = NA))
  }
}

# Dispersion check function
check_dispersion <- function(dist_matrix, meta, var) {
  if (is.factor(meta[[var]])) {
    # Categorical variable — use betadisper
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.dist(d)

    # Skip if <2 groups with ≥2 samples
    if (length(unique(sub_meta[[var]])) < 2 || any(table(sub_meta[[var]]) < 2)) {
      return(tibble(disp_p = NA, disp_note = "Skipped: <2 groups with ≥2 samples"))
    }

    model <- betadisper(d, sub_meta[[var]])
    perm <- permutest(model)
    return(tibble(disp_p = perm$tab[1, "Pr(>F)"], disp_note = NA))
    
  } else {
    # Continuous variable — use regression of average distances
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.matrix(d)
    dist_avg <- apply(d, 1, mean, na.rm = TRUE)

    mod <- lm(dist_avg ~ sub_meta[[var]])
    p <- summary(mod)$coefficients[2, 4]
    return(tibble(disp_p = p, disp_note = NA))
  }
}
```
## Bray-curtis matrix
```{r}
# Extract otu table and metadata table per dataset
# metadata already extracted above for alpha diversity and season added
otu_vgo2_np <- as.data.frame(otu_table(ps_vgo2_np_filtered)) %>% t()

otu_vgo2_op <- as.data.frame(otu_table(ps_vgo2_op_filtered)) %>% t()

otu_vgo3_np <- as.data.frame(otu_table(ps_vgo3_np_filtered)) %>% t()

otu_vgo3_op <- as.data.frame(otu_table(ps_vgo3_op_filtered)) %>% t()

# Compute BC matrices per dataset
bc_dist_vgo2_np <- vegdist(otu_vgo2_np, method = "bray")
bray_vgo2_np_matrix <- bc_dist_vgo2_np %>% as.matrix()

bc_dist_vgo2_op <- vegdist(otu_vgo2_op, method = "bray")
bray_vgo2_op_matrix <- bc_dist_vgo2_op %>% as.matrix()

bc_dist_vgo3_np <- vegdist(otu_vgo3_np, method = "bray")
bray_vgo3_np_matrix <- bc_dist_vgo3_np %>% as.matrix()

bc_dist_vgo3_op <- vegdist(otu_vgo3_op, method = "bray")
bray_vgo3_op_matrix <- bc_dist_vgo3_op %>% as.matrix()
```

## Univariable PERMANOVAs 
### Continuous and categorical lung function variables
#### VGO2 NP
```{r}
# Existing variables and typing
variables_vgo2_np <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status",
  "packyears", "asthma_ever", "copd_gold", "copd_lln",
  "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

# N.B no reference categories need setting for the factor variables for PERMANOVA as it compares the distances between groups to determine differences in community structure. 
meta_vgo2_np[variables_vgo2_np] <- lapply(meta_vgo2_np[variables_vgo2_np], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars_vgo2_np <- variables_vgo2_np[sapply(meta_vgo2_np[variables_vgo2_np], is.factor)]
continuous_vars_vgo2_np <- variables_vgo2_np[sapply(meta_vgo2_np[variables_vgo2_np], is.numeric)]

# Filter categorical variables with enough samples per group
meta_summary_vgo2_np <- pivot_longer(mutate_all(meta_vgo2_np, as.character), 
                                     cols = variables_vgo2_np, names_to = "Var", values_to = "Cat")

meta_summary_vgo2_np_cat <- meta_summary_vgo2_np %>%
  filter(Var %in% categorical_vars_vgo2_np) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_summary_vgo2_np_filtered_cat <- meta_summary_vgo2_np_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_summary_vgo2_np_filtered_cont <- data.frame(Var = continuous_vars_vgo2_np)

final_variable_list_vgo2_np <- unique(c(meta_summary_vgo2_np_filtered_cat$Var,
                                        meta_summary_vgo2_np_filtered_cont$Var))

variable_list_vgo2_np <- tibble(Var = final_variable_list_vgo2_np, fit = NA)

# Run PERMANOVA and dispersion
perm_vgo2_np <- variable_list_vgo2_np %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo2_np_matrix, meta_vgo2_np, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo2_np_matrix, meta_vgo2_np, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo2_np <- perm_vgo2_np %>% select(-fit_Var)

names(perm_vgo2_np) <- sub("^fit_", "", names(perm_vgo2_np))

# Add metadata columns
perm_vgo2_np$type <- "Bray Curtis"
perm_vgo2_np$dataset <- "VGO2_NP"
perm_vgo2_np$niche <- "NP"

# Adjust p-values
perm_vgo2_np$q.value <- p.adjust(perm_vgo2_np$p.value, "BH")

# Add p-value and q-value labels
perm_vgo2_np <- perm_vgo2_np %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

# Save results
write.csv(perm_vgo2_np, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo2_np_with_dispersion.csv")
```
#### VGO2 OP
```{r}
# Variables
variables_vgo2_op <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status",
  "packyears", "asthma_ever", "copd_gold", "copd_lln",
  "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo2_op[variables_vgo2_op] <- lapply(meta_vgo2_op[variables_vgo2_op], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo2_op[sapply(meta_vgo2_op[variables_vgo2_op], is.factor)]
continuous_vars <- variables_vgo2_op[sapply(meta_vgo2_op[variables_vgo2_op], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo2_op, as.character), 
                             cols = variables_vgo2_op, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo2_op <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo2_op_matrix, meta_vgo2_op, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo2_op_matrix, meta_vgo2_op, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo2_op <- perm_vgo2_op %>% select(-fit_Var)
names(perm_vgo2_op) <- sub("^fit_", "", names(perm_vgo2_op))

perm_vgo2_op$type <- "Bray Curtis"
perm_vgo2_op$dataset <- "VGO2_OP"
perm_vgo2_op$niche <- "OP"
perm_vgo2_op$q.value <- p.adjust(perm_vgo2_op$p.value, "BH")

perm_vgo2_op <- perm_vgo2_op %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo2_op, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo2_op_with_dispersion.csv")

```
#### VGO3 NP
```{r}
variables_vgo3_np <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "asthma_ever",
  "copd_doc_diag", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo3_np[variables_vgo3_np] <- lapply(meta_vgo3_np[variables_vgo3_np], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo3_np[sapply(meta_vgo3_np[variables_vgo3_np], is.factor)]
continuous_vars <- variables_vgo3_np[sapply(meta_vgo3_np[variables_vgo3_np], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo3_np, as.character), 
                             cols = variables_vgo3_np, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo3_np <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo3_np_matrix, meta_vgo3_np, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo3_np_matrix, meta_vgo3_np, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo3_np <- perm_vgo3_np %>% select(-fit_Var)
names(perm_vgo3_np) <- sub("^fit_", "", names(perm_vgo3_np))

perm_vgo3_np$type <- "Bray Curtis"
perm_vgo3_np$dataset <- "VGO3_NP"
perm_vgo3_np$niche <- "NP"
perm_vgo3_np$q.value <- p.adjust(perm_vgo3_np$p.value, "BH")

perm_vgo3_np <- perm_vgo3_np %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo3_np, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo3_np_with_dispersion.csv")

```
#### VGO3 OP
```{r}
variables_vgo3_op <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "asthma_ever",
  "copd_doc_diag", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo3_op[variables_vgo3_op] <- lapply(meta_vgo3_op[variables_vgo3_op], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo3_op[sapply(meta_vgo3_op[variables_vgo3_op], is.factor)]
continuous_vars <- variables_vgo3_op[sapply(meta_vgo3_op[variables_vgo3_op], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo3_op, as.character), 
                             cols = variables_vgo3_op, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo3_op <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo3_op_matrix, meta_vgo3_op, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo3_op_matrix, meta_vgo3_op, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo3_op <- perm_vgo3_op %>% select(-fit_Var)
names(perm_vgo3_op) <- sub("^fit_", "", names(perm_vgo3_op))

perm_vgo3_op$type <- "Bray Curtis"
perm_vgo3_op$dataset <- "VGO3_OP"
perm_vgo3_op$niche <- "OP"
perm_vgo3_op$q.value <- p.adjust(perm_vgo3_op$p.value, "BH")

perm_vgo3_op <- perm_vgo3_op %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo3_op, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo3_op_with_dispersion.csv")
```
#### Combine and present all univariable PERMANOVA results
```{r}
# ---- Pretty labels for variables (for plotting) ----
pretty_labels <- c(
  "FEV1_pred"        = "FEV1 % predicted (decline)",
  "FVC_pred"         = "FVC % predicted (decline)",
  "FEV1FVC_pred"     = "FEV1/FVC % predicted (decline)",
  "FEF25_75_pred"    = "FEF 25%-75% % predicted (decline)",
  "FEV1_category"    = "FEV1 % predicted category",
  "FVC_category"     = "FVC % predicted category",
  "FEV1FVC_category" = "FEV1/FVC % predicted category",
  "FEF25_75_category"= "FEF 25%-75% % predicted category",
  "age"              = "Age",
  "gender"           = "Gender",
  "bmi"              = "Body Mass Index",
  "smoke_status"     = "Smoking status",
  "packyears"        = "Pack years of cigarettes",
  "season_microbiome_sampling" = "Season of sampling",
  "asthma_ever"      = "Asthma status",
  "copd_gold"        = "COPD GOLD definition",
  "copd_lln"         = "COPD LLN definition",
  "copd_doc_diag"    = "COPD doctor diagnosed",
  "pneumo_lastyr"    = "Pneumonia in last year (number of times)",
  "pneumo_last3yr"   = "Pneumonia in last 3 years (number of times)"
)

# Apply pretty labels
plot_perm <- perm_all %>%
  rename(variable = Var, R2 = R2, significance = label_q.value) %>%
  mutate(variable = dplyr::recode(variable, !!!pretty_labels))


# Combine all univariable PERMANOVA results
perm_all <- bind_rows(perm_vgo2_np, perm_vgo2_op, perm_vgo3_np, perm_vgo3_op) %>%
  mutate(dataset_niche = paste(dataset, niche, sep = "_")) %>%
  filter(term == "Model")

plot_perm <- perm_all %>%
  rename(variable = Var, R2 = R2, significance = label_q.value)

# Split into datasets
plot_vgo2 <- plot_perm %>% filter(dataset %in% c("VGO2_NP", "VGO2_OP"))
plot_vgo3 <- plot_perm %>% filter(dataset %in% c("VGO3_NP", "VGO3_OP"))

# Define correct variable order (top to bottom in heatmap)
ordered_vars <- rev(c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "packyears",
  "asthma_ever", "copd_gold", "copd_lln", "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred","FEV1_category", "FVC_pred","FVC_category", "FEV1FVC_pred", "FEV1FVC_category", "FEF25_75_pred", "FEF25_75_category"
))

# Format and reorder data
plot_vgo2 <- plot_vgo2 %>%
  mutate(variable = factor(variable, levels = ordered_vars),
         R2 = round(R2, 3))

plot_vgo3 <- plot_vgo3 %>%
  mutate(variable = factor(variable, levels = ordered_vars),
         R2 = round(R2, 3))

# Plot theme and setup
base_theme <- theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid = element_blank()
  )



#### Heatmap for VGO2 ####
# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo2 <- min(plot_vgo2$R2, na.rm = TRUE)
r2_max_vgo2 <- max(plot_vgo2$R2, na.rm = TRUE)
r2_mid_vgo2 <- median(plot_vgo2$R2, na.rm = TRUE)

heatmap_vgo2 <- ggplot(plot_vgo2, aes(x = type, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo2,  # center gradient at median R²
    limits = c(r2_min_vgo2, r2_max_vgo2),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2) 
    ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Explanatory variable")

#### Heatmap for VGO3 ####
# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo3 <- min(plot_vgo3$R2, na.rm = TRUE)
r2_max_vgo3 <- max(plot_vgo3$R2, na.rm = TRUE)
r2_mid_vgo3 <- median(plot_vgo3$R2, na.rm = TRUE)

heatmap_vgo3 <- ggplot(plot_vgo3, aes(x = type, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo3,  # center gradient at median R²
    limits = c(r2_min_vgo3, r2_max_vgo3),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2) 
  ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Explanatory variable")

#### Combine and save plot ####
combined_plot <- (heatmap_vgo2 / heatmap_vgo3) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

univar_beta_plot <- combined_plot +
  patchwork::plot_annotation(
    title = "Univariable PERMANOVAs (VGO2 and VGO3)",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  )

print(univar_beta_plot)

ggsave("../Output/Beta_diversity/Univariable_PERMANOVAS/Univariable_PERMANOVA_R2_heatmap.svg",
       univar_beta_plot, width = 5, height = 7)
```

# Multivariable PERMANOVAs
## Prepare data 
```{r}
# These ps objects are already filtered and converted to RA
ps_vgo2_np_filtered
ps_vgo2_op_filtered
ps_vgo3_np_filtered
ps_vgo3_op_filtered

# Convert the otu tables from the above phyloseq objects to dataframes (samples as rows)
otu_vgo2_np <- as.data.frame(otu_table(ps_vgo2_np_filtered)) %>% t()
otu_vgo2_op <- as.data.frame(otu_table(ps_vgo2_op_filtered)) %>% t()
otu_vgo3_np <- as.data.frame(otu_table(ps_vgo3_np_filtered)) %>% t()
otu_vgo3_op <- as.data.frame(otu_table(ps_vgo3_op_filtered)) %>% t()

# metadata already extracted for each ps object
meta_vgo2_np
meta_vgo2_op
meta_vgo3_np
meta_vgo3_op

# Compute BC dissimilarity matrices for each ps
ord_vgo2_np <- vegdist(otu_vgo2_np, method = "bray")
ord_vgo2_op <- vegdist(otu_vgo2_op, method = "bray")
ord_vgo3_np <- vegdist(otu_vgo3_np, method = "bray")
ord_vgo3_op <- vegdist(otu_vgo3_op, method = "bray")

# Compute mean BC dissimilarity
mean(ord_vgo2_np)
mean(ord_vgo2_op)
mean(ord_vgo3_np)
mean(ord_vgo3_op)
```

## Multivariable PERMANOVAs for lung function 
```{r}
run_permanova <- function(ord, meta, formula_vars) {
  # Ensure ord is a dist object
  ord <- as.matrix(ord)
  
  # Intersect sample IDs
  common_samples <- intersect(rownames(ord), rownames(meta))
  ord <- as.dist(ord[common_samples, common_samples])
  meta <- meta[common_samples, , drop = FALSE]

  n_before <- nrow(meta)

  # Drop rows with NA in any of the formula variables
  meta <- meta %>% filter(if_all(all_of(formula_vars), ~ !is.na(.)))
  keep_samples <- rownames(meta)
  ord <- as.dist(as.matrix(ord)[keep_samples, keep_samples])
  
  n_after <- nrow(meta)
  message(glue("Samples before filtering: {n_before} | After filtering: {n_after} | Removed: {n_before - n_after}"))
  
  # Remove categorical vars with <2 levels
  bad_vars <- formula_vars[sapply(meta[formula_vars], function(x) is.factor(x) && nlevels(factor(x)) < 2)]
  if (length(bad_vars) > 0) {
    message(glue("Dropping variables with only one level: {paste(bad_vars, collapse = ', ')}"))
    formula_vars <- setdiff(formula_vars, bad_vars)
  }
  if (length(formula_vars) == 0) stop("No variables with ≥2 levels remain after filtering.")

  # Build and run formula
  formula_str <- glue("ord ~ {paste(formula_vars, collapse = ' + ')}")
  adonis_res <- adonis2(as.formula(formula_str), data = meta, permutations = 999)
  tidy_res <- broom::tidy(adonis_res)
  tidy_res$term <- rownames(adonis_res)
  
  return(tidy_res)
}


# Fix metadata so that rownames are sample_id values (as the permanova function uses rownames) 
meta_vgo2_np; meta_vgo2_op; meta_vgo3_np;meta_vgo3_op
rownames(meta_vgo2_op) <- meta_vgo2_op$sample_id
rownames(meta_vgo3_np) <- meta_vgo3_np$sample_id
rownames(meta_vgo3_op) <- meta_vgo3_op$sample_id

#### FEV1 predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1_pred <- c("FEV1_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")
str(meta_vgo2_np$age)
str(meta_vgo2_np$gender)
str(meta_vgo2_np$season_microbiome_sampling)
str(meta_vgo2_np$smoke_status)

# PERMANOVAs
perm_FEV1_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1_pred)
perm_FEV1_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1_pred)
perm_FEV1_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1_pred)
perm_FEV1_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1_pred)


#### FEV1 categorised predicted multivariable PERMANOVA models ####
# Select required variables
covariates_FEV1_pred <- c("FEV1_category", "age", "season_microbiome_sampling",
                          "gender", "smoke_status")

# PERMANOVAs
perm_FEV1_cat_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1_pred)
perm_FEV1_cat_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1_pred)
perm_FEV1_cat_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1_pred)
perm_FEV1_cat_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1_pred)


#### FVC predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FVC <- c("FVC_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FVC_pred", "age")] <- lapply(meta_vgo2_np[c("FVC_pred", "age")], as.numeric)
meta_vgo2_op[c("FVC_pred", "age")] <- lapply(meta_vgo2_op[c("FVC_pred", "age")], as.numeric)
meta_vgo3_np[c("FVC_pred", "age")] <- lapply(meta_vgo3_np[c("FVC_pred", "age")], as.numeric)
meta_vgo3_op[c("FVC_pred", "age")] <- lapply(meta_vgo3_op[c("FVC_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FVC_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FVC)
perm_FVC_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FVC)
perm_FVC_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FVC)
perm_FVC_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FVC)


#### FVC categorised predicted multivariable PERMANOVA models ####
# Select required variables
covariates_FVC_pred <- c("FVC_category", "age", "season_microbiome_sampling",
                          "gender", "smoke_status")

# Convert FVC_pred and age to numeric
meta_vgo2_np[c("FVC_category", "age")] <- lapply(meta_vgo2_np[c("FVC_category", "age")], as.numeric)
meta_vgo2_op[c("FVC_category", "age")] <- lapply(meta_vgo2_op[c("FVC_category", "age")], as.numeric)
meta_vgo3_np[c("FVC_category", "age")] <- lapply(meta_vgo3_np[c("FVC_category", "age")], as.numeric)
meta_vgo3_op[c("FVC_category", "age")] <- lapply(meta_vgo3_op[c("FVC_category", "age")], as.numeric)

# PERMANOVAs
perm_FVC_cat_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FVC_pred)
perm_FVC_cat_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FVC_pred)
perm_FVC_cat_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FVC_pred)
perm_FVC_cat_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FVC_pred)


#### FEV1FVC predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1FVC <- c("FEV1FVC_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo2_np[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo2_op[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo2_op[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo3_np[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo3_np[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo3_op[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo3_op[c("FEV1FVC_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEV1FVC_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1FVC)
perm_FEV1FVC_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1FVC)
perm_FEV1FVC_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1FVC)
perm_FEV1FVC_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1FVC)


#### FEV1FVC_category_category predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1FVC_category <- c("FEV1FVC_category", "age", "season_microbiome_sampling",
                        "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEV1FVC_category", "age")] <- lapply(meta_vgo2_np[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo2_op[c("FEV1FVC_category", "age")] <- lapply(meta_vgo2_op[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo3_np[c("FEV1FVC_category", "age")] <- lapply(meta_vgo3_np[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo3_op[c("FEV1FVC_category", "age")] <- lapply(meta_vgo3_op[c("FEV1FVC_category", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEV1FVC_category_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1FVC_category)


#### FEF25_75 predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEF25_75 <- c("FEF25_75_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEF25_75_pred", "age")] <- lapply(meta_vgo2_np[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo2_op[c("FEF25_75_pred", "age")] <- lapply(meta_vgo2_op[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo3_np[c("FEF25_75_pred", "age")] <- lapply(meta_vgo3_np[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo3_op[c("FEF25_75_pred", "age")] <- lapply(meta_vgo3_op[c("FEF25_75_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEF25_75_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEF25_75)
perm_FEF25_75_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEF25_75)
perm_FEF25_75_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEF25_75)
perm_FEF25_75_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEF25_75)

#### FEF25_75_category predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEF25_75_category <- c("FEF25_75_category_pred", "age", "season_microbiome_sampling",
                         "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEF25_75_category", "age")] <- lapply(meta_vgo2_np[c("FEF25_75_category", "age")], as.numeric)
meta_vgo2_op[c("FEF25_75_category", "age")] <- lapply(meta_vgo2_op[c("FEF25_75_category", "age")], as.numeric)
meta_vgo3_np[c("FEF25_75_category", "age")] <- lapply(meta_vgo3_np[c("FEF25_75_category", "age")], as.numeric)
meta_vgo3_op[c("FEF25_75_category", "age")] <- lapply(meta_vgo3_op[c("FEF25_75_category", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEF25_75_category_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEF25_75_category)
```

## Combine and present all multivariable PERMANOVA results
```{r}
# Combine all multivariable PERMANOVA results into one dataframe
perm_all_multi <- bind_rows(
  perm_FEV1_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1_pred"),
  perm_FEV1_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1_pred"),
  perm_FEV1_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1_pred"),
  perm_FEV1_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1_pred"),
  
  perm_FEV1_cat_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1_category"),
  
  perm_FVC_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FVC_pred"),
  perm_FVC_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FVC_pred"),
  perm_FVC_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FVC_pred"),
  perm_FVC_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FVC_pred"),
  
  perm_FVC_cat_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FVC_category"),
  perm_FVC_cat_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FVC_category"),
  perm_FVC_cat_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FVC_category"),
  perm_FVC_cat_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FVC_category"),
  
  perm_FEV1FVC_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1FVC_pred"),
  
  perm_FEV1FVC_category_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1FVC_category"),
  
  perm_FEF25_75_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEF25_75_pred"),
  
  perm_FEF25_75_category_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEF25_75_category")
) %>%
  # Keep only the overall model rows
  filter(term == "Model") %>%
  mutate(dataset_niche = paste(dataset, niche, sep = "_"))

# Heatmap
# Reorder variables for heatmap (consistent with univariable plots)
ordered_vars <- rev(c(
  "age", "gender", "season_microbiome_sampling", "smoke_status",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category"
))

# Prepare plotting dataframe
perm_all_multi$p.value
plot_perm_multi <- perm_all_multi %>%
  mutate(variable = outcome,
         variable = factor(variable, levels = ordered_vars),
         significance = case_when(
           p.value < 0.001 ~ "***",
           p.value < 0.01  ~ "**",
           p.value < 0.05  ~ "*",
           TRUE ~ "ns"
         ),
         R2 = round(R2, 3))

# Split into VGO2 and VGO3 panels
plot_vgo2_multi <- plot_perm_multi %>% filter(dataset == "VGO2")
plot_vgo3_multi <- plot_perm_multi %>% filter(dataset == "VGO3")

# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo2 <- min(plot_vgo2_multi$R2, na.rm = TRUE)
r2_max_vgo2 <- max(plot_vgo2_multi$R2, na.rm = TRUE)
r2_mid_vgo2 <- median(plot_vgo2_multi$R2, na.rm = TRUE)

r2_min_vgo3 <- min(plot_vgo3_multi$R2, na.rm = TRUE)
r2_max_vgo3 <- max(plot_vgo3_multi$R2, na.rm = TRUE)
r2_mid_vgo3 <- median(plot_vgo3_multi$R2, na.rm = TRUE)

# Heatmap for VGO2
heatmap_vgo2_multi <- ggplot(plot_vgo2_multi, aes(x = niche, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo2,  # center gradient at median R²
    limits = c(r2_min_vgo2, r2_max_vgo2),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2)
  ) +
  facet_wrap(~ dataset, ncol = 1) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(y = "Lung Function Variable")

# Heatmap for VGO3
heatmap_vgo3_multi <- ggplot(plot_vgo3_multi, aes(x = niche, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo3,  # center gradient at median R²
    limits = c(r2_min_vgo3, r2_max_vgo3),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2)
  ) +
  facet_wrap(~ dataset, ncol = 1) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(y = "Lung Function Variable")

# Combine both heatmaps
multi_beta_plot <- (heatmap_vgo2_multi / heatmap_vgo3_multi) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

multi_beta_plot <- multi_beta_plot +
  patchwork::plot_annotation(
    title = "Multivariable PERMANOVAs (VGO2 and VGO3)",
    caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant",
    theme = theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      plot.caption = element_text(size = 8, hjust = 0.5)
    )
  )

print(multi_beta_plot)

# Save figure
ggsave("../Output/Beta_diversity/Multivariable_PERMANOVAS/Multivariable_PERMANOVA_R2_heatmap.svg",
       multi_beta_plot, width = 4.5, height = 5)

```
# Differential abundance analysis
## Functions
```{r}
# Function to conduct MaAsLin2
# do_maaslin2 <- function(ps, var, other_var, ref = NULL) { 
#   #pre-processing
#   sample_data(ps) <- sample_data(ps)[!is.na(sample_data(ps)[,var])]
#   ps <- prune_taxa(taxa_sums(ps) > 0, ps)
#   otu <- ps@otu_table %>% as.data.frame() %>% t() %>% as.data.frame()
#   meta <- meta_to_df(ps)
#   rownames(meta) <- meta$sample_id
#   meta <- as.data.frame(meta)
#   otu <- otu[rownames(otu) %in% rownames(meta),]
#   #Maaslin2
#   ms = Maaslin2(
#     input_data = otu, 
#     input_metadata = meta, 
#     output = var, 
#     fixed_effects = c(var, other_var), 
#     normalization = "NONE",
#     transform = "LOG",
#     min_abundance = 0,
#     min_prevalence = 0,
#     plot_heatmap = FALSE,
#     plot_scatter = FALSE,
#     reference = ref)
#   #Edit dataframe
#   ms <- ms$results
#   ms$var <- var
#   
#   ms <- ms[ms$metadata == var,]
#   
#   ms <- ms %>% mutate(conf.low = coef-1.96*stderr,
#                       conf.high = coef+1.96*stderr,
#                       label = ifelse(qval < 0.1, "*", NA)
#                       )
#   return(ms)
# }

do_maaslin2 <- function(ps, var, other_var, ref = NULL, out_dir = NULL) {
  stopifnot(inherits(ps, "phyloseq"))
  if (is.null(other_var)) other_var <- character(0)
  fixed <- unique(c(var, other_var))

  ## 1) Drop samples with NA in model vars (based on a clean, flattened meta first)
  meta0 <- data.frame(as(phyloseq::sample_data(ps), "data.frame"), check.names = FALSE)
  miss <- setdiff(fixed, colnames(meta0))
  if (length(miss)) stop("Missing columns in metadata: ", paste(miss, collapse = ", "))

  keep <- stats::complete.cases(meta0[, fixed, drop = FALSE])
  ps <- phyloseq::prune_samples(keep, ps)

  ## 2) Rebuild and sanitize metadata (NO sample_id column)
  meta <- data.frame(as(phyloseq::sample_data(ps), "data.frame"), check.names = FALSE)

  # Drop/flatten non-atomic columns
  is_bad <- vapply(meta, function(x) is.list(x) || is.data.frame(x) || is.matrix(x), logical(1))
  if (any(is_bad)) {
    for (nm in names(meta)[is_bad]) meta[[nm]] <- as.character(meta[[nm]])
  }

  # Make ordered factors plain factors; characters -> factors; logical -> factors
  for (nm in names(meta)) {
    x <- meta[[nm]]
    if (is.ordered(x))             meta[[nm]] <- factor(x, ordered = FALSE)
    else if (is.character(x))      meta[[nm]] <- factor(x)
    else if (is.logical(x))        meta[[nm]] <- factor(x, levels = c(FALSE, TRUE))
    # leave numeric/integer as-is
  }

  # Strongly expected numerics -> numeric
  num_candidates <- intersect(c("age","bmi","packyears","FEV1_pred","FVC_pred",
                                "FEV1FVC_pred","FEF25_75_pred"), fixed)
  for (nm in num_candidates) if (nm %in% names(meta))
    meta[[nm]] <- suppressWarnings(as.numeric(as.character(meta[[nm]])))

  # Drop rows with NA in fixed effects after coercion
  keep2 <- stats::complete.cases(meta[, fixed, drop = FALSE])
  if (!all(keep2)) {
    meta <- meta[keep2, , drop = FALSE]
    ps   <- phyloseq::prune_samples(rownames(meta), ps)
  }

  # Drop zero-variance predictors (single-level factors or constant numerics)
  drop_fx <- vapply(fixed, function(nm) {
    x <- meta[[nm]]
    if (is.factor(x)) nlevels(x) < 2
    else if (is.numeric(x)) sd(x, na.rm = TRUE) == 0
    else FALSE
  }, logical(1))
  if (any(drop_fx)) {
    msg <- paste(fixed[drop_fx], collapse = ", ")
    warning("Dropping zero-variance predictors: ", msg)
    fixed <- fixed[!drop_fx]
  }
  if (!length(fixed)) stop("All predictors dropped (zero variance).")

  ## 3) Build OTU (samples as rows) and align strictly
  otu <- phyloseq::otu_table(ps)
  otu_df <- as.data.frame(otu)
  if (phyloseq::taxa_are_rows(ps)) otu_df <- t(otu_df)
  otu_df <- as.data.frame(otu_df, check.names = FALSE)
  # numeric ensure
  otu_df[] <- lapply(otu_df, function(x) suppressWarnings(as.numeric(x)))

  common <- intersect(rownames(meta), rownames(otu_df))
  meta   <- meta[common, , drop = FALSE]
  otu_df <- otu_df[common, , drop = FALSE]
  stopifnot(identical(rownames(meta), rownames(otu_df)))

  ## 4) Output dir
  if (is.null(out_dir)) out_dir <- file.path(tempdir(), paste0("Maaslin2_", var))

  ## 5) Run MaAsLin2
  ms <- Maaslin2::Maaslin2(
    input_data     = otu_df,
    input_metadata = meta,
    output         = out_dir,
    fixed_effects  = fixed,
    normalization  = "NONE",
    transform      = "LOG",
    min_abundance  = 0,
    min_prevalence = 0,
    plot_heatmap   = FALSE,
    plot_scatter   = FALSE,
    reference      = ref
  )

  ## 6) Post-process
  res <- ms$results
  if (is.null(res) || !nrow(res)) return(res)
  # ensure numeric
  for (nm in c("coef","stderr","qval")) if (nm %in% names(res))
    res[[nm]] <- suppressWarnings(as.numeric(res[[nm]]))
  res$var <- var
  res <- res[res$metadata == var, , drop = FALSE]
  if (!nrow(res)) return(res)
  res$conf.low  <- res$coef - 1.96 * res$stderr
  res$conf.high <- res$coef + 1.96 * res$stderr
  res$label     <- ifelse(res$qval < 0.1, "*", NA)
  res
}

# Maaslin2:::LOG  # The internal MaAsLin log transformation takes the lowest species count and divides by 2 

```
## Preparing data
```{r}
# These datasets were created above - they have already been filtered and converted to relative abundance
ps_vgo2_np_filtered
ps_vgo2_op_filtered
ps_vgo3_np_filtered
ps_vgo3_op_filtered

# Count number of ASVs and genera left in ps objects
length(unique(as.data.frame(tax_table(ps_vgo2_np_filtered))$ASV)) # 893
length(unique(as.data.frame(tax_table(ps_vgo2_np_filtered))$genus)) # 264

# Count number of ASVs and genera left in ps objects
length(unique(as.data.frame(tax_table(ps_vgo2_op_filtered))$ASV)) # 337
length(unique(as.data.frame(tax_table(ps_vgo2_op_filtered))$genus)) # 50

# Count number of ASVs and genera left in ps objects
length(unique(as.data.frame(tax_table(ps_vgo3_np_filtered))$ASV)) # 883
length(unique(as.data.frame(tax_table(ps_vgo3_np_filtered))$genus)) # 210

# Count number of ASVs and genera left in ps objects
length(unique(as.data.frame(tax_table(ps_vgo3_op_filtered))$ASV)) # 1118
length(unique(as.data.frame(tax_table(ps_vgo3_op_filtered))$genus)) # 95
```
### Convert to absolute abundances 
```{r}
# Convert RA of ASVs by multiplying by the 16S qPCR measurements
vgo2_np_otu <- otu_tab_to_df(ps_vgo2_np_filtered)
vgo2_np_meta <- meta_to_df(ps_vgo2_np_filtered, sample_name = ".row_id")
# Multiply RA by qPCR 
vgo2_np_otu <- vgo2_np_otu %>%
  mutate(across(-sample_id, ~.x*meta_vgo2_np$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo2_np_otu) <- vgo2_np_otu$sample_id
# Remove the first column, transpose and convert to df
vgo2_np_otu <- vgo2_np_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo2_np_ps_abs <- phyloseq(otu_table(vgo2_np_otu, 
                             taxa_are_rows = T), 
                   tax_table(ps_vgo2_np_filtered), 
                   sample_data(ps_vgo2_np_filtered))

#### VGO2 OP ####
# Convert RA of ASVs by multiplying by the 16S qPCR measurements
vgo2_op_otu <- otu_tab_to_df(ps_vgo2_op_filtered)
vgo2_op_meta <- meta_to_df(ps_vgo2_op_filtered, sample_name = ".row_id")
# Multiply RA by qPCR 
vgo2_op_otu <- vgo2_op_otu %>%
  mutate(across(-sample_id, ~.x*meta_vgo2_op$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo2_op_otu) <- vgo2_op_otu$sample_id
# Remove the first column, transpose and convert to df
vgo2_op_otu <- vgo2_op_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo2_op_ps_abs <- phyloseq(otu_table(vgo2_op_otu, 
                                     taxa_are_rows = T), 
                           tax_table(ps_vgo2_op_filtered), 
                           sample_data(ps_vgo2_op_filtered))


#### VGO3 NP ####
# Convert RA of ASVs by multiplying by the 16S qPCR measurements
vgo3_np_otu <- otu_tab_to_df(ps_vgo3_np_filtered)
vgo3_np_meta <- meta_to_df(ps_vgo3_np_filtered, sample_name = ".row_id")
# Multiply RA by qPCR 
vgo3_np_otu <- vgo3_np_otu %>%
  mutate(across(-sample_id, ~.x*meta_vgo3_np$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo3_np_otu) <- vgo3_np_otu$sample_id
# Remove the first column, transpose and convert to df
vgo3_np_otu <- vgo3_np_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo3_np_ps_abs <- phyloseq(otu_table(vgo3_np_otu, 
                                     taxa_are_rows = T), 
                           tax_table(ps_vgo3_np_filtered), 
                           sample_data(ps_vgo3_np_filtered))

#### VGO3 OP ####
# Convert RA of ASVs by multiplying by the 16S qPCR measurements
vgo3_op_otu <- otu_tab_to_df(ps_vgo3_op_filtered)
vgo3_op_meta <- meta_to_df(ps_vgo3_op_filtered, sample_name = ".row_id")
# Multiply RA by qPCR 
vgo3_op_otu <- vgo3_op_otu %>%
  mutate(across(-sample_id, ~.x*meta_vgo3_op$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo3_op_otu) <- vgo3_op_otu$sample_id
# Remove the first column, transpose and convert to df
vgo3_op_otu <- vgo3_op_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo3_op_ps_abs <- phyloseq(otu_table(vgo3_op_otu, 
                                     taxa_are_rows = T), 
                           tax_table(ps_vgo3_op_filtered), 
                           sample_data(ps_vgo3_op_filtered))
```
### Agglomerate data to genus level
```{r}
#### VGO2 NP ####
vgo2_np_ps_RA_genus <- vgo2_np_ps_abs %>%
  tax_glom(taxrank = "genus") %>% 
  to_RA() # 262 genera in the CP
# Change names to the genus-level names
taxa_names(vgo2_np_ps_RA_genus) <- vgo2_np_ps_RA_genus@tax_table[,6] 
# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
vgo2_np_genus_otu <- otu_tab_to_df(vgo2_np_ps_RA_genus)
vgo2_np_genus_meta <- meta_to_df(vgo2_np_ps_RA_genus, sample_name = ".row_id")
# Multiple RA by qPCR 
vgo2_np_genus_otu <- vgo2_np_genus_otu %>%
  mutate(across(-sample_id, ~.x*vgo2_np_genus_meta$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo2_np_genus_otu) <- vgo2_np_genus_otu$sample_id
# Remove the first column, transpose and convert to df
vgo2_np_genus_otu <- vgo2_np_genus_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo2_np_ps_genus_abs <- phyloseq(otu_table(vgo2_np_genus_otu, 
                             taxa_are_rows = T), 
                   tax_table(vgo2_np_ps_RA_genus), 
                   sample_data(vgo2_np_ps_RA_genus))


#### VGO2 OP ####
vgo2_op_ps_RA_genus <- vgo2_op_ps_abs %>%
  tax_glom(taxrank = "genus") %>% 
  to_RA() # 262 genera in the CP
# Change names to the genus-level names
taxa_names(vgo2_op_ps_RA_genus) <- vgo2_op_ps_RA_genus@tax_table[,6] 
# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
vgo2_op_genus_otu <- otu_tab_to_df(vgo2_op_ps_RA_genus)
vgo2_op_genus_meta <- meta_to_df(vgo2_op_ps_RA_genus, sample_name = ".row_id")
# Multiple RA by qPCR 
vgo2_op_genus_otu <- vgo2_op_genus_otu %>%
  mutate(across(-sample_id, ~.x*vgo2_op_genus_meta$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo2_op_genus_otu) <- vgo2_op_genus_otu$sample_id
# Remove the first column, transpose and convert to df
vgo2_op_genus_otu <- vgo2_op_genus_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo2_op_ps_genus_abs <- phyloseq(otu_table(vgo2_op_genus_otu, 
                                           taxa_are_rows = T), 
                                 tax_table(vgo2_op_ps_RA_genus), 
                                 sample_data(vgo2_op_ps_RA_genus))

#### VGO3 NP ####
vgo3_np_ps_RA_genus <- vgo3_np_ps_abs %>%
  tax_glom(taxrank = "genus") %>% 
  to_RA() # 262 genera in the CP
# Change names to the genus-level names
taxa_names(vgo3_np_ps_RA_genus) <- vgo3_np_ps_RA_genus@tax_table[,6] 
# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
vgo3_np_genus_otu <- otu_tab_to_df(vgo3_np_ps_RA_genus)
vgo3_np_genus_meta <- meta_to_df(vgo3_np_ps_RA_genus, sample_name = ".row_id")
# Multiple RA by qPCR 
vgo3_np_genus_otu <- vgo3_np_genus_otu %>%
  mutate(across(-sample_id, ~.x*vgo3_np_genus_meta$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo3_np_genus_otu) <- vgo3_np_genus_otu$sample_id
# Remove the first column, transpose and convert to df
vgo3_np_genus_otu <- vgo3_np_genus_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo3_np_ps_genus_abs <- phyloseq(otu_table(vgo3_np_genus_otu, 
                                           taxa_are_rows = T), 
                                 tax_table(vgo3_np_ps_RA_genus), 
                                 sample_data(vgo3_np_ps_RA_genus))


#### VGO3 OP ####
vgo3_op_ps_RA_genus <- vgo3_op_ps_abs %>%
  tax_glom(taxrank = "genus") %>% 
  to_RA() # 262 genera in the CP
# Change names to the genus-level names
taxa_names(vgo3_op_ps_RA_genus) <- vgo3_op_ps_RA_genus@tax_table[,6] 
# Convert relative abundances to absolute abundances by multiplying the ASV table with the qPCR results
vgo3_op_genus_otu <- otu_tab_to_df(vgo3_op_ps_RA_genus)
vgo3_op_genus_meta <- meta_to_df(vgo3_op_ps_RA_genus, sample_name = ".row_id")
# Multiple RA by qPCR 
vgo3_op_genus_otu <- vgo3_op_genus_otu %>%
  mutate(across(-sample_id, ~.x*vgo3_op_genus_meta$qPCR)) %>% 
  as.data.frame()
# Make sample IDs the row names of the otu table
rownames(vgo3_op_genus_otu) <- vgo3_op_genus_otu$sample_id
# Remove the first column, transpose and convert to df
vgo3_op_genus_otu <- vgo3_op_genus_otu[,-1] %>% 
  t() %>%
  as.data.frame() 
# Create ps object with absolute abundance otu table
vgo3_op_ps_genus_abs <- phyloseq(otu_table(vgo3_op_genus_otu, 
                                           taxa_are_rows = T), 
                                 tax_table(vgo3_op_ps_RA_genus), 
                                 sample_data(vgo3_op_ps_RA_genus))
```

### Additional filtering per dataset
```{r}
#### ASV level - VGO2 NP ####
# Set up a prevalence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# First determine what is 10% of samples
filter_vgo2_np <- nrow(vgo2_np_ps_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo2_np_ps_abs_filt <- vgo2_np_ps_abs %>%
  pres_abund_filter(pres = filter_vgo2_np, abund = 0.005)
#A total of 27 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 21 samples (n = 862 ASVs excluded).
# Count number of ASVs left in ps object

#### ASV level - VGO2 OP ####
# Set up a prevalence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# First determine what is 10% of samples
filter_vgo2_op <- nrow(vgo2_op_ps_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo2_op_ps_abs_filt <- vgo2_op_ps_abs %>%
  pres_abund_filter(pres = filter_vgo2_op, abund = 0.005)
# A total of 49 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 17.9 samples (n = 288 ASVs excluded).
# Count number of ASVs left in ps object

#### ASV level - VGO3 NP ####
# Set up a prevalence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# First determine what is 10% of samples
filter_vgo3_np <- nrow(vgo3_np_ps_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo3_np_ps_abs_filt <- vgo3_np_ps_abs %>%
  pres_abund_filter(pres = filter_vgo3_np, abund = 0.005)
# A total of 13 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 79.9 samples (n = 867 ASVs excluded).

#### ASV level - VGO3 OP ####
# Set up a prevalence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# First determine what is 10% of samples
filter_vgo3_op <- nrow(vgo3_op_ps_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo3_op_ps_abs_filt <- vgo3_op_ps_abs %>%
  pres_abund_filter(pres = filter_vgo3_op, abund = 0.005)
# A total of 52 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 95.2 samples (n = 1066 ASVs excluded).
```
### Additional filtering per dataset on genus level 
```{r}
#### VGO2 NP ####
# Set up a presence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_vgo2_np_genus <- nrow(vgo2_np_ps_genus_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo2_np_ps_genus_abs_filt <- vgo2_np_ps_genus_abs %>%
  pres_abund_filter(pres = filter_vgo2_np_genus, abund = 0.005)
# A total of 21 Genera were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 21 samples (n = 222 ASVs excluded).

#### VGO2 OP ####
# Set up a presence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_vgo2_op_genus <- nrow(vgo2_op_ps_genus_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo2_op_ps_genus_abs_filt <- vgo2_op_ps_genus_abs %>%
  pres_abund_filter(pres = filter_vgo2_op_genus, abund = 0.005)
# A total of 23 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 17.9 samples (n = 23 ASVs excluded).

#### VGO3 NP ####
# Set up a presence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_vgo3_np_genus <- nrow(vgo3_np_ps_genus_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo3_np_ps_genus_abs_filt <- vgo3_np_ps_genus_abs %>%
  pres_abund_filter(pres = filter_vgo3_np_genus, abund = 0.005)
# A total of 9 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 79.9 samples (n = 191 ASVs excluded).

#### VGO3 OP ####
# Set up a presence abundance filter: Keep samples that have a presence in at least 10% of all samples with an abundance of 0.01
# Determine what is 10% of samples
filter_vgo3_op_genus <- nrow(vgo3_op_ps_genus_abs@sam_data) * 0.1 # 10% of total samples = 21
# Apply the filter
vgo3_op_ps_genus_abs_filt <- vgo3_op_ps_genus_abs %>%
  pres_abund_filter(pres = filter_vgo3_op_genus, abund = 0.005)
# A total of 25 ASVs were found to be present at or above a level of confident detection (0.5% relative abundance) in at least 95.2 samples (n = 60 ASVs excluded).
```
## MaAsLin2 analysis
```{r}
# Use your already-filtered + RA phyloseq objects:
ps_list_DA_ASV <- list(
  VGO2_NP = vgo2_np_ps_abs_filt,
  VGO2_OP = vgo2_op_ps_abs_filt,
  VGO3_NP = vgo2_np_ps_abs_filt,
  VGO3_OP = vgo3_op_ps_abs_filt
)

ps_list_DA_genus <- list(
  VGO2_NP_genus = vgo2_np_ps_genus_abs_filt,
  VGO2_OP_genus = vgo2_op_ps_genus_abs_filt,
  VGO3_NP_genus = vgo3_np_ps_genus_abs_filt,
  VGO3_OP_genus = vgo3_op_ps_genus_abs_filt
)


# Function to process the metadata of a single phyloseq object
process_phyloseq_metadata <- function(ps_obj) {
  df <- as.data.frame(sample_data(ps_obj))

  # Binary
  if ("gender"        %in% names(df)) df$gender        <- set_binary(df$gender,        ref = "male")
  if ("asthma_ever"   %in% names(df)) df$asthma_ever   <- set_binary(df$asthma_ever,   ref = "no")
  if ("copd_doc_diag" %in% names(df)) df$copd_doc_diag <- set_binary(df$copd_doc_diag, ref = "no")
  if ("pneumo_lastyr" %in% names(df)) df$pneumo_lastyr <- set_binary(df$pneumo_lastyr, ref = "no")

  # Ordinal
  if ("smoke_status" %in% names(df))
    df$smoke_status <- set_ordered(df$smoke_status, ord_levels$smoke_status)
  if ("season_microbiome_sampling" %in% names(df))
    df$season_microbiome_sampling <- set_ordered(df$season_microbiome_sampling, ord_levels$season_microbiome_sampling)

  for (v in c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")) {
    if (v %in% names(df)) df[[v]] <- set_ordered(df[[v]], ord_levels[[v]])
  }

  if ("copd_gold" %in% names(df)) df$copd_gold <- set_gold_ordered(df$copd_gold)

  # Numeric
  num_names <- intersect(c("age","bmi","packyears","pneumo_last3yr",
                           "FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred"),
                         names(df))
  for (v in num_names) df[[v]] <- suppressWarnings(as.numeric(as.character(df[[v]])))

  # Set back to phyloseq object
  sample_data(ps_obj) <- sample_data(df)
  return(ps_obj)
}

# Apply to ASV-level data
ps_list_DA_ASV <- lapply(ps_list_DA_ASV, process_phyloseq_metadata)
# Apply to Genus-level data
ps_list_DA_genus <- lapply(ps_list_DA_genus, process_phyloseq_metadata)

# MULTIVARIABLE for lung function variables 
# Batch MaAsLin2 runner
safe_name <- function(x) gsub("[^A-Za-z0-9._-]+", "_", x)

ensure_dirs <- function(out_root, datasets, levels) {
  for (ds in datasets) for (lvl in levels) {
    subdir <- file.path(out_root, ds, lvl)       # <- ASV / Genus only
    dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
    if (!dir.exists(subdir)) stop("Cannot create output dir: ", subdir)
  }
}

run_maaslin_batch <- function(
  ps_lists,
  var_sets,
  covars,
  out_root,
  refs = NULL,
  arrange_by = "coef",
  cont_decline = c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred")
) {
  all_out <- list()

  # all dataset names across levels
  all_ds <- unlist(lapply(ps_lists, names), use.names = FALSE)
  # standardize folder names for levels
  level_folders <- names(ps_lists)               # should be c("ASV","Genus")
  stopifnot(all(level_folders %in% c("ASV","Genus")))

  # 0) pre-create all dirs
  ensure_dirs(out_root, all_ds, level_folders)

  for (level in names(ps_lists)) {               # "ASV" / "Genus"
    ps_list <- ps_lists[[level]]

    for (dataset in names(ps_list)) {
      ps <- ps_list[[dataset]]

      # 1) create decline versions in sample_data (if present)
      meta <- as.data.frame(sample_data(ps), check.names = FALSE)
      for (v in cont_decline) {
        if (v %in% names(meta)) {
          meta[[paste0(v, "_decline")]] <- -1 * suppressWarnings(as.numeric(meta[[v]]))
        }
      }
      sample_data(ps) <- sample_data(meta)

      for (set_name in names(var_sets)) {
        var_vec <- var_sets[[set_name]]

        for (v in var_vec) {
          ref_spec <- if (!is.null(refs) && v %in% names(refs)) refs[[v]] else NULL

          res <- try(do_maaslin2(
            ps  = ps,
            var = v,
            other_var = covars,
            ref = ref_spec
          ), silent = TRUE)

          if (inherits(res, "try-error") || is.null(res) || !nrow(res)) {
            message("No results for: ", dataset, " | ", level, " | ", v, " (", set_name, ")")
            next
          }

          if (arrange_by %in% colnames(res)) {
            res <- dplyr::arrange(res, dplyr::desc(.data[[arrange_by]]))
          }

          res$dataset <- dataset
          res$level   <- level
          res$var_set <- set_name

          key <- paste(level, dataset, set_name, v, sep = "___")
          all_out[[key]] <- res

          # 2) write to standardized folder (ASV / Genus)
          subdir <- file.path(out_root, dataset, level)
          if (!dir.exists(subdir)) dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
          if (!dir.exists(subdir)) stop("Output dir still missing: ", subdir)

          fn <- paste0(safe_name(dataset), "_", level, "_", safe_name(v), "_", set_name, ".csv")
          utils::write.csv(res, file.path(subdir, fn), row.names = FALSE)
        }
      }
    }
  }

  dplyr::bind_rows(all_out)
}


# What to run
ps_list_DA_ASV <- list(
  VGO2_NP = vgo2_np_ps_abs_filt,
  VGO2_OP = vgo2_op_ps_abs_filt,
  VGO3_NP = vgo3_np_ps_abs_filt,
  VGO3_OP = vgo3_op_ps_abs_filt
)

ps_list_DA_genus <- list(
  VGO2_NP_genus = vgo2_np_ps_genus_abs_filt,
  VGO2_OP_genus = vgo2_op_ps_genus_abs_filt,
  VGO3_NP_genus = vgo3_np_ps_genus_abs_filt,
  VGO3_OP_genus = vgo3_op_ps_genus_abs_filt
)

str(vgo3_op_ps_genus_abs_filt@sam_data$FVC_category)
# Variables
vars_cont <- c("FEV1_pred_decline","FVC_pred_decline","FEV1FVC_pred_decline","FEF25_75_pred_decline")
vars_cat  <- c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")

# Covariates (fixed across all runs)
covars <- c("age","gender","season_microbiome_sampling","smoke_status")

# Optional factor reference specs for MaAsLin2 (if you want control)
# refs follows MaAsLin2's 'reference' format, e.g. list(metadata = "gender", value = "male")
# Provide per-variable if needed; leave NULL to skip.
refs <- NULL

# Run all (ASV + Genus; continuous + categorical) and collect a single summary table
# maaslin_summary <- run_maaslin_batch(
#   ps_lists  = list(ASV = ps_list_DA_ASV, Genus = ps_list_DA_genus),
#   var_sets  = list(continuous = vars_cont, categorical = vars_cat),
#   covars    = covars,
#   out_root  = "../Output/Differential_abundance/MaAsLin",
#   refs      = refs
# )

maaslin_summary <- run_maaslin_batch(
  ps_lists  = list(ASV = ps_list_DA_ASV, Genus = ps_list_DA_genus),
  var_sets  = list(
    continuous = c("FEV1_pred_decline","FVC_pred_decline","FEV1FVC_pred_decline","FEF25_75_pred_decline"),
    categorical = c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")
  ),
  covars    = c("age","gender","season_microbiome_sampling","smoke_status"),
  out_root  = "../Output/Differential_abundance/MaAsLin"
)
```

### MaAsLin2 heatmaps (ASV + Genus)
```{r}
# Settings
root_dir <- "../Output/Differential_abundance/MaAsLin"

# ASV folders
asv_folders <- list(
  VGO2_NP = file.path(root_dir, "VGO2_NP", "ASV"),
  VGO2_OP = file.path(root_dir, "VGO2_OP", "ASV"),
  VGO3_NP = file.path(root_dir, "VGO3_NP", "ASV"),
  VGO3_OP = file.path(root_dir, "VGO3_OP", "ASV")
)

# Genus folders (datasets end with _genus)
genus_folders <- list(
  VGO2_NP = file.path(root_dir, "VGO2_NP_genus", "Genus"),
  VGO2_OP = file.path(root_dir, "VGO2_OP_genus", "Genus"),
  VGO3_NP = file.path(root_dir, "VGO3_NP_genus", "Genus"),
  VGO3_OP = file.path(root_dir, "VGO3_OP_genus", "Genus")
)

# Order on the x-axis (raw metadata names in CSV)
custom_metadata_order <- c(
  "FEV1_pred_decline", "FEV1_category",
  "FVC_pred_decline",  "FVC_category",
  "FEV1FVC_pred_decline", "FEV1FVC_category",
  "FEF25_75_pred_decline", "FEF25_75_category"
)

# Pretty names to display on x-axis (same order)
pretty_labels <- c(
  "FEV1 % predicted (decline)",
  "FEV1 category",
  "FVC % predicted (decline)",
  "FVC category",
  "FEV1/FVC % predicted (decline)",
  "FEV1/FVC category",
  "FEF 25%-75% % predicted (decline)",
  "FEF 25%-75% category"
)
names(pretty_labels) <- custom_metadata_order

# Helper functions
# Read + tidy one dataset/level folder
read_heatmap_data <- function(folder) {
  files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
  if (!length(files)) return(NULL)

  df <- lapply(files, function(f) {
    tryCatch(readr::read_csv(f, show_col_types = FALSE), error = function(e) NULL)
  }) |> bind_rows()

  if (is.null(df) || !nrow(df)) return(NULL)

  # Keep required columns; tolerate missing 'label'
  if (!"label" %in% names(df)) df$label <- NA_character_
  df <- df |> select(feature, metadata, coef, label)

  # x-axis handling: keep raw + pretty factor
  df$metadata_raw <- factor(df$metadata, levels = custom_metadata_order)
  # map to pretty labels with same order
  df$metadata_plot <- factor(pretty_labels[as.character(df$metadata_raw)],
                             levels = pretty_labels)

  df
}


#  Y-axis label formatter:
#    - replace underscores with spaces
#    - italicize the text part
#    - put trailing numbers at the end in parentheses, non-italic
format_feature_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    if (grepl(" [0-9]+$", lab)) {
      text <- sub("([0-9]+)$", "", lab)   # remove trailing number
      num  <- sub(".* ([0-9]+)$", "\\1", lab)  # extract trailing number
      bquote(italic(.(text)) ~ "(" * .(num) * ")")
    } else {
      bquote(italic(.(lab)))
    }
  })
}


# Generic heatmap builder (works for both ASV/Genus)
make_heatmap <- function(df, title, italicize_features = TRUE) {
  if (is.null(df) || !nrow(df)) {
    return(ggplot() + theme_void() + ggtitle(paste0(title, " (no data)")))
  }

  p <- ggplot(df, aes(x = metadata_plot, y = feature, fill = coef)) +
    geom_tile(color = "white") +
    # EXACT palette requested:
    scale_fill_gradient2(
      low = rcartocolor::carto_pal(7, "Geyser")[1],
      mid = rcartocolor::carto_pal(7, "Geyser")[4],
      high = rcartocolor::carto_pal(7, "Geyser")[7],
      midpoint = 0,
      name = "Coefficient"
    ) +
    # centered significance stars
    geom_text(aes(label = label), color = "black", size = 5,
              hjust = 0.5, vjust = 0.8, na.rm = TRUE) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      axis.text.y  = element_text(),   # base; we format with scale_y_discrete
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5),
      plot.margin  = unit(c(6, 6, 6, 12), "pt")
    ) +
    ggtitle(title)

  if (italicize_features) {
    p + scale_y_discrete(labels = format_feature_labels)
  } else {
    # If you ever want Genus plain: change italicize_features=FALSE
    # Here we still italicize + remove underscores (same formatter)
    p + scale_y_discrete(labels = format_feature_labels)
  }
}




# Formatter for ASV labels: italics + trailing numbers in parentheses
format_asv_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    if (grepl(" [0-9]+$", lab)) {
      text <- sub("([0-9]+)$", "", lab)   # remove trailing number
      num  <- sub(".* ([0-9]+)$", "\\1", lab)  # extract trailing number
      bquote(italic(.(text)) ~ "(" * .(num) * ")")
    } else {
      bquote(italic(.(lab)))
    }
  })
}

# Formatter for genus labels: italics only, no special number handling
format_genus_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    bquote(italic(.(lab)))
  })
}

# Generic heatmap builder (choose formatter based on level)
make_heatmap <- function(df, title, level = c("ASV","Genus")) {
  level <- match.arg(level)

  if (is.null(df) || !nrow(df)) {
    return(ggplot() + theme_void() + ggtitle(paste0(title, " (no data)")))
  }

  p <- ggplot(df, aes(x = metadata_plot, y = feature, fill = coef)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = rcartocolor::carto_pal(7, "Geyser")[1],
      mid = rcartocolor::carto_pal(7, "Geyser")[4],
      high = rcartocolor::carto_pal(7, "Geyser")[7],
      midpoint = 0,
      name = "Coefficient"
    ) +
    geom_text(aes(label = label), color = "black", size = 5,
              hjust = 0.5, vjust = 0.8, na.rm = TRUE) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      axis.text.y  = element_text(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5),
      plot.margin  = unit(c(6, 6, 6, 12), "pt")
    ) +
    ggtitle(title)

  if (level == "ASV") {
    p + scale_y_discrete(labels = format_asv_labels)
  } else {
    p + scale_y_discrete(labels = format_genus_labels)
  }
}


# Build individual plots

# ASV heatmaps
p_VGO2_NP_ASV <- make_heatmap(read_heatmap_data(asv_folders$VGO2_NP), "VGO2 NP (ASV)", "ASV")
p_VGO2_OP_ASV <- make_heatmap(read_heatmap_data(asv_folders$VGO2_OP), "VGO2 OP (ASV)", "ASV")
p_VGO3_NP_ASV <- make_heatmap(read_heatmap_data(asv_folders$VGO3_NP), "VGO3 NP (ASV)", "ASV")
p_VGO3_OP_ASV <- make_heatmap(read_heatmap_data(asv_folders$VGO3_OP), "VGO3 OP (ASV)", "ASV")

# Genus heatmaps
p_VGO2_NP_Gen <- make_heatmap(read_heatmap_data(genus_folders$VGO2_NP), "VGO2 NP (Genus)", "Genus")
p_VGO2_OP_Gen <- make_heatmap(read_heatmap_data(genus_folders$VGO2_OP), "VGO2 OP (Genus)", "Genus")
p_VGO3_NP_Gen <- make_heatmap(read_heatmap_data(genus_folders$VGO3_NP), "VGO3 NP (Genus)", "Genus")
p_VGO3_OP_Gen <- make_heatmap(read_heatmap_data(genus_folders$VGO3_OP), "VGO3 OP (Genus)", "Genus")


# Combine into 2×2 grids and save
asv_grid <- (p_VGO2_NP_ASV | p_VGO2_OP_ASV) /
            (p_VGO3_NP_ASV | p_VGO3_OP_ASV) +
            patchwork::plot_annotation(title = "MaAsLin2 Heatmaps — ASV Level")

genus_grid <- (p_VGO2_NP_Gen | p_VGO2_OP_Gen) /
              (p_VGO3_NP_Gen | p_VGO3_OP_Gen) +
              patchwork::plot_annotation(title = "MaAsLin2 Heatmaps — Genus Level")

ggsave(file.path(root_dir, "Heatmap_Maaslin_results_ASV_all_datasets.svg"),
       asv_grid, width = 12, height = 16)

ggsave(file.path(root_dir, "Heatmap_Maaslin_results_Genus_all_datasets.svg"),
       genus_grid, width = 12, height = 16)
```
# Check sequences of interest 
```{r}
vgo3_op_seq <- refseq(ps_vgo3_op_filtered) %>%
  as.data.frame()

# Check in tax_table of phyloseq first (classification by SILVA)
ps_vgo3_op_filtered@tax_table["TM7x_90",]
which(rownames(seq) == "TM7x_90")
print(seq[55, ])

# Check in tax_table of phyloseq first (classification by SILVA)
ps_vgo3_op_filtered@tax_table["Lachnospiraceae_70",]
which(rownames(seq) == "Lachnospiraceae_70")
print(seq[46, ])

# Check in tax_table of phyloseq first (classification by SILVA)
ps_vgo3_op_filtered@tax_table["Fusobacterium_nucleatum_91",]
which(rownames(seq) == "Fusobacterium_nucleatum_91")
print(seq[56, ])

# Check in tax_table of phyloseq first (classification by SILVA)
ps_vgo3_op_filtered@tax_table["Alloprevotella_84",]
which(rownames(seq) == "Alloprevotella_84")
print(seq[52, ])

# Check in tax_table of phyloseq first (classification by SILVA)
ps_vgo3_op_filtered@tax_table["Corynebacterium_6",]
which(rownames(seq) == "Corynebacterium_6")
print(seq[6, ]) # Most likely to be Corynebacterium pseudodiphtheriticum (according to BLAST results)
```

# Cluster analysis
## Functions
```{r}
clust_ind_plot <- function(dist, n, clust_method="average") {
  
  clust_ind_fnx <- function(dist, k, method = clust_method) {
    stats_k <- c(k=k, fpc::cluster.stats(dist, cutree(hclust(dist, method=method), k))[c("ch", "avg.silwidth")])
    return(stats_k)
  }
  clust_ind <- lapply(n, clust_ind_fnx, dist=dist) %>% 
    bind_rows %>%
    setNames(c("k", "Calinski-Harabasz", "Silhouette"))
  
  p <- clust_ind %>%
    gather(index, value, -k) %>%
    ggplot(aes(x=k, y=value)) +
    geom_point() +
    geom_line() +
    facet_wrap(vars(index), scales="free_y")
  return(list(plot=p, data=clust_ind))
}


noLabel <- function(x) {
  if (stats::is.leaf(x)) {
    attr(x, "label") <- NULL }
  return(x)
}


prep_hm_data <- function(otu_ordered) {
  hm_data <- otu_ordered %>%
    data.frame(check.names = F) %>%
    rownames_to_column("OTU") %>% 
    pivot_longer(-OTU, names_to = "sample_id", values_to = "RA") %>%
    mutate(RA = if_else(RA == 0, NA_real_, RA),
           OTU = format_OTU(OTU) %>% fct_inorder() %>% fct_rev(),
           sample_id = fct_inorder(sample_id))
  return(hm_data)
}


prep_hm_data <- function(otu_ordered) {
  otu_ordered <- otu_ordered %>%
    mutate(across(-OTU, as.character))  # Convert all columns except OTU to character
  
  hm_data <- otu_ordered %>%
    data.frame(check.names = FALSE) %>%
    rownames_to_column("OTU") %>% 
    pivot_longer(cols = -OTU, names_to = "sample_id", values_to = "RA") %>%
    mutate(RA = if_else(RA == "0", NA_real_, as.numeric(RA)),  # Ensure RA is numeric with NA for zero values
           OTU = format_OTU(OTU) %>% fct_inorder() %>% fct_rev(),
           sample_id = fct_inorder(sample_id))
  
  return(hm_data)
}


create_dendro_gg <- function(hc, hang_height=0.05, size = 0.4) {
  suppressPackageStartupMessages(library(dendextend))
  library(ggdendro)
  
  dendro_data_bl <- hc %>% as.dendrogram %>% hang.dendrogram(hang_height=hang_height) %>% dendro_data
  dendro_data_bl$segments$yend[dendro_data_bl$segments$yend<0] <- 0
  
  dendro_data_gr <- hc %>% as.dendrogram %>% dendro_data
  hc_order <- dendro_data_gr$labels$label
  
  plot <- ggplot(segment(dendro_data_gr)) +
    geom_segment(aes(x=x, y=y, xend=xend, yend=yend),colour="white", size = size) + 
    geom_segment(data=segment(dendro_data_bl),aes(x=x, y=y, xend=xend, yend=yend),  size = size) +
    scale_x_continuous(expand = rep(1/length(hc_order)/2, 2)) + 
    scale_y_continuous(expand=c(0,0.02)) +
    #theme(plot.margin=unit(c(0,0,0,0),"lines")) +
    theme_void()
  return(list(plot=plot, hc_order=hc_order)) 
}
```
## VGO2 NP clustering
```{r}
# Check number of taxa
ntaxa(ps_vgo2_np_filtered) # 893 ASVs 
# Extract metadata
meta_vgo2_np <- meta_to_df(ps_vgo2_np_filtered, sample_name = ".row_id")
# Extract otu table
otu_vgo2_np <- as(otu_table(ps_vgo2_np_filtered), "matrix")
# Standardise orientation so that downstream analyses always have rows = samples and columns = ASVs
if(taxa_are_rows(ps_vgo2_np_filtered)){otu_vgo2_np <- t(otu_vgo2_np)} 
otu_vgo2_np <- as.data.frame(otu_vgo2_np)
# Compute BC distances
ord_vgo2_np <- vegdist(otu_vgo2_np, method = "bray") 

# We used complete linkage to create the phylogenetic tree (hierarchical clustering). The method is based on maximum distance; the similarity of any two clusters is the similarity of their most dissimilar pair.
hc_vgo2_np_clust <- hclust(ord_vgo2_np, method = "complete")
# Reorder 'otu_np' based on the clustering order
otu_hc_vgo2_np_order_clust <- otu_vgo2_np[hc_vgo2_np_clust$order,]

# The Calinski-Harabasz and Silhouette methods were used to determine the appropriate location to cut the tree. Higher value of indices means the clusters are dense and well separated. Here, we looked for a peak or at least an abrupt elbow on the line plot of the indices.
CalHara_sil_indices_vgo2_np <- clust_ind_plot(ord_vgo2_np, seq(2, 25, by = 1), clust_method="complete")
write.csv(CalHara_sil_indices_vgo2_np$data, "../Output/Cluster_analysis/VGO2_NP/CalHara_silhouette_indices_vgo2_np.csv")

# Calinski-Harabasz Index: A higher value indicates well-separated clusters with high intra-cluster similarity.
# Silhouette: The optimal number of clusters typically shows the highest silhouette score, suggesting well-separated and compact clusters.
# look for the points where there are peaks or elbows in the curve

# Create dendogram
dend_vgo2_np <- as.dendrogram(hc_vgo2_np_clust)
 
# 4 clusters
dend_4clusters <- color_branches(dend_vgo2_np, k = 4, groupLabels=T)
(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_vgo2_np, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# 5 clusters
dend_5clusters <- color_branches(dend_vgo2_np, k = 5, groupLabels=T)
plot(stats::dendrapply(dend_5clusters, noLabel))
cluster_assignments_5 <- cutree(dend_vgo2_np, k = 5)
cluster_counts_5 <- table(cluster_assignments_5)
print(cluster_counts_5)

# 6 clusters
dend_6clusters <- color_branches(dend_vgo2_np, k = 6, groupLabels=T)
plot(stats::dendrapply(dend_6clusters, noLabel))
cluster_assignments_6 <- cutree(dend_vgo2_np, k = 6)
cluster_counts_6 <- table(cluster_assignments_6)
print(cluster_counts_6)

# 7 clusters
dend_7clusters <- color_branches(dend_vgo2_np, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_vgo2_np, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# 8 clusters
dend_8clusters <- color_branches(dend_vgo2_np, k = 8, groupLabels=T)
plot(stats::dendrapply(dend_8clusters, noLabel))
cluster_assignments_8 <- cutree(dend_vgo2_np, k = 8)
cluster_counts_8 <- table(cluster_assignments_8)
print(cluster_counts_8)

# 9 clusters
dend_9clusters <- color_branches(dend_vgo2_np, k = 9, groupLabels=T)
plot(stats::dendrapply(dend_9clusters, noLabel))
cluster_assignments_9 <- cutree(dend_vgo2_np, k = 9)
cluster_counts_9 <- table(cluster_assignments_9)
print(cluster_counts_9)

# 10 clusters
dend_10clusters <- color_branches(dend_vgo2_np, k = 10, groupLabels=T)
plot(stats::dendrapply(dend_10clusters, noLabel))
cluster_assignments_10 <- cutree(dend_vgo2_np, k = 10)
cluster_counts_10 <- table(cluster_assignments_10)
print(cluster_counts_10)

# 11 clusters
dend_11clusters <- color_branches(dend_vgo2_np, k = 11, groupLabels=T)
plot(stats::dendrapply(dend_11clusters, noLabel))
cluster_assignments_11 <- cutree(dend_vgo2_np, k = 11)
cluster_counts_11 <- table(cluster_assignments_11)
print(cluster_counts_11)

# 12 clusters
dend_12clusters <- color_branches(dend_vgo2_np, k = 12, groupLabels=T)
plot(stats::dendrapply(dend_12clusters, noLabel))
# check number per cluster
cluster_assignments_12 <- cutree(dend_vgo2_np, k = 12)
cluster_counts_12 <- table(cluster_assignments_12)
print(cluster_counts_12)

# 13 clusters
dend_13clusters <- color_branches(dend_vgo2_np, k = 13, groupLabels=T)
plot(stats::dendrapply(dend_13clusters, noLabel))
cluster_assignments_13 <- cutree(dend_vgo2_np, k = 13)
cluster_counts_13 <- table(cluster_assignments_13)
print(cluster_counts_13)

# 14 clusters
dend_14clusters <- color_branches(dend_vgo2_np, k = 14, groupLabels=T)
plot(stats::dendrapply(dend_14clusters, noLabel))
cluster_assignments_14 <- cutree(dend_vgo2_np, k = 14)
cluster_counts_14 <- table(cluster_assignments_14)
print(cluster_counts_14)
# 15 clusters
dend_15clusters <- color_branches(dend_vgo2_np, k = 15, groupLabels=T)
plot(stats::dendrapply(dend_15clusters, noLabel))
cluster_assignments_15 <- cutree(dend_vgo2_np, k = 15)
cluster_counts_15 <- table(cluster_assignments_15)
print(cluster_counts_15)


# 4 clusters appears optimal
# k = 4
dend_4clusters <- color_branches(dend_vgo2_np, k = 4, groupLabels=T)
plot(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_4clusters, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# The function 'cuttree' is used to cut the tree at k = 4
groups_vgo2_np_c4 <- cutree(hc_vgo2_np_clust, k = 4)
ps_vgo2_np_filtered@sam_data$cluster <- groups_vgo2_np_c4[order(rownames(ps_vgo2_np_filtered@sam_data))]
ps_vgo2_np_filtered@sam_data$cluster <- ps_vgo2_np_filtered@sam_data$cluster %>% as.character()

# Determine the top 3 ASVs for each cluster
otu_vgo2_np <- as(otu_table(ps_vgo2_np_filtered), "matrix")
# if (taxa_are_rows(ps_vgo2_np_filtered)) {
#   otu_hc_vgo2_np_order_clust <- t(otu_hc_vgo2_np_order_clust)  # Ensure ASVs are columns, samples are rows
# }

# Extract sample metadata with cluster information
sample_data_vgo2_np <- data.frame(ps_vgo2_np_filtered@sam_data)

common_samples <- intersect(rownames(sample_data_vgo2_op), rownames(otu_hc_vgo2_op_order_clust))
rownames(otu_hc_vgo2_np_order_clust)
rownames(otu_hc_vgo2_op_order_clust)
rownames(sample_data_vgo2_np)
rownames(sample_data_vgo2_np)


# Combine the OTU table with the sample cluster data
otu_vgo2_np_with_clusters <- cbind(cluster = sample_data_vgo2_np$cluster, otu_hc_vgo2_np_order_clust)
otu_vgo2_np_with_clusters <- as.data.frame(otu_vgo2_np_with_clusters)
otu_vgo2_np_with_clusters$cluster <- as.factor(otu_vgo2_np_with_clusters$cluster)
# Convert all ASV abundances to numeric
otu_vgo2_np_with_clusters[-1] <- lapply(otu_vgo2_np_with_clusters[-1], as.numeric)
# Sum ASV abundances for each cluster
asv_sum_by_cluster_vgo2_np <- otu_vgo2_np_with_clusters %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
# Convert back to data frame for further processing
asv_sum_by_cluster_vgo2_np <- as.data.frame(asv_sum_by_cluster_vgo2_np)
# Find the top 3 most abundant ASVs in each cluster and their percentage
top_3_asvs_vgo2_np <- apply(asv_sum_by_cluster_vgo2_np[, -1], 1, function(x) {
  # Get total abundance in the cluster for calculating percentages
  total_abundance_vgo2_np <- sum(x, na.rm = TRUE)
  # Get names and values of the top 3 ASVs
  top_3_indices_vgo2_np <- order(x, decreasing = TRUE)[1:3]
  top_3_asv_names_vgo2_np <- names(x)[top_3_indices_vgo2_np]
  top_3_asv_values_vgo2_np <- x[top_3_indices_vgo2_np]
  # Calculate percentages
  top_3_asv_percentages_vgo2_np <- (top_3_asv_values_vgo2_np / total_abundance_vgo2_np) * 100
  # Combine ASV names and percentages into a string for each cluster
  result <- paste0(top_3_asv_names_vgo2_np, ": ", round(top_3_asv_percentages_vgo2_np, 2), "%", collapse = "; ")
  return(result)
})

# Combine results with cluster information
top_3_asvs_df_vgo2_np <- data.frame(cluster = asv_sum_by_cluster_vgo2_np$cluster, Top_3_ASVs = top_3_asvs_vgo2_np)
# Display the top 3 ASVs for each cluster along with their percentages
print(top_3_asvs_df_vgo2_np)

#### Rename clusters ####
# If an ASV is >40% of the total abundance in a cluster: Name the cluster after that ASV only, as it is the dominant contributor.
# If the top ASV is between 20-40% of the total abundance: Use the two most abundant ASVs to name the cluster.
# If the top ASV is <20% of the total abundance: Use the top three ASVs to name the cluster, as none of the ASVs dominate significantly.

get_cluster_name <- function(top_asvs_string) {
  asv_info <- strsplit(top_asvs_string, "; ")[[1]]
  asv_names <- sapply(asv_info, function(x) strsplit(x, ": ")[[1]][1])
  asv_percentages <- as.numeric(sapply(asv_info, function(x) sub("%", "", strsplit(x, ": ")[[1]][2])))
  
  if (asv_percentages[1] > 40) {
    return(paste0(asv_names[1], "-dominated cluster"))
  } else if (asv_percentages[1] > 20) {
    return(paste0(asv_names[1], " and ", asv_names[2], "-dominated cluster"))
  } else {
    return(paste0(asv_names[1], ", ", asv_names[2], ", and ", asv_names[3], "-dominated cluster"))
  }
}

# Rename based on these rules
top_3_asvs_df_vgo2_np$Cluster_Name <- sapply(top_3_asvs_df_vgo2_np$Top_3_ASVs, get_cluster_name)
top_3_asvs_df_vgo2_np$Cluster_Name

# [1] "Corynebacterium_3, Staphylococcus_2, and Moraxella_4-dominated cluster"      
# [2] "Corynebacterium_3, Moraxella_4, and Staphylococcus_2-dominated cluster"      
# [3] "Staphylococcus_2, Corynebacterium_6, and Corynebacterium_3-dominated cluster"
# [4] "Staphylococcus_2-dominated cluster" 

# Shorten the names made above for plotting 
# Define the new cluster names in the same order as the original clusters
new_cluster_names_vgo2_np <- c(
  "Cor(3)/Staph(2)/Mor(4) clust",
  "Cor(3)/Mor(4)/Staph(2) clust",
  "Staph(2)/Cor(6)/Cor(3) clust",
  "Staph(2) clust"
)

# Assign the new names to 'Cluster_Name' in 'top_3_asvs_df'
top_3_asvs_df_vgo2_np$Cluster_Name <- new_cluster_names_vgo2_np

# Update the clusters in the sample data to use the new names
ps_vgo2_np_filtered@sam_data$cluster <- top_3_asvs_df_vgo2_np$Cluster_Name[match(ps_vgo2_np_filtered@sam_data$cluster, top_3_asvs_df_vgo2_np$cluster)]
# Convert to character
ps_vgo2_np_filtered@sam_data$cluster <- as.character(ps_vgo2_np_filtered@sam_data$cluster)

# Count the number of samples in each cluster
cluster_counts_vgo2_np <- as.data.frame(table(ps_vgo2_np_filtered@sam_data$cluster))
colnames(cluster_counts_vgo2_np) <- c("Cluster", "Count")

# Calculate the relative proportions of each cluster
cluster_counts_vgo2_np$Proportion <- cluster_counts_vgo2_np$Count / sum(cluster_counts_vgo2_np$Count)

write.csv(cluster_counts_vgo2_np, "../Output/Cluster_analysis/CPall_GF_NP/cluster_counts_vgo2_np.csv")

# Create the plot with geom_bar for vertical stacking
bar_clusters_vgo2_np <- ggplot(data = cluster_counts_vgo2_np, aes(x = Cluster, y = (Proportion*100), fill = Cluster)) +
  geom_bar(stat = "identity") +  # Use stat = "identity" to directly use Proportion
  labs(title = "Bar plot of clusters in VGO2 NP samples",
       x = "Clusters",            # X-axis now represents clusters
       y = "% of samples",          # Y-axis represents proportions
       fill = "Cluster") +
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),      # Remove x-axis labels if needed
        axis.ticks.x = element_blank(),     # Remove x-axis ticks if needed
        legend.position = "right")

bar_clusters_vgo2_np
ggsave("../Output/Cluster_analysis/VGO2_NP/Bar_plot_cluster_proportions_clusters.svg", bar_clusters_vgo2_np, width = 8, height = 5)

```


## VGO2 OP clustering
```{r}
# Check number of taxa
ntaxa(ps_vgo2_op_filtered) # 337 ASVs 
# Extract metadata
meta_vgo2_op <- meta_to_df(ps_vgo2_op_filtered, sample_name = ".row_id")
# Extract otu table
otu_vgo2_op <- as(otu_table(ps_vgo2_op_filtered), "matrix")
# Standardise orientation so that downstream analyses always have rows = samples and columns = ASVs
if(taxa_are_rows(ps_vgo2_op_filtered)){otu_vgo2_op <- t(otu_vgo2_op)} 
otu_vgo2_op <- as.data.frame(otu_vgo2_op)
# Compute BC distances
ord_vgo2_op <- vegdist(otu_vgo2_op, method = "bray") 

# We used complete linkage to create the phylogenetic tree (hierarchical clustering). The method is based on maximum distance; the similarity of any two clusters is the similarity of their most dissimilar pair.
hc_vgo2_op_clust <- hclust(ord_vgo2_op, method = "complete")
# Reorder 'otu_op' based on the clustering order
otu_hc_vgo2_op_order_clust <- otu_vgo2_op[hc_vgo2_op_clust$order,]
# The Calinski-Harabasz and Silhouette methods were used to determine the appropriate location to cut the tree. Higher value of indices means the clusters are dense and well separated. Here, we looked for a peak or at least an abrupt elbow on the line plot of the indices.
CalHara_sil_indices_vgo2_op <- clust_ind_plot(ord_vgo2_op, seq(2, 25, by = 1), clust_method="complete")
write.csv(CalHara_sil_indices_vgo2_op$data, "../Output/Cluster_analysis/VGO2_OP/CalHara_silhouette_indices_vgo2_op.csv")
# Calinski-Harabasz Index: A higher value indicates well-separated clusters with high intra-cluster similarity.
# Silhouette: The optimal number of clusters typically shows the highest silhouette score, suggesting well-separated and compact clusters.
# Look for the points where there are peaks or elbows in the curve

# Create dendogram
dend_vgo2_op <- as.dendrogram(hc_vgo2_op_clust)

# 2 clusters
dend_2clusters <- color_branches(dend_vgo2_op, k = 2, groupLabels=T)
(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_vgo2_op, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# 3 clusters
dend_3clusters <- color_branches(dend_vgo2_op, k = 3, groupLabels=T)
(stats::dendrapply(dend_3clusters, noLabel))
cluster_assignments_3 <- cutree(dend_vgo2_op, k = 3)
cluster_counts_3 <- table(cluster_assignments_3)
print(cluster_counts_3)

# 4 clusters
dend_4clusters <- color_branches(dend_vgo2_op, k = 4, groupLabels=T)
(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_vgo2_op, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# 5 clusters
dend_5clusters <- color_branches(dend_vgo2_op, k = 5, groupLabels=T)
plot(stats::dendrapply(dend_5clusters, noLabel))
cluster_assignments_5 <- cutree(dend_vgo2_op, k = 5)
cluster_counts_5 <- table(cluster_assignments_5)
print(cluster_counts_5)

# 6 clusters
dend_6clusters <- color_branches(dend_vgo2_op, k = 6, groupLabels=T)
plot(stats::dendrapply(dend_6clusters, noLabel))
cluster_assignments_6 <- cutree(dend_vgo2_op, k = 6)
cluster_counts_6 <- table(cluster_assignments_6)
print(cluster_counts_6)

# 7 clusters
dend_7clusters <- color_branches(dend_vgo2_op, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_vgo2_op, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# 8 clusters
dend_8clusters <- color_branches(dend_vgo2_op, k = 8, groupLabels=T)
plot(stats::dendrapply(dend_8clusters, noLabel))
cluster_assignments_8 <- cutree(dend_vgo2_op, k = 8)
cluster_counts_8 <- table(cluster_assignments_8)
print(cluster_counts_8)

# 9 clusters
dend_9clusters <- color_branches(dend_vgo2_op, k = 9, groupLabels=T)
plot(stats::dendrapply(dend_9clusters, noLabel))
cluster_assignments_9 <- cutree(dend_vgo2_op, k = 9)
cluster_counts_9 <- table(cluster_assignments_9)
print(cluster_counts_9)

# 10 clusters
dend_10clusters <- color_branches(dend_vgo2_op, k = 10, groupLabels=T)
plot(stats::dendrapply(dend_10clusters, noLabel))
cluster_assignments_10 <- cutree(dend_vgo2_op, k = 10)
cluster_counts_10 <- table(cluster_assignments_10)
print(cluster_counts_10)

# 11 clusters
dend_11clusters <- color_branches(dend_vgo2_op, k = 11, groupLabels=T)
plot(stats::dendrapply(dend_11clusters, noLabel))
cluster_assignments_11 <- cutree(dend_vgo2_op, k = 11)
cluster_counts_11 <- table(cluster_assignments_11)
print(cluster_counts_11)

# 12 clusters
dend_12clusters <- color_branches(dend_vgo2_op, k = 12, groupLabels=T)
plot(stats::dendrapply(dend_12clusters, noLabel))
# check number per cluster
cluster_assignments_12 <- cutree(dend_vgo2_op, k = 12)
cluster_counts_12 <- table(cluster_assignments_12)
print(cluster_counts_12)

# 13 clusters
dend_13clusters <- color_branches(dend_vgo2_op, k = 13, groupLabels=T)
plot(stats::dendrapply(dend_13clusters, noLabel))
cluster_assignments_13 <- cutree(dend_vgo2_op, k = 13)
cluster_counts_13 <- table(cluster_assignments_13)
print(cluster_counts_13)

# 14 clusters
dend_14clusters <- color_branches(dend_vgo2_op, k = 14, groupLabels=T)
plot(stats::dendrapply(dend_14clusters, noLabel))
cluster_assignments_14 <- cutree(dend_vgo2_op, k = 14)
cluster_counts_14 <- table(cluster_assignments_14)
print(cluster_counts_14)
# 15 clusters
dend_15clusters <- color_branches(dend_vgo2_op, k = 15, groupLabels=T)
plot(stats::dendrapply(dend_15clusters, noLabel))
cluster_assignments_15 <- cutree(dend_vgo2_op, k = 15)
cluster_counts_15 <- table(cluster_assignments_15)
print(cluster_counts_15)

#### 2 clusters appears optimal ####
# k = 2
dend_2clusters <- color_branches(dend_vgo2_op, k = 2, groupLabels=T)
plot(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_2clusters, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# The function 'cuttree' is used to cut the tree at k = 2
groups_vgo2_op_c2 <- cutree(hc_vgo2_op_clust, k = 2)
ps_vgo2_op_filtered@sam_data$cluster <- groups_vgo2_op_c2[order(rownames(ps_vgo2_op_filtered@sam_data))]
ps_vgo2_op_filtered@sam_data$cluster <- ps_vgo2_op_filtered@sam_data$cluster %>% as.character()

# Determine the top 3 ASVs for each cluster
otu_vgo2_op <- as(otu_table(ps_vgo2_op_filtered), "matrix")
# if (taxa_are_rows(ps_vgo2_op_filtered)) {
#   otu_hc_vgo2_op_order_clust <- t(otu_hc_vgo2_op_order_clust)  # Ensure ASVs are columns, samples are rows
# }

head(otu_vgo2_op)
# Extract sample metadata with cluster information
sample_data_vgo2_op <- data.frame(ps_vgo2_op_filtered@sam_data)
# Combine the OTU table with the sample cluster data
otu_vgo2_op_with_clusters <- cbind(cluster = sample_data_vgo2_op$cluster, otu_hc_vgo2_op_order_clust)
otu_vgo2_op_with_clusters <- as.data.frame(otu_vgo2_op_with_clusters)
otu_vgo2_op_with_clusters$cluster <- as.factor(otu_vgo2_op_with_clusters$cluster)
# Convert all ASV abundances to numeric
otu_vgo2_op_with_clusters[-1] <- lapply(otu_vgo2_op_with_clusters[-1], as.numeric)
# Sum ASV abundances for each cluster
asv_sum_by_cluster_vgo2_op <- otu_vgo2_op_with_clusters %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
# Convert back to data frame for further processing
asv_sum_by_cluster_vgo2_op <- as.data.frame(asv_sum_by_cluster_vgo2_op)
# Find the top 3 most abundant ASVs in each cluster and their percentage
top_3_asvs_vgo2_op <- apply(asv_sum_by_cluster_vgo2_op[, -1], 1, function(x) {
  # Get total abundance in the cluster for calculating percentages
  total_abundance_vgo2_op <- sum(x, na.rm = TRUE)
  # Get names and values of the top 3 ASVs
  top_3_indices_vgo2_op <- order(x, decreasing = TRUE)[1:3]
  top_3_asv_names_vgo2_op <- names(x)[top_3_indices_vgo2_op]
  top_3_asv_values_vgo2_op <- x[top_3_indices_vgo2_op]
  # Calculate percentages
  top_3_asv_percentages_vgo2_op <- (top_3_asv_values_vgo2_op / total_abundance_vgo2_op) * 100
  # Combine ASV names and percentages into a string for each cluster
  result <- paste0(top_3_asv_names_vgo2_op, ": ", round(top_3_asv_percentages_vgo2_op, 2), "%", collapse = "; ")
  return(result)
})

# Combine results with cluster information
top_3_asvs_df_vgo2_op <- data.frame(cluster = asv_sum_by_cluster_vgo2_op$cluster, Top_3_ASVs = top_3_asvs_vgo2_op)
# Display the top 3 ASVs for each cluster along with their percentages
print(top_3_asvs_df_vgo2_op)

#### Rename clusters ####
# If an ASV is >40% of the total abundance in a cluster: Name the cluster after that ASV only, as it is the dominant contributor.
# If the top ASV is between 20-40% of the total abundance: Use the two most abundant ASVs to name the cluster.
# If the top ASV is <20% of the total abundance: Use the top three ASVs to name the cluster, as none of the ASVs dominate significantly.

get_cluster_name <- function(top_asvs_string) {
  asv_info <- strsplit(top_asvs_string, "; ")[[1]]
  asv_names <- sapply(asv_info, function(x) strsplit(x, ": ")[[1]][1])
  asv_percentages <- as.numeric(sapply(asv_info, function(x) sub("%", "", strsplit(x, ": ")[[1]][2])))
  
  if (asv_percentages[1] > 40) {
    return(paste0(asv_names[1], "-dominated cluster"))
  } else if (asv_percentages[1] > 20) {
    return(paste0(asv_names[1], " and ", asv_names[2], "-dominated cluster"))
  } else {
    return(paste0(asv_names[1], ", ", asv_names[2], ", and ", asv_names[3], "-dominated cluster"))
  }
}

# Rename based on these rules
top_3_asvs_df_vgo2_op$Cluster_Name <- sapply(top_3_asvs_df_vgo2_op$Top_3_ASVs, get_cluster_name)
top_3_asvs_df_vgo2_op$Cluster_Name

# [1] "Streptococcus_5 and Veillonella_7-dominated cluster"  
# [2] "Streptococcus_8 and Streptococcus_5-dominated cluster"

# Shorten the names made above for plotting 
# Define the new cluster names in the same order as the original clusters
new_cluster_names_vgo2_op <- c(
  "Strep(5)/Veill(7) clust",
  "Strep(8)/Strep(5) clust"
)

# Assign the new names to 'Cluster_Name' in 'top_3_asvs_df'
top_3_asvs_df_vgo2_op$Cluster_Name <- new_cluster_names_vgo2_op

# Update the clusters in the sample data to use the new names
ps_vgo2_op_filtered@sam_data$cluster <- top_3_asvs_df_vgo2_op$Cluster_Name[match(ps_vgo2_op_filtered@sam_data$cluster, top_3_asvs_df_vgo2_op$cluster)]
# Convert to character
ps_vgo2_op_filtered@sam_data$cluster <- as.character(ps_vgo2_op_filtered@sam_data$cluster)

# Count the number of samples in each cluster
cluster_counts_vgo2_op <- as.data.frame(table(ps_vgo2_op_filtered@sam_data$cluster))
colnames(cluster_counts_vgo2_op) <- c("Cluster", "Count")

# Calculate the relative proportions of each cluster
cluster_counts_vgo2_op$Proportion <- cluster_counts_vgo2_op$Count / sum(cluster_counts_vgo2_op$Count)

write.csv(cluster_counts_vgo2_op, "../Output/Cluster_analysis/VGO2_OP/cluster_counts_vgo2_op.csv")

# Create the plot with geom_bar for vertical stacking
bar_clusters_vgo2_op <- ggplot(data = cluster_counts_vgo2_op, aes(x = Cluster, y = (Proportion*100), fill = Cluster)) +
  geom_bar(stat = "identity") +  # Use stat = "identity" to directly use Proportion
  labs(title = "Bar plot of clusters in VGO2 op samples",
       x = "Clusters",            # X-axis now represents clusters
       y = "% of samples",          # Y-axis represents proportions
       fill = "Cluster") +
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),      # Remove x-axis labels if needed
        axis.ticks.x = element_blank(),     # Remove x-axis ticks if needed
        legend.position = "right")

bar_clusters_vgo2_op
ggsave("../Output/Cluster_analysis/VGO2_op/Bar_plot_cluster_proportions_clusters.svg", bar_clusters_vgo2_op, width = 8, height = 5)
```
## VGO3 NP clustering
```{r}
# Check number of taxa
ntaxa(ps_vgo3_np_filtered) # 883 ASVs 
# Extract metadata
meta_vgo3_np <- meta_to_df(ps_vgo3_np_filtered, sample_name = ".row_id")
# Extract otu table
otu_vgo3_np <- as(otu_table(ps_vgo3_np_filtered), "matrix")
# Standardise orientation so that downstream analyses always have rows = samples and columns = ASVs
if(taxa_are_rows(ps_vgo3_np_filtered)){otu_vgo3_np <- t(otu_vgo3_np)} 
otu_vgo3_np <- as.data.frame(otu_vgo3_np)
# Compute BC distances
ord_vgo3_np <- vegdist(otu_vgo3_np, method = "bray") 

# We used complete linkage to create the phylogenetic tree (hierarchical clustering). The method is based on maximum distance; the similarity of any two clusters is the similarity of their most dissimilar pair.
hc_vgo3_np_clust <- hclust(ord_vgo3_np, method = "complete")
# Reorder 'otu_np' based on the clustering order
otu_hc_vgo3_np_order_clust <- otu_vgo3_np[hc_vgo3_np_clust$order,]

# The Calinski-Harabasz and Silhouette methods were used to determine the appropriate location to cut the tree. Higher value of indices means the clusters are dense and well separated. Here, we looked for a peak or at least an abrupt elbow on the line plot of the indices.
CalHara_sil_indices_vgo3_np <- clust_ind_plot(ord_vgo3_np, seq(2, 25, by = 1), clust_method="complete")
write.csv(CalHara_sil_indices_vgo3_np$data, "../Output/Cluster_analysis/VGO3_NP/CalHara_silhouette_indices_vgo3_np.csv")

# Calinski-Harabasz Index: A higher value indicates well-separated clusters with high intra-cluster similarity.
# Silhouette: The optimal number of clusters typically shows the highest silhouette score, suggesting well-separated and compact clusters.
# look for the points where there are peaks or elbows in the curve

# Create dendogram
dend_vgo3_np <- as.dendrogram(hc_vgo3_np_clust)

# 2 clusters
dend_2clusters <- color_branches(dend_vgo3_np, k = 2, groupLabels=T)
(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_vgo3_np, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# 3 clusters
dend_3clusters <- color_branches(dend_vgo3_np, k = 3, groupLabels=T)
(stats::dendrapply(dend_3clusters, noLabel))
cluster_assignments_3 <- cutree(dend_vgo3_np, k = 3)
cluster_counts_3 <- table(cluster_assignments_3)
print(cluster_counts_3)

# 4 clusters
dend_4clusters <- color_branches(dend_vgo3_np, k = 4, groupLabels=T)
(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_vgo3_np, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# 5 clusters
dend_5clusters <- color_branches(dend_vgo3_np, k = 5, groupLabels=T)
plot(stats::dendrapply(dend_5clusters, noLabel))
cluster_assignments_5 <- cutree(dend_vgo3_np, k = 5)
cluster_counts_5 <- table(cluster_assignments_5)
print(cluster_counts_5)

# 6 clusters
dend_6clusters <- color_branches(dend_vgo3_np, k = 6, groupLabels=T)
plot(stats::dendrapply(dend_6clusters, noLabel))
cluster_assignments_6 <- cutree(dend_vgo3_np, k = 6)
cluster_counts_6 <- table(cluster_assignments_6)
print(cluster_counts_6)

# 7 clusters
dend_7clusters <- color_branches(dend_vgo3_np, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_vgo3_np, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# 8 clusters
dend_8clusters <- color_branches(dend_vgo3_np, k = 8, groupLabels=T)
plot(stats::dendrapply(dend_8clusters, noLabel))
cluster_assignments_8 <- cutree(dend_vgo3_np, k = 8)
cluster_counts_8 <- table(cluster_assignments_8)
print(cluster_counts_8)

# 9 clusters
dend_9clusters <- color_branches(dend_vgo3_np, k = 9, groupLabels=T)
plot(stats::dendrapply(dend_9clusters, noLabel))
cluster_assignments_9 <- cutree(dend_vgo3_np, k = 9)
cluster_counts_9 <- table(cluster_assignments_9)
print(cluster_counts_9)

# 10 clusters
dend_10clusters <- color_branches(dend_vgo3_np, k = 10, groupLabels=T)
plot(stats::dendrapply(dend_10clusters, noLabel))
cluster_assignments_10 <- cutree(dend_vgo3_np, k = 10)
cluster_counts_10 <- table(cluster_assignments_10)
print(cluster_counts_10)

# 11 clusters
dend_11clusters <- color_branches(dend_vgo3_np, k = 11, groupLabels=T)
plot(stats::dendrapply(dend_11clusters, noLabel))
cluster_assignments_11 <- cutree(dend_vgo3_np, k = 11)
cluster_counts_11 <- table(cluster_assignments_11)
print(cluster_counts_11)

# 12 clusters
dend_12clusters <- color_branches(dend_vgo3_np, k = 12, groupLabels=T)
plot(stats::dendrapply(dend_12clusters, noLabel))
# check number per cluster
cluster_assignments_12 <- cutree(dend_vgo3_np, k = 12)
cluster_counts_12 <- table(cluster_assignments_12)
print(cluster_counts_12)

# 13 clusters
dend_13clusters <- color_branches(dend_vgo3_np, k = 13, groupLabels=T)
plot(stats::dendrapply(dend_13clusters, noLabel))
cluster_assignments_13 <- cutree(dend_vgo3_np, k = 13)
cluster_counts_13 <- table(cluster_assignments_13)
print(cluster_counts_13)

# 14 clusters
dend_14clusters <- color_branches(dend_vgo3_np, k = 14, groupLabels=T)
plot(stats::dendrapply(dend_14clusters, noLabel))
cluster_assignments_14 <- cutree(dend_vgo3_np, k = 14)
cluster_counts_14 <- table(cluster_assignments_14)
print(cluster_counts_14)
# 15 clusters
dend_15clusters <- color_branches(dend_vgo3_np, k = 15, groupLabels=T)
plot(stats::dendrapply(dend_15clusters, noLabel))
cluster_assignments_15 <- cutree(dend_vgo3_np, k = 15)
cluster_counts_15 <- table(cluster_assignments_15)
print(cluster_counts_15)


# 4 clusters appears optimal
# k = 4
dend_4clusters <- color_branches(dend_vgo3_np, k = 4, groupLabels=T)
plot(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_4clusters, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# The function 'cuttree' is used to cut the tree at k = 4
groups_vgo3_np_c4 <- cutree(hc_vgo3_np_clust, k = 4)
ps_vgo3_np_filtered@sam_data$cluster <- groups_vgo3_np_c4[order(rownames(ps_vgo3_np_filtered@sam_data))]
ps_vgo3_np_filtered@sam_data$cluster <- ps_vgo3_np_filtered@sam_data$cluster %>% as.character()

# Determine the top 3 ASVs for each cluster
otu_vgo3_np <- as(otu_table(ps_vgo3_np_filtered), "matrix")
# if (taxa_are_rows(ps_vgo3_np_filtered)) {
#   otu_hc_vgo3_np_order_clust <- t(otu_hc_vgo3_np_order_clust)  # Ensure ASVs are columns, samples are rows
# }

head(otu_vgo3_np)
# Extract sample metadata with cluster information
sample_data_vgo3_np <- data.frame(ps_vgo3_np_filtered@sam_data)

# Combine the OTU table with the sample cluster data
otu_vgo3_np_with_clusters <- cbind(cluster = sample_data_vgo3_np$cluster, otu_hc_vgo3_np_order_clust)
otu_vgo3_np_with_clusters <- as.data.frame(otu_vgo3_np_with_clusters)
otu_vgo3_np_with_clusters$cluster <- as.factor(otu_vgo3_np_with_clusters$cluster)
# Convert all ASV abundances to numeric
otu_vgo3_np_with_clusters[-1] <- lapply(otu_vgo3_np_with_clusters[-1], as.numeric)
# Sum ASV abundances for each cluster
asv_sum_by_cluster_vgo3_np <- otu_vgo3_np_with_clusters %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
# Convert back to data frame for further processing
asv_sum_by_cluster_vgo3_np <- as.data.frame(asv_sum_by_cluster_vgo3_np)
# Find the top 3 most abundant ASVs in each cluster and their percentage
top_3_asvs_vgo3_np <- apply(asv_sum_by_cluster_vgo3_np[, -1], 1, function(x) {
  # Get total abundance in the cluster for calculating percentages
  total_abundance_vgo3_np <- sum(x, na.rm = TRUE)
  # Get names and values of the top 3 ASVs
  top_3_indices_vgo3_np <- order(x, decreasing = TRUE)[1:3]
  top_3_asv_names_vgo3_np <- names(x)[top_3_indices_vgo3_np]
  top_3_asv_values_vgo3_np <- x[top_3_indices_vgo3_np]
  # Calculate percentages
  top_3_asv_percentages_vgo3_np <- (top_3_asv_values_vgo3_np / total_abundance_vgo3_np) * 100
  # Combine ASV names and percentages into a string for each cluster
  result <- paste0(top_3_asv_names_vgo3_np, ": ", round(top_3_asv_percentages_vgo3_np, 2), "%", collapse = "; ")
  return(result)
})

# Combine results with cluster information
top_3_asvs_df_vgo3_np <- data.frame(cluster = asv_sum_by_cluster_vgo3_np$cluster, Top_3_ASVs = top_3_asvs_vgo3_np)
# Display the top 3 ASVs for each cluster along with their percentages
print(top_3_asvs_df_vgo3_np)

#### Rename clusters ####
# If an ASV is >40% of the total abundance in a cluster: Name the cluster after that ASV only, as it is the dominant contributor.
# If the top ASV is between 20-40% of the total abundance: Use the two most abundant ASVs to name the cluster.
# If the top ASV is <20% of the total abundance: Use the top three ASVs to name the cluster, as none of the ASVs dominate significantly.

get_cluster_name <- function(top_asvs_string) {
  asv_info <- strsplit(top_asvs_string, "; ")[[1]]
  asv_names <- sapply(asv_info, function(x) strsplit(x, ": ")[[1]][1])
  asv_percentages <- as.numeric(sapply(asv_info, function(x) sub("%", "", strsplit(x, ": ")[[1]][2])))
  
  if (asv_percentages[1] > 40) {
    return(paste0(asv_names[1], "-dominated cluster"))
  } else if (asv_percentages[1] > 20) {
    return(paste0(asv_names[1], " and ", asv_names[2], "-dominated cluster"))
  } else {
    return(paste0(asv_names[1], ", ", asv_names[2], ", and ", asv_names[3], "-dominated cluster"))
  }
}

# Rename based on these rules
top_3_asvs_df_vgo3_np$Cluster_Name <- sapply(top_3_asvs_df_vgo3_np$Top_3_ASVs, get_cluster_name)
top_3_asvs_df_vgo3_np$Cluster_Name

# [1] "Dolosigranulum_pigrum_1, Corynebacterium_3, and Staphylococcus_2-dominated cluster"
# [2] "Dolosigranulum_pigrum_1 and Corynebacterium_3-dominated cluster"                   
# [3] "Corynebacterium_3, Dolosigranulum_pigrum_1, and Staphylococcus_2-dominated cluster"
# [4] "Dolosigranulum_pigrum_1 and Staphylococcus_2-dominated cluster"

# Shorten the names made above for plotting 
# Define the new cluster names in the same order as the original clusters
new_cluster_names_vgo3_np <- c(
  "Dolo(1)/Cor(3)/Staph(2) clust",
  "Dolo(1)/Cor(3) clust",
  "Cor(3)/Dolo(1)/Staph(2) clust",
  "Dolo(1)/Staph(2) clust"
)

# Assign the new names to 'Cluster_Name' in 'top_3_asvs_df'
top_3_asvs_df_vgo3_np$Cluster_Name <- new_cluster_names_vgo3_np

# Update the clusters in the sample data to use the new names
ps_vgo3_np_filtered@sam_data$cluster <- top_3_asvs_df_vgo3_np$Cluster_Name[match(ps_vgo3_np_filtered@sam_data$cluster, top_3_asvs_df_vgo3_np$cluster)]
# Convert to character
ps_vgo3_np_filtered@sam_data$cluster <- as.character(ps_vgo3_np_filtered@sam_data$cluster)

# Count the number of samples in each cluster
cluster_counts_vgo3_np <- as.data.frame(table(ps_vgo3_np_filtered@sam_data$cluster))
colnames(cluster_counts_vgo3_np) <- c("Cluster", "Count")

# Calculate the relative proportions of each cluster
cluster_counts_vgo3_np$Proportion <- cluster_counts_vgo3_np$Count / sum(cluster_counts_vgo3_np$Count)

write.csv(cluster_counts_vgo3_np, "../Output/Cluster_analysis/VGO3_NP/cluster_counts_vgo3_np.csv")

# Create the plot with geom_bar for vertical stacking
bar_clusters_vgo3_np <- ggplot(data = cluster_counts_vgo3_np, aes(x = Cluster, y = (Proportion*100), fill = Cluster)) +
  geom_bar(stat = "identity") +  # Use stat = "identity" to directly use Proportion
  labs(title = "Bar plot of clusters in VGO3 NP samples",
       x = "Clusters",            # X-axis now represents clusters
       y = "% of samples",          # Y-axis represents proportions
       fill = "Cluster") +
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),      # Remove x-axis labels if needed
        axis.ticks.x = element_blank(),     # Remove x-axis ticks if needed
        legend.position = "right")

bar_clusters_vgo3_np
ggsave("../Output/Cluster_analysis/VGO3_NP/Bar_plot_cluster_proportions_clusters.svg", bar_clusters_vgo3_np, width = 8, height = 5)

```


## VGO3 OP clustering
```{r}
# Check number of taxa
ntaxa(ps_vgo3_op_filtered) # 337 ASVs 
# Extract metadata
meta_vgo3_op <- meta_to_df(ps_vgo3_op_filtered, sample_name = ".row_id")
# Extract otu table
otu_vgo3_op <- as(otu_table(ps_vgo3_op_filtered), "matrix")
# Standardise orientation so that downstream analyses always have rows = samples and columns = ASVs
if(taxa_are_rows(ps_vgo3_op_filtered)){otu_vgo3_op <- t(otu_vgo3_op)} 
otu_vgo3_op <- as.data.frame(otu_vgo3_op)
# Compute BC distances
ord_vgo3_op <- vegdist(otu_vgo3_op, method = "bray") 

# We used complete linkage to create the phylogenetic tree (hierarchical clustering). The method is based on maximum distance; the similarity of any two clusters is the similarity of their most dissimilar pair.
hc_vgo3_op_clust <- hclust(ord_vgo3_op, method = "complete")
# Reorder 'otu_op' based on the clustering order
otu_hc_vgo3_op_order_clust <- otu_vgo3_op[hc_vgo3_op_clust$order,]
# The Calinski-Harabasz and Silhouette methods were used to determine the appropriate location to cut the tree. Higher value of indices means the clusters are dense and well separated. Here, we looked for a peak or at least an abrupt elbow on the line plot of the indices.
CalHara_sil_indices_vgo3_op <- clust_ind_plot(ord_vgo3_op, seq(2, 25, by = 1), clust_method="complete")
write.csv(CalHara_sil_indices_vgo3_op$data, "../Output/Cluster_analysis/VGO3_OP/CalHara_silhouette_indices_vgo3_op.csv")
# Calinski-Harabasz Index: A higher value indicates well-separated clusters with high intra-cluster similarity.
# Silhouette: The optimal number of clusters typically shows the highest silhouette score, suggesting well-separated and compact clusters.
# Look for the points where there are peaks or elbows in the curve

# Create dendogram
dend_vgo3_op <- as.dendrogram(hc_vgo3_op_clust)

# 2 clusters
dend_2clusters <- color_branches(dend_vgo3_op, k = 2, groupLabels=T)
(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_vgo3_op, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# 3 clusters
dend_3clusters <- color_branches(dend_vgo3_op, k = 3, groupLabels=T)
(stats::dendrapply(dend_3clusters, noLabel))
cluster_assignments_3 <- cutree(dend_vgo3_op, k = 3)
cluster_counts_3 <- table(cluster_assignments_3)
print(cluster_counts_3)

# 4 clusters
dend_4clusters <- color_branches(dend_vgo3_op, k = 4, groupLabels=T)
(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_vgo3_op, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# 5 clusters
dend_5clusters <- color_branches(dend_vgo3_op, k = 5, groupLabels=T)
plot(stats::dendrapply(dend_5clusters, noLabel))
cluster_assignments_5 <- cutree(dend_vgo3_op, k = 5)
cluster_counts_5 <- table(cluster_assignments_5)
print(cluster_counts_5)

# 6 clusters
dend_6clusters <- color_branches(dend_vgo3_op, k = 6, groupLabels=T)
plot(stats::dendrapply(dend_6clusters, noLabel))
cluster_assignments_6 <- cutree(dend_vgo3_op, k = 6)
cluster_counts_6 <- table(cluster_assignments_6)
print(cluster_counts_6)

# 7 clusters
dend_7clusters <- color_branches(dend_vgo3_op, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_vgo3_op, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# 8 clusters
dend_8clusters <- color_branches(dend_vgo3_op, k = 8, groupLabels=T)
plot(stats::dendrapply(dend_8clusters, noLabel))
cluster_assignments_8 <- cutree(dend_vgo3_op, k = 8)
cluster_counts_8 <- table(cluster_assignments_8)
print(cluster_counts_8)

# 9 clusters
dend_9clusters <- color_branches(dend_vgo3_op, k = 9, groupLabels=T)
plot(stats::dendrapply(dend_9clusters, noLabel))
cluster_assignments_9 <- cutree(dend_vgo3_op, k = 9)
cluster_counts_9 <- table(cluster_assignments_9)
print(cluster_counts_9)

# 10 clusters
dend_10clusters <- color_branches(dend_vgo3_op, k = 10, groupLabels=T)
plot(stats::dendrapply(dend_10clusters, noLabel))
cluster_assignments_10 <- cutree(dend_vgo3_op, k = 10)
cluster_counts_10 <- table(cluster_assignments_10)
print(cluster_counts_10)

# 11 clusters
dend_11clusters <- color_branches(dend_vgo3_op, k = 11, groupLabels=T)
plot(stats::dendrapply(dend_11clusters, noLabel))
cluster_assignments_11 <- cutree(dend_vgo3_op, k = 11)
cluster_counts_11 <- table(cluster_assignments_11)
print(cluster_counts_11)

# 12 clusters
dend_12clusters <- color_branches(dend_vgo3_op, k = 12, groupLabels=T)
plot(stats::dendrapply(dend_12clusters, noLabel))
# check number per cluster
cluster_assignments_12 <- cutree(dend_vgo3_op, k = 12)
cluster_counts_12 <- table(cluster_assignments_12)
print(cluster_counts_12)

# 13 clusters
dend_13clusters <- color_branches(dend_vgo3_op, k = 13, groupLabels=T)
plot(stats::dendrapply(dend_13clusters, noLabel))
cluster_assignments_13 <- cutree(dend_vgo3_op, k = 13)
cluster_counts_13 <- table(cluster_assignments_13)
print(cluster_counts_13)

# 14 clusters
dend_14clusters <- color_branches(dend_vgo3_op, k = 14, groupLabels=T)
plot(stats::dendrapply(dend_14clusters, noLabel))
cluster_assignments_14 <- cutree(dend_vgo3_op, k = 14)
cluster_counts_14 <- table(cluster_assignments_14)
print(cluster_counts_14)
# 15 clusters
dend_15clusters <- color_branches(dend_vgo3_op, k = 15, groupLabels=T)
plot(stats::dendrapply(dend_15clusters, noLabel))
cluster_assignments_15 <- cutree(dend_vgo3_op, k = 15)
cluster_counts_15 <- table(cluster_assignments_15)
print(cluster_counts_15)

#### 2 clusters appears optimal ####
# k = 2
dend_2clusters <- color_branches(dend_vgo3_op, k = 2, groupLabels=T)
plot(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_2clusters, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# The function 'cuttree' is used to cut the tree at k = 2
groups_vgo3_op_c2 <- cutree(hc_vgo3_op_clust, k = 2)
ps_vgo3_op_filtered@sam_data$cluster <- groups_vgo3_op_c2[order(rownames(ps_vgo3_op_filtered@sam_data))]
ps_vgo3_op_filtered@sam_data$cluster <- ps_vgo3_op_filtered@sam_data$cluster %>% as.character()

# Determine the top 3 ASVs for each cluster
otu_vgo3_op <- as(otu_table(ps_vgo3_op_filtered), "matrix")
# if (taxa_are_rows(ps_vgo3_op_filtered)) {
#   otu_hc_vgo3_op_order_clust <- t(otu_hc_vgo3_op_order_clust)  # Ensure ASVs are columns, samples are rows
# }

head(otu_vgo3_op)
# Extract sample metadata with cluster information
sample_data_vgo3_op <- data.frame(ps_vgo3_op_filtered@sam_data)
# Combine the OTU table with the sample cluster data
otu_vgo3_op_with_clusters <- cbind(cluster = sample_data_vgo3_op$cluster, otu_hc_vgo3_op_order_clust)
otu_vgo3_op_with_clusters <- as.data.frame(otu_vgo3_op_with_clusters)
otu_vgo3_op_with_clusters$cluster <- as.factor(otu_vgo3_op_with_clusters$cluster)
# Convert all ASV abundances to numeric
otu_vgo3_op_with_clusters[-1] <- lapply(otu_vgo3_op_with_clusters[-1], as.numeric)
# Sum ASV abundances for each cluster
asv_sum_by_cluster_vgo3_op <- otu_vgo3_op_with_clusters %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
# Convert back to data frame for further processing
asv_sum_by_cluster_vgo3_op <- as.data.frame(asv_sum_by_cluster_vgo3_op)
# Find the top 3 most abundant ASVs in each cluster and their percentage
top_3_asvs_vgo3_op <- apply(asv_sum_by_cluster_vgo3_op[, -1], 1, function(x) {
  # Get total abundance in the cluster for calculating percentages
  total_abundance_vgo3_op <- sum(x, na.rm = TRUE)
  # Get names and values of the top 3 ASVs
  top_3_indices_vgo3_op <- order(x, decreasing = TRUE)[1:3]
  top_3_asv_names_vgo3_op <- names(x)[top_3_indices_vgo3_op]
  top_3_asv_values_vgo3_op <- x[top_3_indices_vgo3_op]
  # Calculate percentages
  top_3_asv_percentages_vgo3_op <- (top_3_asv_values_vgo3_op / total_abundance_vgo3_op) * 100
  # Combine ASV names and percentages into a string for each cluster
  result <- paste0(top_3_asv_names_vgo3_op, ": ", round(top_3_asv_percentages_vgo3_op, 2), "%", collapse = "; ")
  return(result)
})

# Combine results with cluster information
top_3_asvs_df_vgo3_op <- data.frame(cluster = asv_sum_by_cluster_vgo3_op$cluster, Top_3_ASVs = top_3_asvs_vgo3_op)
# Display the top 3 ASVs for each cluster along with their percentages
print(top_3_asvs_df_vgo3_op)

#### Rename clusters ####
# If an ASV is >40% of the total abundance in a cluster: Name the cluster after that ASV only, as it is the dominant contributor.
# If the top ASV is between 20-40% of the total abundance: Use the two most abundant ASVs to name the cluster.
# If the top ASV is <20% of the total abundance: Use the top three ASVs to name the cluster, as none of the ASVs dominate significantly.

get_cluster_name <- function(top_asvs_string) {
  asv_info <- strsplit(top_asvs_string, "; ")[[1]]
  asv_names <- sapply(asv_info, function(x) strsplit(x, ": ")[[1]][1])
  asv_percentages <- as.numeric(sapply(asv_info, function(x) sub("%", "", strsplit(x, ": ")[[1]][2])))
  
  if (asv_percentages[1] > 40) {
    return(paste0(asv_names[1], "-dominated cluster"))
  } else if (asv_percentages[1] > 20) {
    return(paste0(asv_names[1], " and ", asv_names[2], "-dominated cluster"))
  } else {
    return(paste0(asv_names[1], ", ", asv_names[2], ", and ", asv_names[3], "-dominated cluster"))
  }
}

# Rename based on these rules
top_3_asvs_df_vgo3_op$Cluster_Name <- sapply(top_3_asvs_df_vgo3_op$Top_3_ASVs, get_cluster_name)
top_3_asvs_df_vgo3_op$Cluster_Name

# [1] "Streptococcus_5, Veillonella_7, and Streptococcus_8-dominated cluster"
# [2] "Veillonella_7, Streptococcus_5, and Streptococcus_8-dominated cluster"

# Shorten the names made above for plotting 
# Define the new cluster names in the same order as the original clusters
new_cluster_names_vgo3_op <- c(
  "Strep(5)/Veill(7)/Strep(8) clust",
  "Veill(7)/Strep(5)/Strep(8) clust"
)

# Assign the new names to 'Cluster_Name' in 'top_3_asvs_df'
top_3_asvs_df_vgo3_op$Cluster_Name <- new_cluster_names_vgo3_op

# Update the clusters in the sample data to use the new names
ps_vgo3_op_filtered@sam_data$cluster <- top_3_asvs_df_vgo3_op$Cluster_Name[match(ps_vgo3_op_filtered@sam_data$cluster, top_3_asvs_df_vgo3_op$cluster)]
# Convert to character
ps_vgo3_op_filtered@sam_data$cluster <- as.character(ps_vgo3_op_filtered@sam_data$cluster)

# Count the number of samples in each cluster
cluster_counts_vgo3_op <- as.data.frame(table(ps_vgo3_op_filtered@sam_data$cluster))
colnames(cluster_counts_vgo3_op) <- c("Cluster", "Count")

# Calculate the relative proportions of each cluster
cluster_counts_vgo3_op$Proportion <- cluster_counts_vgo3_op$Count / sum(cluster_counts_vgo3_op$Count)

write.csv(cluster_counts_vgo3_op, "../Output/Cluster_analysis/VGO3_OP/cluster_counts_vgo3_op.csv")

# Create the plot with geom_bar for vertical stacking
bar_clusters_vgo3_op <- ggplot(data = cluster_counts_vgo3_op, aes(x = Cluster, y = (Proportion*100), fill = Cluster)) +
  geom_bar(stat = "identity") +  # Use stat = "identity" to directly use Proportion
  labs(title = "Bar plot of clusters in vgo3 op samples",
       x = "Clusters",            # X-axis now represents clusters
       y = "% of samples",          # Y-axis represents proportions
       fill = "Cluster") +
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),      # Remove x-axis labels if needed
        axis.ticks.x = element_blank(),     # Remove x-axis ticks if needed
        legend.position = "right")

bar_clusters_vgo3_op
ggsave("../Output/Cluster_analysis/VGO3_OP/Bar_plot_cluster_proportions_clusters.svg", bar_clusters_vgo3_op, width = 8, height = 5)
```
# Visualisations - CONTINUE HERE!
## VGO2 NP 
```{r}
#### Visualisations for manuscript
dendro_vgo2_np <- create_dendro_gg(hc_vgo2_np_clust, size = 0.15)

otu_hm_cpall_gf_np <- as(otu_table(ps_RA_cpall_gf_np), "matrix")
otu_hm_cpall_gf_np_order <- otu_hm_cpall_gf_np[order(rowMeans(otu_hm_cpall_gf_np), decreasing = T)[1:40], hc_cpall_gf_np_clust$order]
meta_hm_cpall_gf_np_order <- ps_RA_cpall_gf_np %>% meta_to_df()
meta_hm_cpall_gf_np_order <- meta_hm_cpall_gf_np_order[hc_cpall_gf_np_clust$order,]
meta_hm_cpall_gf_np_order$cluster

all(colnames(otu_hm_cpall_gf_np_order) == hc_cpall_gf_np_clust$labels[hc_cpall_gf_np_clust$order])

prep_hm_data <- function(otu_ordered) {
  # Convert matrix to data frame and move row names to a column called "OTU"
  otu_ordered <- otu_ordered %>%
    as.data.frame() %>%
    tibble::rownames_to_column("OTU")
  # Format data for heatmap
  hm_data <- otu_ordered %>%
    tidyr::pivot_longer(cols = -OTU, names_to = "sample_id", values_to = "RA") %>%
    dplyr::mutate(
      RA = dplyr::if_else(RA == "0", NA_real_, as.numeric(RA)),  # Convert "0" to NA for heatmap
      OTU = format_OTU(OTU) %>% forcats::fct_inorder() %>% forcats::fct_rev(),  # Format OTUs
      sample_id = forcats::fct_inorder(sample_id)
    )
  return(hm_data)
}

# Call prep_hm_data with the data frame
hm_cpall_gf_np_data <- prep_hm_data(otu_hm_cpall_gf_np_order)

# Ensure OTU data is ordered by hierarchical clustering order
otu_hm_cpall_gf_np_order <- otu_hm_cpall_gf_np[order(rowMeans(otu_hm_cpall_gf_np), decreasing = TRUE), hc_cpall_gf_np_clust$order]
meta_hm_cpall_gf_np_order <- meta_hm_cpall_gf_np_order[hc_cpall_gf_np_clust$order, ]

# Overall heatmap
hm_cpall_gf_np <- hm_cpall_gf_np_data %>%
  ggplot(aes(x = sample_id, y = OTU, fill = RA)) +
  geom_tile() +
  scale_fill_gradientn(name="Relative abundance", 
                       colors = carto_pal(7, "BurgYl"), 
                       na.value = "white", 
                       guide = guide_colorbar(title.position = "top", title.hjust = 0.5)) +
  theme_grey(base_size = 10) + 
  theme(axis.text.x = element_blank(), 
        axis.text.y = element_markdown(size = 10),  # Set y-axis text size
        axis.ticks.x = element_blank(), 
        legend.position = "bottom",
        legend.key.height = grid::unit(0.2, "cm"),
        legend.key.width = grid::unit(0.8, "cm"),
        legend.text = element_text(size = 10),      # Set legend text size
        plot.background = element_blank()) +
  xlab("Sample") + 
  ylab("ASV")
hm_cpall_gf_np

# Cluster plot
# Ensure the metadata (meta_hm_cpall_gf_np_order) is also ordered according to the same clustering order
meta_hm_cpall_gf_np_order <- meta_hm_cpall_gf_np_order[hc_cpall_gf_np_clust$order, ]
# Reorder the Sample IDs for Consistency
meta_hm_cpall_gf_np_order$sample_id <- factor(meta_hm_cpall_gf_np_order$sample_id,
                                               levels = hc_cpall_gf_np_clust$labels[hc_cpall_gf_np_clust$order])

# Reorder data for the Cluster Plot
meta_hm_cpall_gf_np_order$cluster <- factor(meta_hm_cpall_gf_np_order$cluster,
                                             levels =unique(meta_hm_cpall_gf_np_order$cluster[order(meta_hm_cpall_gf_np_order$cluster)]))

hm_cluster_cpall_gf_np <- meta_hm_cpall_gf_np_order %>%
  ggplot(aes(x = sample_id, y = "Cluster", fill = cluster)) +
  theme_void() + 
  geom_tile() + 
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) + 
  labs(fill = "Cluster") + 
  theme(axis.text.y = element_text(size = 10),  # Set y-axis text size
        axis.ticks.length.y = unit(.18, "cm"),
        legend.direction = "vertical", 
        legend.box = "vertical",
        legend.key.size = unit(0.8, "lines"),
        legend.text = element_text(size = 10))    # Set legend text size
hm_cluster_cpall_gf_np

# Population plot 
hm_pop_cpall_gf_np <- meta_hm_cpall_gf_np_order %>%
  ggplot(aes(x = sample_id, y = "Population", fill = population)) +
  theme_void() + 
  geom_tile() + 
  scale_fill_brewer(palette = "Set2") + 
  labs(fill = "Population") + 
  theme(axis.text.y = element_text(size = 10),  # Set y-axis text size
        axis.ticks.length.y = unit(.18, "cm"),
        legend.direction = "vertical", 
        legend.box = "vertical",
        legend.key.size = unit(0.8, "lines"),
        legend.text = element_text(size = 10))    # Set legend text size
hm_pop_cpall_gf_np

# Prevalence plot
# Using original phyloseq (not RA)
df_pres_np <- ps_cpall_gf %>%
  prune_samples(sample_data(.)$niche == "NP", .) %>%
  ps_to_df()

# Check dimensions
colnames(df_pres_np)
dim(df_pres_np)
# Convert counts to binary
df_pres_np[, colnames(df_pres_np)[700:20728]] <- lapply(df_pres_np[, colnames(df_pres_np)[700:20728]], function(x) ifelse(x > 2, 1, 0))
df_pres_np <- df_pres_np[700:20728]

prev <- colMeans(df_pres_np)
prev_np <- data.frame(Taxa = colnames(df_pres_np), Prevalence = prev)
prev_np$Taxa <- format_OTU(prev_np$Taxa)
prev_np <- subset(prev_np, Taxa %in% hm_cpall_gf_np_data$OTU)
prev_np$Taxa <- factor(prev_np$Taxa, levels = levels(hm_cpall_gf_np_data$OTU))

p_prev_np <- ggplot(prev_np, aes(x = Prevalence, y = Taxa)) +
  geom_bar(stat = "identity", fill = "azure4", width = 0.5) +
  labs(x = "Prevalence") +
  theme_minimal() + 
  theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_x_continuous(breaks = c(0, 0.5, 1),
                     labels = scales::percent_format())
p_prev_np


# Make patchwork of plot
layout_cp_gf <- "
AAAAAA#
AAAAAA#
AAAAAA#
BBBBBB#
CCCCCC#
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
DDDDDDE
"

dendro_cpall_gf_np$plot + hm_pop_cpall_gf_np + hm_cluster_cpall_gf_np + hm_cpall_gf_np + p_prev_np + plot_layout(design = layout_cp_gf, guides = "collect") & theme(legend.position = 'bottom')

ggsave("../Output/Cluster_analysis/CPall_GF_NP/heatmap_cluster_cpall_gf_np.svg", height = 10, width = 8)
ggsave("../Output/Cluster_analysis/CPall_GF_NP/heatmap_cluster_cpall_gf_np.png", height = 10, width = 8)


```

```

