---
title: "Longitudinal analysis - T1 microbiome --> T1 lung function"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

```{r}
library(phyloseq); library(vegan); library(tidyr); library(dplyr); library(ggplot2);library(tidyverse); library(microbiomer); library(paletteer);library(microbiome); library(devtools); library(forcats); library(haven); library(openxlsx); library(vegan); library(ggsignif); library(patchwork); library(decontam); library(RColorBrewer); library(forcats); library(lubridate); library(stringr); library(janitor); library(gtsummary); library(gt)

library(phyloseq); library(tidyverse); library(broom);library(dplyr); library(tidyr); library(stringr); library(ggplot2); library(MASS); library(phyloseq); library(microbiomer)

source("0_functions.R")

# Theme for plots in script
theme_set(theme_bw())
theme_update(axis.text.x = element_text(angle = 45, hjust=1),
             strip.text = element_text(colour = 'white'),
             strip.background =element_rect(fill="#2F4858", color = "#2F4858"))
```
# Load data
```{r}
#### VGO2 NP #### 
# These samples were sequenced at RIVM, processed using DADA2 at the RIVM by MLO
VGO2_NP_all <- readRDS("../Input/Phyloseq/VGO2_NP/Received_Mari_Lee/ps_raw.Rds") # all time points 
# Check the number of participants included at each time point in the raw dataset (T1, T2, T3)
time_points_NP <- sample_data(VGO2_NP_all) %>%
  data.frame() %>%
  filter(time_period %in% c("T1", "T2", "T3")) %>%
  group_by(time_period) %>%
  summarise(unique_participants = n_distinct(respnr)) %>%
  arrange(time_period) # T1 = 220, T2 = 12, T3 = 12

# Remove samples where subject_identifier contains "T2" or "T3" - we will only keep time point 1 for analysis 
VGO2_NP <- subset_samples(
  VGO2_NP_all,
  !str_detect(subject_identifier, "T2|T3")
)

# # Remove taxa with 0 sum 
# VGO2_NP <- prune_taxa(taxa_sums(VGO2_NP) > 0, VGO2_NP) # 289 samples, 10892 taxa

# Identify all duplicated `respnr` values (only in NP samples)
# Extract full metadata
meta_all <- sample_data(VGO2_NP) %>% data.frame()
meta_all$SampleID <- sample_names(VGO2_NP)
# Split into NP and non-NP
meta_NP <- meta_all %>% filter(sample_type == "NP")
meta_other <- meta_all %>% filter(sample_type != "NP")
# Deduplicate only NP samples by respnr
meta_NP_dedup <- meta_NP %>%
  filter(!is.na(respnr)) %>%
  distinct(respnr, .keep_all = TRUE)
meta_NP_na <- meta_NP %>% filter(is.na(respnr))
# Combine deduplicated NP with other niches
samples_to_keep <- c(meta_NP_dedup$SampleID, meta_NP_na$SampleID, meta_other$SampleID)
# Prune phyloseq object
VGO2_NP_dedup <- prune_samples(samples_to_keep, VGO2_NP)
# Confirm
cat("Original sample count:", nsamples(VGO2_NP), "\n") # 289
cat("After deduplication of NP samples:", nsamples(VGO2_NP_dedup), "\n") # 288


#### VGO2 OP #### 
# These samples were sequenced at the UMCU. Subsequently the V3 region was trimmed by AB, and run through the RIVM DADA2 pipeline by MLO
VGO2_OP_all <- readRDS("../Input/Phyloseq/VGO2_OP/V4/VGO_OP_v4_50_220_3.rds") 
# Extract and modify sample_data
sample_data(VGO2_OP_all) <- sample_data(VGO2_OP_all) %>%
  data.frame() %>%
  mutate(OP_time = case_when(
    str_detect(sample_identifier, "T1") ~ "T1",
    str_detect(sample_identifier, "T2") ~ "T2",
    str_detect(sample_identifier, "T3") ~ "T3",
    TRUE ~ NA_character_  # fallback in case none matched
  )) %>%
  sample_data()

time_points_VGO2_OP <- sample_data(VGO2_OP_all) %>%
  data.frame() %>%
  filter(OP_time %in% c("T1", "T2", "T3")) %>%
  group_by(OP_time) %>%
  summarise(unique_participants = n_distinct(sample_identifier)) %>%
  arrange(OP_time) # T1 = 281, T2 = 21, T3 = 20

# Only keep time point T1 from VGO2 OP data
VGO2_OP <- subset_samples(
  VGO2_OP_all,
  !str_detect(sample_identifier, "T2|T3"))# 340 samples, 1592 taxa (41 removed)
# # Remove taxa with 0 sum 
# VGO2_OP <- prune_taxa(taxa_sums(VGO2_OP) > 0, VGO2_OP) # 340 samples, 1540 taxa

# create respnr column  
VGO2_OP@sam_data$respnr <- sub("T1$", "", VGO2_OP@sam_data$sample_identifier)
VGO2_OP@sam_data$respnr[!grepl("^\\d", VGO2_OP@sam_data$respnr)] <- NA
# number of VGO2 OP samples
length(unique(na.omit(VGO2_OP@sam_data$respnr))) # 281
# Additional metadata for VGO2_OP dataset
VGO2_OP_meta <- read.xlsx("../Input/Phyloseq/VGO2_OP/V4/sample_sheet_VGO_OP.xlsx")
VGO2_OP@sam_data$Sample
VGO2_OP_meta$assay_sample
# First, ensure both identifiers are character type
VGO2_OP@sam_data$Sample <- as.character(VGO2_OP@sam_data$Sample)
VGO2_OP_meta$assay_sample <- as.character(VGO2_OP_meta$assay_sample)
# Subset only relevant metadata column for merging
# Convert sample_data to data.frame and preserve rownames
sam_df <- data.frame(VGO2_OP@sam_data)
sam_df$Sample <- rownames(sam_df)  # create a 'Sample' column for merging
# Make sure meta has correct column names
meta_subset <- VGO2_OP_meta[, c("assay_sample", "pg/ul")]
colnames(meta_subset) <- c("Sample", "pg_ul")  # rename for easier handling
# Merge metadata
merged_meta <- merge(sam_df, meta_subset, by = "Sample", all.x = TRUE)
# Set rownames back using Sample column (which matches phyloseq sample names)
rownames(merged_meta) <- merged_meta$Sample
# Remove redundant 'Sample' column if needed
merged_meta$Sample <- NULL
# Assign merged data back to phyloseq
sample_data(VGO2_OP) <- sample_data(merged_meta)

# Add "niche" column to the sam_data
# Extract sample data as a data.frame
meta_OP_VGO2 <- sample_data(VGO2_OP) %>% data.frame()
meta_OP_VGO2$sample_identifier
# Create the niche column according to your rules:
meta_OP_VGO2$niche <- ifelse(
  !is.na(meta_OP_VGO2$respnr) & grepl("^\\d{5}$", meta_OP_VGO2$respnr),  # VGO3_ID exists & is 5-digit number
  "OP",
  meta_OP_VGO2$sample_identifier  # Otherwise, use sample_ID
)
# Assign updated metadata back to the phyloseq object (make sure niche is a factor)
sample_data(VGO2_OP) <- sample_data(meta_OP_VGO2)

# Identify all duplicated `respnr` values (only in OP samples)
# Extract full metadata
meta_all <- sample_data(VGO2_OP) %>% data.frame()
meta_all$SampleID <- sample_names(VGO2_OP)
# Split into NP and non-NP
meta_OP <- meta_all %>% filter(niche == "OP")
meta_other <- meta_all %>% filter(niche != "OP")
# Deduplicate only NP samples by respnr
meta_OP_dedup <- meta_OP %>%
  filter(!is.na(respnr)) %>%
  distinct(respnr, .keep_all = TRUE)
meta_OP_na <- meta_OP %>% filter(is.na(respnr))
# Combine deduplicated NP with other niches
samples_to_keep <- c(meta_OP_dedup$SampleID, meta_OP_na$SampleID, meta_other$SampleID)
# Prune phyloseq object
VGO2_OP_dedup <- prune_samples(samples_to_keep, VGO2_OP)
# Confirm
cat("Original sample count:", nsamples(VGO2_OP), "\n") # 340
cat("After deduplication of NP samples:", nsamples(VGO2_OP_dedup), "\n") # 340 i.e. none duplicated



#### VGO3 NP & OP #### 
VGO3_NPOP <- readRDS("../Input/Phyloseq/VGO3_NPOP/1_ps_raw.Rds")

# only keep control control participants - exclude GF and GP samples - keep blanks etc.
VGO3_NPOP <- subset_samples(VGO3_NPOP, !(population %in% c("GF", "GP")))
table(VGO3_NPOP@sam_data$niche) 
 #  NP   OP   BD   BP   FB ZMCB ZMCD  UMD   BB 
 # 978  989  152   17  101   87   29   26    3
table(VGO3_NPOP@sam_data$population) # CP 1941

# Exclude duplicated participants
# Extract metadata
meta <- data.frame(sample_data(VGO3_NPOP))
meta$SampleID <- sample_names(VGO3_NPOP)

# Mark duplicates: only within CP population AND niche is OP or NP
meta <- meta %>%
  mutate(dup = if_else(
    population == "CP" & niche %in% c("OP", "NP") & duplicated(paste(VGO3_ID, niche)),
    TRUE, FALSE
  ))

# Keep non-duplicates or samples outside the CP+OP/NP filter
keep_samples <- meta$SampleID[!meta$dup]
# Prune phyloseq object
VGO3_NPOP_dedup <- prune_samples(keep_samples, VGO3_NPOP)
# Optional: confirm counts
cat("Samples before:", nsamples(VGO3_NPOP), "\n") # 2397 
cat("Samples after deduplication:", nsamples(VGO3_NPOP_dedup), "\n") # 2364 

# Realised that some niche values are assigned as NP or OP but they are actually the blanks for these niches 
# Filter for rows where niche is NP or OP but VGO3_ID is NOT a 5-digit number
meta_check <- data.frame(VGO3_NPOP_dedup@sam_data)
bad_niche_assignments <- meta_check %>%
  filter(niche %in% c("NP", "OP") & !grepl("^\\d{5}$", VGO3_ID))
# Print how many problematic rows there are
cat("Number of samples with niche NP or OP but invalid VGO3_ID:", nrow(bad_niche_assignments), "\n") # 22 
# View these rows 
bad_niche_assignments # there are 2 which are not needed, the rest are blanks for the NP so need to be relabelled to be BL

# Extract sample data
meta <- data.frame(VGO3_NPOP_dedup@sam_data)
meta$SampleID <- sample_names(VGO3_NPOP_dedup)

# Remove two specific samples by sample_identifier
samples_to_remove <- meta %>%
  filter(sample_identifier %in% c("ISO_38_OP_TOP_B1_M", "ISO_38_OP_TOP_B2_M")) %>%
  pull(SampleID)
# Prune to remove those samples
VGO3_NPOP_dedup <- prune_samples(!(sample_names(VGO3_NPOP_dedup) %in% samples_to_remove), VGO3_NPOP_dedup)
# Fix niche values where VGO3_ID is not a 5-digit number but niche is NP or OP
# Re-extract updated metadata
meta <- data.frame(VGO3_NPOP_dedup@sam_data)
# Identify the rows to correct
bad_niche_idx <- which(meta$niche %in% c("NP", "OP") & !grepl("^\\d{5}$", meta$VGO3_ID))
# Convert niche to character before assignment
meta$niche <- as.character(meta$niche)
# Now assign "FB" to the bad indices
meta$niche[bad_niche_idx] <- "FB"
# Optionally convert niche back to factor
meta$niche <- factor(meta$niche)
# Assign the modified metadata back to the phyloseq object
sample_data(VGO3_NPOP_dedup) <- sample_data(meta)# Assign corrected metadata back to phyloseq object
table(VGO3_NPOP_dedup@sam_data$niche)

# How many samples have NA in 'niche' 
cat("Number of samples with NA in niche:", sum(is.na(meta$niche)), "\n") #3 have NA
# View the full rows where niche is still NA
meta_with_na_niche <- meta %>%
  filter(is.na(niche))

# Print or inspect these rows
View(meta_with_na_niche)

# Extract metadata from the phyloseq object
meta <- data.frame(sample_data(VGO3_NPOP_dedup))
meta$SampleID <- sample_names(VGO3_NPOP_dedup)

# Assign "FB" to the two identified samples
fb_samples <- c("ISO_73_NP_TOP_BL_N5_1001", "ISO_73_NP_TOP_BL_N5_1005")
meta$niche[meta$sample_identifier %in% fb_samples] <- "FB"

# Identify the remaining samples that had NA in 'niche'
meta_with_na_niche <- meta %>%
  filter(is.na(niche) & !(sample_identifier %in% fb_samples))

# Set 'niche' to "unknown" for these
# Add "unknown" to the levels of niche if not already present
if (!"unknown" %in% levels(meta$niche)) {
  levels(meta$niche) <- c(levels(meta$niche), "unknown")
}
meta$niche[meta$SampleID %in% meta_with_na_niche$SampleID] <- "unknown"
# Update the phyloseq object with the revised metadata
sample_data(VGO3_NPOP_dedup) <- sample_data(meta)
# Optional: check that no NAs remain
cat("Remaining NAs in niche:", sum(is.na(sample_data(VGO3_NPOP_dedup)$niche)), "\n") # 0 


# Check remaining unique participants in NP and OP niches (population CP)
length(unique(VGO3_NPOP_dedup@sam_data$VGO3_ID[VGO3_NPOP_dedup@sam_data$niche == "NP" & VGO3_NPOP_dedup@sam_data$population == "CP"])) # 956
length(unique(VGO3_NPOP_dedup@sam_data$VGO3_ID[VGO3_NPOP_dedup@sam_data$niche == "OP" & VGO3_NPOP_dedup@sam_data$population == "CP"])) # 954

# Add 'dataset' column to each phyloseq object's sample_data
sample_data(VGO2_NP_dedup)$dataset <- "VGO2_NP"
sample_data(VGO2_OP_dedup)$dataset <- "VGO2_OP"
sample_data(VGO3_NPOP_dedup)$dataset <- "VGO3_NPOP"

# Add 'timepoint' column to each phyloseq object's sample_data
sample_data(VGO2_NP_dedup)$timepoint <- "T0"
sample_data(VGO2_OP_dedup)$timepoint <- "T0"
sample_data(VGO3_NPOP_dedup)$timepoint <- "T1"

# dna_isolation_date is simply the date, convert this to run numbers to match with DI.Run from the VGO3 dataset
sample_data(VGO2_NP_dedup)$DI.Run <- as.factor(as.numeric(factor(sample_data(VGO2_NP_dedup)$dna_isolation_date)))
# DNA isolation run is not known for VGO2_OP dataset so add it as an empty column in sam_data
sample_data(VGO2_OP_dedup)$DI.Run <- NA  # This creates logical NAs

# filter metadata in each dataset
names(VGO2_NP_dedup@sam_data) # keep: subject_identifier, dna_isolation_date, qPCR_pg_ul, respnr, dataset, timepoint
names(VGO2_OP_dedup@sam_data)# keep: sample_identifier, dataset, timepoint
names(VGO3_NPOP_dedup@sam_data) # keep: sample_identifier, DI.Run, qPCR, VGO3_ID,  dataset, niche, timepoint


# For VGO2_NP
sample_data(VGO2_NP_dedup) <- sample_data(VGO2_NP_dedup)[, c("subject_identifier", "qPCR_pg_ul", "respnr", "dataset", "timepoint", "sample_type", "DI.Run")]

# For VGO2_OP
sample_data(VGO2_OP_dedup) <- sample_data(VGO2_OP_dedup)[, c("sample_identifier", "pg_ul", "respnr" ,"dataset", "timepoint", "niche", "DI.Run")]

# For VGO3_NPOP
sample_data(VGO3_NPOP_dedup) <- sample_data(VGO3_NPOP_dedup)[, c("sample_identifier", "qPCR", "VGO3_ID", "dataset", "timepoint", "niche", "DI.Run")]

# rename some of the metadata columns to align with eachother for later merging
# VGO2_NP
colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "subject_identifier"
] <- "sample_ID"

colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "sample_type"
] <- "niche"


colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "qPCR_pg_ul"
] <- "qPCR"

colnames(sample_data(VGO2_NP_dedup))[
  colnames(sample_data(VGO2_NP_dedup)) == "respnr"
] <- "VGO3_ID"

# VGO2_OP
colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "sample_identifier"
] <- "sample_ID"

colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "respnr"
] <- "VGO3_ID"

colnames(sample_data(VGO2_OP_dedup))[
  colnames(sample_data(VGO2_OP_dedup)) == "pg_ul"
] <- "qPCR"

# VGO3_NP
colnames(sample_data(VGO3_NPOP_dedup))[
  colnames(sample_data(VGO3_NPOP_dedup)) == "sample_identifier"
] <- "sample_ID"



####  Check longitudinal data (VGO2 and VGO3) #### 
## NP 
VGO3_NP <- subset_samples(VGO3_NPOP_dedup, niche == "NP")
# VGO3_NP <- prune_taxa(taxa_sums(VGO3_NP) > 0, VGO3_NP) # 957 samples, 11053 taxa

# Extract IDs from sample_data
ids_VGO2 <- sample_data(VGO2_NP_dedup)$VGO3_ID
ids_VGO3 <- sample_data(VGO3_NP)$VGO3_ID

# Find intersecting IDs
shared_ids_NP <- intersect(ids_VGO2, ids_VGO3)
# Number of shared individuals
length(shared_ids_NP) # 112 longitudinal NP samples


## OP 
VGO3_OP <- subset_samples(VGO3_NPOP, niche != "NP")
# VGO3_OP <- prune_taxa(taxa_sums(VGO3_OP) > 0, VGO3_OP) # 1404 samples, 12902 taxa

# Extract IDs from sample_data
ids_VGO2 <- sample_data(VGO2_OP_dedup)$VGO3_ID
ids_VGO3 <- sample_data(VGO3_OP)$VGO3_ID
# Find intersecting IDs
shared_ids_OP <- intersect(ids_VGO2, ids_VGO3)
# Number of shared individuals
length(shared_ids_OP) # 141 longitudinal OP samples
```
# Merge phyloseqs and taxonomy naming 
```{r}
head(taxa_names(VGO2_NP_dedup)) # ASVs named already e.g. "Corynebacterium_1", "Moraxella_2"  etc.
head(taxa_names(VGO2_OP_dedup)) # Named by the sequence
head(taxa_names(VGO3_NPOP_dedup))# ASVs named already e.g. "Dolosigranulum_pigrum_1" "Staphylococcus_2"

ntaxa(VGO2_NP_dedup) # 11668
ntaxa(VGO2_OP_dedup) # 1592
ntaxa(VGO3_NPOP_dedup) # 25639
nsamples(VGO2_NP_dedup) # 288
nsamples(VGO2_OP_dedup) # 340
nsamples(VGO3_NPOP_dedup) # 2362

# Fix VGO2_OP so that the refseq() slot has the sequences and taxa_names are the ASV names
# Get DNA sequences from current taxa_names
seqs <- taxa_names(VGO2_OP_dedup)
# Create ASV names
asv_names <- paste0("ASV", seq_along(seqs))
# Assign ASV names to the phyloseq object
taxa_names(VGO2_OP_dedup) <- asv_names
# Create DNAStringSet with ASV names
dna_seqs <- Biostrings::DNAStringSet(seqs)
names(dna_seqs) <- asv_names
# Add refseq slot
VGO2_OP_dedup <- merge_phyloseq(VGO2_OP_dedup, dna_seqs)
# VGO2_OP_dedup <- prune_taxa(taxa_sums(VGO2_OP_dedup) > 0, VGO2_OP_dedup)
# VGO2_OP_dedup <- prune_samples(sample_sums(VGO2_OP_dedup) > 0, VGO2_OP_dedup)

# Rename ASVs to Sequence Level
VGO2_NP_seq <- rename_ASVs(VGO2_NP_dedup)
VGO2_OP_seq <- rename_ASVs(VGO2_OP_dedup)
VGO3_NPOP_seq <- rename_ASVs(VGO3_NPOP_dedup)

# Check that tax_table is annotated in the same way
colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)

# Change all to lower case
colnames(VGO2_OP_seq@tax_table) <- tolower(colnames(VGO2_OP_seq@tax_table))
colnames(VGO2_NP_seq@tax_table) <- tolower(colnames(VGO2_NP_seq@tax_table))
colnames(VGO3_NPOP_seq@tax_table) <- tolower(colnames(VGO3_NPOP_seq@tax_table))

colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)


table(VGO2_NP_seq@sam_data$niche)
table(VGO2_OP_seq@sam_data$niche)
table(VGO3_NPOP_dedup@sam_data$niche)

# Remove 'Species_one' column from VGO3 phyloseqs
# For VGO3_NP_seq
tax_table(VGO3_NPOP_dedup) <- tax_table(VGO3_NPOP_dedup)[, !colnames(tax_table(VGO3_NPOP_dedup)) %in% c("species_one", "asv")]
colnames(VGO2_NP_seq@tax_table)
colnames(VGO2_OP_seq@tax_table)
colnames(VGO3_NPOP_seq@tax_table)


# Check for mismatches in sam_data before merging phyloseqs
# Sam_data variables between objects:
all.equal(colnames(sample_data(VGO2_NP_seq)),
          colnames(sample_data(VGO2_OP_seq)),
          colnames(sample_data(VGO3_NPOP_seq)))
# Get the column names for each phyloseq object
cols_np   <- colnames(sample_data(VGO2_NP_seq))
cols_op   <- colnames(sample_data(VGO2_OP_seq))
cols_vgo3 <- colnames(sample_data(VGO3_NPOP_seq))
# Compare: what's in VGO2 NP but not in VGO2 OP
setdiff(cols_np, cols_op) # nothing
# Compare: what's in VGO2 OP but not in VGO2 NP
setdiff(cols_op, cols_np)
# Compare: what's in VGO2 NP but not in VGO3 OP/NP
setdiff(cols_np, cols_vgo3)

# Get classes for each column in each phyloseq object
classes_np <- sapply(sample_data(VGO2_NP_seq), class)
classes_op <- sapply(sample_data(VGO2_OP_seq), class)
classes_vgo3 <- sapply(sample_data(VGO3_NPOP_seq), class)

# Combine into a comparison table
compare_classes <- data.frame(
  column_name = names(classes_np),
  NP_class    = unname(classes_np),
  OP_class    = unname(classes_op[names(classes_np)]),
  VGO3_class  = unname(classes_vgo3[names(classes_np)])
)

# Convert the mismatches 
sample_data(VGO2_NP_seq)$VGO3_ID <- as.character(sample_data(VGO2_NP_seq)$VGO3_ID)
sample_data(VGO3_NPOP_seq)$niche <- as.character(sample_data(VGO3_NPOP_seq)$niche)
VGO3_NPOP_seq@sam_data$niche

# Merge datasets now that all variables across the phyloseqs are aligned and the same data class
ps_combined <- merge_phyloseq(
  VGO2_NP_seq,
  VGO2_OP_seq,
  VGO3_NPOP_seq
  )

#  Reorder ASVs by Abundance
ps_reordered <- tax_sum_reorder(ps_combined)

# Rename ASVs Based on Taxonomy
ps_longitudinal <- create_phylo_custom(ps_reordered)
table(ps_longitudinal@sam_data$niche)

# Check tax table
head(ps_longitudinal@tax_table)
head(ps_longitudinal@otu_table)
ps_longitudinal@sam_data
ps_longitudinal
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 33168 taxa and 2990 samples ]
# sample_data() Sample Data:       [ 2990 samples by 7 sample variables ]
# tax_table()   Taxonomy Table:    [ 33168 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 33168 reference sequences ]
```

# Save ps
```{r}
# Save ps 
saveRDS(ps_longitudinal, "../Output/Longitudinal/Phyloseq/ps_raw_longitudinal.rds")

```
# Composition plots per dataset - before pre-processing
```{r}
# Subset ps_longitudinalfor dataset == "VGO3_NP"
ps_vgo2_np <- subset_samples(ps_longitudinal, dataset == "VGO2_NP" & niche == "NP")
ps_vgo2_np_filt <- prune_taxa(taxa_sums(ps_vgo2_np) > 0, ps_vgo2_np)
ps_vgo2_np_filt <- prune_samples(sample_sums(ps_vgo2_np_filt) > 0, ps_vgo2_np_filt)

# stacked bar plot
VGO2_NP_stackedbar <- create_ordered_bar_n(ps_vgo2_np, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 NP microbiome composition per sample")
# Save plot

ggsave("../Output/Longitudinal/Microbiome_descriptives/Before/VGO2_NP_barplot.svg", VGO2_NP_stackedbar, width = 5, height = 6)

# Subset ps_longitudinalfor dataset == "VGO2_OP"
ps_vgo2_op <- subset_samples(ps_longitudinal, dataset == "VGO2_OP" & niche == "OP")
ps_vgo2_op_filt <- prune_taxa(taxa_sums(ps_vgo2_op) > 0, ps_vgo2_op)
ps_vgo2_op_filt <- prune_samples(sample_sums(ps_vgo2_op_filt) > 0, ps_vgo2_op_filt)

# stacked bar plot
VGO2_OP_stackedbar <- create_ordered_bar_n(ps_vgo2_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 OP microbiome composition per sample")

# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/Before_preprocessing_composition/VGO2_OP_barplot.svg", VGO2_OP_stackedbar, width = 5, height = 6)


# Subset ps_longitudinalfor dataset == "VGO3_NP"
ps_vgo3_np <- subset_samples(ps_longitudinal, dataset == "VGO3_NPOP" & niche == "NP")
ps_vgo3_np_filt <- prune_taxa(taxa_sums(ps_vgo3_np) > 0, ps_vgo3_np)
ps_vgo3_np_filt <- prune_samples(sample_sums(ps_vgo3_np_filt) > 0, ps_vgo3_np_filt)

# stacked bar plot
VGO3_NP_stackedbar <- create_ordered_bar_n(ps_vgo3_np_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 NP microbiome composition per sample")

# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/Before_preprocessing_composition/VGO3_NP_barplot.svg", VGO3_NP_stackedbar, width = 5, height = 6)


# Subset ps_longitudinalfor dataset == "VGO3_OP"
ps_vgo3_op <- subset_samples(ps_longitudinal, dataset == "VGO3_NPOP" & niche == "OP")
ps_vgo3_op_filt <- prune_taxa(taxa_sums(ps_vgo3_op) > 0, ps_vgo3_op)
ps_vgo3_op_filt <- prune_samples(sample_sums(ps_vgo3_op_filt) > 0, ps_vgo3_op_filt)

# stacked bar plot
VGO3_OP_stackedbar <- create_ordered_bar_n(ps_vgo3_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 OP microbiome composition per sample")

# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/Before_preprocessing_composition/VGO3_OP_barplot.svg", VGO3_OP_stackedbar, width = 5, height = 6)

#### Combine plots: NP on top row, OP on bottom row ####
composition_combined <- (VGO2_NP_stackedbar | VGO3_NP_stackedbar) / 
                        (VGO2_OP_stackedbar | VGO3_OP_stackedbar) +
                        plot_annotation(tag_levels = 'A')  # Optional: add panel tags A, B, C, D

print(composition_combined)

# Save combined figure
ggsave("../Output/Longitudinal/Microbiome_descriptives/Before_preprocessing_composition/Combined_composition_barplots.svg",
       composition_combined, width = 12, height = 12)

```

# Pre-processing
## Visualise library size
```{r}
df <- meta_to_df(ps_longitudinal)
df$LibrarySize <- sample_sums(ps_longitudinal)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))

# Library sizes - all samples including blanks
libsize_niche <- ggplot(data = df, aes(x=Index, y=LibrarySize, color=niche)) + geom_point()
ggsave("../Output/Longitudinal/Preprocessing/Library_size_niche.svg", libsize_niche)

# Library size of NP and OP samples only 
df_filtered <- df %>% filter(niche %in% c("NP", "OP"))
libsize_niche_NPOP <- ggplot(data = df_filtered, aes(x = Index, y = LibrarySize, color = niche)) +
  geom_point()
ggsave("../Output/Longitudinal/Preprocessing/Library_size_NPvsOP.svg", libsize_niche_NPOP, width = 6, height = 5)

```
## Visualise bacterial density 
```{r}
bacterial_density_NP_OP <- df %>%
  filter(niche %in% c("NP", "OP")) %>%
  ggplot(aes(x = niche, y = qPCR)) +
  geom_boxplot(alpha = 0.4, fill = "white") +
  geom_jitter(aes(color = niche), width = 0.2, size = 1, alpha = 0.7) +
  labs(x = "Niche", y = "Bacterial density (pg/µl)", color = "Niche") +
  scale_y_continuous(trans = 'log10',
                     breaks = scales::trans_breaks('log10', function(x) 10^x),
                     labels = scales::trans_format('log10', scales::math_format(10^.x)))

ggsave("../Output/Longitudinal/Preprocessing/Bacterial_density_NPvsOP.svg", bacterial_density_NP_OP)

```
# Decontaminating the microbiome data
## Preparing data for decontamination
```{r}
# We want to decontaminate per sequencing group - i.e. VGO2_NP, VGO2_OP, VGO3_NP and VGO3_OP (VGO3 data was sequenced/processed together for NP and OP)
# Define decontamination groups 
sum(is.na(ps_longitudinal@sam_data$dataset))# No NAs for dataset

unique(sample_data(ps_longitudinal)$dataset) # "VGO2_NP"   "VGO2_OP"   "VGO3_NPOP"
sum(is.na(sample_data(ps_longitudinal)$decontam_group)) # 0 

sample_data(ps_longitudinal)$decontam_group <- case_when(
  sample_data(ps_longitudinal)$dataset == "VGO3_NPOP" ~ "VGO3_NPOP", 
  sample_data(ps_longitudinal)$dataset == "VGO2_NP" ~ "VGO2_NP",
  sample_data(ps_longitudinal)$dataset == "VGO2_OP" ~ "VGO2_OP",
  TRUE ~ NA_character_
)

# Count zero-sum samples *before* the split
zero_samples <- sample_sums(ps_longitudinal) == 0
table(zero_samples, sample_data(ps_longitudinal)$decontam_group)

table(sample_data(ps_longitudinal)$decontam_group)
# VGO2_NP   VGO2_OP  VGO3_combined 
# 288       340      2362 

# Split phyloseq by group
# Get the unique group names
groups <- unique(na.omit(sample_data(ps_longitudinal)$decontam_group))

# Use lapply to avoid `.x` evaluation issues
ps_longitudinal_groups <- setNames(lapply(groups, function(g) {
  keep_samples <- rownames(sample_data(ps_longitudinal))[sample_data(ps_longitudinal)$decontam_group == g]
  prune_samples(keep_samples, ps_longitudinal)
}), groups)

# Create and assign objects with custom names
for (group in names(ps_longitudinal_groups)) {
  assign(paste0(group, "_ps_for_decontam"), ps_longitudinal_groups[[group]])
}

# Save each phyloseq to an individual object
list2env(ps_longitudinal_groups, envir = .GlobalEnv)

# These are the phyloseqs for decontamination
VGO2_NP_ps_for_decontam
VGO2_OP_ps_for_decontam
VGO3_NPOP_ps_for_decontam

table(VGO2_NP_ps_for_decontam@sam_data$niche) # Niche = Blank, NP, Zymo Mock
table(VGO2_OP_ps_for_decontam@sam_data$niche) # EBL, NEG, NTC, OP, SBL, VBL, ZMCB
table(VGO3_NPOP_ps_for_decontam@sam_data$niche) # BB, BD, BL, BP, FB, NP, OP, UMD, ZMCB, ZMCD

# ZMCD	Zymo mock DNA (miseq PCR)		
# ZMCB	Zymo mock Bacteria (extraction)		
# AMCD	ATCC mock DNA (miseq PCR)		
# UMD	UMCU low density mock DNA (miseq PCR)		
# MSS	Mixed (fecal/NP/OP…) sample control 		
# MSD	Mixed (fecal/NP/OP…) sample control DNA		
# BD	Blank from DNA extraction		
# BP	Blank from PCR		

# For decontam we use the following as negative controls: 
## Blank (n=47)
## NTC, SBL, VBL, EBL (n = 54)
## BL (n=22), BD (n=1), BP (n=1) and FB for VGO3NPOP dataset

sample_data(VGO2_NP_ps_for_decontam)$is_neg <- sample_data(VGO2_NP_ps_for_decontam)$niche == "Blank"
sample_data(VGO2_OP_ps_for_decontam)$is_neg <- sample_data(VGO2_OP_ps_for_decontam)$niche %in% c("SBL", "VBL", "EBL")
sample_data(VGO3_NPOP_ps_for_decontam)$is_neg <- sample_data(VGO3_NPOP_ps_for_decontam)$niche == "BD"

# NP samples were in DI.Run = 39-74
# OP samples were in DI.Run = 1,2, 5-8, 10-12, 33-38, 75-91
## Therefore we only want to select the BD samples from these run to do decontam per niche

# Select BD samples for NP decontam
np_DI_runs <- 39:74
# Subset to NP and the correct BD runs
VGO3_NP_ps_for_decontam <- subset_samples(
  VGO3_NPOP_ps_for_decontam,
  (niche == "NP") | (niche == "BD" & DI.Run %in% np_DI_runs)
)
# Prune taxa with 0 total counts
VGO3_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_NP_ps_for_decontam) > 0, VGO3_NP_ps_for_decontam)

# Define BD runs for OP
op_DI_runs <- c(1, 2, 5:8, 10:12, 33:38, 75:91)
# Subset to OP and the correct BD runs
VGO3_OP_ps_for_decontam <- subset_samples(
  VGO3_NPOP_ps_for_decontam,
  (niche == "OP") | (niche == "BD" & DI.Run %in% op_DI_runs)
)
# Prune taxa with 0 total counts
VGO3_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_OP_ps_for_decontam) > 0, VGO3_OP_ps_for_decontam)


# Remove mocks & other unnecessary samples from the datasets
table(VGO2_NP_ps_for_decontam@sam_data$niche)
table(VGO2_OP_ps_for_decontam@sam_data$niche)
table(VGO3_NP_ps_for_decontam@sam_data$niche)
table(VGO3_OP_ps_for_decontam@sam_data$niche)

VGO2_NP_ps_for_decontam <- subset_samples(VGO2_NP_ps_for_decontam, niche != "Zymo Mock") %>% 
  subset_samples(., qPCR != 0) # 269 samples
table(VGO2_NP_ps_for_decontam@sam_data$niche) # blank = 47, NP = 222

VGO2_OP_ps_for_decontam <- subset_samples(VGO2_OP_ps_for_decontam, niche != "NEG") %>%  subset_samples(., niche != "NTC") %>%  subset_samples(., niche != "ZMCB") %>%
  subset_samples(., qPCR != 0) # 313   samples 
table(VGO2_OP_ps_for_decontam@sam_data$niche) # EBL  = 16, OP = 281, SBL= 8, VBL = 8

VGO3_NP_ps_for_decontam <- subset_samples(VGO3_NP_ps_for_decontam, niche != "BB") %>%  subset_samples(., niche != "BP") %>% subset_samples(., niche != "FB") %>% subset_samples(., niche != "UMD") %>%  subset_samples(., niche != "ZMCB") %>% subset_samples(., niche != "ZMCD") %>% subset_samples(., niche != "unknown") %>%  
  subset_samples(., qPCR != 0) # 1026  samples
table(VGO3_NP_ps_for_decontam@sam_data$niche)

VGO3_OP_ps_for_decontam <- subset_samples(VGO3_OP_ps_for_decontam, niche != "BB") %>%  subset_samples(., niche != "BP") %>% subset_samples(., niche != "FB") %>% subset_samples(., niche != "UMD") %>%  subset_samples(., niche != "ZMCB") %>% subset_samples(., niche != "ZMCD") %>% subset_samples(., niche != "unknown") %>%  
  subset_samples(., qPCR != 0) # 991 samples
table(VGO3_OP_ps_for_decontam@sam_data$niche)
```
## Run decontam combined method
```{r}
rownames(VGO2_NP_ps_for_decontam@tax_table)
rownames(VGO2_OP_ps_for_decontam@tax_table)
rownames(VGO3_NPOP_ps_for_decontam@tax_table)

# VGO2 NP 
VGO2_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO2_NP_ps_for_decontam) > 0, VGO2_NP_ps_for_decontam)
contaminants_VGO2_NP <- isContaminant(VGO2_NP_ps_for_decontam, method="combined", conc="qPCR", neg="is_neg")
contaminants_VGO2_NP <- subset(contaminants_VGO2_NP, contaminant == TRUE)
dim(contaminants_VGO2_NP)[1] # 64 identified in the NP
write.csv(contaminants_VGO2_NP, "../Output/Longitudinal/Decontamination/Decontam_combined_method/VGO2_NP_contaminants.csv")

# VGO2 OP
VGO2_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO2_OP_ps_for_decontam) > 0, VGO2_OP_ps_for_decontam)
contaminants_VGO2_OP <- isContaminant(VGO2_OP_ps_for_decontam, method="combined", conc="qPCR", neg="is_neg")
contaminants_VGO2_OP <- subset(contaminants_VGO2_OP, contaminant == TRUE)
dim(contaminants_VGO2_OP)[1] # 24 identified in the NP
write.csv(contaminants_VGO2_OP, "../Output/Longitudinal/Decontamination/Decontam_combined_method/VGO2_OP_contaminants.csv")

# VGO3 NP
# Prune taxa with total abundance of 0 across samples
VGO3_NP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_NP_ps_for_decontam) > 0, VGO3_NP_ps_for_decontam)
# Run decontam on NP subset (using method="combined")
contaminants_VGO3_NP <- isContaminant(VGO3_NP_ps_for_decontam, method = "combined", conc = "qPCR", neg = "is_neg")
contaminants_VGO3_NP <- subset(contaminants_VGO3_NP, contaminant == TRUE)
dim(contaminants_VGO3_NP)[1] # 129 identified in the NP
write.csv(contaminants_VGO3_NP, "../Output/Longitudinal/Decontamination/Decontam_combined_method/VGO3_NP_contaminants.csv")

# VGO3 OP
# Prune taxa with total abundance of 0 across samples
VGO3_OP_ps_for_decontam <- prune_taxa(taxa_sums(VGO3_OP_ps_for_decontam) > 0, VGO3_OP_ps_for_decontam)
# Run decontam on NP subset (using method="combined")
contaminants_VGO3_OP <- isContaminant(VGO3_OP_ps_for_decontam, method = "combined", conc = "qPCR", neg = "is_neg")
contaminants_VGO3_OP <- subset(contaminants_VGO3_OP, contaminant == TRUE)
dim(contaminants_VGO3_OP)[1] # 88 identified in the OP
write.csv(contaminants_VGO3_OP, "../Output/Longitudinal/Decontamination/Decontam_combined_method/VGO3_OP_contaminants.csv")
```
## Kruskall-Wallis across DNA isolation runs 
We implement a Kruskal-Wallis test to see which ASVs show discrepancies across the DNA isolation runs. These differences might indicate contamination, so the Kruskal-Wallis test is applied to each ASV to assess whether its relative abundance (RA) differs significantly across the DI.Run groups N.B. DI run groups are only available for VGO2 NP and VGO3 NP/OP dataset. 
```{r}
# We only know the DI.Run for VGO2 NP and VGO3 datasets
VGO2_NP_ps_for_decontam  

VGO2_NP_filt_RA_1 <- VGO2_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%
# A total of 792 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 32376 ASVs excluded).
# A total of 412 ASVs were found to be present at or above a level of confident detection (0.01% relative abundance) in at least 15 samples (n = 380 ASVs excluded).

VGO2_NP_filt_RA_2 <- VGO2_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%
# A total of 792 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 32376 ASVs excluded).
# A total of 27 ASVs were found to be present at or above a level of confident detection (1% relative abundance) in at least 40 samples (n = 765 ASVs excluded).

VGO2_NP_filt_RA_1 <- prune_taxa(!taxa_names(VGO2_NP_filt_RA_1) %in% taxa_names(VGO2_NP_filt_RA_2), VGO2_NP_filt_RA_1) # Exclude the high-abundance/core taxa from VGO2_NP_filt_RA_1

# Reshape phyloseq 
df_VGO2 <- ps_to_df(VGO2_NP_filt_RA_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO2_NP_filt_RA_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_VGO2))

# Select DI.Run and ASV columns
df_VGO2 <- df_VGO2 %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_VGO2_pivot <- pivot_longer(df_VGO2, cols = existing_asvs,
                             names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results <- df_VGO2_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results$contaminant <- ifelse(kruskal_results$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo2_np <- subset(kruskal_results, contaminant == TRUE)
KW_contams_vgo2_np # 94 detected 
KW_contams_vgo2_np$ASV

# Filter for significant ASVs
KW_contams_VGO2NP <- ggplot(subset(df_VGO2_pivot, ASV %in% KW_contams_vgo2_np$ASV), 
            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Longitudinal/Decontamination/Kruskall_Wallis/KW_VGO2NP.svg", KW_contams_VGO2NP, width = 20, height = 20)


#### VGO3 NP ####
VGO3_NP_ps_for_decontam_1 <- VGO3_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%

VGO3_NP_ps_for_decontam_2 <- VGO3_NP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%

VGO3_NP_ps_for_decontam_1 <- prune_taxa(!taxa_names(VGO3_NP_ps_for_decontam_1) %in% taxa_names(VGO3_NP_ps_for_decontam_2), VGO3_NP_ps_for_decontam_1) # Exclude the high-abundance/core taxa from ps_vgo3_np_for_decontam_1

# Reshape phyloseq 
df_both_np <- ps_to_df(VGO3_NP_ps_for_decontam_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO3_NP_ps_for_decontam_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_both_np))

# Select DI.Run and ASV columns
df_both_np <- df_both_np %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_both_np_pivot <- pivot_longer(df_both_np, cols = existing_asvs,
                             names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results_vgo3_np <- df_both_np_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results_vgo3_np$contaminant <- ifelse(kruskal_results_vgo3_np$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo3_np <- subset(kruskal_results_vgo3_np, contaminant == TRUE)
KW_contams_vgo3_np # 246 detected 
KW_contams_vgo3_np$ASV

# Filter for significant ASVs
KW_contams_VGO3NP <- ggplot(subset(df_both_np_pivot, ASV %in% KW_contams_vgo3_np$ASV), 
            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Longitudinal/Decontamination/Kruskall_Wallis/KW_contams_VGO3NP.svg", KW_contams_VGO3NP, width = 25, height = 25)
ggsave("../Output/Longitudinal/Decontamination/Kruskall_Wallis/KW_contams_VGO3NP.png", KW_contams_VGO3NP, width = 25, height = 25)


#### VGO3 OP ####
VGO3_OP_ps_for_decontam_1 <- VGO3_OP_ps_for_decontam %>% 
  pres_abund_filter() %>% # Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>% # convert to relative abundance
  pres_abund_filter(pres = 15, abund = 0.0001) # keep taxa in ≥15 samples at ≥0.01%

VGO3_OP_ps_for_decontam_2 <- VGO3_OP_ps_for_decontam %>% 
  pres_abund_filter() %>% #  Start with taxa present in ≥2 samples at ≥0.1% RA 
  to_RA() %>%  # convert to relative abundance
  pres_abund_filter(pres = 40, abund = 0.01) # keep taxa in ≥40 samples at ≥1%

VGO3_OP_ps_for_decontam_1 <- prune_taxa(!taxa_names(VGO3_OP_ps_for_decontam_1) %in% taxa_names(VGO3_OP_ps_for_decontam_2), VGO3_OP_ps_for_decontam_1) # Exclude the high-abundance/core taxa from ps_vgo3_op_for_decontam_1

# Reshape phyloseq 
df_both_op <- ps_to_df(VGO3_OP_ps_for_decontam_1)

# Identify ASVs from the phyloseq object
asv_names <- taxa_names(VGO3_OP_ps_for_decontam_1)
# Keep only ASVs that actually exist in df_VGO2
existing_asvs <- intersect(asv_names, colnames(df_both_op))

# Select DI.Run and ASV columns
df_both_op <- df_both_op %>%
  select(all_of(c("DI.Run", existing_asvs)))

# Pivot longer to long format (ASV, RA)
df_both_op_pivot <- pivot_longer(df_both_op, cols = existing_asvs,
                                 names_to = "ASV", values_to = "RA")

# Run Kruskal-Wallis test per ASV
kruskal_results_vgo3_op <- df_both_op_pivot %>%
  group_by(ASV) %>%
  summarise(
    kruskal_test = list(
      kruskal.test(RA ~ DI.Run, data = pick(DI.Run, RA)) %>% broom::tidy()
    ),
    .groups = "drop"
  ) %>%
  unnest(cols = kruskal_test)

# Flag significant ASVs as contaminants
kruskal_results_vgo3_op$contaminant <- ifelse(kruskal_results_vgo3_op$p.value < 0.001, TRUE, FALSE)
# Subset only significant (putative contaminants)
KW_contams_vgo3_op <- subset(kruskal_results_vgo3_op, contaminant == TRUE)
KW_contams_vgo3_op #125 detected 
KW_contams_vgo3_op$ASV

# Filter for significant ASVs
KW_contams_VGO3OP <- ggplot(subset(df_both_op_pivot, ASV %in% KW_contams_vgo3_op$ASV), 
                            aes(x = as.character(DI.Run), y = RA)) +
  labs(y = "Relative abundance (%)", x = "DI Run") +
  geom_point(alpha = 0.5, shape = 16, size = 2,
             position = position_jitter(width = 0.25, height = 0)) +
  geom_boxplot(colour = "black", alpha = 0.5) +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "black") +
  facet_wrap(~ASV, scales = "free_y") +
  theme_bw()

ggsave("../Output/Longitudinal/Decontamination/Kruskall_Wallis/KW_contams_VGO3OP.svg", KW_contams_VGO3OP, width = 20, height = 20)
```
## Combine contaminants from both methods per dataset
```{r}
# Check which contaminants are the same across decontam method and kruskall wallis
# Extract ASV names from decontam (rownames)
decontam_VGO2_NP_ASVs <- rownames(contaminants_VGO2_NP)
decontam_VGO2_OP_ASVs <- rownames(contaminants_VGO2_OP)
decontam_VGO3_NP_ASVs <- rownames(contaminants_VGO3_NP)
decontam_VGO3_OP_ASVs <- rownames(contaminants_VGO3_OP)

# Extract ASV names from Kruskal-Wallis results
KW_VGO2_NP_ASVs <- KW_contams_vgo2_np$ASV
KW_VGO3_NP_ASVs <- KW_contams_vgo3_np$ASV
KW_VGO3_OP_ASVs <- KW_contams_vgo3_op$ASV

# Function to get counts of overlaps and uniques
compare_contaminants <- function(decontam_asvs, kw_asvs) {
  decontam_set <- unique(decontam_asvs)
  kw_set <- unique(kw_asvs)
  
  shared <- intersect(decontam_set, kw_set)
  decontam_only <- setdiff(decontam_set, kw_set)
  kw_only <- setdiff(kw_set, decontam_set)
  
  list(
    shared_ASVs = shared,
    shared_count = length(shared),
    decontam_only_count = length(decontam_only),
    kw_only_count = length(kw_only)
  )
}
# Comparisons per niche/dataset
comparison_VGO2_NP <- compare_contaminants(decontam_VGO2_NP_ASVs, KW_VGO2_NP_ASVs)
# NO KW test for VGO2 OP
comparison_VGO3_NP <- compare_contaminants(decontam_VGO3_NP_ASVs, KW_VGO3_NP_ASVs)
comparison_VGO3_OP <- compare_contaminants(decontam_VGO3_OP_ASVs, KW_VGO3_OP_ASVs)

comparison_VGO2_NP
comparison_VGO3_NP
comparison_VGO3_OP


# Combine the contaminants identified by decontam with those detected using Kruskall-Wallis test
# VGO2_NP: both decontam and KW
contaminants_VGO2_NP_combined <- unique(c(decontam_VGO2_NP_ASVs, KW_VGO2_NP_ASVs))
length(contaminants_VGO2_NP_combined)# 151
# VGO2_OP: decontam only
contaminants_VGO2_OP_combined <- unique(decontam_VGO2_OP_ASVs)
length(contaminants_VGO2_OP_combined) # 24
# VGO3_NP: both decontam and KW
contaminants_VGO3_NP_combined <- unique(c(decontam_VGO3_NP_ASVs, KW_VGO3_NP_ASVs))
length(contaminants_VGO3_NP_combined)# 357
# VGO3_OP: both decontam and KW
contaminants_VGO3_OP_combined <- unique(c(decontam_VGO3_OP_ASVs, KW_VGO3_OP_ASVs))
length(contaminants_VGO3_OP_combined)# 187 


# Sort ASVs in teh list by their ranking (i.e. the number suffix at the end of their name)
# Sort ASV list by numeric suffix
sort_asvs_by_index <- function(asvs) {
  index <- as.numeric(sub(".*_", "", asvs))  # extract suffix
  asvs[order(index)]
}

# VGO2 NP
contaminants_VGO2_NP_combined_sorted <- sort_asvs_by_index(contaminants_VGO2_NP_combined)
# VGO2 OP
contaminants_VGO2_OP_combined_sorted <- sort_asvs_by_index(contaminants_VGO2_OP_combined)
# VGO3 NP
contaminants_VGO3_NP_combined_sorted <- sort_asvs_by_index(contaminants_VGO3_NP_combined)
# VGO3 OP
contaminants_VGO3_OP_combined_sorted <- sort_asvs_by_index(contaminants_VGO3_OP_combined)



# Create table of contaminants detected by each method and in each dataset 
# Create a union of all contaminant ASVs from all datasets
all_contaminants <- unique(c(
  contaminants_VGO2_NP_combined_sorted,
  contaminants_VGO2_OP_combined_sorted,
  contaminants_VGO3_NP_combined_sorted,
  contaminants_VGO3_OP_combined_sorted
))

# Initialize data frame
final_contaminant_df <- data.frame(
  ASV = all_contaminants,
  VGO2_NP_decontam = all_contaminants %in% contaminants_VGO2_NP_combined_sorted,
  VGO2_OP_decontam = all_contaminants %in% contaminants_VGO2_OP_combined_sorted,
  VGO3_NP_decontam = all_contaminants %in% contaminants_VGO3_NP_combined_sorted,
  VGO3_OP_decontam = all_contaminants %in% contaminants_VGO3_OP_combined_sorted
)

# Optional: sort by numeric ASV suffix for readability
final_contaminant_df <- final_contaminant_df[order(as.numeric(sub(".*_", "", final_contaminant_df$ASV))), ]

# View final table
head(final_contaminant_df)

# Write to CSV
write.csv(final_contaminant_df, "../Output/Longitudinal/Decontamination/Final_contaminants_combined.csv", row.names = FALSE)
# Now to do manual checks of ASVs flagged as contaminants
```
## Frequency plot of contaminants
We create a frequency plot of the top 100 identified contaminants. For most of these contaminant ASVs there seems to be a correlation between their relative abundance and the bacterial density (for VGO2 OP the pattern isn't so consistent)
```{r}
# Select top 100 ASVs (or fewer if < 100 available)
top100_VGO2_NP <- head(contaminants_VGO2_NP_combined_sorted, 100)
top100_VGO2_OP <- head(contaminants_VGO2_OP_combined_sorted, 100)
top100_VGO3_NP <- head(contaminants_VGO3_NP_combined_sorted, 100)
top100_VGO3_OP <- head(contaminants_VGO3_OP_combined_sorted, 100)

# Frequency plots for each dataset
# VGO2 NP
plot_VGO2_NP <- plot_frequency(subset_samples(VGO2_NP_ps_for_decontam, niche == "NP"), top100_VGO2_NP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")
plot_VGO2_NP

# VGO2 OP
# plot_VGO2_OP <- plot_frequency(subset_samples(VGO2_OP_ps_for_decontam, niche == "OP"), top100_VGO2_OP, 
#                conc="qPCR") +  xlab("Bacterial density (pg/ul)")
# plot_VGO2_OP
# Doesn't work for OP...
# Subset to OP samples
ps_op <- subset_samples(VGO2_OP_ps_for_decontam, niche == "OP")
# Remove samples with zero total counts
ps_op <- prune_samples(sample_sums(ps_op) > 0, ps_op)
# Get ASVs present in this filtered OP set
asvs_in_op <- taxa_names(ps_op)
# Intersect with top100 ASVs
top100_present_op <- intersect(top100_VGO2_OP, asvs_in_op)
# Plot
plot_VGO2_OP <- plot_frequency(
  ps_op, 
  top100_present_op,
  conc = "qPCR"
) + 
  xlab("Bacterial density (pg/µl)") +
  ggtitle("VGO2 OP")


# VGO3 NP
plot_VGO3_NP <- plot_frequency(subset_samples(ps_vgo3_np_for_decontam, niche == "NP"), top100_VGO3_NP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")

# VGO3 OP
plot_VGO3_OP <- plot_frequency(subset_samples(ps_vgo3_op_for_decontam, niche == "OP"), top100_VGO3_OP, 
               conc="qPCR") +  xlab("Bacterial density (pg/ul)")


ggsave("../Output/Longitudinal/Decontamination/freqplot_VGO2_NP_top100.svg", plot_VGO2_NP, width=15, height=15)
ggsave("../Output/Longitudinal/Decontamination/freqplot_VGO2_OP_top100.svg", plot_VGO2_OP, width=15, height=15)
ggsave("../Output/Longitudinal/Decontamination/freqplot_VGO3_NP_top100.svg", plot_VGO3_NP, width=15, height=15)
ggsave("../Output/Longitudinal/Decontamination/freqplot_VGO3_OP_top100.svg", plot_VGO3_OP, width=15, height=15)


# x-axis: Bacterial load (e.g., qPCR concentration, here pg/μl)
# y-axis: Frequency (relative abundance) of an individual ASV

## For true contaminants we expect higher relative abundance at lower bacterial loads, and a declining or flat trend as load increases. There should be a negative correlation between ASV abundance and sample bacterial density. If it is a false contaminant, then there will be no clear correlation with bacterial load. 
```
## Differences in contaminants between niches 
```{r}
# Identify contaminants shared between 
# Compare NP vs OP for VGO2
compare_np_op <- function(np_asvs, op_asvs) {
  shared <- intersect(np_asvs, op_asvs)
  np_only <- setdiff(np_asvs, op_asvs)
  op_only <- setdiff(op_asvs, np_asvs)
  
  list(
    shared_ASVs = shared,
    shared_count = length(shared),
    np_only_ASVs = np_only,
    np_only_count = length(np_only),
    op_only_ASVs = op_only,
    op_only_count = length(op_only)
  )
}

# VGO2: Compare contaminants between NP and OP
np_op_comparison_VGO2 <- compare_np_op(
  contaminants_VGO2_NP_combined_sorted,
  contaminants_VGO2_OP_combined_sorted
)

# VGO3: Compare contaminants between NP and OP
np_op_comparison_VGO3 <- compare_np_op(
  contaminants_VGO3_NP_combined_sorted,
  contaminants_VGO3_OP_combined_sorted
)

# View the results
np_op_comparison_VGO2 # 6 shared, 145 NP only, 18 OP only 
np_op_comparison_VGO3 # 58 shared, 299 NP only, 129 OP only 

```
## Check false contaminants
After screening the list of combined contaminants, we found some that some ASVs may not actually be contaminants.
```{r}
# Specifically we see that there are some high abundant common NP ASVs that are flagged as contaminants in the VGO3 OP dataset. But in the VGO3 sequencing, there were some mixed sequencing runs (i.e. mixing NP and OP niches). Checking in which runs these common NP ASVs appear to see if we can see signs of spill-over from the NP to OP niches.

# Common NP contaminants to check
# Define the ASVs you want to check for spill-over
contams_check_vgo3_OP <- c("Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3", "Moraxella_4")
# Filter to relevant ASVs in VGO3 OP and keep only non-zero abundance
df_both_op_contams_present <- df_both_op_pivot %>%
  filter(ASV %in% contams_check_vgo3_OP, RA > 0)
# For each ASV, list the unique DI.Runs where it's present
asv_run_presence_simple <- df_both_op_contams_present %>%
  group_by(ASV) %>%
  summarise(
    Runs_present = paste(sort(unique(DI.Run)), collapse = ", "),
    .groups = "drop"
  )
# View result
print(asv_run_presence_simple)
# Optionally write to file
write.csv(asv_run_presence_simple, "../Output/Decontamination/VGO3_OP_suspect_ASV_run_presence.csv", row.names = FALSE) # The ASVs seem to be present across a selection of different DNA isolation runs - therefore not probably related to spill-over from NP niche

# Extra check: 
# Check the relative abundance across DNA isolation runs for these suspicious contaminants identified. 
# Define ASVs of interest
asvs_to_check <- c("Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3", "Moraxella_4")
# Filter dataset for ASVs of interest and ensure DI.Run is ordered
plot_data <- df_both_op_pivot %>%
  filter(ASV %in% asvs_to_check) %>%
  mutate(DI.Run = factor(DI.Run, levels = sort(unique(as.numeric(as.character(DI.Run))))))
# Plot
p <- ggplot(plot_data, aes(x = DI.Run, y = RA)) +
  geom_point(alpha = 0.5, shape = 16, size = 2, position = position_jitter(width = 0.2)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 8, size = 2, color = "red") +
  facet_wrap(~ASV, scales = "free_y") +
  labs(
    x = "DNA Isolation Run (DI.Run)",
    y = "Relative Abundance",
    title = "RA of suspect ASVs across runs (VGO3 OP)"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
# Display plot
print(p)
# Save plot 
ggsave("../Output/Longitudinal/Decontamination/RA_by_run_VGO3_OP_suspects.svg", p, width = 10, height = 6)# no clear pattern

```
## Non-contaminants after manual inspection
```{r}
# Define non-contaminants manually verified
non_contam <- c(
  "Dolosigranulum_pigrum_1", "Staphylococcus_2", "Corynebacterium_3",
  "Moraxella_4", "Veillonella_140", "Prevotella_7_melaninogenica_165",
  "Anaerococcus_provencensis_168", "Acinetobacter_lwoffii_177", "Prevotella_7_179",
  "Cutibacterium_187", "Lactococcus_209", "Actinomyces_225"
)

# Filter contaminants by removing known non-contaminants
all_contaminants_combined_filt <- setdiff(all_contaminants, non_contam)
cat("Number of contaminants after filtering non-contaminants:", length(all_contaminants_combined_filt), "\n") # 597 
```
## Exclude all blanks from phyloseq objects 
```{r}
# Subset phyloseq object to samples from niches "NP" and "OP"
ps_longitudinal_NPOP <- subset_samples(ps_longitudinal, niche %in% c("NP", "OP"))

# Remove taxa with zero counts after subsetting samples
ps_longitudinal_NPOP_f <- prune_taxa(taxa_sums(ps_longitudinal_NPOP) > 0, ps_longitudinal_NPOP)
ps_longitudinal_NPOP; ps_longitudinal_NPOP_f # 6980 taxa removed
```
## Remove contaminants 
```{r}
# Remove contaminant taxa from the filtered phyloseq object
ps_longitudinal_NPOP_no_contams <- prune_taxa(!(taxa_names(ps_longitudinal_NPOP_f) %in% all_contaminants_combined_filt), ps_longitudinal_NPOP_f)

cat("Contaminants removed:", length(taxa_names(ps_longitudinal_NPOP_f)) - length(taxa_names(ps_longitudinal_NPOP_no_contams)), "\n") # 597 removed 

# Remove any taxa with zero counts after contaminant removal
ps2_longitudinal <- prune_taxa(taxa_sums(ps_longitudinal_NPOP_no_contams) > 0, ps_longitudinal_NPOP_no_contams)

# Final object: ps2_longitudinal contains no contaminants and no zero-count taxa

# Manual inspection of list - "Salmonella_25", "Listeria_26" seem suspicious - not likely in NP and OP, do not appear in the VGO3 OP data, see where they are in the dataset - looking into these: 
# Define the ASVs of interest
asvs_of_interest <- c("Salmonella_25", "Listeria_26")
# Get the OTU table and transpose so rows are samples
otu_df <- as.data.frame(t(otu_table(ps_longitudinal_NPOP_no_contams)))
# Find samples where either ASV is present (i.e., abundance > 0)
samples_with_asvs <- rownames(otu_df)[rowSums(otu_df[, asvs_of_interest, drop = FALSE] > 0) > 0]
# Extract the sample_data for those samples
sample_data_with_asvs <- sample_data(ps_longitudinal)[samples_with_asvs, ]
# View 
sample_data_with_asvs

```
# Final list of ASVs remaining
```{r}
# Extract ASV names from phyloseq object
asv_names <- taxa_names(ps2_longitudinal)
# Extract numeric suffix (everything after last underscore)
asv_suffix_numbers <- as.numeric(sub(".*_", "", asv_names))
# Order ASVs by numeric suffix
asv_ordered <- asv_names[order(asv_suffix_numbers)]
# Take only the first 100
asv_first_100 <- head(asv_ordered, 100)
# Print
print(asv_first_100)
# save final ASV list
write.csv(asv_first_100, "../Output/Longitudinal/Microbiome_descriptives/Ordered_ASV_names_top100.csv")
```

# Check for duplicated samples
```{r}
# Duplicates were already resolved before but check if any were missed
# Convert sample_data to data.frame
sample_df <- as.data.frame(sample_data(ps2_longitudinal))
# Create a new combined key of VGO3_ID and niche
sample_df$combined_key <- paste(sample_df$VGO3_ID, sample_df$niche, sample_df$study, sep = "_")
# Count how many times each combined_key appears
counts <- table(sample_df$combined_key)
# Keep keys that appear more than once (true duplicates)
dup_keys <- names(counts[counts > 1])
# Filter the data.frame for those duplicated keys
duplicates <- sample_df[sample_df$combined_key %in% dup_keys, ]
# View duplicates
duplicates
```

# Exclude low density samples
Next, we remove all the samples with low bacterial density (lower than 0.095 pg/ul). This includes 20 samples
```{r}
ps3_longitudinal <- prune_samples(sample_data(ps2_longitudinal)$qPCR >= 0.095, ps2_longitudinal)
ps3_longitudinal <- prune_taxa(taxa_sums(ps3_longitudinal) > 0, ps3_longitudinal) # Remove taxa with taxa sum of 0

ps2_longitudinal; ps3_longitudinal # 84 samples were removed 
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25591 taxa and 2415 samples ]
# sample_data() Sample Data:       [ 2415 samples by 84 sample variables ]
# tax_table()   Taxonomy Table:    [ 25591 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25591 reference sequences ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25355 taxa and 2332 samples ]
# sample_data() Sample Data:       [ 2332 samples by 84 sample variables ]
# tax_table()   Taxonomy Table:    [ 25355 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25355 reference sequences ]


```
# Rarefaction curves
```{r}
# First we take a look at the distribution of # reads per sample
read_counts <- sample_sums(ps3_longitudinal) %>% enframe() %>% ggplot(aes(y = 1, x = value)) +
  ggdist::stat_halfeye( adjust = .5, height = .3, justification = -.3,
                        .width = 0, point_colour = NA) + 
  geom_point(size = 1.3, shape = 16, alpha = .2, 
             position = position_jitter(seed = 1, width = 0, height = .05)) +
  geom_boxplot(width = .12, alpha = 0.5, outlier.color = NA) +
  scale_x_log10(breaks = c(1000, 5000, 10000, 30000, 100000, 300000),
                labels = scales::comma) +
  coord_cartesian(ylim = c(0.9, NA), clip = "off") +
  labs(x = "Number of reads/sample", y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  geom_vline(xintercept = 10000, linetype = "dashed", color = "red")
read_counts
ggsave("../Output/Longitudinal/Preprocessing/read_counts.svg", read_counts, width = 5, height = 4)

# Rarefaction curves - helps evaluate whether the sequencing depth (i.e., number of reads per sample) was sufficient to capture the diversity of microbial communities in the samples.
# X-axis: Number of sequencing reads (sampling effort)
# Y-axis: Number of observed species (or ASVs/OTUs, depending on your level of analysis)
## Each line on the curve represents a sample.
otu_tab_rare <- ps3_longitudinal %>% prune_samples(sample_sums(.) < 30000, .) %>% # only keep samples with <30000 reads
  otu_table() %>% # extract otu table
  as(., "matrix") %>% # convert it to a matrix
  t # transpose the matrix where rows = samples and cols = taxa

saveRDS(otu_tab_rare, "../Output/Longitudinal/Preprocessing/otu_tab_rare.rds")

# Generate rarefaction curves
# rarecurve() (from the vegan package) performs individual-based rarefaction for each sample by randomly subsampling reads without replacement at increasing steps (every 10 reads, here) and calculating the number of unique taxa detected at each step.
# rare_curve_data <- rarecurve(otu_tab_rare, step = 10)

# rare_curves <- map2_dfr(rare_curve_data, 
#                         rownames(otu_tab_rare), ~enframe(.x) %>% 
#                           mutate(sample_id = .y)) %>%
#   mutate(name = str_remove(name, "N") %>% as.numeric) %>%
#   ggplot(aes(x = name, y = value, group = sample_id)) +
#     geom_line(alpha = 0.15) +
#     scale_y_log10() +
#     scale_x_continuous(breaks = c(0, 5000, 10000, 15000, 20000, 25000), 
#                        labels = scales::comma) +
#     geom_vline(aes(xintercept = 10000),  colour="grey50", linetype="dotted") +
#     labs(x = "Sample size", y = "Number of species")
# 
# ggsave("../Output/Microbiome_descriptives/rarefaction_curves.svg", rare_curves)
```

# Exclusion of samples with less than 10,000 reads
```{r}
# Based on these figures, we chose a read cut-off at 10,000 reads. In the rarefaction curves, this was the point where the curves started flattening. 
# Remove the samples with <10,000 reads
ps4_longitudinal <- prune_samples(sample_sums(ps3_longitudinal) > 10000, ps3_longitudinal)
ps4_longitudinal <- prune_samples(!is.na(ps4_longitudinal@sam_data$niche), ps4_longitudinal)
ps4_longitudinal <- prune_taxa(taxa_sums(ps4_longitudinal) > 0, ps4_longitudinal)

ps3_longitudinal; ps4_longitudinal
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 25355 taxa and 2332 samples ]
# sample_data() Sample Data:       [ 2332 samples by 84 sample variables ]
# tax_table()   Taxonomy Table:    [ 25355 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 25355 reference sequences ]
# phyloseq-class experiment-level object
# otu_table()   OTU Table:         [ 23557 taxa and 2137 samples ]
# sample_data() Sample Data:       [ 2137 samples by 84 sample variables ]
# tax_table()   Taxonomy Table:    [ 23557 taxa by 9 taxonomic ranks ]
# refseq()      DNAStringSet:      [ 23557 reference sequences ]
```
# Composition plots per dataset - after pre-processing
```{r}
# Subset ps4_longitudinal for dataset == "VGO3_NP"
ps_vgo2_np <- subset_samples(ps4_longitudinal, dataset == "VGO2_NP" & niche == "NP")
ps_vgo2_np_filt <- prune_taxa(taxa_sums(ps_vgo2_np) > 0, ps_vgo2_np)
ps_vgo2_np_filt <- prune_samples(sample_sums(ps_vgo2_np_filt) > 0, ps_vgo2_np_filt)

# stacked bar plot
VGO2_NP_stackedbar <- create_ordered_bar_n(ps_vgo2_np, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 NP microbiome composition per sample (cleaned)")

# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/VGO2_NP_barplot.svg", VGO2_NP_stackedbar, width = 5, height = 6)

# Subset ps_longitudinalfor dataset == "VGO3_OP"
ps_vgo2_op <- subset_samples(ps4_longitudinal, dataset == "VGO2_OP" & niche == "OP")
ps_vgo2_op_filt <- prune_taxa(taxa_sums(ps_vgo2_op) > 0, ps_vgo2_op)
ps_vgo2_op_filt <- prune_samples(sample_sums(ps_vgo2_op_filt) > 0, ps_vgo2_op_filt)

# stacked bar plot
VGO2_OP_stackedbar <- create_ordered_bar_n(ps_vgo2_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO2 OP microbiome composition per sample (cleaned)")

# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/VGO2_OP_barplot.svg", VGO2_OP_stackedbar, width = 5, height = 6)


# Subset ps_longitudinalfor dataset == "VGO3_NP"
ps_vgo3_np <- subset_samples(ps_longitudinal, dataset == "VGO3_NPOP" & niche == "NP")
ps_vgo3_np_filt <- prune_taxa(taxa_sums(ps_vgo3_np) > 0, ps_vgo3_np)
ps_vgo3_np_filt <- prune_samples(sample_sums(ps_vgo3_np_filt) > 0, ps_vgo3_np_filt)

# stacked bar plot
VGO3_NP_stackedbar <- create_ordered_bar_n(ps_vgo3_np_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 NP microbiome composition per sample (cleaned)")
# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/VGO3_NP_barplot.svg", VGO3_NP_stackedbar, width = 5, height = 6)


# Subset ps_longitudinal for dataset == "VGO3_NP"
ps_vgo3_op <- subset_samples(ps_longitudinal, dataset == "VGO3_NPOP" & niche == "OP")
ps_vgo3_op_filt <- prune_taxa(taxa_sums(ps_vgo3_op) > 0, ps_vgo3_op)
ps_vgo3_op_filt <- prune_samples(sample_sums(ps_vgo3_op_filt) > 0, ps_vgo3_op_filt)

# stacked bar plot
VGO3_OP_stackedbar <- create_ordered_bar_n(ps_vgo3_op_filt, 15) + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) + 
  geom_bar(stat = "identity", width = 1) +
  labs(title = "VGO3 OP microbiome composition per sample (cleaned)")
# Save plot
ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/VGO3_OP_barplot.svg", VGO3_OP_stackedbar, width = 5, height = 6)



#### Combine plots: NP on top row, OP on bottom row ####
composition_combined <- (VGO2_NP_stackedbar | VGO3_NP_stackedbar) / 
                        (VGO2_OP_stackedbar | VGO3_OP_stackedbar) +
                        plot_annotation(tag_levels = 'A')  # Optional: add panel tags A, B, C, D

print(composition_combined)

# Save combined figure
ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/Combined_composition_barplots.svg", composition_combined, width = 12, height = 12)

```

# Composition plots - average RA per dataset (genus level)
Make plots which collapse ASVs to genera, calculate mean relative abundances for each genus per dataset, and keep genera that have >10% mean RA, create side-by-side bar plot across the datasets
```{r}
get_top_genera_RA <- function(ps_obj, dataset_name) {
  # Aggregate to genus level
  ps_genus <- tax_glom(ps_obj, taxrank = "genus")
  ps_genus_RA <- transform_sample_counts(ps_genus, function(x) x / sum(x))
  
  # Convert to data frame
  genus_df <- psmelt(ps_genus_RA)
  
  # Compute mean relative abundance and select top 10 genera
  genus_summary <- genus_df %>%
    group_by(genus) %>%
    summarise(Mean_RA = mean(Abundance, na.rm = TRUE)) %>%
    arrange(desc(Mean_RA)) %>%
    slice_head(n = 5) %>%
    mutate(Dataset = dataset_name)
  
  return(genus_summary)
}


# Run for each group
RA_vgo2_np <- get_top_genera_RA(ps_vgo2_np_filt, "VGO2_NP")
RA_vgo2_op <- get_top_genera_RA(ps_vgo2_op_filt, "VGO2_OP")
RA_vgo3_np <- get_top_genera_RA(ps_vgo3_np_filt, "VGO3_NP")
RA_vgo3_op <- get_top_genera_RA(ps_vgo3_op_filt, "VGO3_OP")

# Combine
RA_all <- bind_rows(RA_vgo2_np, RA_vgo2_op, RA_vgo3_np, RA_vgo3_op)

# Make Genus a factor ordered by abundance for cleaner plotting
RA_all <- RA_all %>%
  mutate(Genus = fct_reorder(genus, Mean_RA, .fun = max, .desc = TRUE))


# Add residual to each group to make bars stack to 100%
RA_all_with_resid <- RA_all %>%
  group_by(Dataset) %>%
  mutate(Total_RA = sum(Mean_RA),
         Residual_RA = 1 - Total_RA) %>%
  ungroup()

residual_rows <- RA_all_with_resid %>%
  select(Dataset, Residual_RA) %>%
  distinct() %>%
  mutate(genus = "Residual", Mean_RA = Residual_RA) %>%
  select(genus, Mean_RA, Dataset)

RA_all_final <- RA_all %>%
  select(genus, Mean_RA, Dataset) %>%
  bind_rows(residual_rows)

# Reorder genus so Residual is always on top
RA_all_final <- RA_all_final %>%
  mutate(Genus = ifelse(genus == "Residual", "Residual", as.character(genus))) %>%
  mutate(Genus = fct_reorder(Genus, Mean_RA, .fun = max, .desc = TRUE)) %>%
  mutate(Genus = fct_relevel(Genus, "Residual", after = 0))

# Reorder Dataset manually
RA_all_final <- RA_all_final %>%
  mutate(Dataset = factor(Dataset, levels = c("VGO2_NP", "VGO3_NP", "VGO2_OP", "VGO3_OP")))


# Assign colors
genus_levels <- unique(RA_all_final$genus)
n_non_residual <- length(setdiff(genus_levels, "Residual"))
palette_colors <- brewer.pal(min(12, n_non_residual), "Paired")

genus_colors <- c(
  setNames(palette_colors[1:n_non_residual], setdiff(genus_levels, "Residual")),
  Residual = "grey80"
)

# plot
genus_plot_all <- RA_all_final %>%
  ggplot(aes(x = Dataset, y = Mean_RA, fill = Genus)) +
  geom_bar(stat = "identity", color = "white", size = 0.3) +  # white lines between stacked genera
  scale_fill_manual(
    values = genus_colors,
    guide = guide_legend(title = "Genus", label.theme = element_text(face = "italic"))
  ) +
  labs(
    title = "Top 5 genera per dataset",
    y = "Mean relative abundance",
    x = "Dataset"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1), expand = c(0, 0)) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.border = element_rect(color = "grey40", size = 0.5),
    legend.key.size = unit(0.5, "cm")
  )

# View the plot
genus_plot_all

ggsave("../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/Genus_composition_plot_all.svg", genus_plot_all, height = 5, width =7)


```

# Composition plots
## Top 5 ASVs per niche 
```{r}
# Subset ps4_longitudinal for dataset == "VGO3_NP"
ps_vgo2_np <- subset_samples(ps4_longitudinal, dataset == "VGO2_NP" & niche == "NP")
ps_vgo2_np_filt <- prune_taxa(taxa_sums(ps_vgo2_np) > 0, ps_vgo2_np)
ps_vgo2_np_filt <- prune_samples(sample_sums(ps_vgo2_np_filt) > 0, ps_vgo2_np_filt)

# Subset ps_longitudinalfor dataset == "VGO3_OP"
ps_vgo2_op <- subset_samples(ps4_longitudinal, dataset == "VGO2_OP" & niche == "OP")
ps_vgo2_op_filt <- prune_taxa(taxa_sums(ps_vgo2_op) > 0, ps_vgo2_op)
ps_vgo2_op_filt <- prune_samples(sample_sums(ps_vgo2_op_filt) > 0, ps_vgo2_op_filt)

# Merge NP + OP
ps_combined_long <- merge_phyloseq(ps_vgo2_np_filt, ps_vgo2_op_filt)
ps_combined_long@sam_data$niche
# Transform to relative abundance 
ps_combined_long_rel <- transform_sample_counts(ps_combined_long, function(x) x / sum(x))

# Extract data
otu_df_long <- as.data.frame(otu_table(ps_combined_long_rel))
meta_df_long <- as.data.frame(sample_data(ps_combined_long_rel))

# Ensure sample order alignment
if (!identical(colnames(otu_df_long), rownames(meta_df_long))) {
  otu_df_long <- otu_df_long[, rownames(meta_df_long)]
}

# Average per niche
meta_df_long$niche <- as.factor(meta_df_long$niche)
otu_avg_long <- sapply(levels(meta_df_long$niche), function(niche) {
  rowMeans(otu_df_long[, rownames(meta_df_long[meta_df_long$niche == niche, ]), drop = FALSE])
})
otu_avg_long <- as.data.frame(otu_avg_long)
otu_avg_long$Taxon <- rownames(otu_avg_long)

#  Compute top 5 per niche
top_np_long <- otu_avg_long %>% arrange(desc(NP)) %>% slice_head(n = 5) %>% pull(Taxon)
top_op_long <- otu_avg_long %>% arrange(desc(OP)) %>% slice_head(n = 5) %>% pull(Taxon)

# Reshape for plotting
otu_long <- otu_avg_long %>%
  tidyr::pivot_longer(cols = c("NP", "OP"), names_to = "Niche", values_to = "Abundance") %>%
  mutate(TaxonGroup = case_when(
    Niche == "NP" & Taxon %in% top_np_long ~ Taxon,
    Niche == "OP" & Taxon %in% top_op_long ~ Taxon,
    TRUE ~ "Other"
  ))

# Average proportions within each Niche
df_plot_long <- otu_long %>%
  group_by(Niche, TaxonGroup) %>%
  summarise(MeanRelAbundance = sum(Abundance), .groups = "drop") %>%
  group_by(Niche) %>%
  mutate(Proportion = MeanRelAbundance / sum(MeanRelAbundance))


# Properly order taxa within each niche and move "Other" to top ----
df_plot_long <- df_plot_long %>%
  group_by(Niche) %>%
  arrange(Niche, Proportion) %>%
  mutate(TaxonGroup = forcats::fct_reorder2(
    TaxonGroup, Niche, Proportion, .desc = FALSE
  ))

# Create a unified ordering for plotting and legend:
# Within each niche: from least → most abundant (so bars stack bottom→top)
# Move "Other" to be plotted last (top of bar) and last in legend
unique_taxa_order <- df_plot_long %>%
  arrange(Niche, Proportion) %>%
  distinct(TaxonGroup) %>%
  pull(TaxonGroup)
unique_taxa_order <- c(setdiff(unique_taxa_order, "Other"), "Other")

df_plot_long$TaxonGroup <- factor(df_plot_long$TaxonGroup, levels = unique_taxa_order)

# Plot
palette_colors <- brewer.pal(10, "Paired")
unique_taxa <- unique(df_plot_long$TaxonGroup)
palette_final <- setNames(rep(palette_colors, length.out = length(unique_taxa)), unique_taxa)
palette_final["Other"] <- "grey80"

ggplot(df_plot_long, aes(x = Niche, y = Proportion, fill = TaxonGroup)) +
  geom_col(color = "white") +
  scale_fill_manual(values = palette_final, breaks = unique_taxa_order) +  # ensures legend order
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Top 5 ASVs per niche (each niche independent)",
    y = "Relative abundance",
    x = NULL,
    fill = "ASV"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

# Save
ggsave(
  "../Output/Longitudinal/Microbiome_descriptives/After_preprocessing_composition/Top5_ASV_per_niche_longitudinal.svg",
  width = 5, height = 4
)

```


# Rarefy data and compute alpha diversity indices 
```{r}
m <- ps4_longitudinal %>% meta_to_df()
gc()
set.seed(77)
ps4_longitudinal_rarefied <- phyloseq::rarefy_even_depth(ps4_longitudinal, replace = T, trimOTUs = T)
sample_sums(ps4_longitudinal_rarefied) 
min(sample_sums(ps4_longitudinal_rarefied))# min library size = 10002   
# 2638OTUs were removed because they are no longer present in any sample after random subsampling
# Compute alpha diversity indices: Observed, shannon and simpson
adiv <- data.frame(
  "Observed" = phyloseq::estimate_richness(ps4_longitudinal_rarefied, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(ps4_longitudinal_rarefied, measures = "Shannon"),
  "Simpson" = phyloseq::estimate_richness(ps4_longitudinal_rarefied, measures = "Simpson")
  )

# sample id have been changed to "." instead of "_"
adiv$sample_id <- gsub("\\.", "-", rownames(adiv))
m <- left_join(m, adiv)
m$qPCR_log <- log2(m$qPCR)

# Add alpha diversity values to phyloseq object
rownames(m) <- m$sample_id
m <- select(m, -c("sample_id"))
sample_data(ps4_longitudinal) <- m

# Add read number metadata
ps4_longitudinal@sam_data$final_read_nr <- colSums(ps4_longitudinal@otu_table)
```

# Preprocessing summary
```{r}
stats <- tibble(step = character(),
                samples = numeric(),
                taxa = numeric(),
                reads = numeric())

ps2_longitudinal
stats <- ps_stats(ps_longitudinal, stats, "Raw data")
stats <- ps_stats(ps_longitudinal_NPOP_f, stats, "Removal of blanks and zero sum taxa")
stats <- ps_stats(ps2_longitudinal, stats, "Exclusion of contaminant ASVs")
stats <- ps_stats(ps3_longitudinal, stats, "Exclusion of samples with bacterial density <0.095 pg/uL")
stats <- ps_stats(ps4_longitudinal, stats, "Exclusion of samples with reads <10,000")
stats <- ps_stats(ps4_longitudinal_rarefied, stats, "Rarefied for alpha diversity")
stats

write.csv(stats, "../Output/Longitudinal/Preprocessing/Processing_steps.csv")

```
# Metadata additions and fixes 
## Add participant metadata from VGO2 and VGO3 
```{r}
# Make 2 phyloseq objects - for our 2 longitudinal research questions:
# 1. how is microbiome at T0 associated with lung function at T1 (and the change in lung function from T0 to T1)
# 2. how does the microbiome change over time (from T0 to T1) and how is this associated with changes in lung function? 


# For RQ1 we only need microbiome at T0 (NP and/or OP), AND lung function at both timepoints (T0 and T1)
# For RQ2 we need lung function at both T0 and T1 and at least one niche (NP or OP) sampled at both T0 and T1

# Inputs (don't modify ps4_longitudinal)
ps4_longitudinal_orig <- ps4_longitudinal

# Load and prepare metadata

metadata_all <- read.csv("../Output/Longitudinal/Participant_data/VGO_combined_longitudinal.csv") %>%
  mutate(
    ID             = as.character(ID),
    study          = factor(study),
    timepoint      = factor(timepoint, levels = c("T0", "T1")),
    date_spiro     = ymd(date_spiro),
    age            = as.numeric(age),
    ethnicity      = factor(ethnicity),
    spirometrist   = as.character(spirometrist),
    gender         = factor(gender),
    height         = as.numeric(height),
    weight         = as.numeric(weight),
    bmi            = as.numeric(bmi),
    educationlevel = factor(educationlevel, levels = c("low", "medium", "high"), ordered = TRUE),
    smoke_status   = factor(smoke_status, levels = c("non", "former", "current"), ordered = TRUE),
    pneumo_lastyr  = factor(pneumo_lastyr),
    copd_doc_diag  = factor(copd_doc_diag),
    copd_gold      = factor(copd_gold),
    copd_lln       = factor(copd_lln),
    asthma_current = factor(asthma_current),
    asthma_ever    = factor(asthma_ever),
    asthma_doc_diag= factor(asthma_doc_diag),
    asthma_attk_lastyr = factor(asthma_attk_lastyr),
    asthma_med_current = factor(asthma_med_current),
    atopic         = factor(atopic),
    farmchild      = factor(farmchild),
    FEV1_pred      = as.numeric(FEV1_pred),
    FVC_pred       = as.numeric(FVC_pred),
    FEV1FVC_pred   = as.numeric(FEV1FVC_pred),
    MMEF_pred      = as.numeric(MMEF_pred),
    delta_FEV1_pred     = as.numeric(delta_FEV1_pred),
    delta_FVC_pred      = as.numeric(delta_FVC_pred),
    delta_FEV1FVC_pred = as.numeric(delta_FEV1FVC_pred),
    delta_MMEF_pred     = as.numeric(delta_MMEF_pred),
    fuperiod       = as.numeric(fuperiod),
    obstr_VGO2     = factor(obstr_VGO2),
    obstr_VGO3     = factor(obstr_VGO3),
    new_obstr      = factor(new_obstr)
  )

# -------------------------------------------------------------------------
# 2. Prepare phyloseq object: fix ID and drop zero-count samples
# -------------------------------------------------------------------------

ps4_work <- ps4_longitudinal  # or ps4_longitudinal_orig if using backup

sd <- as(sample_data(ps4_work), "data.frame")
colnames(sd)[colnames(sd) == "VGO3_ID"] <- "ID"
sample_data(ps4_work) <- sample_data(sd)

ps4_work <- prune_samples(sample_sums(ps4_work) > 0, ps4_work)
ps4_work <- prune_taxa(taxa_sums(ps4_work) > 0, ps4_work)

ps_meta <- as(sample_data(ps4_work), "data.frame") %>%
  rownames_to_column("SampleID") %>%
  mutate(
    ID        = as.character(ID),
    timepoint = as.character(timepoint),
    niche     = as.character(niche)
  )

# -------------------------------------------------------------------------
# 3. Deduplication helper function
# -------------------------------------------------------------------------

dedup_by_depth <- function(ps, group_cols = c("ID", "timepoint", "niche")) {
  sd <- as(sample_data(ps), "data.frame") %>% rownames_to_column("SampleID")
  sd$LibrarySize <- sample_sums(ps)[sd$SampleID]
  
  keep <- sd %>%
    group_by(across(all_of(group_cols))) %>%
    slice_max(order_by = LibrarySize, n = 1, with_ties = FALSE) %>%
    pull(SampleID)
  
  ps2 <- prune_samples(keep, ps)
  prune_taxa(taxa_sums(ps2) > 0, ps2)
}

# -------------------------------------------------------------------------
# RQ1: Microbiome at T0 (NP/OP) + Lung Function at both timepoints
# -------------------------------------------------------------------------

ids_LF_both <- metadata_all %>%
  group_by(ID) %>%
  filter(n_distinct(timepoint) == 2) %>%
  pull(ID) %>%
  unique()

ids_micro_T0 <- ps_meta %>%
  filter(timepoint == "T0", niche %in% c("NP", "OP")) %>%
  pull(ID) %>%
  unique()

ids_RQ1 <- intersect(ids_LF_both, ids_micro_T0)

samples_RQ1 <- ps_meta %>%
  filter(ID %in% ids_RQ1, timepoint == "T0", niche %in% c("NP", "OP")) %>%
  pull(SampleID)

ps_RQ1 <- ps4_work %>%
  prune_samples(samples_RQ1, .) %>%
  prune_taxa(taxa_sums(.) > 0, .) %>%
  dedup_by_depth()

# -------------------------------------------------------------------------
# Add metadata to phyloseq object (including delta_* variables)
# -------------------------------------------------------------------------

sd_RQ1 <- as(sample_data(ps_RQ1), "data.frame") %>%
  rownames_to_column("SampleID") %>%
  mutate(ID = as.character(ID), timepoint = as.character(timepoint))

merged_RQ1 <- sd_RQ1 %>%
  left_join(metadata_all, by = c("ID", "timepoint"))

rownames(merged_RQ1) <- merged_RQ1$SampleID
merged_RQ1$SampleID <- NULL
sample_data(ps_RQ1) <- sample_data(merged_RQ1)

ps_RQ1@sam_data$obstr_VGO2

# Save ps
saveRDS(ps_RQ1, "../Output/Longitudinal/Phyloseq/ps_RQ1_T0micro_LFboth.rds")


# RQ2 phyloseq:
#    LF both + at least ONE niche (NP or OP) paired across T0/T1
ids_micro_paired_niche <- ps_meta %>%
  filter(timepoint %in% c("T0","T1"), niche %in% c("NP","OP")) %>%
  group_by(ID, niche) %>%
  summarise(n_tp = n_distinct(timepoint), .groups = "drop") %>%
  filter(n_tp == 2) %>%
  pull(ID) %>% unique()

ids_RQ2 <- intersect(ids_LF_both, ids_micro_paired_niche)

samples_RQ2 <- ps_meta %>%
  filter(ID %in% ids_RQ2, timepoint %in% c("T0","T1"), niche %in% c("NP","OP")) %>%
  pull(SampleID)

ps_RQ2 <- prune_samples(samples_RQ2, ps4_work)
ps_RQ2 <- prune_taxa(taxa_sums(ps_RQ2) > 0, ps_RQ2)

ps_RQ2 <- dedup_by_depth(ps_RQ2, c("ID","timepoint","niche"))

# Join timepoint-specific LF metadata by ID+timepoint (so T0 sample gets T0 covars etc.)
sd_RQ2 <- as(sample_data(ps_RQ2), "data.frame") %>%
  rownames_to_column("SampleID") %>%
  mutate(ID = as.character(ID), timepoint = as.character(timepoint))

meta_for_join <- metadata_all %>%
  mutate(ID = as.character(ID), timepoint = as.character(timepoint))

lf_wide_RQ2 <- make_lf_wide(ids_RQ2)

merged_RQ2 <- sd_RQ2 %>%
  left_join(meta_for_join, by = c("ID","timepoint")) %>%
  left_join(lf_wide_RQ2, by = "ID")

rownames(merged_RQ2) <- merged_RQ2$SampleID
merged_RQ2$SampleID <- NULL
sample_data(ps_RQ2) <- sample_data(merged_RQ2)

saveRDS(ps_RQ2, "../Output/Longitudinal/Phyloseq/ps_RQ2_LFboth_micro_pairedNiche.rds")

# ============================================================
# 7) OPTIONAL: “paired-only within niche” version of RQ2 (cleanest for change models)
#     Keeps only ID+niche combos that truly have BOTH T0 and T1.
# ============================================================

sd2 <- as(sample_data(ps_RQ2), "data.frame") %>%
  rownames_to_column("SampleID") %>%
  mutate(ID = as.character(ID), timepoint = as.character(timepoint), niche = as.character(niche))

paired_id_niche <- sd2 %>%
  distinct(ID, timepoint, niche) %>%
  count(ID, niche, name = "n_tp") %>%
  filter(n_tp == 2) %>%
  select(ID, niche)

keep_samples <- sd2 %>%
  semi_join(paired_id_niche, by = c("ID","niche")) %>%
  pull(SampleID)

ps_RQ2_pairedOnly <- prune_samples(keep_samples, ps_RQ2)
ps_RQ2_pairedOnly <- prune_taxa(taxa_sums(ps_RQ2_pairedOnly) > 0, ps_RQ2_pairedOnly)

saveRDS(ps_RQ2_pairedOnly, "../Output/Longitudinal/Phyloseq/ps_RQ2_pairedNiche_only.rds")

# ============================================================
# 8) Quick checks
# ============================================================

cat("\n=== ORIGINAL ps4_longitudinal stays intact ===\n")
cat("Samples original:", nsamples(ps4_longitudinal_orig), "\n")

cat("\n=== RQ1 ===\n")
sd1 <- as(sample_data(ps_RQ1), "data.frame")
cat("Samples:", nsamples(ps_RQ1), "\n")
cat("Unique IDs:", length(unique(sd1$ID)), "\n")
print(table(sd1$timepoint))
print(table(sd1$niche))

cat("\n=== RQ2 ===\n")
sd2b <- as(sample_data(ps_RQ2), "data.frame")
cat("Samples:", nsamples(ps_RQ2), "\n")
cat("Unique IDs:", length(unique(sd2b$ID)), "\n")
print(table(sd2b$timepoint))
print(table(sd2b$niche))

cat("\n=== RQ2 paired-only (optional) ===\n")
sd2p <- as(sample_data(ps_RQ2_pairedOnly), "data.frame")
cat("Samples:", nsamples(ps_RQ2_pairedOnly), "\n")
cat("Unique IDs:", length(unique(sd2p$ID)), "\n")
print(table(sd2p$timepoint))
print(table(sd2p$niche))
```
# Population summary tables
## Summary of health data at T0 and T1 for longitudinal participants
```{r}
ps_VGO3 <- readRDS("../Output/Cross_sectional/Phyloseq/ps_final_VGO3.rds")
ps_meta_VGO3 <- sample_data(ps_VGO3)
ps_meta_VGO3$Sam_ID <- rownames(ps_VGO3@sam_data)

ids_both # IDs with BOTH lung function and microbiome at both timepoints - previously created (n = 110)

## Subset phyloseq to those IDs of participants with longitudinal LF and microbiome data
samples_keep_VGO3 <- ps_meta_VGO3$Sam_ID[ps_meta_VGO3$VGO3_ID %in% ids_both]

length(samples_keep_VGO3) # 194 microbiome samples kept

ps_VGO3_longitudinal <- prune_samples(samples_keep_VGO3, ps_VGO3)
ps_VGO3_longitudinal # 197 samples


# Pull sample_data and keep only NP/OP samples
sd_longitudinal <- sample_data(ps_VGO3_longitudinal) %>%
  data.frame(check.names = FALSE, stringsAsFactors = FALSE) %>%
  filter(niche %in% c("NP","OP"))

sd_vgo3_longitudinal <- sd_longitudinal %>%
  mutate(niche_rank = ifelse(niche == "NP", 1L, 2L)) %>%   # prefer NP (1) over OP (2)
  arrange(VGO3_ID, niche_rank) %>%                         # put NP first within each ID
  filter(!duplicated(VGO3_ID)) %>%                         # keep first row per VGO3_ID
  select(-niche_rank)

## Just select the variables for Table 1
df_vgo3_longitudinal <- sd_vgo3_longitudinal %>%
  select(
    Age, Gender, Height, Weight, BMI, Ethnicity, Education_level,
    Lung_med, Asthma_med_current, Antibiotics_lastmonth, Antibiotics_lastyr, Pneumonia_lastyr, Asthma_ever, Asthma_doc_diag, Asthma_attk_lastyr, COPD_emphy_doc_diag,
    Smoking_status, Farm_childhood, Farm_visit_lastyr, T0_wkprior_NH3, T1_wkprior_NH3, FEV1, FVC, FEV1_FVC, MMEF, FEV1_pred, FVC_pred, FEV1FVC_pred, MMEF_pred
  )

# Helper functions for summary
cont_summary <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_character_)
  m  <- mean(x)
  s  <- sd(x)
  q  <- stats::quantile(x, probs = c(.25,.5,.75), na.rm = TRUE)
  sprintf("%.1f (%.1f)\n%.1f [%.1f, %.1f]",
          m, s, q[2], q[1], q[3])
}

cat_summary <- function(x) {
  x <- factor(x)
  if (all(is.na(x))) return(NA_character_)
  n_tot <- sum(!is.na(x))
  levs <- levels(x)
  out <- sapply(levs, function(lv) {
    n <- sum(x == lv, na.rm = TRUE)
    pct <- if (n_tot > 0) 100 * n / n_tot else NA_real_
    sprintf("%s: %d (%.1f%%)", lv, n, pct)
  }, USE.NAMES = FALSE)
  paste(out, collapse = "\n")
}


## Build the Table 1 rows in the order you want them to appear
rows_vgo3_longitudinal <- tibble::tibble(
  Variable = c(
    "Age", "Gender", "Height", "Weight", "BMI", "Ethnicity", "Education_level",
    "Lung_med", "Asthma_med_current", "Antibiotics_lastmonth", "Antibiotics_lastyr","Pneumonia_lastyr", "Asthma_ever", "Asthma_doc_diag", "Asthma_attk_lastyr", "COPD_emphy_doc_diag",
    "Smoking_status", "Farm_childhood", "Farm_visit_lastyr","T0_wkprior_NH3", "T1_wkprior_NH3","FEV1", "FVC", "FEV1_FVC", "FEF25_75", 
    "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred"
  ),
  `VGO3` = c(
    cont_summary(df_vgo3_longitudinal$Age),
    cat_summary(df_vgo3_longitudinal$Gender),
    cont_summary(df_vgo3_longitudinal$Height),
    cont_summary(df_vgo3_longitudinal$Weight),
    cont_summary(df_vgo3_longitudinal$BMI),
    cat_summary(df_vgo3_longitudinal$Ethnicity),
    cat_summary(df_vgo3_longitudinal$Education_level),
    cat_summary(df_vgo3_longitudinal$Lung_med),
    cat_summary(df_vgo3_longitudinal$Asthma_med_current),
    cat_summary(df_vgo3_longitudinal$Antibiotics_lastmonth),
    cat_summary(df_vgo3_longitudinal$Antibiotics_lastyr),
    cat_summary(df_vgo3_longitudinal$Pneumonia_lastyr),
    cat_summary(df_vgo3_longitudinal$Asthma_ever),
    cat_summary(df_vgo3_longitudinal$Asthma_doc_diag),
    cat_summary(df_vgo3_longitudinal$Asthma_attk_lastyr),
    cat_summary(df_vgo3_longitudinal$COPD_emphy_doc_diag),
    cat_summary(df_vgo3_longitudinal$Smoking_status),
    cat_summary(df_vgo3_longitudinal$Farm_childhood),
    cat_summary(df_vgo3_longitudinal$Farm_visit_lastyr),
    cont_summary(df_vgo3_longitudinal$T0_wkprior_NH3),
    cont_summary(df_vgo3_longitudinal$T1_wkprior_NH3),
    cont_summary(df_vgo3_longitudinal$FEV1),
    cont_summary(df_vgo3_longitudinal$FVC),
    cont_summary(df_vgo3_longitudinal$FEV1_FVC),
    cont_summary(df_vgo3_longitudinal$MMEF),
    cont_summary(df_vgo3_longitudinal$FEV1_pred),
    cont_summary(df_vgo3_longitudinal$FVC_pred),
    cont_summary(df_vgo3_longitudinal$FEV1FVC_pred),
    cont_summary(df_vgo3_longitudinal$MMEF_pred)
  )
)

# Write to Excel
writexl::write_xlsx(
  list("Table 1 - VGO3" = rows_vgo3_longitudinal),
  path = "../Output/Longitudinal/Participant_data/Table1_longitudinal_T1.xlsx"
)




#### VGO2 longitudinal data ####
ps_meta_VGO2 <- read.csv("../Output/Cross_sectional/Participant_data/VGO2_data.csv")

ids_both # IDs with BOTH lung function and microbiome at both timepoints - previously created

# Ensure same structure
str(ps_meta_VGO2$ID)
str(ids_both)
ids_both <- as.integer(ids_both)

## Subset metadata to those IDs of participants with longitudinal LF and microbiome data
samples_keep_VGO2 <- ps_meta_VGO2[ps_meta_VGO2$ID %in% ids_both,]

nrow(samples_keep_VGO2) # 110 participants kept - correct

## Just select the variables needed for Table 1
samples_keep_VGO2

##
df_vgo2_longitudinal <- samples_keep_VGO2 %>%
  select(
    age, gender, height, weight, bmi, ethnicity, educationlevel, asthma_med_current, pneumo_lastyr, asthma_ever, asthma_doc_diag, asthma_attk_lastyr, copd_doc_diag,
    smoke_status, farmchild, FEV1, FVC, FEV1_FVC, MMEF, FEV1_pred, FVC_pred, FEV1FVC_pred, MMEF_pred
  )

# Helper functions for summary
cont_summary <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_character_)
  m  <- mean(x)
  s  <- sd(x)
  q  <- stats::quantile(x, probs = c(.25,.5,.75), na.rm = TRUE)
  sprintf("%.1f (%.1f)\n%.1f [%.1f, %.1f]",
          m, s, q[2], q[1], q[3])
}

cat_summary <- function(x) {
  x <- factor(x)
  if (all(is.na(x))) return(NA_character_)
  n_tot <- sum(!is.na(x))
  levs <- levels(x)
  out <- sapply(levs, function(lv) {
    n <- sum(x == lv, na.rm = TRUE)
    pct <- if (n_tot > 0) 100 * n / n_tot else NA_real_
    sprintf("%s: %d (%.1f%%)", lv, n, pct)
  }, USE.NAMES = FALSE)
  paste(out, collapse = "\n")
}


## Build the Table 1 rows in the order you want them to appear
rows_vgo2_longitudinal <- tibble::tibble(
  Variable = c("age", "gender", "height", "weight", "bmi", "ethnicity", "educationlevel", "asthma_med_current", "pneumo_lastyr", "asthma_ever", "asthma_doc_diag", "asthma_attk_lastyr", "copd_doc_diag",
    "smoke_status", "farmchild", "FEV1", "FVC", "FEV1_FVC", "MMEF", "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "MMEF_pred"
  ),
  `VGO2` = c(
    cont_summary(df_vgo2_longitudinal$age),
    cat_summary(df_vgo2_longitudinal$gender),
    cont_summary(df_vgo2_longitudinal$height),
    cont_summary(df_vgo2_longitudinal$weight),
    cont_summary(df_vgo2_longitudinal$bmi),
    cat_summary(df_vgo2_longitudinal$ethnicity),
    cat_summary(df_vgo2_longitudinal$educationlevel),
    cat_summary(df_vgo2_longitudinal$asthma_med_current),
    cat_summary(df_vgo2_longitudinal$pneumo_lastyr),
    cat_summary(df_vgo2_longitudinal$asthma_ever),
    cat_summary(df_vgo2_longitudinal$asthma_doc_diag),
    cat_summary(df_vgo2_longitudinal$asthma_attk_lastyr),
    cat_summary(df_vgo2_longitudinal$copd_doc_diag),
    cat_summary(df_vgo2_longitudinal$smoke_status),
    cat_summary(df_vgo2_longitudinal$farmchild),
    cont_summary(df_vgo2_longitudinal$FEV1),
    cont_summary(df_vgo2_longitudinal$FVC),
    cont_summary(df_vgo2_longitudinal$FEV1_FVC),
    cont_summary(df_vgo2_longitudinal$MMEF),
    cont_summary(df_vgo2_longitudinal$FEV1_pred),
    cont_summary(df_vgo2_longitudinal$FVC_pred),
    cont_summary(df_vgo2_longitudinal$FEV1FVC_pred),
    cont_summary(df_vgo2_longitudinal$MMEF_pred)
  )
)

# Write to Excel
writexl::write_xlsx(
  list("Table 1 - VGO2" = rows_vgo2_longitudinal),
  path = "../Output/Longitudinal/Participant_data/Table1_longitudinal_T0.xlsx"
)
```

## Summary of lung function changes
```{r}
RQ1_meta <- data.frame(ps_RQ1@sam_data)

# Helper functions for summary
cont_summary <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_character_)
  m  <- mean(x)
  s  <- sd(x)
  q  <- stats::quantile(x, probs = c(.25,.5,.75), na.rm = TRUE)
  sprintf("%.1f (%.1f)\n%.1f [%.1f, %.1f]",
          m, s, q[2], q[1], q[3])
}

cat_summary <- function(x) {
  x <- factor(x)
  if (all(is.na(x))) return(NA_character_)
  n_tot <- sum(!is.na(x))
  levs <- levels(x)
  out <- sapply(levs, function(lv) {
    n <- sum(x == lv, na.rm = TRUE)
    pct <- if (n_tot > 0) 100 * n / n_tot else NA_real_
    sprintf("%s: %d (%.1f%%)", lv, n, pct)
  }, USE.NAMES = FALSE)
  paste(out, collapse = "\n")
}


## Build the Table 1 rows in the order you want them to appear
rows_both <- tibble::tibble(
  Variable = c("fuperiod", "delta_FEV1_pred", "delta_FVC_pred", "delta_FEV1FVC_pred", "delta_MMEF_pred", "obstr_VGO2", "obstr_VGO3", "new_obstr"
  ),
   c(
    cont_summary(RQ1_meta$fuperiod),
    cont_summary(RQ1_meta$delta_FEV1_pred),
    cont_summary(RQ1_meta$delta_FVC_pred),
    cont_summary(RQ1_meta$delta_FEV1FVC_pred),
    cont_summary(RQ1_meta$delta_MMEF_pred),
    cat_summary(RQ1_meta$obstr_VGO2),
    cat_summary(RQ1_meta$obstr_VGO3),
    cat_summary(RQ1_meta$new_obstr)
    
  )
)

# Write to Excel
writexl::write_xlsx(
  list("Table 1 - change" = rows_both),
  path = "../Output/Longitudinal/Participant_data/Table1_VGO2toVGO3_change.xlsx"
)


## Plot change
# Lung function change per person (unique row per VGO3_ID)
# pick one row per person with the most non-missing change values
RQ1_meta <- data.frame(sample_data(ps_RQ1))

delta_vars  <- c("delta_FEV1_pred", "delta_FVC_pred", "delta_FEV1FVC_pred", "delta_MMEF_pred")

RQ1_meta_unique <- RQ1_meta %>%
  mutate(nonmiss_lf = rowSums(across(all_of(delta_vars), ~ !is.na(.x)))) %>%
  arrange(ID, desc(nonmiss_lf)) %>%
  group_by(ID) %>% slice_head(n = 1) %>% ungroup() %>%
  select(ID, all_of(lung_change_cols))


plot_df <- RQ1_meta_unique %>%
  select(all_of(delta_vars)) %>%
  pivot_longer(everything(), names_to = "Measure", values_to = "Delta") %>%
  filter(!is.na(Delta)) %>%
  mutate(
    Measure = recode(
      Measure,
      delta_FEV1_pred    = "ΔFEV1",
      delta_FVC_pred     = "ΔFVC",
      delta_FEV1FVC_pred = "ΔFEV1/FVC",
      delta_MMEF_pred    = "ΔFEF25–75"
    ),
    Measure = factor(Measure, levels = c("ΔFEF25–75","ΔFEV1/FVC","ΔFVC","ΔFEV1")),  # match your example order (top→bottom)
    Direction = if_else(Delta >= 0, "Improved", "Declined")
  )


summ_df <- plot_df %>%
  group_by(Measure) %>%
  summarise(
    n = n(),
    med = median(Delta, na.rm = TRUE),
    q1  = quantile(Delta, 0.25, na.rm = TRUE),
    q3  = quantile(Delta, 0.75, na.rm = TRUE),
    p_improved = mean(Delta >= 0, na.rm = TRUE) * 100,
    p_declined = mean(Delta <  0, na.rm = TRUE) * 100,
    .groups = "drop"
  )


plot_delta <- ggplot(plot_df, aes(x = Delta, y = Measure)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  geom_violin(fill = "grey70", color = NA, width = 0.85, trim = FALSE) +
  geom_errorbarh(data = summ_df,
                 aes(xmin = q1, xmax = q3, y = Measure),
                 inherit.aes = FALSE, height = 0, linewidth = 1.0) +
  geom_point(data = summ_df,
             aes(x = med, y = Measure),
             inherit.aes = FALSE, size = 2.2) +
  geom_point(aes(color = Direction),
             position = position_jitter(height = 0.12, width = 0),
             size = 2, alpha = 0.75) +
    # add % improved 5 reduced
  geom_text(
  data = summ_df,
  aes(
    x = Inf, y = Measure,
    label = sprintf("+%.0f%%  /  −%.0f%%", p_improved, p_declined)
  ),
  inherit.aes = FALSE,
  hjust = 1.05, size = 3.6
)+
  labs(
    x = "Change per year \n(+ improvement, − decline beyond expected, 0 ≈ expected age decline)",
    y = NULL,
    color = "Direction"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    # legend
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14),
    # axes
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16)
  )



# % decreased on the left, % increased on the right
plot_delta <- ggplot(plot_df, aes(x = Delta, y = Measure)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.8) +
  geom_violin(fill = "grey70", color = NA, width = 0.85, trim = FALSE) +
  geom_errorbarh(
    data = summ_df,
    aes(xmin = q1, xmax = q3, y = Measure),
    inherit.aes = FALSE, height = 0, linewidth = 1.0
  ) +
  geom_point(
    data = summ_df,
    aes(x = med, y = Measure),
    inherit.aes = FALSE, size = 2.2
  ) +
  geom_point(
    aes(color = Direction),
    position = position_jitter(height = 0.12, width = 0),
    size = 2, alpha = 0.75
  ) +

  # left side: % decreased
  geom_text(
    data = summ_df,
    aes(x = -Inf, y = Measure, label = sprintf("%.0f%% decreased", p_declined)),
    inherit.aes = FALSE,
    hjust = -0.05,  # nudges text slightly inside plotting area
    size = 3.6
  ) +
  # right side: % increased
  geom_text(
    data = summ_df,
    aes(x = Inf, y = Measure, label = sprintf("%.0f%% increased", p_improved)),
    inherit.aes = FALSE,
    hjust = 1.05,
    size = 3.6
  ) +

  labs(
    x = "Change per year",
    y = NULL,
    color = "Direction"
  ) +
  theme_bw() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 16),
    legend.text  = element_text(size = 14),
    axis.text.x  = element_text(size = 14),
    axis.text.y  = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16)
  ) 


ggsave(
  "../Output/Longitudinal/Participant_data/Deltas_VGO2toVGO3.svg",
  plot_delta, width = 8.5, height = 4.8
)
```



# ANALYSES
## Read in data
```{r}
ps_l <- readRDS("../Output/Longitudinal/Phyloseq/ps_RQ1_T0micro_LFboth.rds")
```

# How is microbiome at T0 associated with lung function at T1?
## Data 
```{r}
get_season <- function(date_vec) {
  mo <- month(date_vec)
  out <- ifelse(mo %in% 3:5,  "Spring",
         ifelse(mo %in% 6:8,  "Summer",
         ifelse(mo %in% 9:11, "Autumn", "Winter")))
  factor(out, levels = c("Winter","Spring","Summer","Autumn"))
}

sd_l <- as(sample_data(ps_l), "data.frame") %>%
  rownames_to_column("SampleID") %>%
  mutate(
    date_spiro_T0 = ymd(date_spiro_T0),
    date_spiro_T1 = ymd(date_spiro_T1),

    # pick the spirometry date corresponding to the sample timepoint
    date_spiro_this_tp = case_when(
      timepoint == "T0" ~ date_spiro_T0,
      timepoint == "T1" ~ date_spiro_T1,
      TRUE ~ as.Date(NA)
    ),

    season_spiro = get_season(date_spiro_this_tp)
  )

# If you want your analyses to use a generic "season_microbiome_sampling" covariate name:
sd_l$season_microbiome_sampling <- sd_l$season_spiro

# Put back into phyloseq
rownames(sd_l) <- sd_l$SampleID
sd_l$SampleID <- NULL
sample_data(ps_l) <- sample_data(sd_l)

# checks
table(sample_data(ps_l)$timepoint, useNA = "ifany")
table(sample_data(ps_l)$season_microbiome_sampling, useNA = "ifany")

saveRDS(ps_l, "../Output/Longitudinal/Phyloseq/ps_RQ1_T0micro_LFboth_withSeason.rds")
```
# Separate microbiome data per dataset and niche
```{r}
# Subset per niche
# VGO3_NPOP dataset, NP niche
T0_NP <- subset_samples(ps_l, niche == "NP")
nsamples(T0_NP) # 95 NP samples

# VGO3_NPOP dataset, OP niche
T0_OP <- subset_samples(ps_l, niche == "OP")
nsamples(T0_OP) # 73 OP samples
```
# Filter taxa and convert to relative abundance
```{r}
# VGO3 NP 
# Remove low abundant ASVs and convert counts to relative abundances
T0_NP_filtered <- T0_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 477 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 5543 ASVs excluded).

# VGO3 OP 
# Remove low abundant ASVs and convert counts to relative abundances
T0_OP_filtered <- T0_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 215 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 5805 ASVs excluded).
```
# Extract OTU and metadata 
```{r}
otu_t0_np <- t(as.data.frame(otu_table(T0_NP_filtered)))
meta_t0_np <- meta_to_df(T0_NP_filtered, sample_name = ".row_id")

otu_t0_op <- t(as.data.frame(otu_table(T0_OP_filtered)))
meta_t0_op <- meta_to_df(T0_OP_filtered, sample_name = ".row_id")
```

# Alpha diversity and lung function at T1
```{r}
### Multivariable
dataset_list <- list(T0_NP = meta_t0_np, T0_OP = meta_t0_op)
alpha_vars   <- c("Observed","Shannon","Simpson")
lung_fun_cont <- c("FEV1_pred_T1","FVC_pred_T1","FEV1FVC_pred_T1","MMEF_pred_T1","delta_FEV1_pred","delta_FVC_pred" , "delta_FEV1FVC_pred", "delta_MMEF_pred")
core_covariates <- c("age_T0","gender_T0","smoke_status_T0","season_microbiome_sampling")
pretty_alpha <- c("Observed_z"="Observed richness","Shannon_z"="Shannon diversity","Simpson_z"="Simpson diversity")
pretty_lung  <- c(
  "FEV1_pred_T1_decline_z"      = "FEV1 % predicted (T1)",
  "FVC_pred_T1_decline_z"       = "FVC % predicted (T1)",
  "FEV1FVC_pred_T1_decline_z"   = "FEV1/FVC % predicted (T1)",
  "MMEF_pred_T1_decline_z"      = "MMEF % predicted (T1)",
  "delta_FEV1_pred_decline_z"   = "Delta FEV1 % predicted (T1-T0)",
  "delta_FVC_pred_decline_z"    = "Delta FVC % predicted (T1-T0)",
  "delta_FEV1FVC_pred_decline_z"= "Delta FEV1FVC % predicted (T1-T0)",
  "delta_MMEF_pred_decline_z"   = "Delta MMEF % predicted (T1-T0)"
)
meta_t0_np$delta_FEV1FVC_pred

preprocess_dataset <- function(df){
  df <- df %>%
    dplyr::mutate(
      Age = suppressWarnings(as.numeric(as.character(age_T0))),
      FEV1_pred = suppressWarnings(as.numeric(as.character(FEV1_pred_T1))),
      FVC_pred  = suppressWarnings(as.numeric(as.character(FVC_pred_T1))),
      FEV1FVC_pred = suppressWarnings(as.numeric(as.character(FEV1FVC_pred_T1))),
      MMEF_pred = suppressWarnings(as.numeric(as.character(MMEF_pred_T1))),
      Observed  = suppressWarnings(as.numeric(as.character(Observed))),
      Shannon   = suppressWarnings(as.numeric(as.character(Shannon))),
      Simpson   = suppressWarnings(as.numeric(as.character(Simpson))),
      Gender = factor(as.character(gender_T0), levels=c("Male","Female")),
      Smoking_status = factor(tolower(as.character(smoke_status_T0)),
                              levels=c("never","former","current"), ordered=TRUE),
      season_microbiome_sampling = factor(as.character(season_microbiome_sampling),
                              levels=c("Winter","Spring","Summer","Autumn"), ordered=TRUE)
    )
  
  df <- df %>%
    dplyr::mutate(
      Observed_z = as.numeric(scale(Observed)),
      Shannon_z  = as.numeric(scale(Shannon)),
      Simpson_z  = as.numeric(scale(Simpson))
    )

  # z-scored "decline" lung vars (higher = worse)
  for (lf in lung_fun_cont) if (lf %in% names(df))
    df[[paste0(lf, "_decline_z")]] <- as.numeric(scale(-1 * df[[lf]]))

  df
}

fit_pooled_mv <- function(df, dataset_name){
  df <- preprocess_dataset(df)
  alpha_z <- c("Observed_z","Shannon_z","Simpson_z")
  lung_z  <- paste0(lung_fun_cont, "_decline_z")

  out <- list()
  for (a in alpha_z){
    for (lz in lung_z){
      need <- c(a, lz, core_covariates)
      if (!all(need %in% names(df))) next
      dat <- df %>% dplyr::select(dplyr::all_of(need)) %>% tidyr::drop_na()
      if (nrow(dat) < 20) next

      fml <- as.formula(paste(a, "~", paste(c(lz, core_covariates), collapse = " + ")))
      fit <- tryCatch(stats::lm(fml, data = dat), error=function(e) NULL)
      if (is.null(fit)) next
      row <- broom::tidy(fit) %>% dplyr::filter(term == lz)
      if (!nrow(row)) next

      out[[length(out)+1]] <- tibble::tibble(
        dataset = dataset_name,
        niche = dplyr::recode(dataset_name, T0_NP="NP", T0_OP="OP"),
        alpha_metric = a,
        lung_var = lz,
        estimate = row$estimate, std.error = row$std.error, p.value = row$p.value,
        n_used = stats::nobs(fit)
      )
    }
  }
  dplyr::bind_rows(out)
}


mv_all <- dplyr::bind_rows(
  fit_pooled_mv(dataset_list$T0_NP, "T0_NP"),
  fit_pooled_mv(dataset_list$T0_OP, "T0_OP")
) %>%
  dplyr::mutate(
    alpha_metric = dplyr::recode(alpha_metric, !!!pretty_alpha),
    lung_var     = dplyr::recode(lung_var, !!!pretty_lung)
  ) %>%
  dplyr::group_by(dataset) %>%
  dplyr::mutate(q.value = p.adjust(p.value, method="BH"),
                stars   = dplyr::case_when(q.value<=0.05~"***", q.value<=0.10~"**", q.value<=0.25~"*", TRUE~"")) %>%
  dplyr::ungroup()

# Plot heatmap
pal <- try(rcartocolor::carto_pal(7, "Geyser"), silent=TRUE)
if (inherits(pal, "try-error")) pal <- c("#2166ac","#67a9cf","#d1e5f0","#f7f7f7","#fddbc7","#ef8a62","#b2182b")
lim <- max(abs(mv_all$estimate), na.rm=TRUE); if (!is.finite(lim)) lim <- 1

plot_df <- mv_all %>%
  dplyr::mutate(
    lung_var = factor(lung_var, levels = rev(c("FEV1 % predicted (T1)",
                                               "FVC % predicted (T1)",
                                               "FEV1/FVC % predicted (T1)",
                                               "MMEF % predicted (T1)",
                                               "Delta FEV1 % predicted (T1-T0)",
                                               "Delta FVC % predicted (T1-T0)",
                                               "Delta FEV1FVC % predicted (T1-T0)",
                                               "Delta MMEF % predicted (T1-T0)"))),
    alpha_metric = factor(alpha_metric, levels=c("Observed richness","Shannon diversity","Simpson diversity")),
    niche = factor(dplyr::recode(dataset, T0_NP="NP", T0_OP="OP"), levels=c("NP","OP"))
  )

p_mv <- ggplot2::ggplot(plot_df, aes(x = alpha_metric, y = lung_var, fill = estimate)) +
  ggplot2::geom_tile(color="white") +
  ggplot2::geom_text(aes(label = stars), size=4) +
  ggplot2::scale_fill_gradient2(low=pal[1], mid=pal[4], high=pal[7], midpoint=0, limits=c(-lim,lim),
                                name="Std. β (per 1 SD worse)") +
  ggplot2::facet_wrap(~ niche, nrow=1) +
  ggplot2::theme_minimal(base_size=12) +
  ggplot2::theme(panel.grid=element_blank(), axis.text.x=element_text(angle=30,hjust=1),
                 strip.text=element_text(face="bold")) +
  ggplot2::labs(x="Alpha diversity metric", y="Lung function metric (decline)",
                title="Multivariable models (pooled; adjusted for Age, Gender, Smoking, Season)")

ggplot2::ggsave("../Output/Longitudinal/Alpha_diversity/Multivariable/Non_stratified/Multivariable_alpha_heatmap.svg", p_mv, width=7.0, height=4)

readr::write_csv(mv_all,"../Output/Longitudinal/Alpha_diversity/Multivariable/Non_stratified/Multivariable_alpha_results.csv")

```

# Beta diversity 
### Multivariable PERMANOVA
```{r}
lung_outcomes <- c("FEV1_pred_T1","FVC_pred_T1","FEV1FVC_pred_T1","MMEF_pred_T1", "delta_FEV1_pred","delta_FVC_pred" , "delta_FEV1FVC_pred", "delta_MMEF_pred")
covars_adj    <- c("age_T0","gender_T0","smoke_status_T0","season_microbiome_sampling")
perms         <- 999
set.seed(202501)

# Helper: samples x taxa matrix
otu_samples_x_taxa <- function(ps) {
  M <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) M <- t(M)
  mode(M) <- "numeric"
  M
}

# Main runner for one niche
run_permanova_niche <- function(ps_rq1, niche_keep, outcomes, covars, perms = 999) {

  sd0 <- as(sample_data(ps_rq1), "data.frame")
  keep_samp <- as.character(sd0$niche) == niche_keep

  ps_n <- prune_samples(keep_samp, ps_rq1)
  ps_n <- prune_taxa(taxa_sums(ps_n) > 0, ps_n)

  # relative abundance
  ps_ra <- transform_sample_counts(ps_n, function(x) if (sum(x) == 0) x else x / sum(x))

  # samples x taxa
  otu <- as(otu_table(ps_ra), "matrix")
  if (taxa_are_rows(ps_ra)) otu <- t(otu)
  mode(otu) <- "numeric"

  meta <- as(sample_data(ps_ra), "data.frame")

  bray <- vegan::vegdist(otu, method = "bray")

  out <- lapply(outcomes, function(y) {

    if (!(y %in% names(meta))) return(NULL)

    need <- c(y, covars)
    need <- need[need %in% names(meta)]

    df <- meta %>%
      dplyr::select(dplyr::all_of(need)) %>%
      dplyr::mutate(
        y_num = suppressWarnings(as.numeric(as.character(.data[[y]]))),
        y_z   = as.numeric(scale(y_num))
      ) %>%
      tidyr::drop_na()

    ids <- intersect(rownames(df), attr(bray, "Labels"))
    if (length(ids) < 20) return(NULL)

    df <- df[ids, , drop = FALSE]
    br <- as.dist(as.matrix(bray)[ids, ids, drop = FALSE])

    cov_use <- covars[covars %in% names(df)]
    cov_use <- cov_use[vapply(cov_use, function(v) {
      x <- df[[v]]
      x <- x[!is.na(x)]
      if (!length(x)) return(FALSE)
      if (is.numeric(x)) return(stats::sd(x) > 0)
      length(unique(x)) > 1
    }, logical(1))]

    fml <- as.formula(paste(
      "br ~ y_z",
      if (length(cov_use)) paste("+", paste(cov_use, collapse = " + ")) else ""
    ))

    fit <- vegan::adonis2(fml, data = df, permutations = perms, by = "margin")

    tibble::tibble(
      niche   = niche_keep,
      outcome = y,
      n_used  = length(ids),
      R2      = fit["y_z", "R2"],
      F       = fit["y_z", "F"],
      p       = fit["y_z", "Pr(>F)"]
    )
  })

  res <- dplyr::bind_rows(out)
  if (nrow(res)) res <- res %>% dplyr::mutate(q = p.adjust(p, method = "BH"))
  res
}



# Run PERMANOVA

ps_rq1 <- readRDS("../Output/Longitudinal/Phyloseq/ps_RQ1_T0micro_LFboth.rds")

res_np <- run_permanova_niche(ps_rq1, "NP", lung_outcomes, covars_adj, perms)
res_op <- run_permanova_niche(ps_rq1, "OP", lung_outcomes, covars_adj, perms)

res_all <- bind_rows(res_np, res_op) %>%
  mutate(
    group = ifelse(grepl("^delta_", outcome), "delta", "T1")
  ) %>%
  group_by(niche, group) %>%
  mutate(q = p.adjust(p, method = "BH")) %>%
  ungroup()


res_all
write.csv(res_all, "../Output/Longitudinal/Beta_diversity/RQ1_PERMANOVA_T0composition_vs_T1lung_byNiche.csv",
          row.names = FALSE)


# Plot results in bar plot
pretty_outcomes <- c(
  "FEV1_pred_T1"    = "FEV1 % predicted",
  "FVC_pred_T1"     = "FVC % predicted",
  "FEV1FVC_pred_T1" = "FEV1/FVC % predicted",
  "MMEF_pred_T1"    = "MMEF % predicted",
  "delta_FEV1_pred" = "Delta FEV1 % predicted (T1-T0)",
  "delta_FVC_pred"  = "Delta FVC % predicted (T1-T0)",
  "delta_FEV1FVC_pred" = "Delta FEV1FVC % predicted (T1-T0)",
  "delta_MMEF_pred" = "Delta MMEF % predicted (T1-T0)"
  )

desired_outcome_order <- c(
  "FEV1 % predicted","FVC % predicted","FEV1/FVC % predicted","MMEF % predicted", "Delta FEV1 % predicted (T1-T0)", "Delta FVC % predicted (T1-T0)", "Delta FEV1FVC % predicted (T1-T0)","Delta MMEF % predicted (T1-T0)"
)

q_lung_sig_thresh <- 0.25
nice_ceiling <- function(x, step = 0.0025) step * ceiling(x / step)

plot_df <- res_all %>%
  mutate(
    outcome_lab = recode(outcome, !!!pretty_outcomes, .default = outcome),
    outcome_f   = factor(outcome_lab, levels = rev(desired_outcome_order)),
    niche_f     = factor(niche, levels = c("NP","OP")),
    fill_q      = factor(if_else(!is.na(q) & q <= q_lung_sig_thresh, "q \u2264 0.25", "ns"),
                         levels = c("ns","q \u2264 0.25")),
    stars_q     = case_when(
      is.na(q)   ~ "",
      q <= 0.05  ~ "***",
      q <= 0.10  ~ "**",
      q <= 0.25  ~ "*",
      TRUE       ~ ""
    ),
    label_inside = stars_q
  )

# x-limit
xlim_all <- nice_ceiling(max(plot_df$R2, na.rm = TRUE) * 1.05)
if (!is.finite(xlim_all) || xlim_all <= 0) xlim_all <- 0.01

# dummy df to force a "Significance" legend
stars_legend_df <- tibble(
  stars_level = factor(c("*","**","***"), levels = c("*","**","***")),
  x = 0, y = 0
)

# --- Make the plot ---
p <- ggplot(plot_df, aes(x = R2 * 100, y = outcome_f, fill = fill_q)) +
  geom_col(width = 0.7) +
  geom_text(
    data = filter(plot_df, label_inside != ""),
    aes(x = pmax((R2 * 100) / 2, 0.1), y = outcome_f, label = label_inside),
    inherit.aes = FALSE,
    color = "white", fontface = "bold", size = 3.2, vjust = 0.6
  ) +
  facet_grid(cols = vars(niche_f), scales = "free_y", space = "free_y", drop = FALSE) +
  scale_x_continuous(
    limits = c(0, xlim_all * 100),
    breaks = pretty_breaks(n = 5),
    labels = label_number(accuracy = 0.1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_fill_manual(
    values = c("ns" = "grey30", "q \u2264 0.25" = "#F28E2B"),
    labels = c("Non significant", "Significant (q \u2264 0.25)"),
    name   = NULL
  ) +
  labs(
    x = "Explained variance (% R\u00B2)", y = NULL,
    title = "RQ1: T0 microbiome composition vs T1 lung function (PERMANOVA; adjusted)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    strip.text         = element_text(face = "bold"),
    axis.text.y        = element_text(size = 9),
    plot.title         = element_text(face = "bold", hjust = 0),
    plot.margin        = margin(t = 4, r = 8, b = 4, l = 4)
  ) +
  # legend hack (optional)
  geom_point(
    data = stars_legend_df,
    aes(x = x, y = y, shape = stars_level),
    inherit.aes = FALSE, alpha = 0, size = 0, show.legend = TRUE
  ) +
  scale_shape_manual(
    values = c("*" = 16, "**" = 16, "***" = 16),
    labels = c("*" = "* q \u2264 0.25","**" = "** q \u2264 0.10","***" = "*** q \u2264 0.05"),
    name   = "Significance"
  ) +
  guides(fill = guide_legend(order = 1), shape = guide_legend(order = 2)) +
  coord_cartesian(clip = "off")

ggsave("../Output/Longitudinal/Beta_diversity/T0composition_vs_T1lung_barplot.svg",p, width =7, height = 4)




```
## PCoAs
```{r}
# helper
`%||%` <- function(a, b) if (!is.null(a)) a else b

make_pretty_pcoa_cat <- function(ps, group_var,
                                 covars = NULL,
                                 group_levels = NULL,
                                 pretty_map = NULL,
                                 colors = NULL,   # <-- pass in call
                                 shapes = NULL,   # <-- optional
                                 title = "PCoA — Bray–Curtis",
                                 legend_title = NULL,
                                 perms = 999, seed = 202501) {

  stopifnot(inherits(ps, "phyloseq"))

  meta <- as(sample_data(ps), "data.frame")
  meta$sample_id <- rownames(meta)

  if (!group_var %in% names(meta)) stop(sprintf("'%s' not in metadata.", group_var))

  # coerce grouping var
  meta[[group_var]] <- as.character(meta[[group_var]])
  if (!is.null(group_levels)) meta[[group_var]][!meta[[group_var]] %in% group_levels] <- NA
  meta[[group_var]] <- factor(meta[[group_var]],
                              levels = group_levels %||% sort(unique(meta[[group_var]])))

  # complete cases for group (+ covars if adjusted)
  needed <- c(group_var, covars)
  meta_cc <- meta %>% tidyr::drop_na(dplyr::all_of(needed))

  ps_cc <- prune_samples(meta_cc$sample_id, ps)
  if (nsamples(ps_cc) < 3) stop("Too few samples after pruning.")

  # distance + align
  dist_bc <- phyloseq::distance(ps_cc, method = "bray")
  ids <- intersect(attr(dist_bc, "Labels"), rownames(meta_cc))
  dist_bc <- as.dist(as.matrix(dist_bc)[ids, ids, drop = FALSE])
  meta_cc <- meta_cc[ids, , drop = FALSE]

  # drop singleton levels
  tab <- table(meta_cc[[group_var]])
  keep_lv <- names(tab)[tab >= 2]
  meta_cc <- meta_cc[meta_cc[[group_var]] %in% keep_lv, , drop = FALSE]
  meta_cc[[group_var]] <- droplevels(meta_cc[[group_var]])
  dist_bc <- as.dist(as.matrix(dist_bc)[rownames(meta_cc), rownames(meta_cc), drop = FALSE])

  if (nlevels(meta_cc[[group_var]]) < 2) stop("Need at least 2 groups.")

  # PERMANOVA
  form <- reformulate(c(group_var, covars), response = "dist_bc")
  set.seed(seed)
  fit <- vegan::adonis2(form, data = meta_cc, permutations = perms, by = "margin")
  aov_tbl <- as.data.frame(fit); aov_tbl$term <- rownames(aov_tbl)

  row_term <- aov_tbl[aov_tbl$term == group_var, , drop = FALSE]
  R2_val <- if (nrow(row_term)) row_term$R2[1] else NA_real_
  p_val  <- if (nrow(row_term)) row_term$`Pr(>F)`[1] else NA_real_

  sig <- dplyr::case_when(
    is.na(p_val)  ~ "",
    p_val < 0.001 ~ "***",
    p_val < 0.01  ~ "**",
    p_val < 0.05  ~ "*",
    TRUE          ~ ""
  )

  lab_str <- sprintf("PERMANOVA%s\n%s: R² = %.3f, p = %.3g %s",
                     ifelse(is.null(covars), " (univariable)", " (adjusted)"),
                     ifelse(is.null(legend_title), group_var, legend_title),
                     R2_val, p_val, sig)

  # PCoA
  pco <- ape::pcoa(dist_bc)
  ax1_var <- round(100 * pco$values$Relative_eig[1], 1)
  ax2_var <- round(100 * pco$values$Relative_eig[2], 1)

  scr <- as.data.frame(pco$vectors[, 1:2, drop = FALSE])
  colnames(scr) <- c("PCoA1","PCoA2")
  scr$sample_id <- rownames(scr)

  df <- dplyr::left_join(scr, meta_cc, by = "sample_id")

  surv_lv <- levels(df[[group_var]])
  if (is.null(pretty_map)) pretty_map <- setNames(surv_lv, surv_lv)

  df$group_plot <- factor(as.character(df[[group_var]]),
                          levels = surv_lv,
                          labels = unname(pretty_map[surv_lv]))

  # colors: use what you pass; otherwise default ggplot hues
  if (is.null(colors)) {
    colors <- setNames(scales::hue_pal()(length(surv_lv)), surv_lv)
  } else {
    # allow unnamed vector; assume order matches levels
    if (is.null(names(colors))) colors <- setNames(colors, surv_lv)
  }

  # shapes: binary gets circle/triangle; multi-level all circles unless you pass shapes
  if (is.null(shapes)) {
    if (length(surv_lv) == 2) shapes <- setNames(c(16,17), surv_lv)
    else shapes <- setNames(rep(16, length(surv_lv)), surv_lv)
  } else {
    if (is.null(names(shapes))) shapes <- setNames(shapes, surv_lv)
  }

  # ellipses only if >=3 per group
  cnt <- table(df$group_plot)
  can_ell <- length(cnt) > 1 && all(cnt >= 3)

  p <- ggplot(df, aes(PCoA1, PCoA2,
                      color = .data[[group_var]],
                      shape = .data[[group_var]])) +
    geom_point(size = 3, alpha = 0.7) +
    scale_color_manual(values = colors, aesthetics = "colour") +
    scale_fill_manual(values  = colors, aesthetics = "fill") +
    scale_shape_manual(values = shapes) +
    { if (can_ell)
      stat_ellipse(aes(fill = .data[[group_var]], color = .data[[group_var]]),
                   type = "t", level = 0.95, size = 1,
                   geom = "polygon", alpha = 0.1,
                   show.legend = FALSE)
    } +
    labs(
      title = title,
      x = paste0("PCoA1 (", ax1_var, "%)"),
      y = paste0("PCoA2 (", ax2_var, "%)"),
      color = ifelse(is.null(legend_title), group_var, legend_title)
    ) +
    annotate("text", x = Inf, y = -Inf, label = lab_str,
             hjust = 1.05, vjust = -0.2, size = 4, color = "black") +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
      legend.position = "right", 
      axis.title = element_text(size = 16, face = "bold"),
axis.text  = element_text(size = 13)
    ) +
    guides(
      fill  = "none",
      shape = "none",
      color = guide_legend(override.aes = list(
        shape = unname(shapes[surv_lv]),
        linetype = 0, fill = NA, alpha = 1, size = 3
      ))
    )

  attr(p, "permanova") <- fit
  return(p)
}

out_dir <- "../Output/Longitudinal/Beta_diversity/Ordinations"

# Niche - NP vs OP
OPvsNP_PCoA <- make_pretty_pcoa_cat(
  ps           = ps_rq1,
  group_var    = "niche",
  colors       = c("#7987AF","#7c674d"), 
  title        = "PCoA — Niche (VGO3 NP vs OP)",
  legend_title = "Niche (OP vs NP)"
)

ggsave(file.path(out_dir, "Niche_OPvsNP_PCoA2.svg"), OPvsNP_PCoA, width=7, height=5)

```


# Differential abundance analysis
## MaAsLin2
```{r}
# Differential abundance analysis – MaAsLin2 (Longitudinal)
# Outcomes: T1 lung function + Delta lung function (all treated as declines)
# FDR correction is applied separately per outcome group (T1 vs Delta)
# Results are saved as .csv per outcome × dataset × tax level

# Load & prepare data
# These ps objects are already filtered and converted to RA
T0_NP_filtered
T0_OP_filtered

# Quick counts 
length(unique(as.data.frame(tax_table(T0_NP_filtered))$ASV)) # 477
length(unique(as.data.frame(tax_table(T0_NP_filtered))$genus)) # 164
length(unique(as.data.frame(tax_table(T0_OP_filtered))$ASV)) # 215
length(unique(as.data.frame(tax_table(T0_OP_filtered))$genus)) # 40

## Convert ASV RA -> "absolute" abundance via qPCR 
#### T0 NP ####
# Ensure sample_id is present in the sample_data
sample_data(T0_NP_filtered) <- sample_data(T0_NP_filtered) %>%
  data.frame() %>%
  dplyr::rename(sample_id = sample_ID)

# Convert OTU and metadata to data frames
T0_np_otu  <- otu_tab_to_df(T0_NP_filtered)  # has 'sample_id'
T0_np_meta <- meta_to_df(T0_NP_filtered, sample_name = "sample_id_row")  # rownames column is now 'sample_id_row'

# Merge and multiply by qPCR
T0_np_otu <- T0_np_otu %>%
  dplyr::left_join(T0_np_meta[, c("sample_id_row", "qPCR")], by = c("sample_id" = "sample_id_row")) %>%
  dplyr::mutate(dplyr::across(-c(sample_id, qPCR), ~ .x * qPCR)) %>%
  dplyr::select(-qPCR) %>%
  as.data.frame()

# Convert to wide OTU table format for phyloseq
rownames(T0_np_otu) <- T0_np_otu$sample_id
T0_np_otu <- T0_np_otu[, -1] %>% t() %>% as.data.frame()

# Rebuild phyloseq object with absolute abundances
T0_np_ps_abs <- phyloseq(
  otu_table(T0_np_otu, taxa_are_rows = TRUE),
  tax_table(T0_NP_filtered),
  sample_data(T0_NP_filtered)
)


#### T0 OP ####
# Ensure sample_id is present in the sample_data
sample_data(T0_OP_filtered) <- sample_data(T0_OP_filtered) %>%
  data.frame() %>%
  dplyr::rename(sample_id = sample_ID)

# Convert OTU and metadata to data frames
T0_op_otu  <- otu_tab_to_df(T0_OP_filtered)  # has 'sample_id'
T0_op_meta <- meta_to_df(T0_OP_filtered, sample_name = "sample_id_row")  # rownames column is now 'sample_id_row'

# Merge and multiply by qPCR
T0_op_otu <- T0_op_otu %>%
  dplyr::left_join(T0_op_meta[, c("sample_id_row", "qPCR")], by = c("sample_id" = "sample_id_row")) %>%
  dplyr::mutate(dplyr::across(-c(sample_id, qPCR), ~ .x * qPCR)) %>%
  dplyr::select(-qPCR) %>%
  as.data.frame()

# Convert to wide OTU table format for phyloseq
rownames(T0_op_otu) <- T0_op_otu$sample_id
T0_op_otu <- T0_op_otu[, -1] %>% t() %>% as.data.frame()

# Rebuild phyloseq object with absolute abundances
T0_op_ps_abs <- phyloseq(
  otu_table(T0_op_otu, taxa_are_rows = TRUE),
  tax_table(T0_OP_filtered),
  sample_data(T0_OP_filtered)
)


#### Agglomerate absolute phyloseqs to Genus level #### 
set_genus_names <- function(ps) {
  tt <- as(phyloseq::tax_table(ps), "matrix")
  if ("genus" %in% colnames(tt)) {
    g <- as.character(tt[, "genus"])
    g[is.na(g) | g == ""] <- paste0("Unclassified_genus_", seq_len(sum(is.na(g) | g == "")))
    if (any(duplicated(g))) {
      dup <- duplicated(g) | duplicated(g, fromLast = TRUE)
      g[dup] <- paste0(g[dup], "_", ave(seq_along(g[dup]), g[dup], FUN = seq_along))
    }
    taxa_names(ps) <- g
  }
  ps
}

T0_np_ps_genus_abs <- T0_np_ps_abs %>% tax_glom(taxrank = "genus") %>% set_genus_names()
T0_op_ps_genus_abs <- T0_op_ps_abs %>% tax_glom(taxrank = "genus") %>% set_genus_names()



#### Additional filtering ####
rel_thresh <- 0.005
prev_frac  <- 0.10

# Helper: apply RA-based prevalence filter, return ABS pruned
filter_abs_by_RA_rule <- function(ps_abs, rel_thresh, prev_frac) {
  ps_ra <- phyloseq::transform_sample_counts(ps_abs, function(x) { s <- sum(x); if (s == 0) x else x/s })
  ns <- phyloseq::nsamples(ps_ra)
  need <- ceiling(ns * prev_frac)
  otu <- as(otu_table(ps_ra), "matrix")
  if (!taxa_are_rows(ps_ra)) otu <- t(otu)
  keep_taxa <- rownames(otu)[rowSums(otu >= rel_thresh) >= need]
  phyloseq::prune_taxa(keep_taxa, ps_abs)
}

#### ASV level ####
T0_np_ps_abs_filt <- filter_abs_by_RA_rule(T0_np_ps_abs, rel_thresh, prev_frac)
T0_op_ps_abs_filt <- filter_abs_by_RA_rule(T0_op_ps_abs, rel_thresh, prev_frac)

#### Genus level ####
T0_np_ps_genus_abs_filt <- filter_abs_by_RA_rule(T0_np_ps_genus_abs, rel_thresh, prev_frac)
T0_op_ps_genus_abs_filt <- filter_abs_by_RA_rule(T0_op_ps_genus_abs, rel_thresh, prev_frac)

# Per-niche lists (names are *niche only*, so folders will be <out_root>/<niche>/{ASV,Genus})
ps_list_DA_ASV <- list(
  T0_NP = T0_np_ps_abs_filt,
  T0_OP = T0_op_ps_abs_filt
)
ps_list_DA_genus <- list(
  T0_NP = T0_np_ps_genus_abs_filt,
  T0_OP = T0_op_ps_genus_abs_filt
)


#### MaAsLin function ####
# Define outcome variables (treated as decline)
outcomes_decline <- c(
  "FEV1_pred_T1", "FVC_pred_T1", "FEV1FVC_pred_T1", "MMEF_pred_T1",
  "delta_FEV1_pred", "delta_FVC_pred", "delta_FEV1FVC_pred", "delta_MMEF_pred"
)

# Define covariates
covariates <- c("age_T0", "gender_T0", "smoke_status_T0", "season_microbiome_sampling")

# Safe file-naming helper
safe_name <- function(x) gsub("[^A-Za-z0-9._-]+", "_", x)

# Output directory
out_root <- "../Output/Longitudinal/DA/MaAsLin2"

# MaAsLin2 wrapper (non-stratified)
run_maaslin2_longitudinal <- function(ps, outcome_var, covariates, out_dir) {
  stopifnot(inherits(ps, "phyloseq"))
  
  meta <- data.frame(sample_data(ps), check.names = FALSE)
  
  if (!outcome_var %in% colnames(meta)) return(NULL)
  
  # Convert to decline direction
  meta[[paste0(outcome_var, "_decline")]] <- -1 * suppressWarnings(as.numeric(meta[[outcome_var]]))
  outcome_use <- paste0(outcome_var, "_decline")

  sample_data(ps) <- sample_data(meta)

  # Prepare OTU table
  otu <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) otu <- t(otu)

  otu_df <- as.data.frame(otu, check.names = FALSE)
  meta_df <- data.frame(sample_data(ps), check.names = FALSE)

  common <- intersect(rownames(meta_df), rownames(otu_df))
  otu_df  <- otu_df[common, , drop = FALSE]
  meta_df <- meta_df[common, , drop = FALSE]

  if (!identical(rownames(otu_df), rownames(meta_df))) return(NULL)

  # Run MaAsLin2
  fit <- Maaslin2(
    input_data     = otu_df,
    input_metadata = meta_df,
    output         = out_dir,
    fixed_effects  = c(outcome_use, covariates),
    normalization  = "NONE",
    transform      = "LOG",
    min_abundance  = 0,
    min_prevalence = 0,
    plot_heatmap   = FALSE,
    plot_scatter   = FALSE
  )

  # Post-process result
  res <- fit$results
  if (is.null(res) || !nrow(res)) return(NULL)
  
  res <- res[res$metadata == outcome_use, ]
  res$outcome <- outcome_var
  res$conf.low  <- res$coef - 1.96 * res$stderr
  res$conf.high <- res$coef + 1.96 * res$stderr
  res$label <- dplyr::case_when(
    is.na(res$qval)        ~ NA_character_,
    res$qval <= 0.05       ~ "***",
    res$qval <= 0.10       ~ "**",
    res$qval <= 0.25       ~ "*",
    TRUE                   ~ NA_character_
  )

  return(res)
}

# Batch runner
run_all_maaslin_models <- function(ps_lists, outcomes, covars, out_root) {
  results_all <- list()
  
  for (level in names(ps_lists)) {
    for (niche in names(ps_lists[[level]])) {
      ps <- ps_lists[[level]][[niche]]
      for (v in outcomes) {
        subdir <- file.path(out_root, niche, level)
        dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
        
        out_dir <- file.path(subdir, safe_name(v))
        dir.create(out_dir, showWarnings = FALSE)
        
        cat("Running:", niche, "|", level, "|", v, "\n")
        res <- try(run_maaslin2_longitudinal(ps, v, covars, out_dir), silent = TRUE)
        
        if (inherits(res, "try-error") || is.null(res) || !nrow(res)) {
          message("No results for:", niche, level, v)
          next
        }
        
        res$niche <- niche
        res$level <- level
        results_all[[paste(niche, level, v, sep = "_")]] <- res

        # Save individual file
        fn <- paste0(niche, "_", level, "_", safe_name(v), ".csv")
        readr::write_csv(res, file.path(subdir, fn))
      }
    }
  }

  # Save full summary
  res_df <- dplyr::bind_rows(results_all)
  readr::write_csv(res_df, file.path(out_root, "MaAsLin2_all_results_summary.csv"))
  return(res_df)
}

# Run all (using your prepared filtered objects)
ps_lists_longitudinal <- list(
  ASV   = ps_list_DA_ASV,
  Genus = ps_list_DA_genus
)

maaslin_results_longitudinal <- run_all_maaslin_models(
  ps_lists = ps_lists_longitudinal,
  outcomes = outcomes_decline,
  covars   = covariates,
  out_root = out_root
)




```
#### MaAsLin2 heatmaps 
```{r}
# Settings
root_dir <- "../Output/Longitudinal/DA/MaAsLin2/"

# ASV folders
asv_folders <- list(
  NP = file.path(root_dir, "T0_NP", "ASV"),
  OP = file.path(root_dir, "T0_OP", "ASV")
)

# Genus folders (datasets end with _genus)
genus_folders <- list(
  NP = file.path(root_dir, "T0_NP", "Genus"),
  OP = file.path(root_dir, "T0_OP", "Genus")
)

# Order on the x-axis (raw metadata names in CSV)
custom_metadata_order <- c(
  "FEV1_pred_T1_decline","FVC_pred_T1_decline", "FEV1FVC_pred_T1_decline",  "MMEF_pred_T1_decline", "delta_FEV1_pred_decline", "delta_FVC_pred_decline", "delta_FEV1FVC_pred_decline", "delta_MMEF_pred_decline"
)

# Pretty names to display on x-axis (same order)
pretty_labels <- c(
  "FEV1",
  "FVC",
  "FEV1/FVC",
  "MMEF",
  "Delta FEV1",
  "Delta FVC",
  "Delta FEV1/FVC",
  "Delta MMEF"
)

names(pretty_labels) <- custom_metadata_order

# Helper functions
# Read + tidy one dataset/level folder
read_heatmap_data <- function(folder) {
  files <- list.files(folder, pattern = "\\.csv$", full.names = TRUE)
  if (!length(files)) return(NULL)

  df <- lapply(files, function(f) {
    tryCatch(readr::read_csv(f, show_col_types = FALSE), error = function(e) NULL)
  }) |> bind_rows()

  if (is.null(df) || !nrow(df)) return(NULL)

  # Keep required columns; tolerate missing 'label'
  if (!"label" %in% names(df)) df$label <- NA_character_
  df <- df |> select(feature, metadata, coef, label)

  # x-axis handling: keep raw + pretty factor
  df$metadata_raw <- factor(df$metadata, levels = custom_metadata_order)
  # map to pretty labels with same order
  df$metadata_plot <- factor(pretty_labels[as.character(df$metadata_raw)],
                             levels = pretty_labels)

  df
}


#  Y-axis label formatter:
#    - replace underscores with spaces
#    - italicize the text part
#    - put trailing numbers at the end in parentheses, non-italic
format_feature_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    if (grepl(" [0-9]+$", lab)) {
      text <- sub("([0-9]+)$", "", lab)   # remove trailing number
      num  <- sub(".* ([0-9]+)$", "\\1", lab)  # extract trailing number
      bquote(italic(.(text)) ~ "(" * .(num) * ")")
    } else {
      bquote(italic(.(lab)))
    }
  })
}


# Generic heatmap builder (works for both ASV/Genus)
make_heatmap <- function(df, title, italicize_features = TRUE) {
  if (is.null(df) || !nrow(df)) {
    return(ggplot() + theme_void() + ggtitle(paste0(title, " (no data)")))
  }

  p <- ggplot(df, aes(x = metadata_plot, y = feature, fill = coef)) +
    geom_tile(color = "white") +
    # EXACT palette requested:
    scale_fill_gradient2(
      low = rcartocolor::carto_pal(7, "Geyser")[1],
      mid = rcartocolor::carto_pal(7, "Geyser")[4],
      high = rcartocolor::carto_pal(7, "Geyser")[7],
      midpoint = 0,
      name = "Coefficient"
    ) +
    # centered significance stars
    geom_text(aes(label = label), color = "black", size = 5,
              hjust = 0.5, vjust = 0.8, na.rm = TRUE) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      axis.text.y  = element_text(),   # base; we format with scale_y_discrete
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5),
      plot.margin  = unit(c(6, 6, 6, 12), "pt")
    ) +
    ggtitle(title)

  if (italicize_features) {
    p + scale_y_discrete(labels = format_feature_labels)
  } else {
    # If you ever want Genus plain: change italicize_features=FALSE
    # Here we still italicize + remove underscores (same formatter)
    p + scale_y_discrete(labels = format_feature_labels)
  }
}

# Formatter for ASV labels: italics + trailing numbers in parentheses
format_asv_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    if (grepl(" [0-9]+$", lab)) {
      text <- sub("([0-9]+)$", "", lab)   # remove trailing number
      num  <- sub(".* ([0-9]+)$", "\\1", lab)  # extract trailing number
      bquote(italic(.(text)) ~ "(" * .(num) * ")")
    } else {
      bquote(italic(.(lab)))
    }
  })
}

# Formatter for genus labels: italics only, no special number handling
format_genus_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)  # underscores -> spaces
    bquote(italic(.(lab)))
  })
}

# Generic heatmap builder (choose formatter based on level)
make_heatmap <- function(df, title, level = c("ASV","Genus")) {
  level <- match.arg(level)

  if (is.null(df) || !nrow(df)) {
    return(ggplot() + theme_void() + ggtitle(paste0(title, " (no data)")))
  }

  p <- ggplot(df, aes(x = metadata_plot, y = feature, fill = coef)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = rcartocolor::carto_pal(7, "Geyser")[1],
      mid = rcartocolor::carto_pal(7, "Geyser")[4],
      high = rcartocolor::carto_pal(7, "Geyser")[7],
      midpoint = 0,
      name = "Coefficient"
    ) +
    geom_text(aes(label = label), color = "black", size = 5,
              hjust = 0.5, vjust = 0.8, na.rm = TRUE) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1),
      axis.text.y  = element_text(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid   = element_blank(),
      plot.title   = element_text(hjust = 0.5),
      plot.margin  = unit(c(6, 6, 6, 12), "pt")
    ) +
    ggtitle(title)

  if (level == "ASV") {
    p + scale_y_discrete(labels = format_asv_labels)
  } else {
    p + scale_y_discrete(labels = format_genus_labels)
  }
}


# Build individual plots
# ASV heatmaps
p_T0_NP_ASV <- make_heatmap(read_heatmap_data(asv_folders$NP), "T0 NP (ASV)", "ASV")
p_T0_OP_ASV <- make_heatmap(read_heatmap_data(asv_folders$OP), "T0 OP (ASV)", "ASV")

# Genus heatmaps
p_T0_NP_Gen <- make_heatmap(read_heatmap_data(genus_folders$NP), "T0 NP (Genus)", "Genus")
p_T0_OP_Gen <- make_heatmap(read_heatmap_data(genus_folders$OP), "T0 OP (Genus)", "Genus")

# Combine into 2×2 grids and save
asv_grid <- (p_T0_NP_ASV | p_T0_OP_ASV) +
            patchwork::plot_annotation(title = "MaAsLin2 Heatmaps — ASV Level")

genus_grid <- (p_T0_NP_Gen | p_T0_OP_Gen) +
              patchwork::plot_annotation(title = "MaAsLin2 Heatmaps — Genus Level")

ggsave(file.path(root_dir, "Heatmap_Maaslin_results_ASV.svg"),
       asv_grid, width = 10, height = 8)

ggsave(file.path(root_dir, "Heatmap_Maaslin_results_Genus.svg"),
       genus_grid, width = 10, height = 8)
```

## ANCOMBC2 
```{r}
# Make tree summarised experiment data
T0_np_asv_tse <- mia::makeTreeSummarizedExperimentFromPhyloseq(T0_np_ps_abs_filt)
T0_op_asv_tse <- mia::makeTreeSummarizedExperimentFromPhyloseq(T0_op_ps_abs_filt)
T0_np_genus_tse <- mia::makeTreeSummarizedExperimentFromPhyloseq(T0_np_ps_genus_abs_filt)
T0_op_genus_tse <- mia::makeTreeSummarizedExperimentFromPhyloseq(T0_op_ps_genus_abs_filt)


# Order on the x-axis (raw metadata names in CSV)
custom_metadata_order <- c(
  "FEV1_pred_T1_decline","FVC_pred_T1_decline", "FEV1FVC_pred_T1_decline",  "MMEF_pred_T1_decline", "delta_FEV1_pred_decline", "delta_FVC_pred_decline", "delta_FEV1FVC_pred_decline", "delta_MMEF_pred_decline"
)

# Pretty names to display on x-axis (same order)
pretty_labels <- c(
  "FEV1 % predicted",
  "FVC % predicted",
  "FEV1/FVC % predicted",
  "MMEF % predicted",
  "Delta FEV1 % predicted (T1-T0)",
  "Delta FVC % predicted (T1-T0)",
  "Delta FEV1/FVC % predicted (T1-T0)",
  "Delta MMEF % predicted (T1-T0)"
)

outcomes_decline <- c(
  "FEV1_pred_T1", "FVC_pred_T1", "FEV1FVC_pred_T1", "MMEF_pred_T1",
  "delta_FEV1_pred", "delta_FVC_pred", "delta_FEV1FVC_pred", "delta_MMEF_pred"
)

# Define covariates
covariates <- c("age_T0", "gender_T0", "smoke_status_T0", "season_microbiome_sampling")


models <- list(
  FEV1 = "FEV1_pred_T1 + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  FVC = "FVC_pred_T1 + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  FEV1FVC = "FEV1FVC_pred_T1 + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  MMEF = "MMEF_pred_T1 + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  DeltaFEV1 = "delta_FEV1_pred + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  DeltaFVC = "delta_FVC_pred + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  DeltaFEV1FVC = "delta_FEV1FVC_pred + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling",
  DeltaMMEF = "delta_MMEF_pred + age_T0 + gender_T0 + smoke_status_T0 + season_microbiome_sampling"
)


# create *_decline variables in your metadata 
# vars to convert
decline_vars <- c("FEV1_pred_T1","FVC_pred_T1","FEV1FVC_pred_T1", "MMEF_pred_T1", "delta_FEV1_pred", "delta_FVC_pred", "delta_FEV1FVC_pred", "delta_MMEF_pred")

add_decline_vars <- function(obj, vars = decline_vars) {
  if (inherits(obj, c("TreeSummarizedExperiment","SummarizedExperiment"))) {
    cd <- SummarizedExperiment::colData(obj)
    for (v in vars) {
      if (v %in% colnames(cd)) {
        cd[[paste0(v, "_decline")]] <- -1 * suppressWarnings(as.numeric(as.character(cd[[v]])))
      } else {
        warning("Column not found in colData: ", v)
      }
    }
    SummarizedExperiment::colData(obj) <- cd
    return(obj)
  }

  # phyloseq
  if (inherits(obj, "phyloseq")) {
    md <- data.frame(phyloseq::sample_data(obj), check.names = FALSE)
    for (v in vars) {
      if (v %in% names(md)) {
        md[[paste0(v, "_decline")]] <- -1 * suppressWarnings(as.numeric(as.character(md[[v]])))
      } else {
        warning("Column not found in sample_data: ", v)
      }
    }
    phyloseq::sample_data(obj) <- phyloseq::sample_data(md)
    return(obj)
  }

  stop("Unsupported object class: ", class(obj)[1])
}

# Apply to the tse objects
T0_np_asv_tse <- add_decline_vars(T0_np_asv_tse)
T0_op_asv_tse <- add_decline_vars(T0_op_asv_tse)
T0_np_genus_tse <- add_decline_vars(T0_np_genus_tse)
T0_op_genus_tse <- add_decline_vars(T0_op_genus_tse)


fix_factors <- function(tse) {
  colData(tse) <- colData(tse) %>%
    as.data.frame() %>%
    mutate(
      gender_T0 = factor(gender_T0, levels = c("male", "female")),
      season_microbiome_sampling = factor(
        season_microbiome_sampling,
        levels = c("Winter", "Spring", "Summer", "Autumn")
      ),
      smoke_status_T0 = factor(
        smoke_status_T0,
        levels = c("non", "former", "current")
      )
    ) %>%
    S4Vectors::DataFrame()
  tse
}

T0_np_asv_tse   <- fix_factors(T0_np_asv_tse)
T0_op_asv_tse   <- fix_factors(T0_op_asv_tse)
T0_np_genus_tse <- fix_factors(T0_np_genus_tse)
T0_op_genus_tse <- fix_factors(T0_op_genus_tse)

# references
refs <- list(
  gender_0 = "male",
  season_microbiome_sampling = "Winter") # smoke_status is already an ordered factor variable (non< former < current)

# Minimal helper: relevel selected columns if they exist
relevel_tse <- function(tse, refs) {
  cd <- as.data.frame(SummarizedExperiment::colData(tse), check.names = FALSE)
  for (nm in names(refs)) {
    if (nm %in% names(cd)) cd[[nm]] <- stats::relevel(factor(cd[[nm]]), ref = refs[[nm]])
  }
  SummarizedExperiment::colData(tse) <- S4Vectors::DataFrame(cd, row.names = rownames(cd))
  tse
}

# Apply to each object
T0_np_asv_tse <- relevel_tse(T0_np_asv_tse, refs)
T0_op_asv_tse <- relevel_tse(T0_op_asv_tse, refs)
T0_np_genus_tse <- relevel_tse(T0_np_genus_tse, refs)
T0_op_genus_tse <- relevel_tse(T0_op_genus_tse, refs)


# Run ANCOM-BC2 analyses for all models
ancom_np_asv <- map(models, ~ ANCOMBC::ancombc2(
  data = T0_np_asv_tse,
  fix_formula = .x,
  rand_formula = NULL,
  p_adj_method = "BH",
  prv_cut = 0,
  lib_cut = 0,
  struc_zero = FALSE,
  alpha = 0.05,
  global = FALSE,
  pairwise = FALSE
))

# Process and save results 
ancom_np_asv <- map(ancom_np_asv, ~ mutate_if(.x$res, is.numeric, round, 2))

# Define output paths
output_dir <- "../Output/Longitudinal/DA/ANCOMBC2/NP/ASV"
file_names <- c("FEV1", "FVC", "FEV1FVC", "MMEF", "Delta_FEV1", "Delta_FVC","Delta_FEV1FVC","Delta_MMEF")

# Save results as CSV
walk2(ancom_np_asv, file_names, ~ write_csv(.x, file.path(output_dir, paste0("ANCOMBC2_NP_ASV_", .y, ".csv"))))


#### OP ASV level ####
# Run ANCOM-BC2 analyses for all models
ancom_op_asv <- map(models, ~ ANCOMBC::ancombc2(
  data = T0_op_asv_tse,
  fix_formula = .x,
  rand_formula = NULL,
  p_adj_method = "BH",
  prv_cut = 0,
  lib_cut = 0,
  struc_zero = FALSE,
  alpha = 0.05,
  global = FALSE,
  pairwise = FALSE
))

# Process and save results 
ancom_op_asv <- map(ancom_op_asv, ~ mutate_if(.x$res, is.numeric, round, 2))

# Define output paths
output_dir <- "../Output/Longitudinal/DA/ANCOMBC2/OP/ASV"
file_names <- c("FEV1", "FVC", "FEV1FVC", "MMEF", "Delta_FEV1", "Delta_FVC","Delta_FEV1FVC","Delta_MMEF")

# Save results as CSV
walk2(ancom_op_asv, file_names, ~ write_csv(.x, file.path(output_dir, paste0("ANCOMBC2_OP_ASV_", .y, ".csv"))))



#### NP Genus level ####
ancom_np_genus <- map(models, ~ ANCOMBC::ancombc2(
  data = T0_np_genus_tse,
  fix_formula = .x,
  rand_formula = NULL,
  p_adj_method = "BH",
  prv_cut = 0,
  lib_cut = 0,
  struc_zero = FALSE,
  alpha = 0.05,
  global = FALSE,
  pairwise = FALSE
))

# Process and save results 
ancom_np_genus <- map(ancom_np_genus, ~ mutate_if(.x$res, is.numeric, round, 2))

# Define output paths
output_dir <- "../Output/Longitudinal/DA/ANCOMBC2/NP/Genus"
file_names <- c("FEV1", "FVC", "FEV1FVC", "MMEF", "Delta_FEV1", "Delta_FVC","Delta_FEV1FVC","Delta_MMEF")

# Save results as CSV
walk2(ancom_np_genus, file_names, ~ write_csv(.x, file.path(output_dir, paste0("ANCOMBC2_NP_genus_", .y, ".csv"))))


#### OP Genus level ####
ancom_op_genus <- map(models, ~ ANCOMBC::ancombc2(
  data = T0_op_genus_tse,
  fix_formula = .x,
  rand_formula = NULL,
  p_adj_method = "BH",
  prv_cut = 0,
  lib_cut = 0,
  struc_zero = FALSE,
  alpha = 0.05,
  global = FALSE,
  pairwise = FALSE
))

# Process and save results 
ancom_op_genus <- map(ancom_op_genus, ~ mutate_if(.x$res, is.numeric, round, 2))

# Define output paths
output_dir <- "../Output/Longitudinal/DA/ANCOMBC2/OP/Genus"
file_names <- c("FEV1", "FVC", "FEV1FVC", "MMEF", "Delta_FEV1", "Delta_FVC","Delta_FEV1FVC","Delta_MMEF")

# Save results as CSV
walk2(ancom_op_genus, file_names, ~ write_csv(.x, file.path(output_dir, paste0("ANCOMBC2_OP_genus_", .y, ".csv"))))
```
#### ANCOM-BC2 heatmaps 
```{r}
# ANCOM-BC2 heatmaps — NP & OP combined, ASV & Genus
root_ancom <- "../Output/Longitudinal/DA/ANCOMBC2"
paths <- list(
  NP_ASV    = file.path(root_ancom, "NP", "ASV"),
  OP_ASV    = file.path(root_ancom, "OP", "ASV"),
  NP_Genus  = file.path(root_ancom, "NP", "Genus"),
  OP_Genus  = file.path(root_ancom, "OP", "Genus")
)

lung_vars <- tibble::tribble(
  ~file_suffix,        ~plot_var,        ~model_var,
  "FEV1",              "FEV1",            "FEV1_pred_T1",
  "FVC",               "FVC",             "FVC_pred_T1",
  "FEV1FVC",           "FEV1FVC",          "FEV1FVC_pred_T1",
  "MMEF",              "MMEF",             "MMEF_pred_T1",
  "Delta_FEV1",        "Delta_FEV1",       "delta_FEV1_pred",
  "Delta_FVC",         "Delta_FVC",        "delta_FVC_pred",
  "Delta_FEV1FVC",     "Delta_FEV1FVC",    "delta_FEV1FVC_pred",
  "Delta_MMEF",        "Delta_MMEF",       "delta_MMEF_pred"
)


pretty_labels <- c(
  FEV1 = "FEV1",
  FVC = "FVC",
  FEV1FVC = "FEV1/FVC",
  MMEF = "MMEF",
  Delta_FEV1 = "Delta FEV1",
  Delta_FVC = "Delta FVC",
  Delta_FEV1FVC = "Delta FEV1/FVC",
  Delta_MMEF = "Delta MMEF"
)


# Change Y axis labels of ASVs and genera
# ASV: italics + trailing number in parentheses
format_asv_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)
    if (grepl(" [0-9]+$", lab)) {
      text <- sub("([0-9]+)$", "", lab)
      num  <- sub(".* ([0-9]+)$", "\\1", lab)
      bquote(italic(.(trimws(text))) ~ "(" * .(num) * ")")
    } else {
      bquote(italic(.(lab)))
    }
  })
}

# Genus: just italics
format_genus_labels <- function(labels) {
  sapply(labels, function(lab) {
    lab <- gsub("_", " ", lab)
    bquote(italic(.(lab)))
  })
}


read_ancom_manual_NP_ASV <- function(folder, lung_vars) {

  all_out <- list()

  for (i in seq_len(nrow(lung_vars))) {

    file <- file.path(
      folder,
      paste0("ANCOMBC2_NP_ASV_", lung_vars$file_suffix[i], ".csv")
    )

    if (!file.exists(file)) {
      warning("Missing file: ", basename(file))
      next
    }

    df <- readr::read_csv(file, show_col_types = FALSE)

    lfc_col <- paste0("lfc_", lung_vars$model_var[i])
    q_col   <- paste0("q_", lung_vars$model_var[i])
    ps_col  <- paste0("passed_ss_", lung_vars$model_var[i])

    if (!all(c("taxon", lfc_col, q_col, ps_col) %in% names(df))) {
      warning("Missing columns in ", basename(file))
      next
    }

    all_out[[i]] <- tibble::tibble(
      feature   = df$taxon,
      metadata  = lung_vars$plot_var[i],
      coef      = as.numeric(df[[lfc_col]]),
      q_value   = as.numeric(df[[q_col]]),
      passed_ss = as.logical(df[[ps_col]]),
      label = dplyr::case_when(
        is.na(df[[q_col]])        ~ "",
        df[[q_col]] <= 0.05       ~ "***",
        df[[q_col]] <= 0.10       ~ "**",
        df[[q_col]] <= 0.25       ~ "*",
        TRUE                      ~ ""
      )
    )
  }

  d <- dplyr::bind_rows(all_out)

  d$metadata <- factor(d$metadata, levels = names(pretty_labels))
  d$metadata_plot <- factor(pretty_labels[d$metadata], levels = pretty_labels)

  d
}


read_ancom_manual_OP_ASV <- function(folder, lung_vars) {

  all_out <- list()

  for (i in seq_len(nrow(lung_vars))) {

    file <- file.path(
      folder,
      paste0("ANCOMBC2_OP_ASV_", lung_vars$file_suffix[i], ".csv")
    )

    if (!file.exists(file)) {
      warning("Missing file: ", basename(file))
      next
    }

    df <- readr::read_csv(file, show_col_types = FALSE)

    lfc_col <- paste0("lfc_", lung_vars$model_var[i])
    q_col   <- paste0("q_", lung_vars$model_var[i])
    ps_col  <- paste0("passed_ss_", lung_vars$model_var[i])

    if (!all(c("taxon", lfc_col, q_col, ps_col) %in% names(df))) {
      warning("Missing columns in ", basename(file))
      next
    }

    all_out[[i]] <- tibble::tibble(
      feature   = df$taxon,
      metadata  = lung_vars$plot_var[i],
      coef      = as.numeric(df[[lfc_col]]),
      q_value   = as.numeric(df[[q_col]]),
      passed_ss = as.logical(df[[ps_col]]),
      label = dplyr::case_when(
        is.na(df[[q_col]])        ~ "",
        df[[q_col]] <= 0.05       ~ "***",
        df[[q_col]] <= 0.10       ~ "**",
        df[[q_col]] <= 0.25       ~ "*",
        TRUE                      ~ ""
      )
    )
  }

  d <- dplyr::bind_rows(all_out)

  d$metadata <- factor(d$metadata, levels = names(pretty_labels))
  d$metadata_plot <- factor(pretty_labels[d$metadata], levels = pretty_labels)

  d
}

read_ancom_manual_NP_Genus <- function(folder, lung_vars) {

  all_out <- list()

  for (i in seq_len(nrow(lung_vars))) {

    file <- file.path(
      folder,
      paste0("ANCOMBC2_NP_Genus_", lung_vars$file_suffix[i], ".csv")
    )

    if (!file.exists(file)) {
      warning("Missing file: ", basename(file))
      next
    }

    df <- readr::read_csv(file, show_col_types = FALSE)

    lfc_col <- paste0("lfc_", lung_vars$model_var[i])
    q_col   <- paste0("q_", lung_vars$model_var[i])
    ps_col  <- paste0("passed_ss_", lung_vars$model_var[i])

    if (!all(c("taxon", lfc_col, q_col, ps_col) %in% names(df))) {
      warning("Missing columns in ", basename(file))
      next
    }

    all_out[[i]] <- tibble::tibble(
      feature   = df$taxon,
      metadata  = lung_vars$plot_var[i],
      coef      = as.numeric(df[[lfc_col]]),
      q_value   = as.numeric(df[[q_col]]),
      passed_ss = as.logical(df[[ps_col]]),
      label = dplyr::case_when(
        is.na(df[[q_col]])        ~ "",
        df[[q_col]] <= 0.05       ~ "***",
        df[[q_col]] <= 0.10       ~ "**",
        df[[q_col]] <= 0.25       ~ "*",
        TRUE                      ~ ""
      )
    )
  }

  d <- dplyr::bind_rows(all_out)

  d$metadata <- factor(d$metadata, levels = names(pretty_labels))
  d$metadata_plot <- factor(pretty_labels[d$metadata], levels = pretty_labels)

  d
}



read_ancom_manual_OP_Genus <- function(folder, lung_vars) {

  all_out <- list()

  for (i in seq_len(nrow(lung_vars))) {

    file <- file.path(
      folder,
      paste0("ANCOMBC2_OP_Genus_", lung_vars$file_suffix[i], ".csv")
    )

    if (!file.exists(file)) {
      warning("Missing file: ", basename(file))
      next
    }

    df <- readr::read_csv(file, show_col_types = FALSE)

    lfc_col <- paste0("lfc_", lung_vars$model_var[i])
    q_col   <- paste0("q_", lung_vars$model_var[i])
    ps_col  <- paste0("passed_ss_", lung_vars$model_var[i])

    if (!all(c("taxon", lfc_col, q_col, ps_col) %in% names(df))) {
      warning("Missing columns in ", basename(file))
      next
    }

    all_out[[i]] <- tibble::tibble(
      feature   = df$taxon,
      metadata  = lung_vars$plot_var[i],
      coef      = as.numeric(df[[lfc_col]]),
      q_value   = as.numeric(df[[q_col]]),
      passed_ss = as.logical(df[[ps_col]]),
      label = dplyr::case_when(
        is.na(df[[q_col]])        ~ "",
        df[[q_col]] <= 0.05       ~ "***",
        df[[q_col]] <= 0.10       ~ "**",
        df[[q_col]] <= 0.25       ~ "*",
        TRUE                      ~ ""
      )
    )
  }

  d <- dplyr::bind_rows(all_out)

  d$metadata <- factor(d$metadata, levels = names(pretty_labels))
  d$metadata_plot <- factor(pretty_labels[d$metadata], levels = pretty_labels)

  d
}


# Do for all datasets
np_asv <- read_ancom_manual_NP_ASV(paths$NP_ASV, lung_vars)
op_asv <- read_ancom_manual_OP_ASV(paths$OP_ASV, lung_vars)
np_genus <- read_ancom_manual_NP_Genus(paths$NP_Genus, lung_vars)
op_genus <- read_ancom_manual_OP_Genus(paths$OP_Genus, lung_vars)


make_heatmap <- function(df, title, level = c("ASV", "Genus")) {
  level <- match.arg(level)

  pal <- rcartocolor::carto_pal(7, "Geyser")

  p <- ggplot(df, aes(x = metadata_plot, y = feature, fill = coef)) +
    geom_tile(color = "white") +
    geom_text(aes(label = label,
                  color = ifelse(passed_ss, "black", "white")),
              size = 4) +
    scale_color_identity() +
    scale_fill_gradient2(
      low = pal[1], mid = pal[4], high = pal[7],
      midpoint = 0, name = "Coefficient"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5)
    ) +
    ggtitle(title)

  if (level == "ASV") {
    p + scale_y_discrete(labels = format_asv_labels)
  } else {
    p + scale_y_discrete(labels = format_genus_labels)
  }
}


p_np_asv <- make_heatmap(np_asv, "T0 NP (ASV)")
p_op_asv <- make_heatmap(op_asv, "T0 OP (ASV)")
p_np_genus <- make_heatmap(np_genus, "T0 NP (Genus)")
p_op_genus <- make_heatmap(op_genus, "T0 NP (Genus)")

# ASV level heatmaps 
p_asv_grid <- (p_np_asv | p_op_asv) +
  patchwork::plot_annotation(title = "ANCOM-BC2 Heatmaps — ASV level")

ggsave("../Output/Longitudinal/DA/ANCOMBC2/Heatmap_ANCOMBC2_ASV_combined.svg",
       p_asv_grid, width = 10, height = 8)

# Genus level heatmaps 
p_gen_grid <- (p_np_genus | p_op_genus) +
  patchwork::plot_annotation(title = "ANCOM-BC2 Heatmaps — Genus level")

ggsave("../Output/Longitudinal/DA/ANCOMBC2/Heatmap_ANCOMBC2_Genus_combined.svg",
       p_gen_grid, width = 10, height = 8)


```



# RF to predict lung function - CONTINUE 

```{r}
library(tidyverse)
library(phyloseq)
library(mikropml)
library(furrr)
library(glue)
library(here)
library(vegan)
library(caret)
library(ranger)
library(xgboost)
library(patchwork)

plan("multisession")  # parallel processing

out_dir <- "../Output/Longitudinal/Machine_learning/"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Select your baseline data (T0)
ps_np_t0 <- T0_np_ps_abs_filt  # or ps_vgo3_op_filtered, or combined

# Use only ASVs found significant in DA analysis
sig_asvs <- read.csv("../Output/Longitudinal/DA/MaAsLin2/T0_NP/ASV/T0_NP_ASV_delta_FEV1_pred.csv")$feature
ps_np_t0_sig <- prune_taxa(sig_asvs, ps_np_t0)
nrow(ps_np_t0_sig@tax_table) # 27 remaining as for DAA

# Transform to relative abundance and merge metadata
ps_rel <- transform_sample_counts(ps_np_t0_sig, function(x) x / sum(x))
otu_df <- as.data.frame(t(otu_table(ps_rel)))
meta_df <- as.data.frame(sample_data(ps_rel))

# Merge OTU + metadata
df <- cbind(meta_df, otu_df)

# Keep only samples with follow-up lung function
df <- df %>% filter(!is.na(FEV1_FVC))
ps_np_t0_sig@sam_data$rev
# Create binary obstruction variable
df <- df %>%
  mutate(Obstruction = ifelse(FEV1_FVC < 0.7, "Obstructive", "Normal")) %>%
  mutate(Obstruction = factor(Obstruction, levels = c("Normal","Obstructive")))


# Preprocess features (mikropml-style)
# Choose outcome type
# For classification:
df_ml <- df %>% select(Obstruction, all_of(sig_asvs))
df
# For regression (predict continuous FEV1/FVC):
# df_ml <- df %>% select(FEV1_FVC, all_of(sig_asvs))

# Scale and center features
preproc <- preprocess_data(df_ml, outcome_colname = names(df_ml)[1])
m_pre <- preproc$dat_transformed


# Define helper functions
run_ml_models <- function(seed, method, pre_df = m_pre, feat = FALSE) {
  mikropml::run_ml(pre_df,
                   method = method,
                   outcome_colname = names(pre_df)[1],
                   kfold = 5, cv_times = 50,
                   training_frac = 0.8,
                   hyperparameters = test_hp,
                   seed = seed,
                   find_feature_importance = feat)
}

extract_performance <- function(model_list, method_name) {
  map_dfr(seq_along(model_list),
          ~ model_list[[.x]]$performance %>%
            mutate(seed = .x, method = method_name))
}


#Train multiple models
# Logistic regression (classification)
test_hp <- list(alpha = 0, lambda = c(0.001, 0.005, 0.01, 0.05))
it_glmnet <- future_map(1:20, ~run_ml_models(.x, "glmnet"), .options = furrr_options(seed = TRUE))

# Random Forest
test_hp <- list(mtry = c(2, 4, 6))
it_rf <- future_map(1:20, ~run_ml_models(.x, "rf", feat = TRUE), .options = furrr_options(seed = TRUE))

# XGBoost
test_hp <- list(nrounds = 200, eta = 0.01, max_depth = 6, subsample = 0.7)
it_xgb <- future_map(1:20, ~run_ml_models(.x, "xgbTree"), .options = furrr_options(seed = TRUE))

# Evaluate model performance

# Combine performance results
perf_glmnet <- extract_performance(it_glmnet, "Logistic regression")
perf_rf <- extract_performance(it_rf, "Random forest")
perf_xgb <- extract_performance(it_xgb, "XGBoost")

perf_all <- bind_rows(perf_glmnet, perf_rf, perf_xgb)

# Reshape for plotting
perf_trans <- perf_all %>%
  pivot_longer(c(cv_metric_AUC, AUC), names_to = "AUC_type", values_to = "AUC_val") %>%
  mutate(AUC_type = recode(AUC_type, cv_metric_AUC = "Cross-validation", AUC = "Testing"))

# Compare mean AUC
perf_summary <- perf_trans %>%
  group_by(method, AUC_type) %>%
  summarise(mean_AUC = mean(AUC_val, na.rm = TRUE),
            sd_AUC = sd(AUC_val, na.rm = TRUE),
            .groups = "drop")

# Plot
ggplot(perf_trans, aes(AUC_type, AUC_val, color = method)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.6) +
  labs(y = "Area Under Curve (AUC)", x = "", title = "ML model performance") +
  theme_bw()


# Feature importance (for Random Forest)

feat_imp <- map_dfr(1:20, ~it_rf[[.x]]$feature_importance %>%
                      mutate(seed = .x), .id = "model")

feat_summary <- feat_imp %>%
  group_by(feat) %>%
  summarise(mean_imp = mean(perf_metric_diff, na.rm = TRUE),
            sd_imp = sd(perf_metric_diff, na.rm = TRUE)) %>%
  arrange(desc(mean_imp))

ggplot(feat_summary %>% top_n(20, mean_imp),
       aes(x = reorder(feat, mean_imp), y = mean_imp)) +
  geom_col(fill = "#4B8BBE") +
  coord_flip() +
  labs(x = "ASV", y = "Mean Importance (ΔAUC)", title = "Top predictors (Random Forest)") +
  theme_bw()


# Save models and results 

saveRDS(it_rf, file.path(out_dir, "rf_models_T0toT1.rds"))
saveRDS(it_glmnet, file.path(out_dir, "glmnet_models_T0toT1.rds"))
saveRDS(it_xgb, file.path(out_dir, "xgb_models_T0toT1.rds"))
write.csv(feat_summary, file.path(out_dir, "RF_feature_importance.csv"), row.names = FALSE)




```


