---
title: "Longitudinal lung function analysis"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
---

# Data
```{r}
ps_final <- readRDS("../Output/Phyloseq/ps4.rds")
```
# Packages 
```{r}
library(phyloseq); library(tidyverse); library(broom);library(dplyr)
library(broom)
library(tidyr)
library(stringr)
library(ggplot2)

```

```{r}
############################################################
# Longitudinal changes (VGO2 → VGO3) and simple summaries  #
# - Lung function change per person (unique VGO3_ID)       #
# - Alpha-diversity change per person × niche              #
# - Both joined back into phyloseq sample_data             #
# - Simple violin plots for quick interpretation           #
############################################################

## 1) Settings and small helpers 
# Variables expected in sample_data(ps_final)
lung_num_pred <- c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred") # continuous (%-pred)
lung_cat_vars <- c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category") # ordinal categories
alpha_vars    <- c("Observed","Shannon","Simpson")# α-diversity indices

# Map dataset -> visit label
visit_from_dataset <- function(x) ifelse(grepl("VGO2", x), "VGO2",
                                  ifelse(grepl("VGO3", x), "VGO3", NA))

# Safe mode (for collapsing categories within a visit)
mode_chr <- function(x) {
  x <- x[!is.na(x) & nzchar(trimws(as.character(x)))]
  if (length(x) == 0) return(NA_character_)
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Ordinal severity (0 = Normal; higher = worse)
severity_order <- list(
  FEV1_category      = c("Normal","Mild","Moderate","Severe"),
  FEF25_75_category  = c("Normal","Mild","Moderate","Severe"),
  FVC_category       = c("Normal","Severely reduced"),
  FEV1FVC_category   = c("Normal","Obstructive")
)
to_severity_rank <- function(var, x) {
  lv <- severity_order[[var]]
  if (is.null(lv)) return(rep(NA_real_, length(x)))
  idx <- match(tolower(trimws(as.character(x))), tolower(lv))
  ifelse(is.na(idx), NA_real_, idx - 1L)
}


## 2) Pull sample_data as a data frame (once) 
sd0 <- sample_data(ps_final) %>% data.frame(check.names = FALSE) %>%
  rownames_to_column("SampleID")

# Ensure useful columns exist
stopifnot(all(c("VGO3_ID", "dataset", "niche") %in% names(sd0)))
if (!"date_spiro" %in% names(sd0)) {
  stop("Column 'date_spiro' (spirometry date) not found in sample_data.")
}

# Add visit label and coerce date
sd0 <- sd0 %>%
  mutate(
    visit = visit_from_dataset(dataset),
    date_spiro = suppressWarnings(as.Date(date_spiro))
  )
```

# Lung function change (VGO2 to VGO3)
## Compute 
```{r}
## Change in lung function (per person)
# Keep minimal columns for lung-function change
keep_lf <- c("SampleID","VGO3_ID","visit","date_spiro", lung_num_pred, lung_cat_vars)
sd_lf <- sd0 %>% select(any_of(keep_lf)) %>% filter(visit %in% c("VGO2","VGO3"))

# Collapse to ONE row per (person × visit), averaging numeric & modal categorical across niches
lf_by_visit <- sd_lf %>%
  group_by(VGO3_ID, visit) %>%
  summarise(
    across(all_of(lung_num_pred), ~ mean(suppressWarnings(as.numeric(.x)), na.rm = TRUE)),
    across(all_of(lung_cat_vars), mode_chr),
    date_spiro = suppressWarnings(min(date_spiro, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  filter(!is.na(VGO3_ID))

# Keep people with both visits
ids_both <- lf_by_visit %>% count(VGO3_ID) %>% filter(n == 2) %>% pull(VGO3_ID)
lf_both  <- lf_by_visit %>% filter(VGO3_ID %in% ids_both)

# Wide format
lf_wide <- lf_both %>%
  tidyr::pivot_wider(names_from = visit,
                     values_from = c(all_of(lung_num_pred), all_of(lung_cat_vars), date_spiro),
                     names_sep = "_")

# Years between visits
lf_wide <- lf_wide %>%
  mutate(
    years_between = as.numeric(difftime(date_spiro_VGO3, date_spiro_VGO2, units = "days"))/365.25
  ) %>%
  filter(is.finite(years_between), years_between > 0)

# Continuous change: (VGO3 − VGO2) / years_between
for (v in lung_num_pred) {
  lf_wide[[paste0(v, "_per_year")]] <- (lf_wide[[paste0(v, "_VGO3")]] - lf_wide[[paste0(v, "_VGO2")]]) / lf_wide$years_between
}

# Categorical change: delta in severity (+ worse, − better) and labels
for (v in lung_cat_vars) {
  sev0 <- to_severity_rank(v, lf_wide[[paste0(v, "_VGO2")]])
  sev1 <- to_severity_rank(v, lf_wide[[paste0(v, "_VGO3")]])
  dnm  <- paste0(v, "_delta")
  cnm  <- paste0(v, "_change")
  lf_wide[[dnm]] <- sev1 - sev0
  lf_wide[[cnm]] <- factor(
    ifelse(is.na(lf_wide[[dnm]]), NA_character_,
           ifelse(lf_wide[[dnm]] > 0, "worse",
                  ifelse(lf_wide[[dnm]] < 0, "better", "no_change"))),
    levels = c("no_change","worse","better")
  )
}

lung_change_by_person <- lf_wide %>%
  select(
    VGO3_ID, years_between,
    tidyselect::ends_with("_per_year"),
    tidyselect::ends_with("_delta"),
    tidyselect::ends_with("_change")
  )

# Join back to EVERY sample of that person (replicates subject-level changes)
sd1 <- sd0 %>%
  left_join(lung_change_by_person, by = "VGO3_ID")
```
## Plot change
```{r}
# Lung function change per person (unique row per VGO3_ID)
# pick one row per person with the most non-missing change values
lung_change_cols <- grep("_per_year$", names(sd2), value = TRUE)
stopifnot(length(lung_change_cols) > 0)

df_unique <- sd2 %>%
  mutate(nonmiss_lf = rowSums(across(all_of(lung_change_cols), ~ !is.na(.x)))) %>%
  arrange(VGO3_ID, desc(nonmiss_lf)) %>%
  group_by(VGO3_ID) %>% slice_head(n = 1) %>% ungroup() %>%
  select(VGO3_ID, all_of(lung_change_cols))

lung_long <- df_unique %>%
  pivot_longer(-VGO3_ID, names_to = "lf", values_to = "per_year") %>%
  mutate(
    lf_lab = dplyr::recode(lf,
      FEV1_pred_per_year     = "ΔFEV1",
      FVC_pred_per_year      = "ΔFVC",
      FEV1FVC_pred_per_year  = "ΔFEV1/FVC",
      FEF25_75_pred_per_year = "ΔFEF25–75"
    ),
    # <<< set the desired order here
    lf_lab = factor(lf_lab,
      levels = c("ΔFEF25–75","ΔFEV1/FVC", "ΔFVC", "ΔFEV1"))
  )

lung_summ <- lung_long %>%
  group_by(lf_lab) %>%
  summarise(
    n    = sum(is.finite(per_year)),
    mean = mean(per_year, na.rm = TRUE),
    sd   = sd(per_year,   na.rm = TRUE),
    se   = sd / sqrt(n),
    lo95 = mean - qt(0.975, df = pmax(n - 1, 1)) * se,
    hi95 = mean + qt(0.975, df = pmax(n - 1, 1)) * se,
    Direction = case_when(
      is.na(mean) ~ "No data",
      mean >  0   ~ "Improved",
      mean <  0   ~ "Declined",
      TRUE        ~ "No change"
    ),
    .groups = "drop"
  )

lung_long <- lung_long %>%
  left_join(lung_summ %>% select(lf_lab, Direction), by = "lf_lab")

rng_lung <- max(abs(lung_long$per_year), na.rm = TRUE)

p_lung <- ggplot(lung_long, aes(x = lf_lab, y = per_year, fill = Direction)) +
  geom_violin(trim = FALSE, width = 0.9, color = NA) +
  geom_jitter(width = 0.08, alpha = 0.25, size = 0.6) +
  stat_summary(fun = mean, geom = "point", size = 2.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip(ylim = c(-rng_lung, rng_lung)) +
  scale_fill_manual(values = c("Improved"="#44A3AA","Declined"="#AA4A44",
                               "No change"="grey80","No data"="grey90")) +
  labs(
    x = NULL,
    y = "Change per year\n(+ improvement, − decline beyond expected, \n0 ≈ expected age decline)",
    title = "Lung function change per year"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

print(p_lung)

ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Lung_change_violin_plot.svg",
       p_lung, width = 5, height = 3.6)

```
# Alpha diversity change (VGO2 to VGO3)
## Compute
```{r}
# Change in α-diversity (per person × niche)
keep_alpha <- c("SampleID","VGO3_ID","niche","dataset","date_spiro", alpha_vars)
sd_a <- sd0 %>%
  select(any_of(keep_alpha)) %>%
  mutate(visit = visit_from_dataset(dataset)) %>%
  filter(visit %in% c("VGO2","VGO3"))

# Collapse to ONE row per (person × niche × visit)
alpha_by_visit <- sd_a %>%
  group_by(VGO3_ID, niche, visit) %>%
  summarise(
    across(all_of(alpha_vars), ~ mean(suppressWarnings(as.numeric(.x)), na.rm = TRUE)),
    date_spiro = suppressWarnings(min(date_spiro, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  filter(!is.na(VGO3_ID), !is.na(niche))

# Keep pairs within each person × niche
alpha_pairs <- alpha_by_visit %>%
  group_by(VGO3_ID, niche) %>%
  filter(n_distinct(visit) == 2) %>%
  ungroup()

# Wide and deltas (VGO3 − VGO2)
alpha_wide <- alpha_pairs %>%
  tidyr::pivot_wider(names_from = visit,
                     values_from = c(all_of(alpha_vars), date_spiro),
                     names_sep = "_") %>%
  mutate(
    d_Observed = Observed_VGO3 - Observed_VGO2,
    d_Shannon  = Shannon_VGO3  - Shannon_VGO2,
    d_Simpson  = Simpson_VGO3  - Simpson_VGO2
  ) %>%
  select(VGO3_ID, niche, d_Observed, d_Shannon, d_Simpson)

# Join back to rows of that person × niche (replicates panel-level changes)
sd2 <- sd1 %>% left_join(alpha_wide, by = c("VGO3_ID","niche"))
```
## Plot change
```{r}
# α-diversity change per person × niche (free y by index)
alpha_change_cols <- c("d_Observed","d_Shannon","d_Simpson")
alpha_long <- sd2 %>%
  select(VGO3_ID, niche, all_of(alpha_change_cols)) %>%
  pivot_longer(-c(VGO3_ID, niche), names_to = "alpha", values_to = "delta") %>%
  mutate(alpha_lab = dplyr::recode(alpha,
                            d_Observed = "Δ Observed",
                            d_Shannon  = "Δ Shannon",
                            d_Simpson  = "Δ Simpson"))

alpha_summ <- alpha_long %>%
  group_by(niche, alpha_lab) %>%
  summarise(
    n    = sum(is.finite(delta)),
    mean = mean(delta, na.rm = TRUE),
    sd   = sd(delta, na.rm = TRUE),
    se   = sd / sqrt(n),
    lo95 = mean - qt(0.975, df = pmax(n - 1, 1)) * se,
    hi95 = mean + qt(0.975, df = pmax(n - 1, 1)) * se,
    direction = case_when(
      is.na(mean) ~ "No data",
      mean >  0   ~ "Increased",
      mean <  0   ~ "Decreased",
      TRUE        ~ "No change"
    ),
    .groups = "drop"
  )

alpha_long <- alpha_long %>%
  left_join(alpha_summ %>% select(niche, alpha_lab, direction), by = c("niche","alpha_lab"))

p_alpha <- ggplot(alpha_long, aes(x = 1, y = delta, fill = direction)) +
  geom_violin(color = "grey65", trim = FALSE, width = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.25, size = 0.6) +
  geom_point(data = alpha_summ, aes(x = 1, y = mean), inherit.aes = FALSE, size = 2.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(alpha_lab ~ niche, scales = "free_y", switch = "y") +
  scale_x_continuous(breaks = NULL) +
  scale_fill_manual(values = c("Increased"="#44A3AA","Decreased"="#AA4A44","No change"="grey80","No data"="grey90")) +
  labs(x = NULL,
       y = "α-diversity change (VGO3 − VGO2)\n(+ increase, − decrease)",
       title = "α-diversity change by niche") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(fill = "grey90", color = NA))
print(p_alpha)
ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_change_violin.svg",
       p_alpha, width = 6.5, height = 4.5)
```

# Add lung function and alpha diversity changes to phyloseq metadata
```{r}
sd2_out <- sd2
rownames(sd2_out) <- sd2_out$SampleID
sd2_out$SampleID <- NULL
sample_data(ps_final) <- sample_data(sd2_out)

saveRDS(ps_final, "../Output/Phyloseq/ps4_with_changes.rds")
```

# Alpha diversity longitudinal analysis
```{r}
# Longitudinal models: Δ alpha-diversity ~ change in lung function
# (per person × niche; adjust for baseline alpha, baseline lung, and VGO2 covariates)

# Labels for plots
alpha_labs_map <- c(
  d_Observed = "Δ Observed",
  d_Shannon  = "Δ Shannon",
  d_Simpson  = "Δ Simpson"
)

cont_pred_labs_map <- c(
  FEV1_pred_per_year     = "Δ FEV1",
  FVC_pred_per_year      = "Δ FVC",
  FEV1FVC_pred_per_year  = "Δ FEV1/FVC",
  FEF25_75_pred_per_year = "Δ FEF25–75"
)

cat_pred_labs_map <- c(
  FEV1_category_change     = "Δ FEV1 (category)",
  FVC_category_change      = "Δ FVC (category)",
  FEV1FVC_category_change  = "Δ FEV1/FVC (category)",
  FEF25_75_category_change = "Δ FEF25–75 (category)"
)

# Variables / helpers used for the models
id_var       <- "VGO3_ID"
niche_var    <- "niche"
dataset_var  <- "dataset"
date_var     <- "date_spiro"
alpha_vars   <- c("Observed","Shannon","Simpson")
lung_num_pred <- c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred")
lung_cat_vars <- c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")
# covars_base   <- c("age","gender","smoke_status","season_microbiome_sampling")

visit_from_dataset <- function(x) ifelse(grepl("VGO2", x), "VGO2",
                                  ifelse(grepl("VGO3", x), "VGO3", NA))

severity_order <- list(
  FEV1_category      = c("Normal","Mild","Moderate","Severe"),
  FEF25_75_category  = c("Normal","Mild","Moderate","Severe"),
  FVC_category       = c("Normal","Severely reduced"),
  FEV1FVC_category   = c("Normal","Obstructive")
)
to_severity_rank <- function(varname, x) {
  lv <- severity_order[[varname]]
  if (is.null(lv)) return(rep(NA_real_, length(x)))
  idx <- match(tolower(trimws(as.character(x))), tolower(lv))
  ifelse(is.na(idx), NA_real_, idx - 1L)
}
change_label_from_delta <- function(delta) {
  dplyr::case_when(
    is.na(delta) ~ NA_character_,
    delta > 0    ~ "worse",
    delta < 0    ~ "better",
    TRUE         ~ "no_change"
  )
}

# Build paired (VGO2/VGO3) analysis frame per person × niche
sam_data_all <- sd0 %>%
  dplyr::select(
    SampleID, all_of(c(id_var, niche_var, dataset_var, date_var,
                       covars_base, alpha_vars, lung_num_pred, lung_cat_vars))
  ) %>%
  dplyr::mutate(
    visit = visit_from_dataset(.data[[dataset_var]]),
    !!date_var := suppressWarnings(as.Date(.data[[date_var]]))
  ) %>%
  dplyr::filter(visit %in% c("VGO2","VGO3"))

# collapse duplicates within (id × niche × visit)
sam_data_pair <- sam_data_all %>%
  dplyr::group_by(.data[[id_var]], .data[[niche_var]], visit) %>%
  dplyr::summarise(
    dplyr::across(all_of(c(alpha_vars, lung_num_pred)),
                  ~ mean(suppressWarnings(as.numeric(.x)), na.rm = TRUE)),
    dplyr::across(all_of(lung_cat_vars), ~ {
      x <- .x; x <- x[!is.na(x) & nzchar(trimws(as.character(x)))]
      if (length(x) == 0) NA_character_ else {
        ux <- unique(x); ux[which.max(tabulate(match(x, ux)))]
      }
    }),
    dplyr::across(all_of(covars_base), ~ dplyr::first(.x)),
    !!date_var := suppressWarnings(min(.data[[date_var]], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::group_by(.data[[id_var]], .data[[niche_var]]) %>%
  dplyr::filter(dplyr::n_distinct(visit) == 2) %>%  # keep paired only
  dplyr::ungroup()

# wide: *_VGO2 / *_VGO3
df_wide_models <- sam_data_pair %>%
  tidyr::pivot_wider(
    names_from  = visit,
    values_from = c(all_of(alpha_vars), all_of(lung_num_pred), all_of(lung_cat_vars),
                    all_of(covars_base), all_of(date_var)),
    names_sep   = "_"
  ) %>%
  dplyr::mutate(
    years_between = as.numeric(
      difftime(.data[[paste0(date_var, "_VGO3")]],
               .data[[paste0(date_var, "_VGO2")]], units = "days")
    )/365.25
  ) %>%
  dplyr::filter(is.finite(years_between), years_between > 0) %>%
  dplyr::mutate(
    d_Observed = Observed_VGO3 - Observed_VGO2,
    d_Shannon  = Shannon_VGO3  - Shannon_VGO2,
    d_Simpson  = Simpson_VGO3  - Simpson_VGO2
  )

# annualised change in %-pred (continuous)
for (v in lung_num_pred) {
  v_T0 <- paste0(v, "_VGO2"); v_T1 <- paste0(v, "_VGO3")
  df_wide_models[[paste0(v, "_per_year")]] <-
    (df_wide_models[[v_T1]] - df_wide_models[[v_T0]]) / df_wide_models$years_between
}

# categorical change: delta + labels
for (v in lung_cat_vars) {
  v_T0 <- paste0(v, "_VGO2"); v_T1 <- paste0(v, "_VGO3")
  sev0 <- to_severity_rank(v, df_wide_models[[v_T0]])
  sev1 <- to_severity_rank(v, df_wide_models[[v_T1]])
  df_wide_models[[paste0(v, "_delta")]]  <- sev1 - sev0
  df_wide_models[[paste0(v, "_change")]] <- factor(
    change_label_from_delta(df_wide_models[[paste0(v, "_delta")]]),
    levels = c("no_change","worse","better")
  )
}

# ensure baseline covariates at VGO2 are factors where appropriate
fac_vgo2 <- c("gender_VGO2","smoke_status_VGO2","season_microbiome_sampling_VGO2")
fac_vgo2 <- intersect(fac_vgo2, names(df_wide_models))
df_wide_models[fac_vgo2] <- lapply(df_wide_models[fac_vgo2], factor)


# Create season-change (VGO2 to VGO3) and smoking status change (from VGO2 to VGO3)
same_or_diff <- function(a, b) ifelse(is.na(a) | is.na(b), NA_character_,
                                      ifelse(a == b, "same", "different"))

df_wide_models <- df_wide_models %>%
  mutate(
    season_change = factor(
      same_or_diff(season_microbiome_sampling_VGO3, season_microbiome_sampling_VGO2),
      levels = c("same","different")
    ),
    smoke_change  = factor(
      same_or_diff(smoke_status_VGO3, smoke_status_VGO2),
      levels = c("same","different")
    )
  )


# Check number of NP and OP longitudinal samples
sum(df_wide_models$niche == "NP")# 83 matched NP samples
sum(df_wide_models$niche == "OP")# 71 matched OP samples

#### CONTINUOUS MODELS #### 
alpha_deltas <- c("d_Observed","d_Shannon","d_Simpson")
lung_slopes  <- paste0(lung_num_pred, "_per_year")

fit_one_lm <- function(data, outcome, slope_var) {
  alpha0 <- sub("^d_(.*)$", "\\1_VGO2", outcome)          # baseline α
  lung0  <- sub("_per_year$", "_VGO2", slope_var)         # baseline lung
  rhs    <- c(
    slope_var,           # Δ lung (annualised)
    alpha0,              # baseline α
    lung0,               # baseline lung
    "age_VGO2",          # baseline age
    "season_change",     # same/different between visits
    "smoke_change"       # same/different between visits
  )
  fml <- reformulate(rhs, response = outcome)

  dat <- data %>% dplyr::select(dplyr::all_of(c(outcome, rhs))) %>% tidyr::drop_na()
  if (nrow(dat) < 20) return(NULL)

  fit <- lm(fml, data = dat)
  broom::tidy(fit, conf.int = TRUE) %>%
    dplyr::filter(term == slope_var) %>%
    dplyr::mutate(n = nobs(fit), r.sq = summary(fit)$r.squared)
}

run_continuous_block <- function(d) {
  combos <- tidyr::expand_grid(outcome = alpha_deltas, predictor = lung_slopes)
  purrr::pmap_dfr(combos, function(outcome, predictor) {
    out <- fit_one_lm(d, outcome, predictor)
    if (is.null(out)) return(tibble())
    out %>% dplyr::mutate(outcome = outcome, predictor = predictor)
  }) %>%
    dplyr::select(outcome, predictor, estimate, conf.low, conf.high, p.value, n, r.sq)
}

res_cont_np <- run_continuous_block(dplyr::filter(df_wide_models, niche == "NP")) %>% dplyr::mutate(niche = "NP")
res_cont_op <- run_continuous_block(dplyr::filter(df_wide_models, niche == "OP")) %>% dplyr::mutate(niche = "OP")

# FDR per niche × alpha metric
res_continuous <- dplyr::bind_rows(res_cont_np, res_cont_op) %>%
  dplyr::group_by(niche, outcome) %>%
  dplyr::mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::ungroup() %>%
  dplyr::select(niche, outcome, predictor, estimate, conf.low, conf.high, p.value, p_adj, n, r.sq)

write.csv(res_continuous, "../Output/Alpha_diversity/Longitudinal_analysis/Alpha_vs_lung_continuous.csv", row.names = FALSE)


#### CATEGORICAL MODELS ####
cat_changes <- paste0(lung_cat_vars, "_change")

fit_one_lm_cat <- function(data, outcome, cat_var) {
  alpha0    <- sub("^d_(.*)$", "\\1_VGO2", outcome)       # baseline α
  lung0_cat <- sub("_change$", "_VGO2", cat_var)          # baseline lung category (optional)
  rhs       <- c(cat_var, 
                 alpha0, 
                 lung0_cat, 
                 "season_change", 
                 "years_between",
                 "smoke_change") 

  dat <- data %>% dplyr::select(dplyr::all_of(c(outcome, rhs))) %>% tidyr::drop_na()
  if (nrow(dat) < 20) return(NULL)

  dat[[cat_var]] <- stats::relevel(factor(dat[[cat_var]]), ref = "no_change")

  # Drop any factor covariates that collapsed to one level
  single_lvl <- rhs[sapply(dat[rhs], function(x) is.factor(x) && nlevels(factor(x)) < 2)]
  rhs2 <- setdiff(rhs, single_lvl)

  fml <- reformulate(rhs2, response = outcome)
  fit <- tryCatch(lm(fml, data = dat), error = function(e) NULL)
  if (is.null(fit)) return(NULL)

  broom::tidy(fit, conf.int = TRUE) %>%
    dplyr::filter(grepl(paste0("^", cat_var), term)) %>%
    dplyr::mutate(n = nobs(fit), r.sq = summary(fit)$r.squared)
}


run_categorical_block <- function(d) {
  combos <- tidyr::expand_grid(outcome = alpha_deltas, cat = cat_changes)
  purrr::pmap_dfr(combos, function(outcome, cat) {
    out <- fit_one_lm_cat(d, outcome, cat)
    if (is.null(out) || nrow(out) == 0) return(tibble())
    out %>% dplyr::mutate(outcome = outcome, predictor = cat)
  }) %>%
    dplyr::filter(grepl("(worse|better)$", term)) %>%
    dplyr::mutate(
      contrast = dplyr::case_when(
        grepl("worse$", term)  ~ "worse_vs_no_change",
        grepl("better$", term) ~ "better_vs_no_change",
        TRUE ~ NA_character_
      )
    ) %>%
    dplyr::select(outcome, predictor, contrast, estimate, conf.low, conf.high, p.value, n, r.sq)
}

res_cat_np <- run_categorical_block(dplyr::filter(df_wide_models, .data[[niche_var]] == "NP")) %>% dplyr::mutate(niche = "NP")
res_cat_op <- run_categorical_block(dplyr::filter(df_wide_models, .data[[niche_var]] == "OP")) %>% dplyr::mutate(niche = "OP")

res_categorical <- dplyr::bind_rows(res_cat_np, res_cat_op) %>%
  dplyr::group_by(niche, outcome) %>%                 # <= per niche × alpha
  dplyr::mutate(p_adj = p.adjust(p.value, method = "BH")) %>%
  dplyr::ungroup() %>%
  dplyr::select(niche, dplyr::everything())

write.csv(res_categorical, "../Output/Alpha_diversity/Longitudinal_analysis/Alpha_vs_lung_categorical.csv", row.names = FALSE)
```

## Bubble plots
```{r}
# For continuous models
res_cont_plot <- res_continuous %>%
  dplyr::mutate(
    alpha = factor(outcome,   levels = names(alpha_labs_map), labels = unname(alpha_labs_map)),
    lung  = factor(predictor, levels = names(cont_pred_labs_map), labels = unname(cont_pred_labs_map)),
    stars = dplyr::case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    ),
    se   = (conf.high - conf.low) / (2 * 1.96),
    z    = estimate / se,
    zabs = pmin(abs(z), 4),
    sign = dplyr::case_when(estimate > 0 ~ "Positive β",
                            estimate < 0 ~ "Negative β",
                            TRUE ~ "≈ 0")
  )


# choose the same z-range for sizing across facets
size_limits  <- c(0, 4)                   # cap at |z|=4
size_breaks  <- c(0.5, 1, 2, 3, 4)

p_cont_bubble <- ggplot(res_cont_plot, aes(x = lung, y = alpha)) +
  geom_point(aes(size = zabs, fill = sign),
             shape = 21, color = "white", stroke = 0.4) +
  geom_text(aes(label = stars), vjust = 0.5, fontface = "bold", size = 3) +
  facet_wrap(~ niche, nrow = 1) +
  scale_fill_manual(
    name   = "Direction",
    values = c("Negative β" = "#CA562C", "Positive β" = "#2ca0ca", "≈ 0" = "grey80"),
    guide  = guide_legend(order = 1, override.aes = list(size = 6))  # bigger dots
  ) +
  scale_size_continuous(
    name   = "Effect size |β/SE|",
    limits = size_limits,
    breaks = size_breaks,
    range  = c(3, 12),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = 21, fill = "grey70", color = "grey70", alpha = 1) # draw circles
    )
  ) +
  labs(x = "Lung function change", y = "α-diversity change",
       caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        panel.grid  = element_blank(),
        legend.key.height = grid::unit(6, "mm"),
        legend.key.width  = grid::unit(8, "mm"))



ggplot2::ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_diversity_change_lung_func_change.svg",
                p_cont_bubble, width = 7, height = 4)






# ============================================================
# Bubble plot — categorical models (αΔ ~ lung category change)
# ============================================================

# Contrast labels + desired x-axis order:
contrast_labs_map <- c(
  worse_vs_no_change  = "Worse",
  better_vs_no_change = "Better"
)

lung_order_cat <- c(
  "Δ FEV1 (category)",
  "Δ FVC (category)",
  "Δ FEV1/FVC (category)",
  "Δ FEF25–75 (category)"
)
contrast_order <- c("Worse","Better")
combo_levels <- unlist(lapply(lung_order_cat, function(lv) paste(lv, contrast_order, sep = "\n")))



# ---- Labels & desired x-axis order (flipped) ----
contrast_labs_map <- c(
  worse_vs_no_change  = "Worse",
  better_vs_no_change = "Better"
)

lung_order_cat <- c(
  "Δ FEV1 (category)",
  "Δ FVC (category)",
  "Δ FEV1/FVC (category)",
  "Δ FEF25–75 (category)"
)

# We want: "Worse Δ FEV1 (category)", then "Better Δ FEV1 (category)" for each lung metric
combo_levels <- unlist(lapply(lung_order_cat, function(lv) c(paste("Worse", lv), paste("Better", lv))))


# Color palette (robust in print and for most color-vision deficiencies)
pal_dir <- c(
  "Negative β" = "#CA562C",  # deep blue
  "Positive β" = "#2ca0ca",  # deep red
  "≈ 0"       = "grey80"
)

# Prepare plotting frame
res_cat_plot <- res_categorical %>%
  dplyr::mutate(
    alpha = factor(outcome, levels = names(alpha_labs_map), labels = unname(alpha_labs_map)),
    lung  = factor(predictor, levels = names(cat_pred_labs_map), labels = unname(cat_pred_labs_map)),
    contrast_lab = dplyr::recode(contrast, !!!contrast_labs_map),

    # single-line text to match combo_levels
    xlab = factor(paste(contrast_lab, lung), levels = combo_levels),

    stars = dplyr::case_when(p_adj < 0.001 ~ "***", p_adj < 0.01 ~ "**", p_adj < 0.05 ~ "*", TRUE ~ ""),
    se    = (conf.high - conf.low) / (2 * 1.96),
    z     = estimate / se,
    zabs  = abs(z),
    zabs_cap = pmin(pmax(zabs, 0), 4),
    sign  = dplyr::case_when(estimate > 0 ~ "Positive β",
                             estimate < 0 ~ "Negative β",
                             TRUE         ~ "≈ 0")
  )



p_cat_bubble <- ggplot2::ggplot(res_cat_plot, ggplot2::aes(x = xlab, y = alpha)) +
  ggplot2::geom_point(
    ggplot2::aes(size = zabs_cap, fill = sign),   # <- use clamped size
    shape = 21, color = "white", stroke = 0.5
  ) +
  ggplot2::geom_text(ggplot2::aes(label = stars), vjust = 0.5, fontface = "bold", size = 3) +
  ggplot2::facet_wrap(~ niche, nrow = 1) +
  ggplot2::scale_fill_manual(
    values = pal_dir,
    name   = "Direction",
    guide  = ggplot2::guide_legend(
      title.position = "top",
      override.aes = list(shape = 21, size = 7, colour = "white", stroke = 0.6)
    )
  ) +
  ggplot2::scale_size_continuous(
    name   = expression(paste("Effect size  |", beta, "/SE|")),
    limits = c(0, 4),
    breaks = c(0.5, 1, 2, 3, 4),
    labels = c("0.5", "1.0", "2.0", "3.0", "4.0"),
    range  = c(3, 11),
    guide  = ggplot2::guide_legend(
      title.position = "top",
      override.aes = list(fill = "grey60", shape = 21, colour = "white", stroke = 0.6)
    )
  ) +
  ggplot2::labs(
    x = "Lung function category change (vs no change)",
    y = "α-diversity change",
    caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x  = ggplot2::element_text(angle = 40, hjust = 1),
    panel.grid   = ggplot2::element_blank(),
    legend.box   = "vertical",
    legend.title = ggplot2::element_text(face = "bold"),
    legend.key.height = grid::unit(0.7, "cm"),
    legend.key.width  = grid::unit(1.1, "cm")
  ) 


print(p_cat_bubble)

ggplot2::ggsave(
  "../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_diversity_change_lung_cat_change.svg",
  p_cat_bubble, width = 9, height = 4
)

```

# Beta diversity
## NP within vs between individual composition
```{r}
id_var <- "VGO3_ID"

# Subset & merge (fix the typo) 
VGO2_NP_ps <- subset_samples(ps_final, study == "VGO2" & niche == "NP")
VGO3_NP_ps <- subset_samples(ps_final, study == "VGO3" & niche == "NP")  # fixed
VGO_NP_combined <- merge_phyloseq(VGO2_NP_ps, VGO3_NP_ps)

# Filter & RA, then Bray–Curtis 
VGO_NP_RA <- VGO_NP_combined %>% pres_abund_filter() %>% to_RA()
np_dist   <- phyloseq::distance(VGO_NP_RA, method = "bray")
meta_np   <- sample_data(VGO_NP_RA) %>% data.frame(check.names = FALSE)

# Long pairwise distances + subject IDs 
np_pairs <- as.matrix(np_dist) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample1") %>%
  tidyr::pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
  dplyr::filter(Sample1 < Sample2) %>%                        # unique pairs
  dplyr::mutate(
    ID1 = meta_np[Sample1, id_var],
    ID2 = meta_np[Sample2, id_var]
  ) %>%
  dplyr::filter(!is.na(ID1) & !is.na(ID2)) %>%                # drop NA IDs
  dplyr::mutate(Comparison = ifelse(ID1 == ID2, "Within", "Between"))

# Sanity check: need both groups 
if (!all(c("Within","Between") %in% unique(np_pairs$Comparison))) {
  stop("You don't have both 'Within' and 'Between' pairs. Check that VGO2 & VGO3 NP samples exist and IDs are populated.")
}

# Balance the groups (downsample larger to the size of the smaller)
set.seed(123)
n_to_sample <- min(table(np_pairs$Comparison))
np_bal <- np_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::slice_sample(n = n_to_sample) %>%
  dplyr::ungroup()

# Y limits from boxplot whiskers (avoid NA)
whisk <- np_bal %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    whisker_low  = boxplot.stats(Distance)$stats[1],
    whisker_high = boxplot.stats(Distance)$stats[5],
    .groups = "drop"
  )

y_min <- min(whisk$whisker_low,  na.rm = TRUE)
y_max <- min(1, max(whisk$whisker_high, na.rm = TRUE))
y_max_ext <- y_max + 0.03 * (y_max - y_min)
p_y <- y_max_ext - 0.01 * (y_max_ext - y_min)

# --- Test & label ---
wt <- wilcox.test(Distance ~ Comparison, data = np_bal, exact = FALSE)
label <- ifelse(is.finite(wt$p.value) && wt$p.value < 0.001,
                "italic(p) < 0.001",
                paste0("italic(p) == ", sprintf('%.3f', wt$p.value)))

# --- Plot ---
np_plot <- ggplot(np_bal, aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_jitter(aes(color = Comparison), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("Within" = "#ad7c8a", "Between" = "#d9b9c2")) +
  scale_color_manual(values = c("Within" = "#ad7c8a", "Between" = "#d9b9c2")) +
  coord_cartesian(ylim = c(y_min, y_max_ext)) +
  labs(
    title = "NP: Within vs Between-Individual Bray–Curtis\n (VGO2 vs VGO3)",
    x = "Comparison Type", y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = label, y_position = p_y,
    xmin = 1, xmax = 2, tip_length = 0.01, parse = TRUE
  )

ggsave("../Output/Beta_diversity/Longitudinal_analysis/NP_jitter_boxplot_balanced.svg",
       np_plot, width = 5, height = 5)

```
## OP within vs between individual composition
```{r}
id_var <- "VGO3_ID"

# --- Subset & merge (fix the typo) ---
VGO2_OP_ps <- subset_samples(ps_final, study == "VGO2" & niche == "OP")
VGO3_OP_ps <- subset_samples(ps_final, study == "VGO3" & niche == "OP")  # fixed
VGO_OP_combined <- merge_phyloseq(VGO2_OP_ps, VGO3_OP_ps)

# --- Filter & RA, then Bray–Curtis ---
VGO_OP_RA <- VGO_OP_combined %>% pres_abund_filter() %>% to_RA()
op_dist   <- phyloseq::distance(VGO_OP_RA, method = "bray")
meta_op   <- sample_data(VGO_OP_RA) %>% data.frame(check.names = FALSE)

#  Long pairwise distances + subject IDs 
op_pairs <- as.matrix(op_dist) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample1") %>%
  tidyr::pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
  dplyr::filter(Sample1 < Sample2) %>%                        # unique pairs
  dplyr::mutate(
    ID1 = meta_op[Sample1, id_var],
    ID2 = meta_op[Sample2, id_var]
  ) %>%
  dplyr::filter(!is.na(ID1) & !is.na(ID2)) %>%                # drop NA IDs
  dplyr::mutate(Comparison = ifelse(ID1 == ID2, "Within", "Between"))

#  Sanity check: need both groups 
if (!all(c("Within","Between") %in% unique(op_pairs$Comparison))) {
  stop("You don't have both 'Within' and 'Between' pairs. Check that VGO2 & VGO3 NP samples exist and IDs are populated.")
}

#  Balance the groups (downsample larger to the size of the smaller) 
set.seed(123)
n_to_sample <- min(table(op_pairs$Comparison))
op_bal <- op_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::slice_sample(n = n_to_sample) %>%
  dplyr::ungroup()

# Y limits from boxplot whiskers (avoid NA)
whisk <- op_bal %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    whisker_low  = boxplot.stats(Distance)$stats[1],
    whisker_high = boxplot.stats(Distance)$stats[5],
    .groups = "drop"
  )

y_min <- min(whisk$whisker_low,  na.rm = TRUE)
y_max <- min(1, max(whisk$whisker_high, na.rm = TRUE))
y_max_ext <- y_max + 0.03 * (y_max - y_min)
p_y <- y_max_ext - 0.01 * (y_max_ext - y_min)

# Test & label 
wt <- wilcox.test(Distance ~ Comparison, data = op_bal, exact = FALSE)
label <- ifelse(is.finite(wt$p.value) && wt$p.value < 0.001,
                "italic(p) < 0.001",
                paste0("italic(p) == ", sprintf('%.3f', wt$p.value)))

# Plot
op_plot <- ggplot(op_bal, aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_jitter(aes(color = Comparison), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("Within" = "#3f7566", "Between" = "#b2c8c2")) +
  scale_color_manual(values = c("Within" = "#3f7566", "Between" = "#b2c8c2")) +
  coord_cartesian(ylim = c(y_min, y_max_ext)) +
  labs(
    title = "OP: Within vs Between-Individual Bray–Curtis\n (VGO2 vs VGO3)",
    x = "Comparison Type", y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = label, y_position = p_y,
    xmin = 1, xmax = 2, tip_length = 0.01, parse = TRUE
  )

ggsave("../Output/Beta_diversity/Longitudinal_analysis/OP_jitter_boxplot_balanced.svg",
       op_plot, width = 5, height = 5)
```

## Combined plots NP and OP within vs between composition
```{r}
combined_plot <- np_plot + op_plot +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = "A")

# Save combined plot
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NP_OP_jitter_boxplot_balanced.svg",
       combined_plot, width = 10, height = 5)
```
# NP vs OP 
## BC distances and visualisation
```{r}
# Helper function 
get_between_individual_distances <- function(ps_obj, niche_label) {
  # Filter and convert to relative abundance
  ps_RA <- ps_obj %>% pres_abund_filter() %>% to_RA()
  dist_mat <- phyloseq::distance(ps_RA, method = "bray")
  
  meta <- data.frame(sample_data(ps_RA))
  meta$SampleID <- rownames(meta)
  
  dist_df <- as.matrix(dist_mat) %>%
    as.data.frame() %>%
    rownames_to_column("Sample1") %>%
    pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
    filter(Sample1 < Sample2) %>%  # Unique pairs only
    mutate(
      ID1 = meta[Sample1, "VGO3_ID"],
      ID2 = meta[Sample2, "VGO3_ID"],
      Comparison = ifelse(ID1 == ID2, "Within", "Between")
    ) %>%
    filter(Comparison == "Between") %>%
    mutate(Niche = niche_label)
  
  return(dist_df)
}

# Get distances
# Subset and prepare NP and OP datasets (VGO2+VGO3)
VGO_NP <- subset_samples(ps_final, niche == "NP" & dataset %in% c("VGO2_NP", "VGO3_NPOP"))
VGO_OP <- subset_samples(ps_final, niche == "OP" & dataset %in% c("VGO2_OP", "VGO3_NPOP"))

np_between <- get_between_individual_distances(VGO_NP, "NP")
op_between <- get_between_individual_distances(VGO_OP, "OP")

combined_between <- bind_rows(np_between, op_between)

# Optional: Downsample to balance group sizes
set.seed(123)
min_n <- min(table(combined_between$Niche))
combined_between_bal <- combined_between %>%
  group_by(Niche) %>%
  slice_sample(n = min_n) %>%
  ungroup()

# -- Wilcoxon test -- #
wt <- wilcox.test(Distance ~ Niche, data = combined_between_bal)
p_label <- ifelse(
  wt$p.value < 0.001,
  "italic(p) < 0.001",
  paste0("italic(p) == ", sprintf("%.3f", wt$p.value))
)

# -- Plot -- #
y_min <- min(combined_between_bal$Distance)
y_max <- max(combined_between_bal$Distance)
y_max_extended <- y_max + 0.03 * (y_max - y_min)
pval_y_pos <- y_max_extended - 0.01 * (y_max_extended - y_min)

between_plot <- ggplot(combined_between_bal, aes(x = Niche, y = Distance, fill = Niche)) +
  geom_jitter(aes(color = Niche), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
  scale_color_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
  coord_cartesian(ylim = c(y_min, y_max_extended)) +
  labs(
    title = "Between-Individual Bray–Curtis Dissimilarity",
    subtitle = "Comparison between NP and OP samples (VGO2 + VGO3)",
    x = "Niche",
    y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = p_label,
    y_position = pval_y_pos,
    xmin = 1, xmax = 2,
    tip_length = 0.01,
    parse = TRUE
  )

# Save
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NPvsOP/Between_individual_NP_OP_comparison.png", between_plot, width = 6, height = 5)

# Show
print(between_plot)

```
## PERMANOVA NP vs OP
```{r}
# Subset VGO2 and VGO3 data
vgo2_ps <- subset_samples(ps_final, dataset %in% c("VGO2_NP", "VGO2_OP") & niche == c("NP", "OP"))
vgo3_ps <- subset_samples(ps_final, dataset == "VGO3_NPOP" & niche == c("NP", "OP"))

compare_NP_OP <- function(ps_obj, label) {
  ps_filt <- ps_obj %>%
    pres_abund_filter() %>%
    to_RA()

  dist <- phyloseq::distance(ps_filt, method = "bray")
  meta <- data.frame(sample_data(ps_filt))

  # Run PERMANOVA
  ad <- adonis2(dist ~ niche, data = meta, permutations = 999)
  pval <- ad$`Pr(>F)`[1]
  r2 <- round(ad$R2[1], 3)
  p_label <- ifelse(pval < 0.001, "italic(p)<0.001", paste0("italic(p)==", sprintf("%.3f", pval)))

  # Ordination plot with manual colors
  ord <- ordinate(ps_filt, method = "PCoA", distance = dist)
  plot <- plot_ordination(ps_filt, ord, color = "niche") +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
    labs(
      title = paste0(label, ": NP vs OP (Bray-Curtis)"),
      subtitle = paste0("PERMANOVA R² = ", r2),
      color = "Niche"
    ) +
    annotate("text", x = Inf, y = Inf, label = p_label, parse = TRUE, hjust = 1.1, vjust = 1.5) +
    theme_minimal(base_size = 12)

  return(plot)
}

# Create both plots
vgo2_plot <- compare_NP_OP(vgo2_ps, "VGO2")
vgo3_plot <- compare_NP_OP(vgo3_ps, "VGO3")

# Combine plots side-by-side
combined_np_op_plot <- vgo2_plot + vgo3_plot + 
  plot_layout(ncol = 2) + 
  plot_annotation(tag_levels = "A")

# Save
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NPvsOP/NP_OP_PERMANOVA_ordination.svg",
       combined_np_op_plot, width = 10, height = 5)

# Show
print(combined_np_op_plot)
```
