---
title: "Longitudinal analysis - change in microbiome --> change in lung function"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

## Plot change
```{r}
# Lung function change per person (unique row per VGO3_ID)
# pick one row per person with the most non-missing change values
lung_change_cols <- c("delta_FEV1", delta_FVC, )


df_unique <- sd2 %>%
  mutate(nonmiss_lf = rowSums(across(all_of(lung_change_cols), ~ !is.na(.x)))) %>%
  arrange(VGO3_ID, desc(nonmiss_lf)) %>%
  group_by(VGO3_ID) %>% slice_head(n = 1) %>% ungroup() %>%
  select(VGO3_ID, all_of(lung_change_cols))

lung_long <- df_unique %>%
  pivot_longer(-VGO3_ID, names_to = "lf", values_to = "per_year") %>%
  mutate(
    lf_lab = dplyr::recode(lf,
      FEV1_pred_per_year     = "ΔFEV1",
      FVC_pred_per_year      = "ΔFVC",
      FEV1FVC_pred_per_year  = "ΔFEV1/FVC",
      FEF25_75_pred_per_year = "ΔFEF25–75"
    ),
    # <<< set the desired order here
    lf_lab = factor(lf_lab,
      levels = c("ΔFEF25–75","ΔFEV1/FVC", "ΔFVC", "ΔFEV1"))
  )

lung_summ <- lung_long %>%
  group_by(lf_lab) %>%
  summarise(
    n    = sum(is.finite(per_year)),
    mean = mean(per_year, na.rm = TRUE),
    sd   = sd(per_year,   na.rm = TRUE),
    se   = sd / sqrt(n),
    lo95 = mean - qt(0.975, df = pmax(n - 1, 1)) * se,
    hi95 = mean + qt(0.975, df = pmax(n - 1, 1)) * se,
    Direction = case_when(
      is.na(mean) ~ "No data",
      mean >  0   ~ "Improved",
      mean <  0   ~ "Declined",
      TRUE        ~ "No change"
    ),
    .groups = "drop"
  )

lung_long <- lung_long %>%
  left_join(lung_summ %>% select(lf_lab, Direction), by = "lf_lab")

rng_lung <- max(abs(lung_long$per_year), na.rm = TRUE)

p_lung <- ggplot(lung_long, aes(x = lf_lab, y = per_year, fill = Direction)) +
  geom_violin(trim = FALSE, width = 0.9, color = NA) +
  geom_jitter(width = 0.08, alpha = 0.25, size = 0.6) +
  stat_summary(fun = mean, geom = "point", size = 2.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip(ylim = c(-rng_lung, rng_lung)) +
  scale_fill_manual(values = c("Improved"="#44A3AA","Declined"="#AA4A44",
                               "No change"="grey80","No data"="grey90")) +
  labs(
    x = NULL,
    y = "Change per year\n(+ improvement, − decline beyond expected, \n0 ≈ expected age decline)",
    title = "Lung function change per year"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

print(p_lung)

ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Lung_change_violin_plot.svg",
       p_lung, width = 5, height = 3.6)

```
# Alpha diversity change (VGO2 to VGO3)
## Compute
```{r}
# Change in α-diversity (per person × niche)
keep_alpha <- c("SampleID","VGO3_ID","niche","dataset","date_spiro", alpha_vars)
sd_a <- sd0 %>%
  select(any_of(keep_alpha)) %>%
  mutate(visit = visit_from_dataset(dataset)) %>%
  filter(visit %in% c("VGO2","VGO3"))

# Collapse to ONE row per (person × niche × visit)
alpha_by_visit <- sd_a %>%
  group_by(VGO3_ID, niche, visit) %>%
  summarise(
    across(all_of(alpha_vars), ~ mean(suppressWarnings(as.numeric(.x)), na.rm = TRUE)),
    date_spiro = suppressWarnings(min(date_spiro, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  filter(!is.na(VGO3_ID), !is.na(niche))

# Keep pairs within each person × niche
alpha_pairs <- alpha_by_visit %>%
  group_by(VGO3_ID, niche) %>%
  filter(n_distinct(visit) == 2) %>%
  ungroup()

# Wide and deltas (VGO3 − VGO2)
alpha_wide <- alpha_pairs %>%
  tidyr::pivot_wider(names_from = visit,
                     values_from = c(all_of(alpha_vars), date_spiro),
                     names_sep = "_") %>%
  mutate(
    d_Observed = Observed_VGO3 - Observed_VGO2,
    d_Shannon  = Shannon_VGO3  - Shannon_VGO2,
    d_Simpson  = Simpson_VGO3  - Simpson_VGO2
  ) %>%
  select(VGO3_ID, niche, d_Observed, d_Shannon, d_Simpson)

# Join back to rows of that person × niche (replicates panel-level changes)
sd2 <- sd1 %>% left_join(alpha_wide, by = c("VGO3_ID","niche"))

```
## Plot change
```{r}
# α-diversity change per person × niche (free y by index)
alpha_change_cols <- c("d_Observed","d_Shannon","d_Simpson")
alpha_long <- sd2 %>%
  select(VGO3_ID, niche, all_of(alpha_change_cols)) %>%
  pivot_longer(-c(VGO3_ID, niche), names_to = "alpha", values_to = "delta") %>%
  mutate(alpha_lab = dplyr::recode(alpha,
                            d_Observed = "Δ Observed",
                            d_Shannon  = "Δ Shannon",
                            d_Simpson  = "Δ Simpson"))

alpha_summ <- alpha_long %>%
  group_by(niche, alpha_lab) %>%
  summarise(
    n    = sum(is.finite(delta)),
    mean = mean(delta, na.rm = TRUE),
    sd   = sd(delta, na.rm = TRUE),
    se   = sd / sqrt(n),
    lo95 = mean - qt(0.975, df = pmax(n - 1, 1)) * se,
    hi95 = mean + qt(0.975, df = pmax(n - 1, 1)) * se,
    direction = case_when(
      is.na(mean) ~ "No data",
      mean >  0   ~ "Increased",
      mean <  0   ~ "Decreased",
      TRUE        ~ "No change"
    ),
    .groups = "drop"
  )

alpha_long <- alpha_long %>%
  left_join(alpha_summ %>% select(niche, alpha_lab, direction), by = c("niche","alpha_lab"))

p_alpha <- ggplot(alpha_long, aes(x = 1, y = delta, fill = direction)) +
  geom_violin(color = "grey65", trim = FALSE, width = 0.9) +
  geom_jitter(width = 0.08, alpha = 0.25, size = 0.6) +
  geom_point(data = alpha_summ, aes(x = 1, y = mean), inherit.aes = FALSE, size = 2.2, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_grid(alpha_lab ~ niche, scales = "free_y", switch = "y") +
  scale_x_continuous(breaks = NULL) +
  scale_fill_manual(values = c("Increased"="#44A3AA","Decreased"="#AA4A44","No change"="grey80","No data"="grey90")) +
  labs(x = NULL,
       y = "α-diversity change (VGO3 − VGO2)\n(+ increase, − decrease)",
       title = "α-diversity change by niche") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(fill = "grey90", color = NA))
print(p_alpha)
ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_change_violin.svg",
       p_alpha, width = 6.5, height = 4.5)
```

# Add lung function and alpha diversity changes to phyloseq metadata
```{r}
sd2_out <- sd2
rownames(sd2_out) <- sd2_out$SampleID
sd2_out$SampleID <- NULL
sample_data(ps_final) <- sample_data(sd2_out)

saveRDS(ps_final, "../Output/Phyloseq/ps4_with_changes.rds")
```

# Alpha diversity longitudinal analysis
```{r}
# Longitudinal models: Δ alpha-diversity ~ change in lung function
# (per person × niche; adjust for baseline alpha, baseline lung, and VGO2 covariates)

# Labels for plots
alpha_labs_map <- c(
  d_Observed = "Δ Observed",
  d_Shannon  = "Δ Shannon",
  d_Simpson  = "Δ Simpson"
)

cont_pred_labs_map <- c(
  FEV1_pred_per_year     = "Δ FEV1",
  FVC_pred_per_year      = "Δ FVC",
  FEV1FVC_pred_per_year  = "Δ FEV1/FVC",
  FEF25_75_pred_per_year = "Δ FEF25–75"
)

cat_pred_labs_map <- c(
  FEV1_category_change     = "Δ FEV1 (category)",
  FVC_category_change      = "Δ FVC (category)",
  FEV1FVC_category_change  = "Δ FEV1/FVC (category)",
  FEF25_75_category_change = "Δ FEF25–75 (category)"
)

# Variables / helpers used for the models
id_var       <- "VGO3_ID"
niche_var    <- "niche"
dataset_var  <- "dataset"
date_var     <- "date_spiro"
alpha_vars   <- c("Observed","Shannon","Simpson")
lung_num_pred <- c("FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred")
lung_cat_vars <- c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")
# covars_base   <- c("age","gender","smoke_status","season_microbiome_sampling")

visit_from_dataset <- function(x) ifelse(grepl("VGO2", x), "VGO2",
                                  ifelse(grepl("VGO3", x), "VGO3", NA))

severity_order <- list(
  FEV1_category      = c("Normal","Mild","Moderate","Severe"),
  FEF25_75_category  = c("Normal","Mild","Moderate","Severe"),
  FVC_category       = c("Normal","Severely reduced"),
  FEV1FVC_category   = c("Normal","Obstructive")
)
to_severity_rank <- function(varname, x) {
  lv <- severity_order[[varname]]
  if (is.null(lv)) return(rep(NA_real_, length(x)))
  idx <- match(tolower(trimws(as.character(x))), tolower(lv))
  ifelse(is.na(idx), NA_real_, idx - 1L)
}
change_label_from_delta <- function(delta) {
  dplyr::case_when(
    is.na(delta) ~ NA_character_,
    delta > 0    ~ "worse",
    delta < 0    ~ "better",
    TRUE         ~ "no_change"
  )
}

# Build paired (VGO2/VGO3) analysis frame per person × niche
sam_data_all <- sd0 %>%
  dplyr::select(
    SampleID, all_of(c(id_var, niche_var, dataset_var, date_var,
                       covars_base, alpha_vars, lung_num_pred, lung_cat_vars))
  ) %>%
  dplyr::mutate(
    visit = visit_from_dataset(.data[[dataset_var]]),
    !!date_var := suppressWarnings(as.Date(.data[[date_var]]))
  ) %>%
  dplyr::filter(visit %in% c("VGO2","VGO3"))

# collapse duplicates within (id × niche × visit)
sam_data_pair <- sam_data_all %>%
  dplyr::group_by(.data[[id_var]], .data[[niche_var]], visit) %>%
  dplyr::summarise(
    dplyr::across(all_of(c(alpha_vars, lung_num_pred)),
                  ~ mean(suppressWarnings(as.numeric(.x)), na.rm = TRUE)),
    dplyr::across(all_of(lung_cat_vars), ~ {
      x <- .x; x <- x[!is.na(x) & nzchar(trimws(as.character(x)))]
      if (length(x) == 0) NA_character_ else {
        ux <- unique(x); ux[which.max(tabulate(match(x, ux)))]
      }
    }),
    dplyr::across(all_of(covars_base), ~ dplyr::first(.x)),
    !!date_var := suppressWarnings(min(.data[[date_var]], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::group_by(.data[[id_var]], .data[[niche_var]]) %>%
  dplyr::filter(dplyr::n_distinct(visit) == 2) %>%  # keep paired only
  dplyr::ungroup()

# wide: *_VGO2 / *_VGO3
df_wide_models <- sam_data_pair %>%
  tidyr::pivot_wider(
    names_from  = visit,
    values_from = c(all_of(alpha_vars), all_of(lung_num_pred), all_of(lung_cat_vars),
                    all_of(covars_base), all_of(date_var)),
    names_sep   = "_"
  ) %>%
  dplyr::mutate(
    years_between = as.numeric(
      difftime(.data[[paste0(date_var, "_VGO3")]],
               .data[[paste0(date_var, "_VGO2")]], units = "days")
    )/365.25
  ) %>%
  dplyr::filter(is.finite(years_between), years_between > 0) %>%
  dplyr::mutate(
    d_Observed = Observed_VGO3 - Observed_VGO2,
    d_Shannon  = Shannon_VGO3  - Shannon_VGO2,
    d_Simpson  = Simpson_VGO3  - Simpson_VGO2
  )

# annualised change in %-pred (continuous)
for (v in lung_num_pred) {
  v_T0 <- paste0(v, "_VGO2"); v_T1 <- paste0(v, "_VGO3")
  df_wide_models[[paste0(v, "_per_year")]] <-
    (df_wide_models[[v_T1]] - df_wide_models[[v_T0]]) / df_wide_models$years_between
}

# categorical change: delta + labels
for (v in lung_cat_vars) {
  v_T0 <- paste0(v, "_VGO2"); v_T1 <- paste0(v, "_VGO3")
  sev0 <- to_severity_rank(v, df_wide_models[[v_T0]])
  sev1 <- to_severity_rank(v, df_wide_models[[v_T1]])
  df_wide_models[[paste0(v, "_delta")]]  <- sev1 - sev0
  df_wide_models[[paste0(v, "_change")]] <- factor(
    change_label_from_delta(df_wide_models[[paste0(v, "_delta")]]),
    levels = c("no_change","worse","better")
  )
}

# ensure baseline covariates at VGO2 are factors where appropriate
fac_vgo2 <- c("gender_VGO2","smoke_status_VGO2","season_microbiome_sampling_VGO2")
fac_vgo2 <- intersect(fac_vgo2, names(df_wide_models))
df_wide_models[fac_vgo2] <- lapply(df_wide_models[fac_vgo2], factor)


# Create season-change (VGO2 to VGO3) and smoking status change (from VGO2 to VGO3)
same_or_diff <- function(a, b) ifelse(is.na(a) | is.na(b), NA_character_,
                                      ifelse(a == b, "same", "different"))

df_wide_models <- df_wide_models %>%
  mutate(
    season_change = factor(
      same_or_diff(season_microbiome_sampling_VGO3, season_microbiome_sampling_VGO2),
      levels = c("same","different")
    ),
    smoke_change  = factor(
      same_or_diff(smoke_status_VGO3, smoke_status_VGO2),
      levels = c("same","different")
    )
  )


# Check number of NP and OP longitudinal samples
sum(df_wide_models$niche == "NP")# 83 matched NP samples
sum(df_wide_models$niche == "OP")# 71 matched OP samples

#### CONTINUOUS MODELS #### 
alpha_deltas <- c("d_Observed","d_Shannon","d_Simpson")
lung_slopes  <- paste0(lung_num_pred, "_per_year")

# fit_one_lm <- function(data, outcome, slope_var) {
#   alpha0 <- sub("^d_(.*)$", "\\1_VGO2", outcome)          # baseline α
#   lung0  <- sub("_per_year$", "_VGO2", slope_var)         # baseline lung
#   rhs    <- c(
#     slope_var,           # Δ lung (annualised)
#     alpha0,              # baseline α
#     lung0,               # baseline lung
#     "age_VGO2",          # baseline age
#     "season_change",     # same/different between visits
#     "smoke_change"       # same/different between visits
#   )
#   fml <- reformulate(rhs, response = outcome)
# 
#   dat <- data %>% dplyr::select(dplyr::all_of(c(outcome, rhs))) %>% tidyr::drop_na()
#   if (nrow(dat) < 20) return(NULL)
# 
#   fit <- lm(fml, data = dat)
#   broom::tidy(fit, conf.int = TRUE) %>%
#     dplyr::filter(term == slope_var) %>%
#     dplyr::mutate(n = nobs(fit), r.sq = summary(fit)$r.squared)
# }





fit_lung_on_alpha <- function(data, lung_slope, alpha_delta) {
  lung0  <- sub("_per_year$", "_VGO2", lung_slope)      # baseline lung
  alpha0 <- sub("^d_(.*)$", "\\1_VGO2", alpha_delta)    # baseline alpha (Observed_VGO2 etc.)
  rhs <- c(alpha_delta, alpha0, lung0, "age_VGO2", "season_change", "smoke_change")
  fml <- reformulate(rhs, response = lung_slope)
  dat <- data %>% dplyr::select(dplyr::all_of(c(lung_slope, rhs))) %>% tidyr::drop_na()
  if (nrow(dat) < 20) return(NULL)
  broom::tidy(lm(fml, dat), conf.int = TRUE) %>%
    dplyr::filter(term == alpha_delta) %>%
    dplyr::mutate(n = nobs(lm(fml, dat)),
                  outcome = lung_slope, predictor = alpha_delta)
}

alpha_deltas <- c("d_Observed","d_Shannon","d_Simpson")
lung_slopes  <- paste0(lung_num_pred, "_per_year")

res_np <- tidyr::expand_grid(lung = lung_slopes, a = alpha_deltas) %>%
  purrr::pmap_dfr(~fit_lung_on_alpha(dplyr::filter(df_wide_models, niche=="NP"), ..1, ..2)) %>%
  dplyr::mutate(niche="NP")

res_op <- tidyr::expand_grid(lung = lung_slopes, a = alpha_deltas) %>%
  purrr::pmap_dfr(~fit_lung_on_alpha(dplyr::filter(df_wide_models, niche=="OP"), ..1, ..2)) %>%
  dplyr::mutate(niche="OP")

res_alpha_to_lung <- dplyr::bind_rows(res_np, res_op) %>%
  dplyr::group_by(niche, outcome) %>%
  dplyr::mutate(p_adj = p.adjust(p.value, "BH")) %>%
  dplyr::ungroup()

write.csv(res_alpha_to_lung, "../Output/Alpha_diversity/Longitudinal_analysis/Alpha_vs_lung_continuous.csv", row.names = FALSE)



#### CATEGORICAL MODELS ####
alpha_deltas <- c("d_Observed","d_Shannon","d_Simpson")
lung_cat_outcomes <- c("FEV1_category_change","FVC_category_change",
                       "FEV1FVC_category_change","FEF25_75_category_change")

safe_tidy_polr_wald <- function(fit, term_keep) {
  s <- summary(fit)
  co <- coef(fit)                       # slope coefficients only (not thresholds)
  V  <- vcov(fit)
  se <- sqrt(diag(V))[names(co)]
  est <- co[term_keep]
  se1 <- se[term_keep]
  if (is.na(est) || is.na(se1)) return(NULL)
  z  <- est / se1
  tibble(
    term      = term_keep,
    estimate  = est,
    std.error = se1,
    statistic = z,
    p.value   = 2*pnorm(abs(z), lower.tail = FALSE),
    conf.low  = est - 1.96*se1,
    conf.high = est + 1.96*se1
  )
}

fit_cat_ordinal <- function(dat, lung_cat, alpha_delta) {
  dd <- dat %>%
    dplyr::select(all_of(c(lung_cat, alpha_delta,
                           "age_VGO2","season_change","smoke_change"))) %>%
    tidyr::drop_na()

  if (nrow(dd) < 30) return(NULL)

  # Outcome must be ordered with >= 2 levels present
  dd[[lung_cat]] <- factor(dd[[lung_cat]],
                           levels = c("no_change","better","worse"),
                           ordered = TRUE)
  lvl_tab <- table(dd[[lung_cat]])
  if (nlevels(dd[[lung_cat]]) < 2 || sum(lvl_tab > 0) < 2) return(NULL)

  # Predictor must vary
  if (stats::sd(dd[[alpha_delta]]) == 0 || !is.finite(stats::sd(dd[[alpha_delta]]))) return(NULL)

  # Fit with Hessian; catch separation problems
  fml <- as.formula(paste(
    lung_cat, "~", paste(c(alpha_delta,"age_VGO2","season_change","smoke_change"), collapse=" + ")
  ))
  fit <- tryCatch(MASS::polr(fml, data = dd, Hess = TRUE, method = "logistic"),
                  error = function(e) NULL)
  if (is.null(fit)) return(NULL)

  # Wald CI for the alpha term (avoids profile/approx)
  out <- tryCatch(
    safe_tidy_polr_wald(fit, term_keep = alpha_delta),
    error = function(e) NULL
  )
  if (is.null(out)) return(NULL)

  out %>% mutate(outcome = lung_cat, predictor = alpha_delta, n = nrow(dd))
}

run_cat_block <- function(d) {
  cross_df(list(lung = lung_cat_outcomes, a = alpha_deltas)) %>%
    pmap_dfr(~ fit_cat_ordinal(d, ..1, ..2))
}

# Example:
res_cat_np <- run_cat_block(df_wide_models %>% filter(niche=="NP")) %>% mutate(niche="NP")
res_cat_op <- run_cat_block(df_wide_models %>% filter(niche=="OP")) %>% mutate(niche="OP")

res_cat_ordinal <- bind_rows(res_cat_np, res_cat_op) %>%
  group_by(niche, outcome) %>% mutate(p_adj = p.adjust(p.value, "BH")) %>% ungroup()


write.csv(res_cat_ordinal, "../Output/Alpha_diversity/Longitudinal_analysis/Alpha_vs_lung_categorical.csv", row.names = FALSE)
```

## Bubble plots
```{r}
# For continuous models
res_alpha_to_lung$co
res_cont_plot <- res_alpha_to_lung %>%
  dplyr::mutate(
    lung = factor(outcome, levels = names(cont_pred_labs_map), labels = unname(cont_pred_labs_map)),
    alpha  = factor(predictor, levels = names(alpha_labs_map), labels = unname(alpha_labs_map)),
    stars = dplyr::case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    ),
    se   = (conf.high - conf.low) / (2 * 1.96),
    z    = estimate / se,
    zabs = pmin(abs(z), 4),
    sign = dplyr::case_when(estimate > 0 ~ "Positive β",
                            estimate < 0 ~ "Negative β",
                            TRUE ~ "≈ 0")
  )

# choose the same z-range for sizing across facets
size_limits  <- c(0, 4)                   # cap at |z|=4
size_breaks  <- c(0.5, 1, 2, 3, 4)

p_cont_bubble <- ggplot(res_cont_plot, aes(x = alpha, y = lung)) +
  geom_point(aes(size = zabs, fill = sign),
             shape = 21, color = "white", stroke = 0.4) +
  geom_text(aes(label = stars), vjust = 0.5, fontface = "bold", size = 3) +
  facet_wrap(~ niche, nrow = 1) +
  scale_fill_manual(
    name   = "Direction",
    values = c("Negative β" = "#CA562C", "Positive β" = "#2ca0ca", "≈ 0" = "grey80"),
    guide  = guide_legend(order = 1, override.aes = list(size = 6))  # bigger dots
  ) +
  scale_size_continuous(
    name   = "Effect size |β/SE|",
    limits = size_limits,
    breaks = size_breaks,
    range  = c(3, 12),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = 21, fill = "grey70", color = "grey70", alpha = 1) # draw circles
    )
  ) +
  labs(x = "Lung function change", y = "α-diversity change",
       caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        panel.grid  = element_blank(),
        legend.key.height = grid::unit(6, "mm"),
        legend.key.width  = grid::unit(8, "mm"))


ggplot2::ggsave("../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_diversity_change_lung_func_change2.svg",
                p_cont_bubble, width = 7, height = 4)






# ============================================================
# Bubble plot — categorical models (lung category change ~ αΔ)
# ============================================================

# Contrast labels + desired x-axis order:
contrast_labs_map <- c(
  worse_vs_no_change  = "Worse",
  better_vs_no_change = "Better"
)

lung_order_cat <- c(
  "Δ FEV1 (category)",
  "Δ FVC (category)",
  "Δ FEV1/FVC (category)",
  "Δ FEF25–75 (category)"
)
contrast_order <- c("Worse","Better")
combo_levels <- unlist(lapply(lung_order_cat, function(lv) paste(lv, contrast_order, sep = "\n")))



# ---- Labels & desired x-axis order (flipped) ----
contrast_labs_map <- c(
  worse_vs_no_change  = "Worse",
  better_vs_no_change = "Better"
)

lung_order_cat <- c(
  "Δ FEV1 (category)",
  "Δ FVC (category)",
  "Δ FEV1/FVC (category)",
  "Δ FEF25–75 (category)"
)

# We want: "Worse Δ FEV1 (category)", then "Better Δ FEV1 (category)" for each lung metric
combo_levels <- unlist(lapply(lung_order_cat, function(lv) c(paste("Worse", lv), paste("Better", lv))))


# Color palette (robust in print and for most color-vision deficiencies)
pal_dir <- c(
  "Negative β" = "#CA562C",  # deep blue
  "Positive β" = "#2ca0ca",  # deep red
  "≈ 0"       = "grey80"
)

# Prepare plotting frame
res_cat_plot <- res_categorical %>%
  dplyr::mutate(
    alpha = factor(outcome, levels = names(alpha_labs_map), labels = unname(alpha_labs_map)),
    lung  = factor(predictor, levels = names(cat_pred_labs_map), labels = unname(cat_pred_labs_map)),
    contrast_lab = dplyr::recode(contrast, !!!contrast_labs_map),

    # single-line text to match combo_levels
    ylab = factor(paste(contrast_lab, lung), levels = combo_levels),

    stars = dplyr::case_when(p_adj < 0.001 ~ "***", p_adj < 0.01 ~ "**", p_adj < 0.05 ~ "*", TRUE ~ ""),
    se    = (conf.high - conf.low) / (2 * 1.96),
    z     = estimate / se,
    zabs  = abs(z),
    zabs_cap = pmin(pmax(zabs, 0), 4),
    sign  = dplyr::case_when(estimate > 0 ~ "Positive β",
                             estimate < 0 ~ "Negative β",
                             TRUE         ~ "≈ 0")
  )



p_cat_bubble <- ggplot2::ggplot(res_cat_plot, ggplot2::aes(x = alpha, y = ylab )) +
  ggplot2::geom_point(
    ggplot2::aes(size = zabs_cap, fill = sign),   # <- use clamped size
    shape = 21, color = "white", stroke = 0.5
  ) +
  ggplot2::geom_text(ggplot2::aes(label = stars), vjust = 0.5, fontface = "bold", size = 3) +
  ggplot2::facet_wrap(~ niche, nrow = 1) +
  ggplot2::scale_fill_manual(
    values = pal_dir,
    name   = "Direction",
    guide  = ggplot2::guide_legend(
      title.position = "top",
      override.aes = list(shape = 21, size = 7, colour = "white", stroke = 0.6)
    )
  ) +
  ggplot2::scale_size_continuous(
    name   = expression(paste("Effect size  |", beta, "/SE|")),
    limits = c(0, 4),
    breaks = c(0.5, 1, 2, 3, 4),
    labels = c("0.5", "1.0", "2.0", "3.0", "4.0"),
    range  = c(3, 11),
    guide  = ggplot2::guide_legend(
      title.position = "top",
      override.aes = list(fill = "grey60", shape = 21, colour = "white", stroke = 0.6)
    )
  ) +
  ggplot2::labs(
    x = "Lung function category change (vs no change)",
    y = "α-diversity change",
    caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    axis.text.x  = ggplot2::element_text(angle = 40, hjust = 1),
    panel.grid   = ggplot2::element_blank(),
    legend.box   = "vertical",
    legend.title = ggplot2::element_text(face = "bold"),
    legend.key.height = grid::unit(0.7, "cm"),
    legend.key.width  = grid::unit(1.1, "cm")
  ) 


print(p_cat_bubble)

ggplot2::ggsave(
  "../Output/Alpha_diversity/Longitudinal_analysis/Figures/Alpha_diversity_change_lung_cat_change.svg",
  p_cat_bubble, width = 8, height = 5
)

```

# Beta diversity
## NP within vs between individual composition
```{r}
id_var <- "VGO3_ID"

# Subset & merge (fix the typo) 
VGO2_NP_ps <- subset_samples(ps_final, study == "VGO2" & niche == "NP")
VGO3_NP_ps <- subset_samples(ps_final, study == "VGO3" & niche == "NP") 
VGO_NP_combined <- merge_phyloseq(VGO2_NP_ps, VGO3_NP_ps)

# Filter & RA, then Bray–Curtis 
VGO_NP_RA <- VGO_NP_combined %>% pres_abund_filter() %>% to_RA()
np_dist   <- phyloseq::distance(VGO_NP_RA, method = "bray")
meta_np   <- sample_data(VGO_NP_RA) %>% data.frame(check.names = FALSE)

# Long pairwise distances + subject IDs 
np_pairs <- as.matrix(np_dist) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample1") %>%
  tidyr::pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
  dplyr::filter(Sample1 < Sample2) %>%                        # unique pairs
  dplyr::mutate(
    ID1 = meta_np[Sample1, id_var],
    ID2 = meta_np[Sample2, id_var]
  ) %>%
  dplyr::filter(!is.na(ID1) & !is.na(ID2)) %>%                # drop NA IDs
  dplyr::mutate(Comparison = ifelse(ID1 == ID2, "Within", "Between"))

# Sanity check: need both groups 
if (!all(c("Within","Between") %in% unique(np_pairs$Comparison))) {
  stop("You don't have both 'Within' and 'Between' pairs. Check that VGO2 & VGO3 NP samples exist and IDs are populated.")
}

# Balance the groups (downsample larger to the size of the smaller)
set.seed(123)
n_to_sample <- min(table(np_pairs$Comparison))
np_bal <- np_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::slice_sample(n = n_to_sample) %>%
  dplyr::ungroup()

# Y limits from boxplot whiskers (avoid NA)
whisk <- np_bal %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    whisker_low  = boxplot.stats(Distance)$stats[1],
    whisker_high = boxplot.stats(Distance)$stats[5],
    .groups = "drop"
  )

y_min <- min(whisk$whisker_low,  na.rm = TRUE)
y_max <- min(1, max(whisk$whisker_high, na.rm = TRUE))
y_max_ext <- y_max + 0.03 * (y_max - y_min)
p_y <- y_max_ext - 0.01 * (y_max_ext - y_min)

# --- Test & label ---
wt <- wilcox.test(Distance ~ Comparison, data = np_bal, exact = FALSE)
label <- ifelse(is.finite(wt$p.value) && wt$p.value < 0.001,
                "italic(p) < 0.001",
                paste0("italic(p) == ", sprintf('%.3f', wt$p.value)))

# --- Plot ---
np_plot <- ggplot(np_bal, aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_jitter(aes(color = Comparison), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("Within" = "#ad7c8a", "Between" = "#d9b9c2")) +
  scale_color_manual(values = c("Within" = "#ad7c8a", "Between" = "#d9b9c2")) +
  coord_cartesian(ylim = c(y_min, y_max_ext)) +
  labs(
    title = "NP: Within vs Between-Individual Bray–Curtis\n (VGO2 vs VGO3)",
    x = "Comparison Type", y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = label, y_position = p_y,
    xmin = 1, xmax = 2, tip_length = 0.01, parse = TRUE
  )

ggsave("../Output/Beta_diversity/Longitudinal_analysis/NP_jitter_boxplot_balanced.svg",
       np_plot, width = 5, height = 5)

```
## OP within vs between individual composition
```{r}
id_var <- "VGO3_ID"

# --- Subset & merge (fix the typo) ---
VGO2_OP_ps <- subset_samples(ps_final, study == "VGO2" & niche == "OP")
VGO3_OP_ps <- subset_samples(ps_final, study == "VGO3" & niche == "OP")  # fixed
VGO_OP_combined <- merge_phyloseq(VGO2_OP_ps, VGO3_OP_ps)

# --- Filter & RA, then Bray–Curtis ---
VGO_OP_RA <- VGO_OP_combined %>% pres_abund_filter() %>% to_RA()
op_dist   <- phyloseq::distance(VGO_OP_RA, method = "bray")
meta_op   <- sample_data(VGO_OP_RA) %>% data.frame(check.names = FALSE)

#  Long pairwise distances + subject IDs 
op_pairs <- as.matrix(op_dist) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Sample1") %>%
  tidyr::pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
  dplyr::filter(Sample1 < Sample2) %>%                        # unique pairs
  dplyr::mutate(
    ID1 = meta_op[Sample1, id_var],
    ID2 = meta_op[Sample2, id_var]
  ) %>%
  dplyr::filter(!is.na(ID1) & !is.na(ID2)) %>%                # drop NA IDs
  dplyr::mutate(Comparison = ifelse(ID1 == ID2, "Within", "Between"))

#  Sanity check: need both groups 
if (!all(c("Within","Between") %in% unique(op_pairs$Comparison))) {
  stop("You don't have both 'Within' and 'Between' pairs. Check that VGO2 & VGO3 NP samples exist and IDs are populated.")
}

#  Balance the groups (downsample larger to the size of the smaller) 
set.seed(123)
n_to_sample <- min(table(op_pairs$Comparison))
op_bal <- op_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::slice_sample(n = n_to_sample) %>%
  dplyr::ungroup()

# Y limits from boxplot whiskers (avoid NA)
whisk <- op_bal %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    whisker_low  = boxplot.stats(Distance)$stats[1],
    whisker_high = boxplot.stats(Distance)$stats[5],
    .groups = "drop"
  )

y_min <- min(whisk$whisker_low,  na.rm = TRUE)
y_max <- min(1, max(whisk$whisker_high, na.rm = TRUE))
y_max_ext <- y_max + 0.03 * (y_max - y_min)
p_y <- y_max_ext - 0.01 * (y_max_ext - y_min)

# Test & label 
wt <- wilcox.test(Distance ~ Comparison, data = op_bal, exact = FALSE)
label <- ifelse(is.finite(wt$p.value) && wt$p.value < 0.001,
                "italic(p) < 0.001",
                paste0("italic(p) == ", sprintf('%.3f', wt$p.value)))

# Plot
op_plot <- ggplot(op_bal, aes(x = Comparison, y = Distance, fill = Comparison)) +
  geom_jitter(aes(color = Comparison), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("Within" = "#3f7566", "Between" = "#b2c8c2")) +
  scale_color_manual(values = c("Within" = "#3f7566", "Between" = "#b2c8c2")) +
  coord_cartesian(ylim = c(y_min, y_max_ext)) +
  labs(
    title = "OP: Within vs Between-Individual Bray–Curtis\n (VGO2 vs VGO3)",
    x = "Comparison Type", y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = label, y_position = p_y,
    xmin = 1, xmax = 2, tip_length = 0.01, parse = TRUE
  )

ggsave("../Output/Beta_diversity/Longitudinal_analysis/OP_jitter_boxplot_balanced.svg",
       op_plot, width = 5, height = 5)
```

## Combined plots NP and OP within vs between composition
```{r}
combined_plot <- np_plot + op_plot +
  plot_layout(ncol = 2) +
  plot_annotation(tag_levels = "A")

# Save combined plot
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NP_OP_jitter_boxplot_balanced.svg",
       combined_plot, width = 10, height = 5)
```
# NP vs OP 
## BC distances and visualisation
```{r}
# Helper function 
get_between_individual_distances <- function(ps_obj, niche_label) {
  # Filter and convert to relative abundance
  ps_RA <- ps_obj %>% pres_abund_filter() %>% to_RA()
  dist_mat <- phyloseq::distance(ps_RA, method = "bray")
  
  meta <- data.frame(sample_data(ps_RA))
  meta$SampleID <- rownames(meta)
  
  dist_df <- as.matrix(dist_mat) %>%
    as.data.frame() %>%
    rownames_to_column("Sample1") %>%
    pivot_longer(-Sample1, names_to = "Sample2", values_to = "Distance") %>%
    filter(Sample1 < Sample2) %>%  # Unique pairs only
    mutate(
      ID1 = meta[Sample1, "VGO3_ID"],
      ID2 = meta[Sample2, "VGO3_ID"],
      Comparison = ifelse(ID1 == ID2, "Within", "Between")
    ) %>%
    filter(Comparison == "Between") %>%
    mutate(Niche = niche_label)
  
  return(dist_df)
}

# Get distances
# Subset and prepare NP and OP datasets (VGO2+VGO3)
VGO_NP <- subset_samples(ps_final, niche == "NP" & dataset %in% c("VGO2_NP", "VGO3_NPOP"))
VGO_OP <- subset_samples(ps_final, niche == "OP" & dataset %in% c("VGO2_OP", "VGO3_NPOP"))

np_between <- get_between_individual_distances(VGO_NP, "NP")
op_between <- get_between_individual_distances(VGO_OP, "OP")

combined_between <- bind_rows(np_between, op_between)

# Optional: Downsample to balance group sizes
set.seed(123)
min_n <- min(table(combined_between$Niche))
combined_between_bal <- combined_between %>%
  group_by(Niche) %>%
  slice_sample(n = min_n) %>%
  ungroup()

# -- Wilcoxon test -- #
wt <- wilcox.test(Distance ~ Niche, data = combined_between_bal)
p_label <- ifelse(
  wt$p.value < 0.001,
  "italic(p) < 0.001",
  paste0("italic(p) == ", sprintf("%.3f", wt$p.value))
)

# -- Plot -- #
y_min <- min(combined_between_bal$Distance)
y_max <- max(combined_between_bal$Distance)
y_max_extended <- y_max + 0.03 * (y_max - y_min)
pval_y_pos <- y_max_extended - 0.01 * (y_max_extended - y_min)

between_plot <- ggplot(combined_between_bal, aes(x = Niche, y = Distance, fill = Niche)) +
  geom_jitter(aes(color = Niche), width = 0.2, alpha = 0.5, size = 1) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_fill_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
  scale_color_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
  coord_cartesian(ylim = c(y_min, y_max_extended)) +
  labs(
    title = "Between-Individual Bray–Curtis Dissimilarity",
    subtitle = "Comparison between NP and OP samples (VGO2 + VGO3)",
    x = "Niche",
    y = "Bray–Curtis Dissimilarity"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ggsignif::geom_signif(
    annotations = p_label,
    y_position = pval_y_pos,
    xmin = 1, xmax = 2,
    tip_length = 0.01,
    parse = TRUE
  )

# Save
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NPvsOP/Between_individual_NP_OP_comparison.png", between_plot, width = 6, height = 5)

# Show
print(between_plot)

```
## PERMANOVA NP vs OP
```{r}
# Subset VGO2 and VGO3 data
vgo2_ps <- subset_samples(ps_final, dataset %in% c("VGO2_NP", "VGO2_OP") & niche == c("NP", "OP"))
vgo3_ps <- subset_samples(ps_final, dataset == "VGO3_NPOP" & niche == c("NP", "OP"))

compare_NP_OP <- function(ps_obj, label) {
  ps_filt <- ps_obj %>%
    pres_abund_filter() %>%
    to_RA()

  dist <- phyloseq::distance(ps_filt, method = "bray")
  meta <- data.frame(sample_data(ps_filt))

  # Run PERMANOVA
  ad <- adonis2(dist ~ niche, data = meta, permutations = 999)
  pval <- ad$`Pr(>F)`[1]
  r2 <- round(ad$R2[1], 3)
  p_label <- ifelse(pval < 0.001, "italic(p)<0.001", paste0("italic(p)==", sprintf("%.3f", pval)))

  # Ordination plot with manual colors
  ord <- ordinate(ps_filt, method = "PCoA", distance = dist)
  plot <- plot_ordination(ps_filt, ord, color = "niche") +
    geom_point(size = 2, alpha = 0.7) +
    scale_color_manual(values = c("NP" = "#ad7c8a", "OP" = "#3f7566")) +
    labs(
      title = paste0(label, ": NP vs OP (Bray-Curtis)"),
      subtitle = paste0("PERMANOVA R² = ", r2),
      color = "Niche"
    ) +
    annotate("text", x = Inf, y = Inf, label = p_label, parse = TRUE, hjust = 1.1, vjust = 1.5) +
    theme_minimal(base_size = 12)

  return(plot)
}

# Create both plots
vgo2_plot <- compare_NP_OP(vgo2_ps, "VGO2")
vgo3_plot <- compare_NP_OP(vgo3_ps, "VGO3")

# Combine plots side-by-side
combined_np_op_plot <- vgo2_plot + vgo3_plot + 
  plot_layout(ncol = 2) + 
  plot_annotation(tag_levels = "A")

# Save
ggsave("../Output/Beta_diversity/Longitudinal_analysis/NPvsOP/NP_OP_PERMANOVA_ordination.svg",
       combined_np_op_plot, width = 10, height = 5)

# Show
print(combined_np_op_plot)
```
# Inter-individual differences 
```{r}
# --- NP: mean (SD) within vs between ---
np_desc <- np_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    n    = dplyr::n(),
    mean = mean(Distance, na.rm = TRUE),
    sd   = sd(Distance,   na.rm = TRUE),
    .groups = "drop"
  )

# --- OP: mean (SD) within vs between ---
op_desc <- op_pairs %>%
  dplyr::group_by(Comparison) %>%
  dplyr::summarise(
    n    = dplyr::n(),
    mean = mean(Distance, na.rm = TRUE),
    sd   = sd(Distance,   na.rm = TRUE),
    .groups = "drop"
  )

np_desc; op_desc
```

# Change in beta vs change in lung function
```{r}
# From my existing dataframes np_pairs / op_pairs, extract each person’s within-person Bray–Curtis (distance between their own VGO2 and VGO3 samples) separately for NP and OP
bc_np <- np_pairs %>%
  dplyr::filter(Comparison == "Within") %>%
  dplyr::transmute(VGO3_ID = ID1, niche = "NP", BC_within = Distance)

bc_op <- op_pairs %>%
  dplyr::filter(Comparison == "Within") %>%
  dplyr::transmute(VGO3_ID = ID1, niche = "OP", BC_within = Distance)

bc_within <- dplyr::bind_rows(bc_np, bc_op)

# Join to dataframe containing lung function change variables etc. 
df_beta <- df_wide_models %>%
  dplyr::left_join(bc_within, by = c("VGO3_ID","niche"))



# Continuous lung function variables - association with BC distances
lung_slopes <- c("FEV1_pred_per_year","FVC_pred_per_year",
                 "FEV1FVC_pred_per_year","FEF25_75_pred_per_year")

fit_beta_cont <- function(dat, slope, robust_se = FALSE) {
  # Prepare data
  dat2 <- dat %>%
    mutate(
      season_change = factor(season_change, levels = c("same","different")),
      smoke_change  = factor(smoke_change,  levels = c("same","different")),
      BC_centered   = scale(BC_within, center = TRUE, scale = FALSE)[,1]
    )

  lung0 <- sub("_per_year$", "_VGO2", slope)

  fml <- as.formula(paste(
    slope, "~ BC_centered +",
    paste(c(lung0, "age_VGO2", "season_change", "smoke_change"), collapse = " + ")
  ))

  dd <- dat2 %>% dplyr::select(dplyr::all_of(all.vars(fml))) %>% tidyr::drop_na()
  if (nrow(dd) < 30) return(NULL)

  fit <- lm(fml, dd)

  if (!robust_se) {
    out <- broom::tidy(fit, conf.int = TRUE)
  } else {
    # HC3 robust SEs (optional)
    library(sandwich); library(lmtest)
    ct <- lmtest::coeftest(fit, vcov = sandwich::vcovHC(fit, type = "HC3"))
    out <- tibble::tibble(
      term = rownames(ct),
      estimate = ct[,1],
      std.error = ct[,2],
      statistic = ct[,3],
      p.value   = ct[,4]
    ) %>%
      mutate(
        conf.low  = estimate - 1.96 * std.error,
        conf.high = estimate + 1.96 * std.error
      )
  }

  out %>%
    dplyr::filter(term == "BC_centered") %>%
    dplyr::mutate(n = nrow(dd), outcome = slope)
}

# Run per niche
res_beta_np <- lapply(lung_slopes, \(v) fit_beta_cont(df_beta %>% filter(niche=="NP"), v)) %>%
  bind_rows() %>% mutate(niche = "NP")

res_beta_op <- lapply(lung_slopes, \(v) fit_beta_cont(df_beta %>% filter(niche=="OP"), v)) %>%
  bind_rows() %>% mutate(niche = "OP")

res_beta_cont <- bind_rows(res_beta_np, res_beta_op) %>%
  group_by(niche) %>% mutate(p_adj = p.adjust(p.value, "BH")) %>% ungroup()

readr::write_csv(res_beta_cont,
                 "../Output/Beta_diversity/Longitudinal_analysis/BC_within_vs_lung_continuous.csv")








# Categorical lung function variables - outcome = lung category change
lung_cats <- c("FEV1_category_change","FVC_category_change",
               "FEV1FVC_category_change","FEF25_75_category_change")

fit_beta_cat_ord <- function(dat, catvar) {
  dd <- dat %>%
    dplyr::mutate(
      season_change = factor(season_change, levels = c("same","different")),
      smoke_change  = factor(smoke_change,  levels = c("same","different")),
      BC_centered   = scale(BC_within, center = TRUE, scale = FALSE)[,1],
      !!catvar := factor(.data[[catvar]], levels = c("no_change","better","worse"), ordered = TRUE)
    ) %>%
    dplyr::select(BC_centered, dplyr::all_of(catvar), age_VGO2, season_change, smoke_change) %>%
    tidyr::drop_na()

  if (nrow(dd) < 30 || nlevels(dd[[catvar]]) < 2) return(NULL)

  fml <- stats::as.formula(paste(
    catvar, "~ BC_centered + age_VGO2 + season_change + smoke_change"
  ))

  fit <- tryCatch(MASS::polr(fml, data = dd, Hess = TRUE, method = "logistic"),
                  error = function(e) NULL)
  if (is.null(fit)) return(NULL)

  V  <- stats::vcov(fit); co <- coef(fit)
  if (!"BC_centered" %in% names(co)) return(NULL)
  se <- sqrt(diag(V))["BC_centered"]
  est <- co["BC_centered"]; z <- est / se

  tibble::tibble(
    predictor = catvar,
    term      = "BC_centered",
    estimate  = est,
    std.error = se,
    statistic = z,
    p.value   = 2*pnorm(abs(z), lower.tail = FALSE),
    conf.low  = est - 1.96*se,
    conf.high = est + 1.96*se,
    n         = nrow(dd)
  )
}

res_beta_cat_np <- lapply(lung_cats, \(v) fit_beta_cat_ord(df_beta %>% dplyr::filter(niche=="NP"), v)) %>%
  dplyr::bind_rows() %>% dplyr::mutate(niche="NP")

res_beta_cat_op <- lapply(lung_cats, \(v) fit_beta_cat_ord(df_beta %>% dplyr::filter(niche=="OP"), v)) %>%
  dplyr::bind_rows() %>% dplyr::mutate(niche="OP")

res_beta_cat <- dplyr::bind_rows(res_beta_cat_np, res_beta_cat_op) %>%
  dplyr::group_by(niche, predictor) %>% dplyr::mutate(p_adj = p.adjust(p.value, "BH")) %>%
  dplyr::ungroup()


readr::write_csv(res_beta_cat,
                 "../Output/Beta_diversity/Longitudinal_analysis/BC_within_vs_lung_categorical.csv")







# Visual check of associations between lung fucntion variables (continuous and BC distance)
# NP 
## ΔFEV1 
p_np_fev1 <- ggplot(df_beta %>% dplyr::filter(niche=="NP"), aes(BC_within, FEV1_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEV1 per year", title="NP")

## ΔFVC
p_np_fvc <- ggplot(df_beta %>% dplyr::filter(niche=="NP"), aes(BC_within, FVC_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFVC per year", title="NP")

## ΔFEV1/FVC
p_np_fev1fvc <- ggplot(df_beta %>% dplyr::filter(niche=="NP"), aes(BC_within, FEV1FVC_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEV1/FVC per year", title="NP")

## ΔFEF25-75% 
p_np_fef <- ggplot(df_beta %>% dplyr::filter(niche=="NP"), aes(BC_within, FEF25_75_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEF25-75% per year", title="NP")

# OP 
## ΔFEV1 
p_op_fev1 <- ggplot(df_beta %>% dplyr::filter(niche=="OP"), aes(BC_within, FEV1_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEV1 per year", title="OP")

## ΔFVC
p_op_fvc <- ggplot(df_beta %>% dplyr::filter(niche=="OP"), aes(BC_within, FVC_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFVC per year", title="OP")

## ΔFEV1/FVC
p_op_fev1fvc <- ggplot(df_beta %>% dplyr::filter(niche=="OP"), aes(BC_within, FEV1FVC_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEV1/FVC per year", title="OP")

## ΔFEF25-75% 
p_op_fef <- ggplot(df_beta %>% dplyr::filter(niche=="OP"), aes(BC_within, FEF25_75_pred_per_year)) +
  geom_point(alpha=.5) + geom_smooth(method="lm") + theme_bw() +
  labs(x="Within-person Bray–Curtis (VGO2–VGO3)", y="ΔFEF25-75% per year", title="OP")


combined <-
  (p_np_fev1 | p_np_fvc | p_np_fev1fvc | p_np_fef) /
  (p_op_fev1 | p_op_fvc | p_op_fev1fvc | p_op_fef) +
  plot_annotation(
    title = "BC distance vs lung function change",
    subtitle = "Top: NP, Bottom: OP"
  ) &
  theme(legend.position = "bottom")

combined
ggsave("../Output/Beta_diversity/Longitudinal_analysis/BC_vs_lung_change_plots.svg", combined, width = 14, height = 7)





# VIsualisation of categorical results

plot_empirical_props <- function(dat, catvar, niche_title, nbins = 6) {
  dd <- dat %>%
    dplyr::mutate(
      !!catvar := factor(.data[[catvar]],
                         levels = c("no_change","better","worse"),
                         ordered = TRUE)
    ) %>%
    dplyr::select(BC_within, dplyr::all_of(catvar)) %>%
    tidyr::drop_na()

  if (nrow(dd) < 30 || nlevels(dd[[catvar]]) < 2) {
    return(ggplot() + labs(title = paste(niche_title, "-", catvar, "(insufficient data)")))
  }

  dd <- dd %>% dplyr::mutate(bin = cut(BC_within, breaks = nbins, include.lowest = TRUE))

  props <- dd %>%
    dplyr::group_by(bin, !!rlang::sym(catvar)) %>%
    dplyr::summarise(
      n   = dplyr::n(),
      x   = mean(BC_within),
      .groups = "drop_last"
    ) %>%
    dplyr::mutate(Nbin = sum(n), prop = n / Nbin) %>%
    dplyr::ungroup() %>%
    dplyr::rename(category = !!rlang::sym(catvar))

  ggplot(props, aes(x = x, y = prop, color = category)) +
    geom_point() +
    geom_line() +
    theme_bw() +
    labs(x = "Within-person Bray–Curtis (VGO2–VGO3)",
         y = "Observed proportion",
         title = paste(niche_title, "(empirical) -", catvar),
         color = "Category")
}

lung_cats <- c("FEV1_category_change","FVC_category_change",
               "FEV1FVC_category_change","FEF25_75_category_change")

plots_np_emp <- lapply(lung_cats, function(v)
  plot_empirical_props(df_beta %>% dplyr::filter(niche=="NP"), v, "NP"))

plots_op_emp <- lapply(lung_cats, function(v)
  plot_empirical_props(df_beta %>% dplyr::filter(niche=="OP"), v, "OP"))

# No clear pattern in the NP 
# No clear pattern in the OP either 
```

# Change in cluster composition
## Data 

```{r}
# Separate microbiome data per niche (not separate per dataset as before)
# Subset per study and niche
# NP niche
microbiome_NP <- subset_samples(microbiome_all, dataset == c("VGO2_NP", "VGO3_NPOP") & niche == "NP")
nsamples(microbiome_NP) # 503 NP samples

# OP niche
microbiome_OP <- subset_samples(microbiome_all, dataset == c("VGO2_OP","VGO3_NPOP") & niche == "OP")
nsamples(microbiome_OP) # 568 OP samples
```
## Filter taxa and convert to relative abundance
```{r}
# VGO2 NP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_np_filtered <- microbiome_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 967 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22669 ASVs excluded).

# VGO2 OP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_op_filtered <- microbiome_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 879 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22757 ASVs excluded).
```
## NP clustering VGO2 and VGO3 combined
```{r}
# Check number of taxa
ntaxa(ps_np_filtered) # 967 ASVs 
# Extract metadata
meta_np <- meta_to_df(ps_np_filtered, sample_name = ".row_id")
# Extract otu table
otu_np <- as(otu_table(ps_np_filtered), "matrix")
# Standardise orientation so that downstream analyses always have rows = samples and columns = ASVs
if(taxa_are_rows(ps_np_filtered)){otu_np <- t(otu_np)} 
otu_np <- as.data.frame(otu_np)
# Compute BC distances
ord_np <- vegdist(otu_np, method = "bray") 

# We used complete linkage to create the phylogenetic tree (hierarchical clustering). The method is based on maximum distance; the similarity of any two clusters is the similarity of their most dissimilar pair.
hc_np_clust <- hclust(ord_np, method = "complete")
# Reorder 'otu_np' based on the clustering order
otu_hc_np_order_clust <- otu_np[hc_np_clust$order,]

# The Calinski-Harabasz and Silhouette methods were used to determine the appropriate location to cut the tree. Higher value of indices means the clusters are dense and well separated. Here, we looked for a peak or at least an abrupt elbow on the line plot of the indices.
CalHara_sil_indices_np <- clust_ind_plot(ord_np, seq(2, 25, by = 1), clust_method="complete")
write.csv(CalHara_sil_indices_np$data, "../Output/Cluster_analysis/NP_all/CalHara_silhouette_indices_vgo2_np.csv")

# Calinski-Harabasz Index: A higher value indicates well-separated clusters with high intra-cluster similarity.
# Silhouette: The optimal number of clusters typically shows the highest silhouette score, suggesting well-separated and compact clusters.
# look for the points where there are peaks or elbows in the curve

# Create dendogram
dend_np <- as.dendrogram(hc_np_clust)

# 2 clusters
dend_2clusters <- color_branches(dend_np, k = 2, groupLabels=T)
(stats::dendrapply(dend_2clusters, noLabel))
cluster_assignments_2 <- cutree(dend_np, k = 2)
cluster_counts_2 <- table(cluster_assignments_2)
print(cluster_counts_2)

# 3 clusters
dend_3clusters <- color_branches(dend_np, k = 3, groupLabels=T)
(stats::dendrapply(dend_3clusters, noLabel))
cluster_assignments_3 <- cutree(dend_np, k = 3)
cluster_counts_3 <- table(cluster_assignments_3)
print(cluster_counts_3)
 
# 4 clusters
dend_4clusters <- color_branches(dend_np, k = 4, groupLabels=T)
(stats::dendrapply(dend_4clusters, noLabel))
cluster_assignments_4 <- cutree(dend_np, k = 4)
cluster_counts_4 <- table(cluster_assignments_4)
print(cluster_counts_4)

# 5 clusters
dend_5clusters <- color_branches(dend_np, k = 5, groupLabels=T)
plot(stats::dendrapply(dend_5clusters, noLabel))
cluster_assignments_5 <- cutree(dend_np, k = 5)
cluster_counts_5 <- table(cluster_assignments_5)
print(cluster_counts_5)

# 6 clusters
dend_6clusters <- color_branches(dend_np, k = 6, groupLabels=T)
plot(stats::dendrapply(dend_6clusters, noLabel))
cluster_assignments_6 <- cutree(dend_np, k = 6)
cluster_counts_6 <- table(cluster_assignments_6)
print(cluster_counts_6)

# 7 clusters
dend_7clusters <- color_branches(dend_np, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_np, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# 8 clusters
dend_8clusters <- color_branches(dend_np, k = 8, groupLabels=T)
plot(stats::dendrapply(dend_8clusters, noLabel))
cluster_assignments_8 <- cutree(dend_np, k = 8)
cluster_counts_8 <- table(cluster_assignments_8)
print(cluster_counts_8)

# 9 clusters
dend_9clusters <- color_branches(dend_np, k = 9, groupLabels=T)
plot(stats::dendrapply(dend_9clusters, noLabel))
cluster_assignments_9 <- cutree(dend_np, k = 9)
cluster_counts_9 <- table(cluster_assignments_9)
print(cluster_counts_9)

# 10 clusters
dend_10clusters <- color_branches(dend_np, k = 10, groupLabels=T)
plot(stats::dendrapply(dend_10clusters, noLabel))
cluster_assignments_10 <- cutree(dend_np, k = 10)
cluster_counts_10 <- table(cluster_assignments_10)
print(cluster_counts_10)

# 11 clusters
dend_11clusters <- color_branches(dend_np, k = 11, groupLabels=T)
plot(stats::dendrapply(dend_11clusters, noLabel))
cluster_assignments_11 <- cutree(dend_np, k = 11)
cluster_counts_11 <- table(cluster_assignments_11)
print(cluster_counts_11)

# 12 clusters
dend_12clusters <- color_branches(dend_np, k = 12, groupLabels=T)
plot(stats::dendrapply(dend_12clusters, noLabel))
# check number per cluster
cluster_assignments_12 <- cutree(dend_np, k = 12)
cluster_counts_12 <- table(cluster_assignments_12)
print(cluster_counts_12)

# 13 clusters
dend_13clusters <- color_branches(dend_np, k = 13, groupLabels=T)
plot(stats::dendrapply(dend_13clusters, noLabel))
cluster_assignments_13 <- cutree(dend_np, k = 13)
cluster_counts_13 <- table(cluster_assignments_13)
print(cluster_counts_13)

# 14 clusters
dend_14clusters <- color_branches(dend_np, k = 14, groupLabels=T)
plot(stats::dendrapply(dend_14clusters, noLabel))
cluster_assignments_14 <- cutree(dend_np, k = 14)
cluster_counts_14 <- table(cluster_assignments_14)
print(cluster_counts_14)

# 15 clusters
dend_15clusters <- color_branches(dend_np, k = 15, groupLabels=T)
plot(stats::dendrapply(dend_15clusters, noLabel))
cluster_assignments_15 <- cutree(dend_np, k = 15)
cluster_counts_15 <- table(cluster_assignments_15)
print(cluster_counts_15)


# 7 clusters appears optimal
# k = 7
dend_7clusters <- color_branches(dend_np, k = 7, groupLabels=T)
plot(stats::dendrapply(dend_7clusters, noLabel))
cluster_assignments_7 <- cutree(dend_7clusters, k = 7)
cluster_counts_7 <- table(cluster_assignments_7)
print(cluster_counts_7)

# The function 'cuttree' is used to cut the tree at k = 4
groups_np_c7 <- cutree(hc_vgo2_np_clust, k = 7)
ps_np_filtered@sam_data$cluster <- groups_np_c7[order(rownames(ps_np_filtered@sam_data))]
ps_np_filtered@sam_data$cluster <- ps_np_filtered@sam_data$cluster %>% as.character()

# Determine the top 3 ASVs for each cluster
otu_np <- as(otu_table(ps_np_filtered), "matrix")
# if (taxa_are_rows(ps_vgo2_np_filtered)) {
#   otu_hc_vgo2_np_order_clust <- t(otu_hc_vgo2_np_order_clust)  # Ensure ASVs are columns, samples are rows
# }

# Extract sample metadata with cluster information
sample_data_np <- data.frame(ps_np_filtered@sam_data)

# Combine the OTU table with the sample cluster data
otu_np_with_clusters <- cbind(cluster = sample_data_np$cluster, otu_hc_np_order_clust)
otu_np_with_clusters <- as.data.frame(otu_np_with_clusters)
otu_np_with_clusters$cluster <- as.factor(otu_np_with_clusters$cluster)
# Convert all ASV abundances to numeric
otu_np_with_clusters[-1] <- lapply(otu_np_with_clusters[-1], as.numeric)
# Sum ASV abundances for each cluster
asv_sum_by_cluster_np <- otu_np_with_clusters %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))
# Convert back to data frame for further processing
asv_sum_by_cluster_np <- as.data.frame(asv_sum_by_cluster_np)
# Find the top 3 most abundant ASVs in each cluster and their percentage
top_3_asvs_np <- apply(asv_sum_by_cluster_np[, -1], 1, function(x) {
  # Get total abundance in the cluster for calculating percentages
  total_abundance_np <- sum(x, na.rm = TRUE)
  # Get names and values of the top 3 ASVs
  top_3_indices_np <- order(x, decreasing = TRUE)[1:3]
  top_3_asv_names_np <- names(x)[top_3_indices_np]
  top_3_asv_values_np <- x[top_3_indices_np]
  # Calculate percentages
  top_3_asv_percentages_np <- (top_3_asv_values_np / total_abundance_np) * 100
  # Combine ASV names and percentages into a string for each cluster
  result <- paste0(top_3_asv_names_np, ": ", round(top_3_asv_percentages_np, 2), "%", collapse = "; ")
  return(result)
})

# Combine results with cluster information
top_3_asvs_df_np <- data.frame(cluster = asv_sum_by_cluster_np$cluster, Top_3_ASVs = top_3_asvs_np)
# Display the top 3 ASVs for each cluster along with their percentages
print(top_3_asvs_df_np)

#### Rename clusters ####
# If an ASV is >40% of the total abundance in a cluster: Name the cluster after that ASV only, as it is the dominant contributor.
# If the top ASV is between 20-40% of the total abundance: Use the two most abundant ASVs to name the cluster.
# If the top ASV is <20% of the total abundance: Use the top three ASVs to name the cluster, as none of the ASVs dominate significantly.

get_cluster_name <- function(top_asvs_string) {
  asv_info <- strsplit(top_asvs_string, "; ")[[1]]
  asv_names <- sapply(asv_info, function(x) strsplit(x, ": ")[[1]][1])
  asv_percentages <- as.numeric(sapply(asv_info, function(x) sub("%", "", strsplit(x, ": ")[[1]][2])))
  
  if (asv_percentages[1] > 40) {
    return(paste0(asv_names[1], "-dominated cluster"))
  } else if (asv_percentages[1] > 20) {
    return(paste0(asv_names[1], " and ", asv_names[2], "-dominated cluster"))
  } else {
    return(paste0(asv_names[1], ", ", asv_names[2], ", and ", asv_names[3], "-dominated cluster"))
  }
}

# Rename based on these rules
top_3_asvs_df_np$Cluster_Name <- sapply(top_3_asvs_df_np$Top_3_ASVs, get_cluster_name)
top_3_asvs_df_np$Cluster_Name

# [1] "Corynebacterium_3, Staphylococcus_2, and Moraxella_4-dominated cluster"      
# [2] "Corynebacterium_3, Moraxella_4, and Staphylococcus_2-dominated cluster"      
# [3] "Staphylococcus_2, Corynebacterium_6, and Corynebacterium_3-dominated cluster"
# [4] "Staphylococcus_2-dominated cluster" 

# Shorten the names made above for plotting 
# Define the new cluster names in the same order as the original clusters
new_cluster_names_np <- c(
  "Cor(3)/Staph(2)/Mor(4) clust",
  "Cor(3)/Mor(4)/Staph(2) clust",
  "Staph(2)/Cor(6)/Cor(3) clust",
  "Staph(2) clust"
)

# Assign the new names to 'Cluster_Name' in 'top_3_asvs_df'
top_3_asvs_df_np$Cluster_Name <- new_cluster_names_np

# Update the clusters in the sample data to use the new names
ps_np_filtered@sam_data$cluster <- top_3_asvs_df_np$Cluster_Name[match(ps_np_filtered@sam_data$cluster, top_3_asvs_df_np$cluster)]
# Convert to character
ps_np_filtered@sam_data$cluster <- as.character(ps_np_filtered@sam_data$cluster)

# Count the number of samples in each cluster
cluster_counts_np <- as.data.frame(table(ps_np_filtered@sam_data$cluster))
colnames(cluster_counts_np) <- c("Cluster", "Count")

# Calculate the relative proportions of each cluster
cluster_counts_np$Proportion <- cluster_counts_np$Count / sum(cluster_counts_np$Count)

write.csv(cluster_counts_np, "../Output/Cluster_analysis/NP/cluster_counts_np.csv")

# Create the plot with geom_bar for vertical stacking
bar_clusters_np <- ggplot(data = cluster_counts_np, aes(x = Cluster, y = (Proportion*100), fill = Cluster)) +
  geom_bar(stat = "identity") +  # Use stat = "identity" to directly use Proportion
  labs(title = "Bar plot of clusters in VGO2 NP samples",
       x = "Clusters",            # X-axis now represents clusters
       y = "% of samples",          # Y-axis represents proportions
       fill = "Cluster") +
  scale_fill_manual(values = c(paletteer_d("pals::stepped")[c(13,20,10,16,1,18,21,5,12)])) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),      # Remove x-axis labels if needed
        axis.ticks.x = element_blank(),     # Remove x-axis ticks if needed
        legend.position = "right")

bar_clusters_np
ggsave("../Output/Cluster_analysis/NP/Bar_plot_cluster_proportions_clusters.svg", bar_clusters_np, width = 8, height = 5)

```

