---
title: "Cross-sectional microbiome-lung function analysis"
author: "Beatrice Cornu Hewitt"
date: "`r Sys.time()`"
output: html_document
---

Firstly we look at the relationship between teh microbiome and lung function cross-sectionally in 2014/15 (VGO2) and 2021/22 (VGO3)
# Packages
```{r}
library(phyloseq); library(microbiomer); library(ggplot2); library(cowplot); library(patchwork); library(scales); library(tidyverse); library(car); library(lmtest); library(ggsignif); library(forcats); library(rcartocolor); library(broom); library(writexl); library(vegan); library(tibble); library(glue); library(xlsx); library(corrplot); library(reshape2); library(RColorBrewer); library(Hmisc)
```
# Data 
```{r}
microbiome_all <- readRDS("../Output/Phyloseq/ps4.rds")
```
# Separate microbiome data per dataset and niche
```{r}
# Subset per study and niche
# VGO2_NP dataset, NP niche
microbiome_VGO2_NP <- subset_samples(microbiome_all, dataset == "VGO2_NP" & niche == "NP")
nsamples(microbiome_VGO2_NP) # 210 NP samples

# VGO2_OP dataset, OP niche
microbiome_VGO2_OP <- subset_samples(microbiome_all, dataset == "VGO2_OP" & niche == "OP")
nsamples(microbiome_VGO2_OP) # 179 OP samples

# VGO3_NPOP dataset, NP niche
microbiome_VGO3_NP <- subset_samples(microbiome_all, dataset == "VGO3_NPOP" & niche == "NP")
nsamples(microbiome_VGO3_NP) # 799 NP samples

# VGO3_NPOP dataset, OP niche
microbiome_VGO3_OP <- subset_samples(microbiome_all, dataset == "VGO3_NPOP" & niche == "OP")
nsamples(microbiome_VGO3_OP) # 952 OP samples
```
# Filter taxa and convert to relative abundance
```{r}
# VGO2 NP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo2_np_filtered <- microbiome_VGO2_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 893 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22743 ASVs excluded).

# VGO2 OP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo2_op_filtered <- microbiome_VGO2_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 337 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 23299 ASVs excluded).

# VGO3 NP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo3_np_filtered <- microbiome_VGO3_NP %>% 
  pres_abund_filter() %>% # 0.1% relative abundance in at least 2 samples is default
  to_RA()
# A total of 883 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22753 ASVs excluded).

# VGO3 OP 
# Remove low abundant ASVs and convert counts to relative abundances
ps_vgo3_op_filtered <- microbiome_VGO3_OP %>% 
  pres_abund_filter() %>% 
  to_RA()
# A total of 1118 ASVs were found to be present at or above a level of confident detection (0.1% relative abundance) in at least 2 samples (n = 22518 ASVs excluded).
```
# Extract OTU and metadata 
```{r}
otu_vgo2_np <- t(as.data.frame(otu_table(ps_vgo2_np_filtered)))
meta_vgo2_np <- meta_to_df(ps_vgo2_np_filtered, sample_name = ".row_id")

otu_vgo2_op <- t(as.data.frame(otu_table(ps_vgo2_op_filtered)))
meta_vgo2_op <- meta_to_df(ps_vgo2_op_filtered, sample_name = ".row_id")

otu_vgo3_np <- t(as.data.frame(otu_table(ps_vgo3_np_filtered)))
meta_vgo3_np <- meta_to_df(ps_vgo3_np_filtered, sample_name = ".row_id")

otu_vgo3_op <- t(as.data.frame(otu_table(ps_vgo3_op_filtered)))
meta_vgo3_op <- meta_to_df(ps_vgo3_op_filtered, sample_name = ".row_id")
```
# Check associations between variables
```{r}
# Select lung function variables
lung_vars <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
df_lung <- df[, lung_vars]

# Calculate correlation matrix with p-values
cor_res <- rcorr(as.matrix(df_lung), type = "pearson")  # Hmisc package
cor_matrix <- round(cor_res$r, 3)
p_matrix <- cor_res$P

# Melt the matrices
melted_cor <- melt(cor_matrix)
melted_p <- melt(p_matrix)

# Merge correlation values and p-values
melted_cor$p_value <- melted_p$value

# Function to assign significance stars
melted_cor$signif <- cut(
  melted_cor$p_value,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "")
)

# Keep only lower triangle (exclude duplicates & diagonal)
melted_cor <- melted_cor[as.numeric(melted_cor$Var1) > as.numeric(melted_cor$Var2), ]

# Plot enhanced lower triangular heatmap
lung_func_corr <- ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(sprintf("%.2f", value), signif)), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#2166AC",   # blue for negative correlations
    mid = "white",     # neutral = 0
    high = "#B2182B",  # red for positive correlations
    midpoint = 0,
    limits = c(-1, 1),
    name = "Pearson r"
  ) +
  theme_minimal() +
  labs(title = "Correlation Matrix of Lung Function\n Outcomes with Significance",
       x = "", y = "") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

ggsave("../Output/Variable_associations/Lung_function_correlations.svg", width = 5, height = 5)


# Test associations between variables to be included in multivariable models
vars <- c("age", "gender", "season_microbiome_sampling", "smoke_status")

# Check data types
sapply(df[vars], class)

# Initialize results dataframe
results <- data.frame(
  var1 = character(),
  var2 = character(),
  test = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through all pairs
for (i in 1:(length(vars)-1)) {
  for (j in (i+1):length(vars)) {
    v1 <- df[[vars[i]]]
    v2 <- df[[vars[j]]]
    
    # Continuous vs continuous
    if (is.numeric(v1) & is.numeric(v2)) {
      test_res <- cor.test(v1, v2)
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = "Pearson",
        p_value = test_res$p.value
      ))
    }
    
    # Continuous vs factor
    else if ((is.numeric(v1) & is.factor(v2)) | (is.factor(v1) & is.numeric(v2))) {
      if (is.factor(v1)) { tmp <- v1; v1 <- v2; v2 <- tmp; }
      if (nlevels(v2) == 2) {
        test_res <- t.test(v1 ~ v2)
        test_name <- "t-test"
        pval <- test_res$p.value
      } else {
        test_res <- aov(v1 ~ v2)
        test_name <- "ANOVA"
        pval <- summary(test_res)[[1]][["Pr(>F)"]][1]
      }
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = test_name,
        p_value = pval
      ))
    }
    
    # Factor vs factor
    else if (is.factor(v1) & is.factor(v2)) {
      test_res <- chisq.test(table(v1, v2))
      results <- rbind(results, data.frame(
        var1 = vars[i],
        var2 = vars[j],
        test = "Chi-squared",
        p_value = test_res$p.value
      ))
    }
  }
}

# Add significance stars
results <- results %>%
  mutate(
    signif = cut(
      p_value,
      breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
      labels = c("***", "**", "*", "")
    ),
    p_value = signif(round(p_value, 4))
  )

# Arrange results nicely
results <- results %>% arrange(var1, var2)

# View the table
print(results)

# Save the results table to Excel
write_xlsx(results, "../Output/Variable_associations/Covariate_association_results.xlsx")

```

# ALPHA DIVERSITY
## Prepare data
We run linear models to assess the association between lung function values (% predicted lung function measures (FEV1, FVC, FEV1/FVC) and these variables categorised) and alpha diversity (Observed, Shannon, Simpson)

<!-- ## Univariable linear regression for alpha diversity  -->
<!-- ```{r} -->
<!-- # Univariable linear regression for standardized alpha diversity - so dependent variable is alpha diversity in SD units -->
<!-- # For continuous numeric variables, we use a simple linear regression - the slope (beta) is a SD change in alpha diversity per 1 unit increase in that variable. Positive = higher diversity with higher variable. p-value is the t-test for that slope.  -->
<!-- # If a factor has ≥3 levels (e.g., FEV1_category, FEF25_75_category in many datasets): the loop uses an overall ANOVA F-test and the tile fill is eta² (variance explained). -->
<!-- # If a factor has exactly 2 levels (e.g., FVC_category, FEV1FVC_category): the loop takes the single coefficient from lm(). i.e. the p-value is the t-test for that single contrast.  -->


<!-- # Load and organise dataset list -->
<!-- dataset_list <- list( -->
<!--   VGO2_NP = meta_vgo2_np, -->
<!--   VGO2_OP = meta_vgo2_op, -->
<!--   VGO3_NP = meta_vgo3_np, -->
<!--   VGO3_OP = meta_vgo3_op -->
<!-- ) -->

<!-- alpha_vars <- c("Shannon", "Simpson", "Observed") -->

<!-- test_vars <- c( -->
<!--   "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", -->
<!--   "packyears", "asthma_ever", "copd_gold", -->
<!--   "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr", -->
<!--   "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category" -->
<!-- ) -->

<!-- categorical_vars <- c( -->
<!--   "gender", "season_microbiome_sampling", "smoke_status", "asthma_ever", -->
<!--   "copd_gold", "copd_doc_diag", "pneumo_lastyr", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category" -->
<!-- ) # N.B "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category" are all 'Ordered factor' variables, so no need to reset the reference category, but will need to specify ref category for gender, season_microbiome_sampling, smoke_status, asthma_ever, copd_gold, copd_lln, pneumo_lastyr -->


<!-- # Recode factors and set reference levels for unordered factor variables  -->
<!-- # N.B. for factor variables with 'no' 'yes' - no is automatically set as the reference  -->
<!-- for (dataset in names(dataset_list)) { -->
<!--   for (var in categorical_vars) { -->
<!--     if (var %in% names(dataset_list[[dataset]])) { -->
<!--       dataset_list[[dataset]][[var]] <- as.factor(dataset_list[[dataset]][[var]]) -->
<!--     } -->
<!--   } -->
<!--   df <- dataset_list[[dataset]] -->
<!--   if ("gender" %in% names(df)) df$gender <- relevel(df$gender, ref = "male")  -->
<!--   if ("season_microbiome_sampling" %in% names(df)) df$season_microbiome_sampling <- relevel(df$season_microbiome_sampling, ref = "Winter")  -->
<!--   if ("smoke_status" %in% names(df)) df$smoke_status <- relevel(df$smoke_status, ref = "never") -->
<!--   # if ("asthma_ever" %in% names(df)) df$asthma_ever <- relevel(df$asthma_ever, ref = "no")  -->
<!--   # if ("copd_gold" %in% names(df)) df$copd_gold <- relevel(df$copd_gold, ref = "no")  -->
<!--   # if ("copd_doc_diag" %in% names(df)) df$copd_doc_diag <- relevel(df$copd_doc_diag, ref = "no") -->
<!--   # if ("pneumo_lastyr" %in% names(df)) df$pneumo_lastyr <- relevel(df$pneumo_lastyr, ref = "no") -->
<!--   dataset_list[[dataset]] <- df -->
<!-- } -->

<!-- # Convert pneumo_last3yr to numeric where needed -->
<!-- for (dataset in names(dataset_list)) { -->
<!--   if ("pneumo_last3yr" %in% names(dataset_list[[dataset]])) { -->
<!--     dataset_list[[dataset]]$pneumo_last3yr <- as.numeric(as.character(dataset_list[[dataset]]$pneumo_last3yr)) -->
<!--   } -->
<!-- } -->

<!-- #### Univariable models with standardized alpha diversity ####  -->
<!-- all_results_std <- list() -->

<!-- for (dataset_name in names(dataset_list)) { -->
<!--   metadata <- dataset_list[[dataset_name]] -->

<!--   # Standardize alpha diversity metrics -->
<!--   metadata_std <- metadata -->
<!--   for (alpha in alpha_vars) { -->
<!--     if (alpha %in% names(metadata_std)) { -->
<!--       metadata_std[[alpha]] <- scale(metadata_std[[alpha]]) -->
<!--     } -->
<!--   } -->

<!--   results_list <- list() -->

<!--   for (alpha in alpha_vars) { -->
<!--   for (var in test_vars) { -->
<!--     if (!all(var %in% names(metadata_std), alpha %in% names(metadata_std))) next -->
<!--     if (all(is.na(metadata_std[[alpha]])) || all(is.na(metadata_std[[var]]))) next -->
<!--     # need at least 2 non-NA unique values/levels -->
<!--     if (length(unique(na.omit(metadata_std[[var]]))) < 2) next -->

<!--     form  <- as.formula(paste(alpha, "~", var)) -->
<!--     model <- tryCatch(lm(form, data = metadata_std), error = function(e) NULL) -->
<!--     if (is.null(model)) next -->

<!--     # Common effect size for all predictors -->
<!--     r2 <- tryCatch(summary(model)$r.squared, error = function(e) NA_real_) -->

<!--     # p-value: overall F for multi-level factors; slope/contrast for others -->
<!--     is_fac <- is.factor(metadata_std[[var]]) -->
<!--     nlvls  <- if (is_fac) nlevels(droplevels(metadata_std[[var]])) else NA -->

<!--     if (is_fac && !is.na(nlvls) && nlvls > 2) { -->
<!--       a <- tryCatch(anova(model), error = function(e) NULL) -->
<!--       if (is.null(a) || !var %in% rownames(a)) next -->
<!--       p_val <- a[var, "Pr(>F)"] -->
<!--     } else { -->
<!--       tid <- tryCatch(broom::tidy(model), error = function(e) NULL) -->
<!--       if (is.null(tid)) next -->
<!--       tid <- tid[tid$term != "(Intercept)", , drop = FALSE] -->
<!--       if (!nrow(tid)) next -->
<!--       p_val <- tid$p.value[1] -->
<!--     } -->

<!--     results_list[[paste(alpha, var, sep = "_")]] <- data.frame( -->
<!--       dataset    = dataset_name, -->
<!--       alpha      = alpha, -->
<!--       variable   = var, -->
<!--       coefficient= r2,       # <— tile colour = R² (0..1) -->
<!--       p_value    = p_val -->
<!--     ) -->
<!--   } -->
<!-- } -->


<!--   dataset_results <- bind_rows(results_list) -->
<!--   all_results_std[[dataset_name]] <- dataset_results -->
<!-- } -->

<!-- # Combine results and annotate -->
<!-- univariable_results_std <- bind_rows(all_results_std) %>% -->
<!--   filter(!is.na(p_value), !is.na(coefficient)) %>% -->
<!--   mutate( -->
<!--     significance = case_when( -->
<!--       p_value < 0.001 ~ "***", -->
<!--       p_value < 0.01  ~ "**", -->
<!--       p_value < 0.05  ~ "*", -->
<!--       TRUE            ~ "" -->
<!--     ), -->
<!--     variable = factor(variable, levels = rev(test_vars)) -->
<!--   ) -->

<!-- #### Save standardised univariable results #### -->
<!-- write_xlsx(univariable_results_std %>% filter(dataset %in% c("VGO2_NP", "VGO2_OP")), -->
<!--            "../Output/Alpha_diversity/Univariable/Univariable_alpha_diversity_model_resultstandardized_univariable_alpha_diversity_vgo2.xlsx") -->

<!-- write_xlsx(univariable_results_std %>% filter(dataset %in% c("VGO3_NP", "VGO3_OP")), -->
<!--            "../Output/Alpha_diversity/Univariable/Standardized_univariable_alpha_diversity_vgo3.xlsx") -->

<!-- #### Plotting standardised coefficients as heatmap #### -->
<!-- plot_vgo2 <- univariable_results_std %>% filter(dataset %in% c("VGO2_NP", "VGO2_OP")) -->
<!-- plot_vgo3 <- univariable_results_std %>% filter(dataset %in% c("VGO3_NP", "VGO3_OP")) -->

<!-- base_theme <- theme_minimal(base_size = 12) + -->
<!--   theme( -->
<!--     axis.text.x = element_text(angle = 30, hjust = 1), -->
<!--     panel.grid = element_blank() -->
<!--   ) -->

<!-- heatmap_vgo2 <- ggplot(plot_vgo2, aes(x = alpha, y = variable, fill = coefficient)) + -->
<!--   geom_tile(color = "white") + -->
<!--   geom_text(aes(label = significance), color = "black", size = 3) + -->
<!--   scale_fill_gradient2( -->
<!--     low = carto_pal(7, "Geyser")[1], -->
<!--     mid = carto_pal(7, "Geyser")[4], -->
<!--     high = carto_pal(7, "Geyser")[7], -->
<!--     midpoint = 0, -->
<!--     name = "Standardised Coefficient" -->
<!--   ) + -->
<!--   facet_wrap(~ dataset, ncol = 2) + -->
<!--   base_theme + -->
<!--   theme( -->
<!--     axis.title.x = element_blank(), -->
<!--     axis.text.x = element_blank(), -->
<!--     axis.ticks.x = element_blank() -->
<!--   ) + -->
<!--   labs(y = "Explanatory variable") -->

<!-- heatmap_vgo3 <- ggplot(plot_vgo3, aes(x = alpha, y = variable, fill = coefficient)) + -->
<!--   geom_tile(color = "white") + -->
<!--   geom_text(aes(label = significance), color = "black", size = 3) + -->
<!--   scale_fill_gradient2( -->
<!--     low = carto_pal(7, "Geyser")[1], -->
<!--     mid = carto_pal(7, "Geyser")[4], -->
<!--     high = carto_pal(7, "Geyser")[7], -->
<!--     midpoint = 0, -->
<!--     name = "Standardised Coefficient" -->
<!--   ) + -->
<!--   facet_wrap(~ dataset, ncol = 2) + -->
<!--   base_theme + -->
<!--   labs(x = "Alpha Diversity Metric", y = "Explanatory variable") -->

<!-- combined_plot <- (heatmap_vgo2 / heatmap_vgo3) + -->
<!--   patchwork::plot_layout(guides = "collect", heights = c(1, 1)) & -->
<!--   theme(legend.position = "right") -->

<!-- univar_alpha_plot <- combined_plot + -->
<!--   patchwork::plot_annotation( -->
<!--     title = "Standardised Univariable Associations with Alpha Diversity (VGO2 and VGO3)", -->
<!--     theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5)) -->
<!--   ) -->

<!-- # Show plot  -->
<!-- print(univar_alpha_plot) -->

<!-- # Save plot -->
<!-- ggsave("../Output/Alpha_diversity/Univariable/Univariable_alpha_diversity_standardised_heatmap.svg", -->
<!--        univar_alpha_plot, width = 8, height = 7) -->
<!-- ``` -->

## Univariable linear regression for alpha diversity 
```{r}
## Univariable linear regression for alpha diversity (signed, standardised coefficients)

# Datasets
dataset_list <- list(
  VGO2_NP = meta_vgo2_np,
  VGO2_OP = meta_vgo2_op,
  VGO3_NP = meta_vgo3_np,
  VGO3_OP = meta_vgo3_op
)

alpha_vars <- c("Shannon", "Simpson", "Observed")

test_vars <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status",
  "packyears", "asthma_ever", "copd_gold",
  "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred",
  "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category"
)

# Variables we will treat as ORDERED (map to 0..K-1 then z-score)
ordinal_vars <- c(
  "FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category",
  "smoke_status",              # never < former < current
  "season_microbiome_sampling",# Winter < Spring < Summer < Autumn
  "copd_gold"                  # if present and meaningful order
)

# Variables we will treat as BINARY (reference first, code 0/1 then z-score)
# N.B. for binary variables with 'no' 'yes' - no is automatically set as the reference 
binary_vars <- c("gender","asthma_ever","copd_doc_diag","pneumo_lastyr")  # ensure values like male/female, no/yes

# --- Recode factors, set references, and enforce orders ---
for (nm in names(dataset_list)) {
  df <- dataset_list[[nm]]

  # Ensure factors
  fac_names <- intersect(
    c("gender","season_microbiome_sampling","smoke_status","asthma_ever",
      "copd_gold","copd_doc_diag","pneumo_lastyr",
      "FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category"),
    names(df)
  )
  for (v in fac_names) df[[v]] <- as.factor(df[[v]])

  # Set explicit orders/refs
  if ("gender" %in% names(df)) df$gender <- relevel(df$gender, ref = "male")
  if ("smoke_status" %in% names(df)) {
    df$smoke_status <- factor(df$smoke_status, levels = c("never","former","current"), ordered = TRUE)
  }
  if ("season_microbiome_sampling" %in% names(df)) {
    df$season_microbiome_sampling <- factor(df$season_microbiome_sampling,
      levels = c("Winter","Spring","Summer","Autumn"), ordered = TRUE)
  }
  for (v in c("FEV1_category","FVC_category","FEV1FVC_category","FEF25_75_category")) {
    if (v %in% names(df)) df[[v]] <- factor(df[[v]],
      levels = c("Normal","Mild","Moderate","Severe")[1:nlevels(df[[v]])], ordered = TRUE)
  }
  # Optional: enforce order for GOLD if present, e.g. 0<1<2<3<4; otherwise comment this line
  if ("copd_gold" %in% names(df)) df$copd_gold <- factor(df$copd_gold, ordered = TRUE)

  # Make sure numeric fields are numeric
  num_names <- intersect(c("age","bmi","packyears","pneumo_last3yr",
                           "FEV1_pred","FVC_pred","FEV1FVC_pred","FEF25_75_pred"),
                         names(df))
  for (v in num_names) df[[v]] <- suppressWarnings(as.numeric(as.character(df[[v]])))

  dataset_list[[nm]] <- df
}

# ----- Helper to extract a single signed, standardised coefficient -----
get_std_coef <- function(y, x, varname, df) {
  # y is already numeric (we will z-score y upstream); x may be factor/numeric
  x_raw <- df[[varname]]

  # If ordered (or declared ordinal), map to integer 0..K-1
  if ((is.factor(x_raw) && is.ordered(x_raw)) || varname %in% ordinal_vars) {
    # preserve declared order; if not ordered, coerce with current levels
    if (!is.ordered(x_raw)) x_raw <- factor(x_raw, levels = levels(x_raw), ordered = TRUE)
    x_num <- as.numeric(x_raw) - 1L
    if (length(na.omit(unique(x_num))) < 2) return(NULL)
    x_std <- as.numeric(scale(x_num))
    fit <- tryCatch(lm(y ~ x_std), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    tt <- broom::tidy(fit)
    return(list(beta = tt$estimate[tt$term == "x_std"], p = tt$p.value[tt$term == "x_std"]))
  }

  # If 2-level factor, set reference to the first level and code 0/1
  if (is.factor(x_raw) && nlevels(droplevels(x_raw)) == 2) {
    # keep current reference (already relevelled above); code ref=0, other=1
    x01 <- as.numeric(x_raw != levels(x_raw)[1])
    if (length(na.omit(unique(x01))) < 2) return(NULL)
    x_std <- as.numeric(scale(x01))
    fit <- tryCatch(lm(y ~ x_std), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    tt <- broom::tidy(fit)
    return(list(beta = tt$estimate[tt$term == "x_std"], p = tt$p.value[tt$term == "x_std"]))
  }

  # If plain numeric
  if (is.numeric(x_raw)) {
    if (sd(x_raw, na.rm = TRUE) == 0 || all(is.na(x_raw))) return(NULL)
    x_std <- as.numeric(scale(x_raw))
    fit <- tryCatch(lm(y ~ x_std), error = function(e) NULL)
    if (is.null(fit)) return(NULL)
    tt <- broom::tidy(fit)
    return(list(beta = tt$estimate[tt$term == "x_std"], p = tt$p.value[tt$term == "x_std"]))
  }

  # If 3+ level unordered factor sneaks in: skip (no unique signed coefficient)
  return(NULL)
}

# ----- Fit models -----
all_results <- list()

for (dataset_name in names(dataset_list)) {
  df <- dataset_list[[dataset_name]]

  # z-score alpha metrics (outcome); keep a copy
  df_std <- df
  for (a in alpha_vars) if (a %in% names(df_std)) df_std[[a]] <- as.numeric(scale(df_std[[a]]))

  res_list <- list()

  for (alpha in alpha_vars) {
    if (!alpha %in% names(df_std)) next
    y <- df_std[[alpha]]

    for (var in test_vars) {
      if (!var %in% names(df_std)) next
      # Need variability
      if (length(na.omit(unique(df_std[[var]]))) < 2) next

      out <- get_std_coef(y, df_std[[var]], var, df_std)
      if (is.null(out)) next

      res_list[[paste(alpha, var, sep = "_")]] <- data.frame(
        dataset    = dataset_name,
        alpha      = alpha,
        variable   = var,
        coefficient= out$beta,     # signed, standardised β
        p_value    = out$p
      )
    }
  }
  all_results[[dataset_name]] <- dplyr::bind_rows(res_list)
}

univariable_results_std <- dplyr::bind_rows(all_results) %>%
  dplyr::filter(!is.na(p_value), !is.na(coefficient)) %>%
  dplyr::mutate(
    significance = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    variable = factor(variable, levels = rev(test_vars))
  )

# (Optional) FDR adjust within dataset x alpha outcome
# univariable_results_std <- univariable_results_std %>%
#   group_by(dataset, alpha) %>%
#   mutate(q_value = p.adjust(p_value, method = "BH")) %>%
#   ungroup()

# ----- Plot -----
plot_vgo2 <- univariable_results_std %>% dplyr::filter(dataset %in% c("VGO2_NP","VGO2_OP"))
plot_vgo3 <- univariable_results_std %>% dplyr::filter(dataset %in% c("VGO3_NP","VGO3_OP"))

# Symmetric colour limits across both panels
lim <- max(abs(c(plot_vgo2$coefficient, plot_vgo3$coefficient)), na.rm = TRUE)

base_theme <- theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())

# after computing `lim <- max(abs(c(plot_vgo2$coefficient, plot_vgo3$coefficient)), na.rm = TRUE)`

heatmap_vgo2 <- ggplot(plot_vgo2, aes(x = alpha, y = variable, fill = coefficient)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), size = 3) +
  scale_fill_gradient2(
    low  = carto_pal(7, "Geyser")[1],
    mid  = carto_pal(7, "Geyser")[4],
    high = carto_pal(7, "Geyser")[7],
    midpoint = 0,
    limits   = c(-lim, lim),
    name     = "Standardised Coefficient"
  ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Explanatory variable")

heatmap_vgo3 <- ggplot(plot_vgo3, aes(x = alpha, y = variable, fill = coefficient)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), size = 3) +
  scale_fill_gradient2(
    low  = carto_pal(7, "Geyser")[1],
    mid  = carto_pal(7, "Geyser")[4],
    high = carto_pal(7, "Geyser")[7],
    midpoint = 0,
    limits   = c(-lim, lim),
    name     = "Standardised Coefficient"
  ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  labs(x = "Alpha Diversity Metric", y = "Explanatory variable")


combined_plot <- (heatmap_vgo2 / heatmap_vgo3) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

univar_alpha_plot <- combined_plot +
  patchwork::plot_annotation(
    title = "Standardised Univariable Associations with Alpha Diversity (signed β)",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  )

print(univar_alpha_plot)

ggsave("../Output/Alpha_diversity/Univariable/Univariable_alpha_diversity_standardised_heatmap.svg",
       univar_alpha_plot, width = 8, height = 7)

```

## Multivariable models with lung function (continuous + categorical) 
```{r}
# We standardise the alpha diversity metrics since observed is much higher than the other two, the coefficients are not comparable

# Define variables
lung_function_vars <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
alpha_vars <- c("Observed", "Shannon", "Simpson")
covariates <- c("age", "gender", "smoke_status", "season_microbiome_sampling")

# Standardize alpha diversity variables within each dataset
for (dataset_name in names(dataset_list)) {
  dataset <- dataset_list[[dataset_name]]
  
  for (alpha_var in alpha_vars) {
    if (alpha_var %in% names(dataset)) {
      dataset[[alpha_var]] <- scale(dataset[[alpha_var]], center = TRUE, scale = TRUE) # standardise the alpha diversity 
    }
  }
  dataset_list[[dataset_name]] <- dataset
}

# Create nested list to store model results
model_results <- list()

# Loop through datasets
for (dataset_name in names(dataset_list)) {
  dataset <- dataset_list[[dataset_name]]
  model_results[[dataset_name]] <- list()
  
  # Loop through alpha diversity metrics
  for (alpha_var in alpha_vars) {
    model_results[[dataset_name]][[alpha_var]] <- list()
    
    # Loop through lung function variables
    for (lung_var in lung_function_vars) {
      # Check if all required variables exist
      vars_needed <- c(alpha_var, lung_var, covariates)
      if (all(vars_needed %in% names(dataset))) {
        
        # Build formula
        formula_str <- paste(alpha_var, "~", paste(c(lung_var, covariates), collapse = " + "))
        model_formula <- as.formula(formula_str)
        
        # Fit model
        model <- lm(model_formula, data = dataset)
        
        # Store model summary
        model_results[[dataset_name]][[alpha_var]][[lung_var]] <- summary(model)
      } else {
        warning(paste("Skipping model for", dataset_name, "-", alpha_var, "-", lung_var, 
                      ": missing variable(s)."))
      }
    }
  }
}

# Build tidy results
result_tables <- list()

for (alpha_var in alpha_vars) {
  results_list <- list()
  
  for (dataset_name in names(model_results)) {
    for (lung_var in lung_function_vars) {
      model <- model_results[[dataset_name]][[alpha_var]][[lung_var]]
      
      if (!is.null(model)) {
        tidy_model <- broom::tidy(model)
        tidy_model$dataset <- dataset_name
        tidy_model$alpha_diversity <- alpha_var
        tidy_model$lung_function <- lung_var
        
        results_list[[paste(dataset_name, alpha_var, lung_var, sep = "_")]] <- tidy_model
      }
    }
  }
  
  result_tables[[alpha_var]] <- do.call(rbind, results_list)
}

# Combine all results
all_results <- bind_rows(result_tables, .id = "Alpha_metric")

# Filter to lung function terms only and add significance stars
multivariable_alpha_diversity_results <- all_results %>%
  filter(term %in% lung_function_vars) %>%
  select(dataset, Alpha_metric, lung_function, term, estimate, std.error, p.value) %>%
  mutate(
    raw_estimate = estimate,  # unrounded
    sig_label = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    estimate_with_sig = paste0(round(estimate, 10), sig_label)
  )


# Print final table
print(multivariable_alpha_diversity_results)

# Optional: write to CSV
write.csv(multivariable_alpha_diversity_results, "../Output/Alpha_diversity/Multivariable/Multivariable_model_results.csv", row.names = FALSE)


# Split results into VGO2 and VGO3
# Prepare datasets for plotting
plot_vgo2_multi <- multivariable_alpha_diversity_results %>%
  filter(dataset %in% c("VGO2_NP", "VGO2_OP")) %>%
  mutate(
    variable = lung_function,
    significance = gsub("[-0-9e\\.]", "", estimate_with_sig),
    estimate = round(estimate, 3)  # ← ROUND HERE
  )

plot_vgo3_multi <- multivariable_alpha_diversity_results %>%
  filter(dataset %in% c("VGO3_NP", "VGO3_OP")) %>%
  mutate(
    variable = lung_function,
    significance = gsub("[-0-9e\\.]", "", estimate_with_sig),
    estimate = round(estimate, 3)  # ← ROUND HERE
  )


# Define base theme
base_theme <- theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid = element_blank()
  )
# Define breaks manually for the legend (e.g. min, 0, max or a sequence)
legend_breaks <- scales::pretty_breaks(n = 5)(c(plot_vgo2_multi$estimate, plot_vgo3_multi$estimate))

# Format labels rounded to 3 decimals
legend_labels <- sprintf("%.3f", legend_breaks)

# Heatmap for VGO2
heatmap_vgo2_multi <- ggplot(plot_vgo2_multi, aes(x = Alpha_metric, y = variable, fill = estimate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), size = 4) +
  scale_fill_gradient2(
    low = carto_pal(7, "Geyser")[1],
    mid = carto_pal(7, "Geyser")[4],
    high = carto_pal(7, "Geyser")[7],
    midpoint = 0,
    name = "Standardised Coefficient",
    breaks = legend_breaks,
    labels = legend_labels
  ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Lung Function Metric")


# Heatmap for VGO3
heatmap_vgo3_multi <- ggplot(plot_vgo3_multi, aes(x = Alpha_metric, y = variable, fill = estimate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), size = 4) +
  scale_fill_gradient2(
  low = carto_pal(7, "Geyser")[1],
  mid = carto_pal(7, "Geyser")[4],
  high = carto_pal(7, "Geyser")[7],
  midpoint = 0,
  name = "Standardised Coefficient",
  breaks = legend_breaks,
  labels = legend_labels) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  labs(x = "Alpha Diversity Metric", y = "Lung Function Metric")

# Combine plots with patchwork
combined_multi_plot <- (heatmap_vgo2_multi / heatmap_vgo3_multi) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

# Add title annotation
final_multivar_plot <- combined_multi_plot +
  patchwork::plot_annotation(
    title = "Standardised Multivariable Associations with Alpha Diversity (VGO2 and VGO3)",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  )

# Show and save the plot
print(final_multivar_plot)

ggsave("../Output/Alpha_diversity/Multivariable/Multivariable_alpha_diversity_standardised_heatmap.svg",
       final_multivar_plot, width = 6, height = 5)

```





# Visualise relationships between lung function (continuous and categorical) and alpha diversity
```{r}
# Define datasets
datasets <- list(
  VGO2_NP = meta_vgo2_np,
  VGO2_OP = meta_vgo2_op,
  VGO3_NP = meta_vgo3_np,
  VGO3_OP = meta_vgo3_op
)

# Variables
lung_vars <- c("FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred")
alpha_metrics <- c("Observed", "Shannon", "Simpson")

# Store final plots
scatter_plots <- list()
box_plots <- list()

# Loop through datasets
for (name in names(datasets)) {
  df <- categorize_lung_function(datasets[[name]])

  scatter_list <- list()
  box_list <- list()

  for (lung_var in lung_vars) {
    lung_label <- switch(lung_var,
                         FEV1_pred = "FEV1 % Predicted",
                         FVC_pred = "FVC % Predicted",
                         FEV1FVC_pred = "FEV1/FVC Ratio",
                         FEF25_75_pred = "FEF25-75 % Predicted")

    lung_cat <- paste0(gsub("_pred", "", lung_var), "_category")

    for (alpha in alpha_metrics) {
      # Scatter plot
      scatter <- ggplot(df, aes_string(x = lung_var, y = alpha)) +
        geom_point(alpha = 0.6, size = 1.5, color = "#2c7fb8") +
        geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
        theme_minimal() +
        labs(
          title = paste(alpha, "vs", lung_label, "(", name, ")"),
          x = lung_label,
          y = alpha
        )
      scatter_list[[paste(lung_var, alpha, sep = "_")]] <- scatter

      # Remove NA for boxplot
      df_clean <- df[!is.na(df[[alpha]]) & !is.na(df[[lung_cat]]), ]

      # Boxplot
      box <- ggplot(df_clean, aes_string(x = lung_cat, y = alpha, fill = lung_cat)) +
        geom_boxplot(outlier.shape = NA, alpha = 0.8) +
        geom_jitter(width = 0.15, alpha = 0.4, size = 1.2) +
        scale_fill_brewer(palette = "Blues") +
        theme_minimal() +
        labs(
          title = paste(alpha, "by", gsub("_", "/", gsub("_category", "", lung_cat)), "category (", name, ")"),
          x = gsub("_", "/", gsub("_category", "", lung_cat)),
          y = alpha
        ) +
        stat_compare_means(method = "kruskal.test", label.y.npc = "top", size = 3)

      box_list[[paste(lung_cat, alpha, sep = "_")]] <- box
    }
  }

  # Combine plots per dataset
  scatter_grid <- wrap_plots(scatter_list, ncol = 3, guides = "collect") +
    plot_annotation(title = paste("Scatter Plots:", name),
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))

  boxplot_grid <- wrap_plots(box_list, ncol = 3, guides = "collect") +
    plot_annotation(title = paste("Boxplots:", name),
                    theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))

  # Store in final plot lists
  scatter_plots[[name]] <- scatter_grid
  box_plots[[name]] <- boxplot_grid
}

# Print and save per dataset
for (name in names(scatter_plots)) {
  print(scatter_plots[[name]])
  print(box_plots[[name]])

  ggsave(paste0("../Output/Alpha_diversity/Univariable/", name, "_scatterplots.svg"),
         scatter_plots[[name]], width = 14, height = 18)

  ggsave(paste0("../Output/Alpha_diversity/Univariable/", name, "_boxplots.svg"),
         box_plots[[name]], width = 14, height = 18)
}
```
# BETA DIVERSITY
## Univariable PERMANOVA functions
```{r}
beta_df <- function(ord, meta, var) { 
  m_sub <- subset(meta, !is.na(meta[var]))
  ord_sub <- ord[rownames(ord) %in% m_sub$sample_id, colnames(ord) %in% 
                   m_sub$sample_id]
  ord_sub <- as.dist(ord_sub)
  return(list(ord_sub = ord_sub, m_sub = m_sub))
}

do_permanova <- function(ord, meta, var) {
  t <- adonis2(as.formula(glue("ord ~ {var}")), 
               permutations = 999, data = meta)
  return(t)
}

do_permanova_com <- function(ord, meta, var) {
  t <- beta_df(ord, meta, var) %>%
    { do_permanova(ord = .$ord_sub, meta = .$m_sub, var) }

  t <- broom::tidy(t)
 
   # (Optional) You might also want to filter out "Total" or "Model" rows if they appear
  t <- t %>% filter(!term %in% c("Residual", "Total"))  

  # Add p-value labels
  t <- t %>% mutate(
    label_p.value = ifelse(p.value < 0.25, "*", NA),
    label_p.value = ifelse(p.value < 0.1, "**", label_p.value),
    label_p.value = ifelse(p.value < 0.05, "***", label_p.value),
    Var = var
  )

  return(t)
}


# Dispersiion check function 
check_dispersion <- function(dist_matrix, meta, var) {
  if (is.factor(meta[[var]])) {
    # Categorical variable — use betadisper
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.dist(d)

    # Skip if <2 groups with ≥2 samples
    if (length(unique(sub_meta[[var]])) < 2 || any(table(sub_meta[[var]]) < 2)) {
      return(tibble(disp_p = NA, disp_note = "Skipped: <2 groups with ≥2 samples"))
    }

    model <- betadisper(d, sub_meta[[var]])
    perm <- permutest(model)
    return(tibble(disp_p = perm$tab[1, "Pr(>F)"], disp_note = NA))
    
  } else {
    # Continuous variable — use regression of average distances
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.matrix(d)
    dist_avg <- apply(d, 1, mean, na.rm = TRUE)

    mod <- lm(dist_avg ~ sub_meta[[var]])
    p <- summary(mod)$coefficients[2, 4]
    return(tibble(disp_p = p, disp_note = NA))
  }
}

# Dispersion check function
check_dispersion <- function(dist_matrix, meta, var) {
  if (is.factor(meta[[var]])) {
    # Categorical variable — use betadisper
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.dist(d)

    # Skip if <2 groups with ≥2 samples
    if (length(unique(sub_meta[[var]])) < 2 || any(table(sub_meta[[var]]) < 2)) {
      return(tibble(disp_p = NA, disp_note = "Skipped: <2 groups with ≥2 samples"))
    }

    model <- betadisper(d, sub_meta[[var]])
    perm <- permutest(model)
    return(tibble(disp_p = perm$tab[1, "Pr(>F)"], disp_note = NA))
    
  } else {
    # Continuous variable — use regression of average distances
    sub_meta <- meta[!is.na(meta[[var]]), ]
    d <- dist_matrix[rownames(sub_meta), rownames(sub_meta)]
    d <- as.matrix(d)
    dist_avg <- apply(d, 1, mean, na.rm = TRUE)

    mod <- lm(dist_avg ~ sub_meta[[var]])
    p <- summary(mod)$coefficients[2, 4]
    return(tibble(disp_p = p, disp_note = NA))
  }
}
```
## Bray-curtis matrix
```{r}
# Extract otu table and metadata table per dataset
# metadata already extracted above for alpha diversity and season added
otu_vgo2_np <- as.data.frame(otu_table(ps_vgo2_np_filtered)) %>% t()

otu_vgo2_op <- as.data.frame(otu_table(ps_vgo2_op_filtered)) %>% t()

otu_vgo3_np <- as.data.frame(otu_table(ps_vgo3_np_filtered)) %>% t()

otu_vgo3_op <- as.data.frame(otu_table(ps_vgo3_op_filtered)) %>% t()

# Compute BC matrices per dataset
bc_dist_vgo2_np <- vegdist(otu_vgo2_np, method = "bray")
bray_vgo2_np_matrix <- bc_dist_vgo2_np %>% as.matrix()

bc_dist_vgo2_op <- vegdist(otu_vgo2_op, method = "bray")
bray_vgo2_op_matrix <- bc_dist_vgo2_op %>% as.matrix()

bc_dist_vgo3_np <- vegdist(otu_vgo3_np, method = "bray")
bray_vgo3_np_matrix <- bc_dist_vgo3_np %>% as.matrix()

bc_dist_vgo3_op <- vegdist(otu_vgo3_op, method = "bray")
bray_vgo3_op_matrix <- bc_dist_vgo3_op %>% as.matrix()
```

## Univariable PERMANOVAs 
### Continuous and categorical lung function variables
#### VGO2 NP
```{r}
# Existing variables and typing
variables_vgo2_np <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status",
  "packyears", "asthma_ever", "copd_gold", "copd_lln",
  "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo2_np[variables_vgo2_np] <- lapply(meta_vgo2_np[variables_vgo2_np], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars_vgo2_np <- variables_vgo2_np[sapply(meta_vgo2_np[variables_vgo2_np], is.factor)]
continuous_vars_vgo2_np <- variables_vgo2_np[sapply(meta_vgo2_np[variables_vgo2_np], is.numeric)]

# Filter categorical variables with enough samples per group
meta_summary_vgo2_np <- pivot_longer(mutate_all(meta_vgo2_np, as.character), 
                                     cols = variables_vgo2_np, names_to = "Var", values_to = "Cat")

meta_summary_vgo2_np_cat <- meta_summary_vgo2_np %>%
  filter(Var %in% categorical_vars_vgo2_np) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_summary_vgo2_np_filtered_cat <- meta_summary_vgo2_np_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_summary_vgo2_np_filtered_cont <- data.frame(Var = continuous_vars_vgo2_np)

final_variable_list_vgo2_np <- unique(c(meta_summary_vgo2_np_filtered_cat$Var,
                                        meta_summary_vgo2_np_filtered_cont$Var))

variable_list_vgo2_np <- tibble(Var = final_variable_list_vgo2_np, fit = NA)

# Run PERMANOVA and dispersion
perm_vgo2_np <- variable_list_vgo2_np %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo2_np_matrix, meta_vgo2_np, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo2_np_matrix, meta_vgo2_np, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo2_np <- perm_vgo2_np %>% select(-fit_Var)

names(perm_vgo2_np) <- sub("^fit_", "", names(perm_vgo2_np))

# Add metadata columns
perm_vgo2_np$type <- "Bray Curtis"
perm_vgo2_np$dataset <- "VGO2_NP"
perm_vgo2_np$niche <- "NP"

# Adjust p-values
perm_vgo2_np$q.value <- p.adjust(perm_vgo2_np$p.value, "BH")

# Add p-value and q-value labels
perm_vgo2_np <- perm_vgo2_np %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

# Save results
write.csv(perm_vgo2_np, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo2_np_with_dispersion.csv")
```
#### VGO2 OP
```{r}
# Variables
variables_vgo2_op <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status",
  "packyears", "asthma_ever", "copd_gold", "copd_lln",
  "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo2_op[variables_vgo2_op] <- lapply(meta_vgo2_op[variables_vgo2_op], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo2_op[sapply(meta_vgo2_op[variables_vgo2_op], is.factor)]
continuous_vars <- variables_vgo2_op[sapply(meta_vgo2_op[variables_vgo2_op], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo2_op, as.character), 
                             cols = variables_vgo2_op, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo2_op <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo2_op_matrix, meta_vgo2_op, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo2_op_matrix, meta_vgo2_op, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo2_op <- perm_vgo2_op %>% select(-fit_Var)
names(perm_vgo2_op) <- sub("^fit_", "", names(perm_vgo2_op))

perm_vgo2_op$type <- "Bray Curtis"
perm_vgo2_op$dataset <- "VGO2_OP"
perm_vgo2_op$niche <- "OP"
perm_vgo2_op$q.value <- p.adjust(perm_vgo2_op$p.value, "BH")

perm_vgo2_op <- perm_vgo2_op %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo2_op, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo2_op_with_dispersion.csv")

```
#### VGO3 NP
```{r}
variables_vgo3_np <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "asthma_ever",
  "copd_doc_diag", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred","FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo3_np[variables_vgo3_np] <- lapply(meta_vgo3_np[variables_vgo3_np], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo3_np[sapply(meta_vgo3_np[variables_vgo3_np], is.factor)]
continuous_vars <- variables_vgo3_np[sapply(meta_vgo3_np[variables_vgo3_np], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo3_np, as.character), 
                             cols = variables_vgo3_np, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo3_np <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo3_np_matrix, meta_vgo3_np, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo3_np_matrix, meta_vgo3_np, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo3_np <- perm_vgo3_np %>% select(-fit_Var)
names(perm_vgo3_np) <- sub("^fit_", "", names(perm_vgo3_np))

perm_vgo3_np$type <- "Bray Curtis"
perm_vgo3_np$dataset <- "VGO3_NP"
perm_vgo3_np$niche <- "NP"
perm_vgo3_np$q.value <- p.adjust(perm_vgo3_np$p.value, "BH")

perm_vgo3_np <- perm_vgo3_np %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo3_np, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo3_np_with_dispersion.csv")

```
#### VGO3 OP
```{r}
variables_vgo3_op <- c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "asthma_ever",
  "copd_doc_diag", "pneumo_lastyr",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category")

meta_vgo3_op[variables_vgo3_op] <- lapply(meta_vgo3_op[variables_vgo3_op], function(x) {
  if (is.character(x) || is.logical(x)) as.factor(x) else x
})

categorical_vars <- variables_vgo3_op[sapply(meta_vgo3_op[variables_vgo3_op], is.factor)]
continuous_vars <- variables_vgo3_op[sapply(meta_vgo3_op[variables_vgo3_op], is.numeric)]

meta_summary <- pivot_longer(mutate_all(meta_vgo3_op, as.character), 
                             cols = variables_vgo3_op, names_to = "Var", values_to = "Cat")

meta_summary_cat <- meta_summary %>%
  filter(Var %in% categorical_vars) %>%
  count(Var, Cat) %>%
  filter(!is.na(Cat))

meta_filtered_cat <- meta_summary_cat %>%
  filter(n > 9) %>%
  count(Var) %>%
  filter(n > 1)

meta_filtered_cont <- data.frame(Var = continuous_vars)

final_vars <- unique(c(meta_filtered_cat$Var, meta_filtered_cont$Var))
variable_list <- tibble(Var = final_vars, fit = NA)

perm_vgo3_op <- variable_list %>%
  mutate(fit = map(Var, ~do_permanova_com(bray_vgo3_op_matrix, meta_vgo3_op, .x)),
         disp = map(Var, ~check_dispersion(bray_vgo3_op_matrix, meta_vgo3_op, .x))) %>%
  unnest(cols = c(fit, disp), names_sep = "_")

perm_vgo3_op <- perm_vgo3_op %>% select(-fit_Var)
names(perm_vgo3_op) <- sub("^fit_", "", names(perm_vgo3_op))

perm_vgo3_op$type <- "Bray Curtis"
perm_vgo3_op$dataset <- "VGO3_OP"
perm_vgo3_op$niche <- "OP"
perm_vgo3_op$q.value <- p.adjust(perm_vgo3_op$p.value, "BH")

perm_vgo3_op <- perm_vgo3_op %>% mutate(
  label_q.value = case_when(
    q.value < 0.05 ~ "***",
    q.value < 0.1 ~ "**",
    q.value < 0.25 ~ "*",
    TRUE ~ NA_character_
  )
)

write.csv(perm_vgo3_op, "../Output/Beta_diversity/Univariable_PERMANOVAs/Univariable_perm_vgo3_op_with_dispersion.csv")
```
#### Combine and present all univariable PERMANOVA results
```{r}
# Combine all univariable PERMANOVA results
perm_all <- bind_rows(perm_vgo2_np, perm_vgo2_op, perm_vgo3_np, perm_vgo3_op) %>%
  mutate(dataset_niche = paste(dataset, niche, sep = "_")) %>%
  filter(term == "Model")

plot_perm <- perm_all %>%
  rename(variable = Var, R2 = R2, significance = label_q.value)

# Split into datasets
plot_vgo2 <- plot_perm %>% filter(dataset %in% c("VGO2_NP", "VGO2_OP"))
plot_vgo3 <- plot_perm %>% filter(dataset %in% c("VGO3_NP", "VGO3_OP"))

# Define correct variable order (top to bottom in heatmap)
ordered_vars <- rev(c(
  "age", "gender", "bmi", "season_microbiome_sampling", "smoke_status", "packyears",
  "asthma_ever", "copd_gold", "copd_lln", "copd_doc_diag", "pneumo_last3yr", "pneumo_lastyr",
  "FEV1_pred","FEV1_category", "FVC_pred","FVC_category", "FEV1FVC_pred", "FEV1FVC_category", "FEF25_75_pred", "FEF25_75_category"
))

# Format and reorder data
plot_vgo2 <- plot_vgo2 %>%
  mutate(variable = factor(variable, levels = ordered_vars),
         R2 = round(R2, 3))

plot_vgo3 <- plot_vgo3 %>%
  mutate(variable = factor(variable, levels = ordered_vars),
         R2 = round(R2, 3))

# Plot theme and setup
base_theme <- theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid = element_blank()
  )



#### Heatmap for VGO2 ####
# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo2 <- min(plot_vgo2$R2, na.rm = TRUE)
r2_max_vgo2 <- max(plot_vgo2$R2, na.rm = TRUE)
r2_mid_vgo2 <- median(plot_vgo2$R2, na.rm = TRUE)

heatmap_vgo2 <- ggplot(plot_vgo2, aes(x = type, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo2,  # center gradient at median R²
    limits = c(r2_min_vgo2, r2_max_vgo2),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2) 
    ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Explanatory variable")

#### Heatmap for VGO3 ####
# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo3 <- min(plot_vgo3$R2, na.rm = TRUE)
r2_max_vgo3 <- max(plot_vgo3$R2, na.rm = TRUE)
r2_mid_vgo3 <- median(plot_vgo3$R2, na.rm = TRUE)

heatmap_vgo3 <- ggplot(plot_vgo3, aes(x = type, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo3,  # center gradient at median R²
    limits = c(r2_min_vgo3, r2_max_vgo3),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2) 
  ) +
  facet_wrap(~ dataset, ncol = 2) +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = "Explanatory variable")

#### Combine and save plot ####
combined_plot <- (heatmap_vgo2 / heatmap_vgo3) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

univar_beta_plot <- combined_plot +
  patchwork::plot_annotation(
    title = "Univariable PERMANOVAs (VGO2 and VGO3)",
    theme = theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  )

print(univar_beta_plot)

ggsave("../Output/Beta_diversity/Univariable_PERMANOVAS/Univariable_PERMANOVA_R2_heatmap.svg",
       univar_beta_plot, width = 5, height = 7)
```



# Multivariable PERMANOVAs
## Filter phyloseq and convert to relative abundance 
```{r}
# These ps objects are already filtered and converted to RA
ps_vgo2_np_filtered
ps_vgo2_op_filtered
ps_vgo3_np_filtered
ps_vgo3_op_filtered

# Convert the otu tables from the above phyloseq objects to dataframes (samples as rows)
otu_vgo2_np <- as.data.frame(otu_table(ps_vgo2_np_filtered)) %>% t()
otu_vgo2_op <- as.data.frame(otu_table(ps_vgo2_op_filtered)) %>% t()
otu_vgo3_np <- as.data.frame(otu_table(ps_vgo3_np_filtered)) %>% t()
otu_vgo3_op <- as.data.frame(otu_table(ps_vgo3_op_filtered)) %>% t()

# metadata already extracted for each ps object
meta_vgo2_np
meta_vgo2_op
meta_vgo3_np
meta_vgo3_op

# Compute BC dissimilarity matrices for each ps
ord_vgo2_np <- vegdist(otu_vgo2_np, method = "bray")
ord_vgo2_op <- vegdist(otu_vgo2_op, method = "bray")
ord_vgo3_np <- vegdist(otu_vgo3_np, method = "bray")
ord_vgo3_op <- vegdist(otu_vgo3_op, method = "bray")

# Compute mean BC dissimilarity
mean(ord_vgo2_np)
mean(ord_vgo2_op)
mean(ord_vgo3_np)
mean(ord_vgo3_op)
```

## Multivariable PERMANOVAs for lung function 
```{r}
run_permanova <- function(ord, meta, formula_vars) {
  # Ensure ord is a dist object
  ord <- as.matrix(ord)
  
  # Intersect sample IDs
  common_samples <- intersect(rownames(ord), rownames(meta))
  ord <- as.dist(ord[common_samples, common_samples])
  meta <- meta[common_samples, , drop = FALSE]

  n_before <- nrow(meta)

  # Drop rows with NA in any of the formula variables
  meta <- meta %>% filter(if_all(all_of(formula_vars), ~ !is.na(.)))
  keep_samples <- rownames(meta)
  ord <- as.dist(as.matrix(ord)[keep_samples, keep_samples])
  
  n_after <- nrow(meta)
  message(glue("Samples before filtering: {n_before} | After filtering: {n_after} | Removed: {n_before - n_after}"))
  
  # Remove categorical vars with <2 levels
  bad_vars <- formula_vars[sapply(meta[formula_vars], function(x) is.factor(x) && nlevels(factor(x)) < 2)]
  if (length(bad_vars) > 0) {
    message(glue("Dropping variables with only one level: {paste(bad_vars, collapse = ', ')}"))
    formula_vars <- setdiff(formula_vars, bad_vars)
  }
  if (length(formula_vars) == 0) stop("No variables with ≥2 levels remain after filtering.")

  # Build and run formula
  formula_str <- glue("ord ~ {paste(formula_vars, collapse = ' + ')}")
  adonis_res <- adonis2(as.formula(formula_str), data = meta, permutations = 999)
  tidy_res <- broom::tidy(adonis_res)
  tidy_res$term <- rownames(adonis_res)
  
  return(tidy_res)
}


# Fix metadata so that rownames are sample_id values (as the permanova function uses rownames) 
meta_vgo2_np; meta_vgo2_op; meta_vgo3_np;meta_vgo3_op
rownames(meta_vgo2_op) <- meta_vgo2_op$sample_id
rownames(meta_vgo3_np) <- meta_vgo3_np$sample_id
rownames(meta_vgo3_op) <- meta_vgo3_op$sample_id

#### FEV1 predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1_pred <- c("FEV1_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")
str(meta_vgo2_np$age)
str(meta_vgo2_np$gender)
str(meta_vgo2_np$season_microbiome_sampling)
str(meta_vgo2_np$smoke_status)

# PERMANOVAs
perm_FEV1_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1_pred)
perm_FEV1_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1_pred)
perm_FEV1_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1_pred)
perm_FEV1_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1_pred)


#### FEV1 categorised predicted multivariable PERMANOVA models ####
# Select required variables
covariates_FEV1_pred <- c("FEV1_category", "age", "season_microbiome_sampling",
                          "gender", "smoke_status")

# PERMANOVAs
perm_FEV1_cat_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1_pred)
perm_FEV1_cat_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1_pred)
perm_FEV1_cat_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1_pred)
perm_FEV1_cat_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1_pred)


#### FVC predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FVC <- c("FVC_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FVC_pred", "age")] <- lapply(meta_vgo2_np[c("FVC_pred", "age")], as.numeric)
meta_vgo2_op[c("FVC_pred", "age")] <- lapply(meta_vgo2_op[c("FVC_pred", "age")], as.numeric)
meta_vgo3_np[c("FVC_pred", "age")] <- lapply(meta_vgo3_np[c("FVC_pred", "age")], as.numeric)
meta_vgo3_op[c("FVC_pred", "age")] <- lapply(meta_vgo3_op[c("FVC_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FVC_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FVC)
perm_FVC_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FVC)
perm_FVC_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FVC)
perm_FVC_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FVC)


#### FVC categorised predicted multivariable PERMANOVA models ####
# Select required variables
covariates_FVC_pred <- c("FVC_category", "age", "season_microbiome_sampling",
                          "gender", "smoke_status")

# Convert FVC_pred and age to numeric
meta_vgo2_np[c("FVC_category", "age")] <- lapply(meta_vgo2_np[c("FVC_category", "age")], as.numeric)
meta_vgo2_op[c("FVC_category", "age")] <- lapply(meta_vgo2_op[c("FVC_category", "age")], as.numeric)
meta_vgo3_np[c("FVC_category", "age")] <- lapply(meta_vgo3_np[c("FVC_category", "age")], as.numeric)
meta_vgo3_op[c("FVC_category", "age")] <- lapply(meta_vgo3_op[c("FVC_category", "age")], as.numeric)

# PERMANOVAs
perm_FVC_cat_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FVC_pred)
perm_FVC_cat_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FVC_pred)
perm_FVC_cat_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FVC_pred)
perm_FVC_cat_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FVC_pred)


#### FEV1FVC predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1FVC <- c("FEV1FVC_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo2_np[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo2_op[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo2_op[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo3_np[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo3_np[c("FEV1FVC_pred", "age")], as.numeric)
meta_vgo3_op[c("FEV1FVC_pred", "age")] <- lapply(meta_vgo3_op[c("FEV1FVC_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEV1FVC_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1FVC)
perm_FEV1FVC_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1FVC)
perm_FEV1FVC_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1FVC)
perm_FEV1FVC_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1FVC)


#### FEV1FVC_category_category predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEV1FVC_category <- c("FEV1FVC_category", "age", "season_microbiome_sampling",
                        "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEV1FVC_category", "age")] <- lapply(meta_vgo2_np[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo2_op[c("FEV1FVC_category", "age")] <- lapply(meta_vgo2_op[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo3_np[c("FEV1FVC_category", "age")] <- lapply(meta_vgo3_np[c("FEV1FVC_category", "age")], as.numeric)
meta_vgo3_op[c("FEV1FVC_category", "age")] <- lapply(meta_vgo3_op[c("FEV1FVC_category", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEV1FVC_category_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEV1FVC_category)
perm_FEV1FVC_category_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEV1FVC_category)


#### FEF25_75 predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEF25_75 <- c("FEF25_75_pred", "age", "season_microbiome_sampling",
                "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEF25_75_pred", "age")] <- lapply(meta_vgo2_np[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo2_op[c("FEF25_75_pred", "age")] <- lapply(meta_vgo2_op[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo3_np[c("FEF25_75_pred", "age")] <- lapply(meta_vgo3_np[c("FEF25_75_pred", "age")], as.numeric)
meta_vgo3_op[c("FEF25_75_pred", "age")] <- lapply(meta_vgo3_op[c("FEF25_75_pred", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEF25_75_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEF25_75)
perm_FEF25_75_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEF25_75)
perm_FEF25_75_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEF25_75)
perm_FEF25_75_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEF25_75)

#### FEF25_75_category predicted multivariable PERMANOVA models #### 
# Select required variables
covariates_FEF25_75_category <- c("FEF25_75_category_pred", "age", "season_microbiome_sampling",
                         "gender", "smoke_status")

# convert to numeric
meta_vgo2_np[c("FEF25_75_category", "age")] <- lapply(meta_vgo2_np[c("FEF25_75_category", "age")], as.numeric)
meta_vgo2_op[c("FEF25_75_category", "age")] <- lapply(meta_vgo2_op[c("FEF25_75_category", "age")], as.numeric)
meta_vgo3_np[c("FEF25_75_category", "age")] <- lapply(meta_vgo3_np[c("FEF25_75_category", "age")], as.numeric)
meta_vgo3_op[c("FEF25_75_category", "age")] <- lapply(meta_vgo3_op[c("FEF25_75_category", "age")], as.numeric)

# Run PERMANOVAs 
perm_FEF25_75_category_vgo2_np <- run_permanova(ord_vgo2_np, meta_vgo2_np, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo2_op <- run_permanova(ord_vgo2_op, meta_vgo2_op, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo3_np <- run_permanova(ord_vgo3_np, meta_vgo3_np, covariates_FEF25_75_category)
perm_FEF25_75_category_vgo3_op <- run_permanova(ord_vgo3_op, meta_vgo3_op, covariates_FEF25_75_category)
```

## Combine and present all multivariable PERMANOVA results
```{r}
# Combine all multivariable PERMANOVA results into one dataframe
perm_all_multi <- bind_rows(
  perm_FEV1_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1_pred"),
  perm_FEV1_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1_pred"),
  perm_FEV1_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1_pred"),
  perm_FEV1_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1_pred"),
  
  perm_FEV1_cat_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1_category"),
  perm_FEV1_cat_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1_category"),
  
  perm_FVC_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FVC_pred"),
  perm_FVC_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FVC_pred"),
  perm_FVC_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FVC_pred"),
  perm_FVC_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FVC_pred"),
  
  perm_FVC_cat_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FVC_category"),
  perm_FVC_cat_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FVC_category"),
  perm_FVC_cat_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FVC_category"),
  perm_FVC_cat_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FVC_category"),
  
  perm_FEV1FVC_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1FVC_pred"),
  perm_FEV1FVC_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1FVC_pred"),
  
  perm_FEV1FVC_category_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEV1FVC_category"),
  perm_FEV1FVC_category_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEV1FVC_category"),
  
  perm_FEF25_75_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEF25_75_pred"),
  perm_FEF25_75_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEF25_75_pred"),
  
  perm_FEF25_75_category_vgo2_np %>% mutate(dataset = "VGO2", niche = "NP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo2_op %>% mutate(dataset = "VGO2", niche = "OP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo3_np %>% mutate(dataset = "VGO3", niche = "NP", outcome = "FEF25_75_category"),
  perm_FEF25_75_category_vgo3_op %>% mutate(dataset = "VGO3", niche = "OP", outcome = "FEF25_75_category")
) %>%
  # Keep only the overall model rows
  filter(term == "Model") %>%
  mutate(dataset_niche = paste(dataset, niche, sep = "_"))

# Heatmap
# Reorder variables for heatmap (consistent with univariable plots)
ordered_vars <- rev(c(
  "age", "gender", "season_microbiome_sampling", "smoke_status",
  "FEV1_pred", "FVC_pred", "FEV1FVC_pred", "FEF25_75_pred", "FEV1_category", "FVC_category", "FEV1FVC_category", "FEF25_75_category"
))

# Prepare plotting dataframe
perm_all_multi$p.value
plot_perm_multi <- perm_all_multi %>%
  mutate(variable = outcome,
         variable = factor(variable, levels = ordered_vars),
         significance = case_when(
           p.value < 0.001 ~ "***",
           p.value < 0.01  ~ "**",
           p.value < 0.05  ~ "*",
           TRUE ~ "ns"
         ),
         R2 = round(R2, 3))

# Split into VGO2 and VGO3 panels
plot_vgo2_multi <- plot_perm_multi %>% filter(dataset == "VGO2")
plot_vgo3_multi <- plot_perm_multi %>% filter(dataset == "VGO3")

# Calculate R² range for VGO2 and VGO3 separately
r2_min_vgo2 <- min(plot_vgo2_multi$R2, na.rm = TRUE)
r2_max_vgo2 <- max(plot_vgo2_multi$R2, na.rm = TRUE)
r2_mid_vgo2 <- median(plot_vgo2_multi$R2, na.rm = TRUE)

r2_min_vgo3 <- min(plot_vgo3_multi$R2, na.rm = TRUE)
r2_max_vgo3 <- max(plot_vgo3_multi$R2, na.rm = TRUE)
r2_mid_vgo3 <- median(plot_vgo3_multi$R2, na.rm = TRUE)

# Heatmap for VGO2
heatmap_vgo2_multi <- ggplot(plot_vgo2_multi, aes(x = niche, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo2,  # center gradient at median R²
    limits = c(r2_min_vgo2, r2_max_vgo2),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2)
  ) +
  facet_wrap(~ dataset, ncol = 1) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(y = "Lung Function Variable")

# Heatmap for VGO3
heatmap_vgo3_multi <- ggplot(plot_vgo3_multi, aes(x = niche, y = variable, fill = R2)) +
  geom_tile(color = "white") +
  geom_text(aes(label = significance), color = "black", size = 4) +
  scale_fill_gradient2(
    low = "#F6EDBD",
    mid = "#E0A275",
    high = "#CA562C",
    midpoint = r2_mid_vgo3,  # center gradient at median R²
    limits = c(r2_min_vgo3, r2_max_vgo3),  # scale based on min and max
    labels = function(x) sprintf("%.3f", x),
    name = expression(R^2)
  ) +
  facet_wrap(~ dataset, ncol = 1) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(y = "Lung Function Variable")

# Combine both heatmaps
multi_beta_plot <- (heatmap_vgo2_multi / heatmap_vgo3_multi) +
  patchwork::plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(legend.position = "right")

multi_beta_plot <- multi_beta_plot +
  patchwork::plot_annotation(
    title = "Multivariable PERMANOVAs (VGO2 and VGO3)",
    caption = "Significance codes: *** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant",
    theme = theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      plot.caption = element_text(size = 8, hjust = 0.5)
    )
  )

print(multi_beta_plot)

# Save figure
ggsave("../Output/Beta_diversity/Multivariable_PERMANOVAS/Multivariable_PERMANOVA_R2_heatmap.svg",
       multi_beta_plot, width = 4.5, height = 5)

```


